---
title: "Table S2"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Setup 

```{r, message=FALSE, warning=FALSE}
R.version.string
library(phyloseq); packageVersion('phyloseq')
library(ggplot2); packageVersion('ggplot2')
library(tableone); packageVersion('tableone')
library(tidyr); packageVersion('tidyr')
library(reshape2); packageVersion('reshape2')

source("Preprocessing/Functions.R")
```


```{r}
theme_set(theme_bw() + theme(plot.title = element_text(hjust = 0.5,face = "bold"), strip.text.x = element_text(size = 11), strip.background = element_rect(colour="white", fill="white"), panel.grid.major = element_blank(), panel.grid.minor = element_blank(), text = element_text(size=10), axis.text = element_text(color="black", size=12), legend.key.size = unit(0.5,"cm")))

mycols30 <- c("#e9fcff","#7435e2","#b9f2b7","#25007a","#ffed5c","#0153d2","#75a800","#93007a", "purple", "#ef0025","#fd6a00","#0191f9","#ffa32b","#004769","#c8ff97","#a3005a","#5d8300","#b5b2ff","#8e7a00","#ff77a1","#004b19","#ff7a5f","#00615b","#9f6000","#715333","#ffedb0","#540500","#00d488", "black","#ffbbd2","#01bfae", "#e82769", "grey")
```

**Data subsets**
```{r} 
data_NG2 <- readRDS("Data_subsets/Main_UnRarefied_NasalGut_Subset.rds")

## total samples, not paired (used for Fig 1B & C stacked bars)
table(sample_data(data_NG2)$SampleType, sample_data(data_NG2)$Visit)


# Data used for downstream trajectory analyses - participants with both 3 month and 1 year data for nasal and/or gut (use for Table S1) 
Traj_data_ls <- readRDS("Data_subsets/ALL_Trajectory_Derived_Data.rds") 
Traj_data_ls_2 <- merge( subset(Traj_data_ls$Trajectories_Nasal_to181_Gut_at182, select = c("Observed_N1y3m", "Observed_G1y3m")), 
                         Traj_data_ls$Metadata, by = "row.names")
N_3m1y <- subset(Traj_data_ls_2, !is.na(Observed_N1y3m))
G_3m1y<- subset(Traj_data_ls_2, !is.na(Observed_G1y3m))
NG_3m1y <- subset(Traj_data_ls_2, !is.na(Observed_G1y3m) & !is.na(Observed_N1y3m)) 

Traj_dfs_Long <- rbind(N_3m1y, G_3m1y, NG_3m1y)

Traj_dfs_Long$Sample_Visit <- c(rep("Nasal Trajectory", nrow(N_3m1y)), rep("Gut Trajectory", nrow(G_3m1y)), rep("Nasal-Gut Trajectory", nrow(NG_3m1y)))

table(Traj_dfs_Long$Sample_Visit)
```


# Participant characteristic table for paired data vs overall data 
```{r}
## Also add column for all participants in CHILD - read in this metadata
CHILD_metadata_all3 <- readRDS(file="Preprocessing/metadata_input/4FINAL_Full_Cleaned_KF_Modified_V2_withRE_forSubj_withNG.rds") 

# variables used (either in main analyses or as covariates/etc.) in analyses
Vars_used <- c("age_1y_months", "age_3m_months", "atbx", "AwayfromHome_1y", "BF_12m_sampling", "BF_status_3m_sampling", "birthmode2", "childgender", "colds_0to3m", "coldsnum_6to12m", "Fever_1y",  "Fever_3m", "gest_days", "mom_age_birth", "mom_asthma_ever",  "mom_atopy",  "mom_bmi_best", "mom_smoke_1y", "Mother_abs_birth_yn",  "older_sibs", "prenatal_smoke", "race_mom4", "Season_1y", "Season_3m", "StudyCenter", "weight_12", "weight_3", "RE_Virus_1y",  "RE_Virus_3m", "WheezeTraj",  "asthma_3y", "asthma_5y" , "atopy3y", "Sample_Visit")

# data_NG2 sample data as dataframe
data_NG_forS1 <- as(sample_data(data_NG2), "data.frame")

# Add some missing varibles to datasets
## add the sample_visit var to other datasets too - this variable will be used to differentiate these datasets after merging 
data_NG_forS1$Sample_Visit <- paste(data_NG_forS1$SampleType, data_NG_forS1$Visit, sep=" ")
CHILD_metadata_all3$Sample_Visit <- "Overall Cohort"

## add the RE_Virus_1y and RE_Virus_3m variable for data_NG2
data_NG_forS1$RE_Virus_1y <- ifelse(as.character(data_NG_forS1$Sample_Visit) %in% c("Gut 1 Year", "Nasal 1 Year"), as.character(data_NG_forS1$RE_Virus), NA) 
### IF Visit == 1 year, then this is equal to RE_Virus var otherwise make NA
data_NG_forS1$RE_Virus_3m <- ifelse(as.character(data_NG_forS1$Sample_Visit) %in% c("Gut 3 Months", "Nasal 3 Months"), as.character(data_NG_forS1$RE_Virus), NA) ### IF Visit == 3 month, then this is equal to RE_Virus var otherwise make NA

# select only variables used for each dataset to include
data_NG_forS1_2 <- subset(data_NG_forS1, select = Vars_used)

Traj_dfs_forS1 <- subset(Traj_dfs_Long, select = Vars_used)

## overall cohort edit - in addition to selecting only variables used, remove participants with not even implied breastfeeding data first (these participants are from the vangard subset)
overall_forS1 <- subset(CHILD_metadata_all3, !is.na(BF_duration_imp), select = Vars_used)

# MERGE datasets

All_datasets_S1 <- rbind(data_NG_forS1_2, Traj_dfs_forS1, overall_forS1)
All_datasets_S1$Sample_Visit <- factor(All_datasets_S1$Sample_Visit, levels = c("Nasal 3 Months", "Nasal 1 Year", "Gut 3 Months", "Gut 1 Year", "Nasal Trajectory", "Gut Trajectory", "Nasal-Gut Trajectory", "Overall Cohort"))
table(All_datasets_S1$Sample_Visit) 
```

```{r}
# Extra edit - want gestational age in weeks for table
All_datasets_S1$gest_weeks <- All_datasets_S1$gest_days/7
All_datasets_S1$gest_days <- NULL

# TABLE 1
colnames(All_datasets_S1) # rename 
colnames(All_datasets_S1) <- c("Age; 1-year (months)","Age; 3-month (months)","Antibiotics; 0-1Y", "Daycare, 1Y", "Breastfeeding; 1-year",  "Breastfeeding status; 3-month", "Birth mode", "Sex", "Colds; 0-3M", "number Colds, 6-12M", "Fever; 6-12M", "Fever; 0-3M", "Maternal age", "Maternal asthma", "Maternal atopy", "Maternal BMI", "Maternal smoke; 1Y", "Antibiotics; birth", "Older siblings", "Prenatal smoke", "Maternal ethnicity", "Season; 1-year", "Season; 3-month","Center", "Weight; 1Y (kg)", "Weight, 3M (kg)", "R/E virus; 1-year", "R/E virus; 3-month", "Wheeze Trajectory", "Asthma; 3Y", "Asthma; 5Y", "Atopy; 3Y", "Sample_Visit", "Gestational age (weeks)") # 3-month and 1-year indicate sampling times. 6M/1Y/3M=assessments/visits

# grouping Transient and intermediate wheeze for table
All_datasets_S1$`Wheeze Trajectory` <- gsub("Intermediate|Transient", "Transient or Intermediate", All_datasets_S1$`Wheeze Trajectory`)
table(All_datasets_S1$`Wheeze Trajectory`)

# reorder some factor levels
All_datasets_S1$`Wheeze Trajectory` <- factor(All_datasets_S1$`Wheeze Trajectory`, levels = c("Persistent", "Never/infrequent", "Transient or Intermediate"))
All_datasets_S1$`Asthma; 3Y` <- factor(All_datasets_S1$`Asthma; 3Y`, levels = c("Probable", "No", "Possible"))
All_datasets_S1$`Asthma; 5Y` <- factor(All_datasets_S1$`Asthma; 5Y`, levels = c("Probable", "No", "Possible")) # marking Possible as excluded
All_datasets_S1$`Breastfeeding status; 3-month` <- factor(All_datasets_S1$`Breastfeeding status; 3-month`, levels = c("Exclusive", "Partial", "No BM"))
All_datasets_S1$`Season; 1-year` <- factor(All_datasets_S1$`Season; 1-year`, levels = c("Winter", "Spring", "Summer", "Fall"))
All_datasets_S1$`Season; 3-month` <- factor(All_datasets_S1$`Season; 3-month`, levels = c("Winter", "Spring", "Summer", "Fall"))



# reorder vars in table func rather than dataframe
reordered_Vars <- c( "Breastfeeding status; 3-month","Breastfeeding; 1-year","R/E virus; 3-month", "R/E virus; 1-year", "Colds; 0-3M", "number Colds, 6-12M", "Fever; 0-3M", "Fever; 6-12M", "Older siblings", "Daycare, 1Y", "Center", "Season; 3-month", "Season; 1-year", "Antibiotics; birth", "Antibiotics; 0-1Y", "Maternal smoke; 1Y",  "Prenatal smoke", "Maternal BMI", "Birth mode", "Maternal age",  "Maternal ethnicity", "Maternal asthma", "Maternal atopy","Sex", "Weight, 3M (kg)", "Weight; 1Y (kg)", "Gestational age (weeks)", "Age; 3-month (months)", "Age; 1-year (months)", "Wheeze Trajectory", "Asthma; 3Y", "Asthma; 5Y", "Atopy; 3Y")

Table1_All_dfs <- CreateTableOne(vars = reordered_Vars, data = All_datasets_S1, strata = "Sample_Visit", includeNA=FALSE, test = TRUE) 
Table1_All_dfs <- print(Table1_All_dfs, quote = FALSE, noSpaces = TRUE, printToggle = FALSE, contDigits=1)

library(knitr)
kable(Table1_All_dfs, format = "pipe")


```


# Genus level relative abundances, overall microbiota composition (Table S2)
Aggregate to genus level and use relative data
```{r}
## aggregate to genus level 
data_NG2_G <- tax_glom(data_NG2, taxrank = "Genus")

NG_relative <- transform_sample_counts(data_NG2_G, function(x) x*100 / sum(x) )  
```

Split data by SampleType and time-point for taxa filter - want to plot genera present in 70% of samples at EITHER time-point
```{r}
genus_n_3m <- subset_samples(NG_relative, SampleType=="Nasal" & Visit == "3 Months")
genus_n_1y <- subset_samples(NG_relative, SampleType=="Nasal" & Visit == "1 Year")

genus_gut_3m <- subset_samples(NG_relative, SampleType=="Gut" & Visit == "3 Months")
genus_gut_1y <- subset_samples(NG_relative, SampleType=="Gut" & Visit == "1 Year")
```

Select taxa to plot
- Genera present in at least 70% of samples in at EITHER time-point will be colored in the plot with the remainder grouped into "Other"  
```{r}
data.p60_n_3m = filter_taxa(genus_n_3m, function(x) sum(x > 0) > (0.70*length(x)), TRUE) 
ntaxa(data.p60_n_3m) 
data.p60_n_1y = filter_taxa(genus_n_1y, function(x) sum(x > 0) > (0.70*length(x)), TRUE) 
ntaxa(data.p60_n_1y)

data.p60_gut_3m = filter_taxa(genus_gut_3m, function(x) sum(x > 0) > (0.70*length(x)), TRUE) 
ntaxa(data.p60_gut_3m)
data.p60_gut_1y = filter_taxa(genus_gut_1y, function(x) sum(x > 0) > (0.70*length(x)), TRUE) 
ntaxa(data.p60_gut_1y)
```

Nasal taxa to plot for 3-month & 1-year data
```{r}
Ntaxa_toPlot <- c(get_taxa_unique(data.p60_n_3m, taxonomic.rank = "Genus"), 
                 get_taxa_unique(data.p60_n_1y, taxonomic.rank = "Genus"))
Ntaxa_toPlot2 <- subset(Ntaxa_toPlot, !duplicated(Ntaxa_toPlot)) # remove dups found in more than 1 sample type
length(Ntaxa_toPlot2) # unique Genera 

# Use subset of all nasal samples
N_relative <- subset_samples(NG_relative, SampleType=="Nasal")

# Make overall nasal dataset of prevalent genera in either time-point 
N_relative_Sub <- subset_taxa(N_relative, Genus %in% Ntaxa_toPlot2)
N_relative_Sub
```

Gut taxa to plot
```{r}
Gtaxa_toPlot <- c(get_taxa_unique(data.p60_gut_3m, taxonomic.rank = "Genus"),
                 get_taxa_unique(data.p60_gut_1y, taxonomic.rank = "Genus"))
Gtaxa_toPlot2 <- subset(Gtaxa_toPlot, !duplicated(Gtaxa_toPlot))
length(Gtaxa_toPlot2) # unique Genera 

# Use subset of all nasal samples
G_relative <- subset_samples(NG_relative, SampleType=="Gut")

# Make overall nasal dataset of prevalent genera in either time-point 
G_relative_Sub <- subset_taxa(G_relative, Genus %in% Gtaxa_toPlot2)
G_relative_Sub # more than 13 b/c of contested [Eubacterium] with different Family level taxonomy - 4 different [Eubacterium] Genera - ok to group for plot.
```


(2) Make summary data for each subset that will be plotted 

Nasal dataset for stacked bars, stratify by Visit
```{r}
n_genus_melt <- psmelt(N_relative_Sub) 

#DATA Summary 
Genus_n_summary <- data_summary(n_genus_melt, varname = "Abundance", groupnames = c("Genus", "Visit"))

## Keep only these columns for plotting
Genus_n_summary2 <-  subset(Genus_n_summary, select = c("Mean", "Genus", "Visit"))

# use this alternative to melting the whole dataset to add 'other' group - to get 100% total, this is much quicker computationally vs melting all data
Genus_n_summary2_S <- split(Genus_n_summary2, Genus_n_summary2$Visit) 

Genus_n_summary3m <- rbind(Genus_n_summary2_S$`3 Months`, c(100-sum(Genus_n_summary2_S$`3 Months`$Abundance), "Other", "3 Months")) # adding extra row for "Other" group, in same order: "Mean", "Genus", "Visit"

Genus_n_summary1y <- rbind(Genus_n_summary2_S$`1 Year`, c(100-sum(Genus_n_summary2_S$`1 Year`$Abundance), "Other", "1 Year")) # adding extra row for "Other" group, in same order: "Mean", "Genus", "Visit"


Genus_N_summary3 <- rbind(Genus_n_summary3m, Genus_n_summary1y)
Genus_N_summary3$Mean <- as.numeric(Genus_N_summary3$Mean)
```

Gut dataset for stacked bars, stratify by Visit
```{r}
g_genus_melt <- psmelt(G_relative_Sub) 

#DATA Summary 
Genus_g_summary <- data_summary(g_genus_melt, varname = "Abundance", groupnames = c("Genus", "Visit"))

## Keep only these columns for plotting
Genus_g_summary2 <-  subset(Genus_g_summary, select = c("Mean", "Genus", "Visit"))


# use this alternative to melting the whole dataset to add 'other' group - to get 100% total, this is much quicker computationally vs melting all data
Genus_g_summary2_S <- split(Genus_g_summary2, Genus_g_summary2$Visit) 

Genus_g_summary3m <- rbind(Genus_g_summary2_S$`3 Months`, c(100-sum(Genus_g_summary2_S$`3 Months`$Abundance), "Other", "3 Months")) # adding extra row for "Other" group, in same order: "Abundance", "Genus", "Visit"

Genus_g_summary1y <- rbind(Genus_g_summary2_S$`1 Year`, c(100-sum(Genus_g_summary2_S$`1 Year`$Abundance), "Other", "1 Year")) # adding extra row for "Other" group, in same order: "Abundance", "Genus", "Visit"


Genus_g_summary3 <- rbind(Genus_g_summary3m, Genus_g_summary1y)
Genus_g_summary3$Mean <- as.numeric(Genus_g_summary3$Mean)
```


## Genrate Table S2
```{r}
library(knitr)

Genus_n_summary$Type <- "Nasal"
Genus_g_summary$Type <- "Gut"

Genus_summary <- rbind(Genus_n_summary, Genus_g_summary)

write.csv(Genus_summary, file="Output/TableS2_Nasal_Gut_Composition_Genus.csv")


Genus_summary_split <- split(Genus_summary, Genus_summary$Type)

Genus_summary_nasal <- Genus_summary_split$Nasal[order(Genus_summary_split$Nasal$Visit, -Genus_summary_split$Nasal$Median),]
Genus_summary_gut <- Genus_summary_split$Gut[order(Genus_summary_split$Gut$Visit, -Genus_summary_split$Gut$Median),]

kable(data.frame(Type  = Genus_summary_nasal$Type, Genus_summary_nasal[1:2], round(Genus_summary_nasal[3:11], 2)))

kable(data.frame(Type  = Genus_summary_gut$Type, Genus_summary_gut[1:2], round(Genus_summary_gut[3:11], 2)))
```



