---
title: "Figure 5. Premature acquisition of specific R.gnavus functions are linked to early breastfeeding cessation, increased asthma risk, and tryptophan metabolites variations."
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```
 
 
```{r}
## Setup 
R.version.string
library(phyloseq)
library(ggplot2)
library(tidyr)
library(reshape2)
library(dplyr)
library(UpSetR)
library(ComplexHeatmap)
library(circlize)
library(ggpubr)
library(cowplot)
library(here)

source(paste0(here::here(), "/Functions.R"))

theme_set(theme_bw() + theme(plot.title = element_text(hjust = 0.5,face = "bold"), strip.text.x = element_text(size = 11), strip.background = element_rect(colour="white", fill="white"), panel.grid.major = element_blank(), panel.grid.minor = element_blank(), text = element_text(size=12), axis.text = element_text(color="black", size=12), legend.key.size = unit(0.5,"cm")))
```

# Figure 5A & 5B - metagenomic results - overlap in functions delayed with asthma and breastfeeding

## Read in Metagenomic PreTCO results

```{r}
# Results from EC permutation tests - per function tests
Res_per_taxa_ec_asthma <- read.csv(file = paste0(path, "Tables/Res_Fig4A_EC_StrataAsthmaPreTCo.csv"), row.names = 1)
Res_per_taxa_ec_bf <- read.csv(file = paste0(path, "Tables/Res_Fig4A_EC_StrataBFPreTCo.csv"), row.names = 1)
```

## EC overlap b/t functions associated with BF and Asthma
```{r} 
p_thresh2 <- 0.01

Res_per_taxa_ec_asthma_late2 <- subset(Res_per_taxa_ec_asthma, p_value < p_thresh2)

Res_per_taxa_ec_bf_late2 <- subset(Res_per_taxa_ec_bf, p_value < p_thresh2)

BF_asth3y_ec_01 <- intersect(row.names(Res_per_taxa_ec_bf_late2), row.names(Res_per_taxa_ec_asthma_late2))
length(BF_asth3y_ec_01) 

BF_asth3y_ec_01_2 <- data.frame(BF_asth3y_ec_01) %>% 
                    separate(BF_asth3y_ec_01, c("Function", "Taxa"), sep = "_")

table(BF_asth3y_ec_01_2$Taxa)
BF_asth3y_ec_01_2
```

## Checks for Figure 5A Venn diagrams & prep for Figure 5B EC heatmap
```{r} 
# making venn diagrams and/or upset plots
asthma_ec_ls1 <- Res_per_taxa_ec_asthma_late2 %>% mutate(variable = row.names(.)) %>% 
    separate(variable, c("Function", "Taxa"), sep = "_") %>% 
    split(., .[["Taxa"]])

bf_ec_ls1 <- Res_per_taxa_ec_bf_late2 %>% mutate(variable = row.names(.)) %>% 
    separate(variable, c("Function", "Taxa"), sep = "_") %>% 
    split(., .[["Taxa"]])

# For Figure 5A
## Upset plots just showing overlap b/t BF and asthma comparison for Function*Taxa strata (use to create venn diagrams) - not showing overlap in functions shared b/t taxa 
# Just making venn diagram of these numbers instead

asthma_ec_funcs <- lapply(asthma_ec_ls1, function(x) row.names(x))
names(asthma_ec_funcs) <- paste0("Asthma - ", names(asthma_ec_funcs))

bf_ec_funcs <- lapply(bf_ec_ls1, function(x) row.names(x))
names(bf_ec_funcs) <- paste0("BF - ", names(bf_ec_funcs))


ec_ast_bf_ls_late <- c(as.list(asthma_ec_funcs), as.list(bf_ec_funcs))

#tiff(file = "Figures/EC_UpSet_FuncTaxa_Overlap_onlyComparison.tiff", width = 1500, height = 1500, units = "px", res = 300)
upset(fromList(ec_ast_bf_ls_late), nsets = length(ec_ast_bf_ls_late), 
      order.by = "freq") 
#dev.off()
# Total late functions (stratified by species)
length(unique(unlist(ec_ast_bf_ls_late))) # 398

## Extra version - Upset plots showing shared functions - both across Taxa (ie same function found in different taxa) - and across comparisons (later with BF and also later with asthma) 

asthma_ec_funcs <- lapply(asthma_ec_ls1, function(x) x[["Function"]])
names(asthma_ec_funcs) <- paste0("Asthma - ", names(asthma_ec_funcs))

bf_ec_funcs <- lapply(bf_ec_ls1, function(x) x[["Function"]])
names(bf_ec_funcs) <- paste0("BF - ", names(bf_ec_funcs))


ec_ast_bf_ls_late <- c(as.list(asthma_ec_funcs), as.list(bf_ec_funcs))

#tiff(file = "Figures/EC_UpSet_Func_Overlap_TaxaandComparison.tiff", width = 1500, height = 1100, units = "px", res = 300)
upset(fromList(ec_ast_bf_ls_late), nsets = length(ec_ast_bf_ls_late), 
      order.by = "freq") 
#dev.off()
# Total late functions (NOT stratified by species - not unique to species necessarily)
length(unique(unlist(ec_ast_bf_ls_late))) # 385

## Also check totals for these function*taxa

## EC
lapply(ec_ast_bf_ls_late, length)
```

## Figure 5B - EC heatmap
```{r} 
Res_per_taxa_ec_bf_late2[grepl("5.3.1.24.phosphoribosylanthranilate.isomerase_ruminococcus.gnavus|4.1.1.48|2.4.2.18.anthranilate.phosphoribosyltransferase_ruminococcus.gnavus|4.1.3.27|4.2.1.20", row.names(Res_per_taxa_ec_bf_late2)),]

# Heatmap now referenced in main text primarily to highlight importance of breastfeeding for tryptophan biosynthesis pathway - focus on functions that are highly associated with BF (p < 0.001 for BF specifically, plus at least p<0.05 for asthma) 
Res_per_taxa_ec_bf_001_late <- subset(Res_per_taxa_ec_bf_late2, p_value < 0.001 & original_effect_diff < -20)
nrow(Res_per_taxa_ec_bf_001_late) # still 313 (of 603 w/t p<0.05) - at least as large of an effect as half of these. 

# EC
# Make heatmap showing which are Late - Change in prevalence in Asthma, no Asthma, BM and No BM
## keep all late for Asthma using all.x true
Res_per_taxa_egg_asthma_BF_late <- merge(Res_per_taxa_ec_asthma_late2, Res_per_taxa_ec_bf_late2, 
                                         all.x = TRUE,
                                         by = "row.names")
dim(Res_per_taxa_egg_asthma_BF_late)
# Add annotation for taxa as well, rather than showing that in the rownames

Res_per_taxa_egg_asthma_BF_late_2 <- Res_per_taxa_egg_asthma_BF_late %>% separate(Row.names, c("Function", "Taxa"), sep = "_")
row.names(Res_per_taxa_egg_asthma_BF_late_2) <- Res_per_taxa_egg_asthma_BF_late_2$Function


## edit row names so that figure looks better
row.names(Res_per_taxa_egg_asthma_BF_late_2)

row.names(Res_per_taxa_egg_asthma_BF_late_2) <- row.names(Res_per_taxa_egg_asthma_BF_late_2) %>%
                                                  gsub("^X|[:alpha:].*", "", .) %>%  ## not rm all alpha for some reason
                                                  gsub("c$|g$|n$|dn$|tr$|m$|di$", "", .) %>%  
                                                  gsub("\\.$", "", .)

# first order by effect size
Res_per_taxa_egg_asthma_BF_late_2 <- Res_per_taxa_egg_asthma_BF_late_2[order(Res_per_taxa_egg_asthma_BF_late_2$original_effect_diff.y),]

# Have 'Taxa' annotations, now add biosynthesis pathway annotations
## searching EC numbers on https://enzyme.expasy.org/enzyme-search-ec.html and entering in pathway 
Res_per_taxa_egg_asthma_BF_late_2$Function ## must match up w/t categories!
Res_per_taxa_egg_asthma_BF_late_2$Function_Category <- c("Glycerol metabolism", "DNA replication", 
                                                        "DNA replication", "Glycerol metabolism",
                                                         "Amino Acid metabolism", "Tryptophan biosynthesis",
                                                         "Siroheme biosynthesis", "Bile hydrolysis",
                                                        "p>0.05 for BM", "p>0.05 for BM",
                                                        "p>0.05 for BM", "p>0.05 for BM") ## not annotating the 4 enzymes that are not overlapping 

## Now that we see related pathways also want ordered by this a bit - re-ordering to make Tryptophan show up first, then the 2 glycerol metabolism functions, keep rest by effect size
Res_per_taxa_egg_asthma_BF_late_2$Fun_order <- seq(1, nrow(Res_per_taxa_egg_asthma_BF_late_2)) #%>% keeping same order with below exceptions
Res_per_taxa_egg_asthma_BF_late_2$Fun_order <- ifelse(Res_per_taxa_egg_asthma_BF_late_2$Function_Category == "Tryptophan biosynthesis", 0,
                                                       ifelse(Res_per_taxa_egg_asthma_BF_late_2$Function_Category == "Glycerol metabolism", 1,
                                                              Res_per_taxa_egg_asthma_BF_late_2$Fun_order))
Res_per_taxa_egg_asthma_BF_late_2$Fun_order
Res_per_taxa_egg_asthma_BF_late_2$Function_Category

Res_per_taxa_egg_asthma_BF_late_3 <- Res_per_taxa_egg_asthma_BF_late_2[order(Res_per_taxa_egg_asthma_BF_late_2$Fun_order),]
Res_per_taxa_egg_asthma_BF_late_3$Function

# make matrix for heatmap
heatmap_2 <- subset(Res_per_taxa_egg_asthma_BF_late_3, select = c("BF_status_3m_sampling_Exclusive", "BF_status_3m_sampling_No.BM", "asthma_3y_No", "asthma_3y_Yes"))
colnames(heatmap_2) <- c("Exclusive BM", "No BM", "No Asthma", "Asthma")

annotations_2 <- subset(Res_per_taxa_egg_asthma_BF_late_3, select = c("Taxa", "Function_Category"))

## for annotations, want legend in specific order too...
annotations_2$Taxa <- factor(annotations_2$Taxa, levels = c("ruminococcus.gnavus", "bifidobacterium.longum", "eubacterium.hallii"))
annotations_2$Function_Category <- factor(annotations_2$Function_Category, levels =unique(annotations_2$Function_Category))

## for taxa colors
cols_sub <- c("#58b6ed","#9eae6a","#ae6a7c") ## match cols of taxa in Fig 4 but heatmap w/t just R. gnavus, B. longum and E. hallii
names(cols_sub) <- levels(annotations_2$Taxa)

## for function colors
length(unique(annotations_2$Function_Category)) ## should be 7, want last one to be nothing/white
cols_func <- c("#f5ab51", "#de7074",
"#0186eb",
"#91e97e",
"#57004d",
"#ff81c0", "grey") ## get 5 other cols 
names(cols_func) <- unique(annotations_2$Function_Category)

cn = colnames(t(heatmap_2))

ha = HeatmapAnnotation(Taxa = annotations_2$Taxa, 
                       Function_Category = annotations_2$Function_Category,
                       col = list(Taxa = cols_sub,
                                  Function_Category = cols_func),
                        text = anno_text(cn, rot = 45, location = unit(1, "npc"), just = "right")
                       )

ht = Heatmap(t(heatmap_2), cluster_rows = FALSE, cluster_columns = FALSE, 
             name = "1 Year - 3 Months\nPrevalence (%)", 
        colorRamp2(c(80, 40, 0, -50), c("red", "pink", "white", "blue")), 
             show_column_names = FALSE, row_names_side = "left",
             bottom_annotation = ha
    ) 

tiff(file = paste0(path, "Figures/EC_heatmap_v02.tiff"), width = 2700, height = 1100, units = "px", res = 300)
draw(ht, heatmap_legend_side = "left", annotation_legend_side = "left")
dev.off()
```


# Figure 5C-5D - Metabolite analyses

## Figure 5C - Tryptophan Metabolites

## Figure 5D - Short Chain Fatty Acids (SCFAs)

```{r}
metab_nmr <- read.csv(paste0(path, "Data/Metagenome_metabolome/BatchCorrect_NMR_matrix.csv"), row.names = 1)

metab_ms <- read.csv(paste0(path, "Data/Metagenome_metabolome/BatchCorrect_LCMS_matrix.csv"), row.names = 1)

metab_sdf_1y <- read.csv(paste0(path, "Data/Metagenome_metabolome/Child_Stool_1 year.csv"))
metab_sdf_3m <- read.csv(paste0(path, "Data/Metagenome_metabolome/Child_Stool_3 months.csv"))

sdf <- read.csv(paste0(path, "Data/Data_subsets/CHILD_metadata_fully_cleaned.csv")) ## see Processing_Metagenomics.Rmd 
```

Extra - Create a data dictionary for metabolite data
```{r}
# Lists of analyte names for future analyses & make metabolite data dictionary - for annotations
ms_metabs <- colnames(metab_ms)
nmr_metabs <- colnames(metab_nmr)
colSums(is.na(metab_nmr))
# Also make annotations for the metabolites:

# SCFA & related: Butyrate, Propionate, Valerate, X4.Aminobutyrate, Isobutyric.acid, Isovaleric.acid
SCFA_rel <- c("X4.Aminobutyrate", "Isobutyric.acid", "Isovaleric.acid")
SCFA_main <- c("Butyrate", "Propionate", "Valerate", "Acetic.acid")

# Other acids: Formate (regulates anaerobic fermentation, pyruvate related), Phenylacetate, X3.Phenylpropionate, Nicotinic.acid, X5.Aminopentanoate 
acids <- c("Formate", "Phenylacetate", "X3.Phenylpropionate", "Nicotinic.acid", "X5.Aminopentanoate")

# Simple fermentation products & solvents: Ethanol, Isopropanol, Acetone, Acetoin, p.Cresol, Methanol, Propylene.glycol, Cadaverine
simp_ferm_solv <- c("Ethanol", "Isopropanol", "Acetone", "Acetoin", "p.Cresol", "Methanol", "Propylene.glycol", "Cadaverine")

# CHO: Fructose, Xylose, D.Galactose, L.Fucose, Myo.inositol, Arabinose
CHO <- c("Fructose", "Xylose", "D.Galactose", "L.Fucose", "Myo.inositol", "Arabinose")

# Lipid related: Glycerol, 
lipid <- "Glycerol"

# Nitrogenous: N6.Acetyllysine, Pyroglutamic.acid, Uracil, Trimethylamine
N <- c("Pyroglutamic.acid", "Uracil")

# Nitrogenous with Acetyl or Methyl group
N_AcetMeth <- c("N6.Acetyllysine", "Trimethylamine")

# Grouping ms metabs more systematically
SCFA_rel_ms <- ms_metabs[grepl("butyric|valeric|acetic|propionic|propionate|acetate", ms_metabs)] ## To prioritize this over 'ms_acids' in ifelse (to separate scfa's from other acids)
ms_acids <- ms_metabs[grepl("acid|ate$|HPHPA", ms_metabs)]
ms_PCs <- ms_metabs[grepl("^PC", ms_metabs)]
ms_LYSOC <- ms_metabs[grepl("^LYSOC", ms_metabs)]
ms_SMs <- ms_metabs[grepl("^SM|SM$", ms_metabs)]
ms_Cer <- ms_metabs[grepl("^Cer|^Hex2Cer|^HexCer", ms_metabs)]
ms_CEs <- ms_metabs[grepl("^CE", ms_metabs)]
ms_DG_TG <- ms_metabs[grepl("^DG|^TG", ms_metabs)]
ms_N <- ms_metabs[grepl("in$|ine$|Diaminopropane|DOPA|Methionine.sulfoxide|Nudifloramide", ms_metabs)] # at least all these will contain nitrogen, if not a full AA
ms_AcMethN <- ms_metabs[grepl("Acetyl|Methyl", ms_metabs)] ## a couple acids mistakenly added - won't be included b/c prioritizing acids in ifelse
ms_shortC <- ms_metabs[grepl("^C4$|^C5$|C7DC$", ms_metabs)]
Tryp_ind <- c("Tryptophan", "Indole") ## also want tryptophan & indole in own group together b/c of potential extra importance, plus these compounds are metabolically related

# Make a metabolite annotation based on above info
metab_annotation <- data.frame(variable = c(nmr_metabs, ms_metabs), 
                               method = c(rep("NMR", length(nmr_metabs)), rep("MS", length(ms_metabs)))) %>% 
                    mutate(., class = ifelse(variable %in% c(SCFA_rel, SCFA_rel_ms), "other SCFAs & related", 
                                          ifelse(variable %in% SCFA_main, "Main SCFAs",
                                                 ifelse(variable %in% c(acids, ms_acids), "Acids (non-SCFA)",
                                                        ifelse(variable %in% simp_ferm_solv, "Solvents/Ferment-products",
                                                               ifelse(variable %in% CHO, "CHO", ifelse(variable %in% c(lipid, ms_DG_TG), "Lipids & related",
                                                                      ifelse(variable %in% c(N, ms_N), "Nitrogenous",
                                                                             ifelse(variable %in% c(N_AcetMeth, ms_AcMethN), "N wt Methyl/Acetyl",
                                                                                    ifelse(variable %in% ms_PCs, "PCs", 
                                                                                           ifelse(variable %in% ms_LYSOC, "LYSOC",
                                                                                                  ifelse(variable %in% ms_SMs, "SMs",
                                                                                                         ifelse(variable %in% ms_Cer, "Ceramides" ,
                                                                                                                ifelse(variable %in% ms_CEs, "CEs",
                                                                                                                       ifelse(variable %in% ms_shortC, "short C chains", 
                                                                                                                       ifelse(variable %in% Tryp_ind, "Tryp. & Indole",  "Other"))
                                                                                                                       ))))))))))))
                                          ) 
                           )

table(metab_annotation$class) # none in 'other' group now, great
# NOTE: p.Cresol is produced from fermentation of tyrosine

## want consistent order of factor levels
levels_metab <- c("Main SCFAs", "other SCFAs & related", "Acids (non-SCFA)", "Tryp. & Indole", "Lipids & related", "CHO",  "Nitrogenous", "N wt Methyl/Acetyl", "Solvents/Ferment-products", "Ceramides", "CEs", "LYSOC", "PCs", "SMs", "short C chains")
metab_annotation$class <- factor(metab_annotation$class, levels = levels_metab)
```

## Quick data checks and cleaning/merging
```{r}
# no na values
sum(is.na(metab_ms))
sum(is.na(metab_nmr))

# sample overlap b/t datasets
nrow(metab_ms) # 1971 samples analzyed
nrow(metab_nmr) # 1709 samples analzyed
length(intersect(row.names(metab_ms), row.names(metab_nmr))) ## 1704, most samples analyzed in both

df_metabs <- merge(metab_ms, metab_nmr, by = "row.names", all = TRUE)
dim(df_metabs)

# Also merge w/t sample data

## get df ready to merge
metab_sdf_1y$CollectionShort.Date <- NULL
metab_sdf_3m$CollectionShortDate <- NULL
metab_sdf_1y$CollectioDate <- NULL
metab_sdf_3m$CollectionDate <- NULL
metab_sdf_1y$Visit <- "1 Year"
metab_sdf_3m$Visit <- "3 Months"

## make Row.names/sample id's match format of other df
df_metabs$Row.names <- df_metabs$Row.names %>% gsub("-", ".", .) %>% gsub("\\.\\d\\.\\d$", "", .)

df_metabs_sdf <- rbind(metab_sdf_1y, metab_sdf_3m) %>% mutate(BarCode = gsub("-", ".", BarCode)) %>% 
                    mutate(BarCode = gsub("\\.\\d\\.\\d$", "", BarCode)) %>% 
                    merge(., df_metabs, by.x = "BarCode", by.y = "Row.names")
nrow(df_metabs_sdf)

sum(duplicated(df_metabs_sdf$BarCode))

table(df_metabs_sdf$Visit)

# Also merge w/t all sample data - sdf
df_metabs_sdf_2 <- merge(df_metabs_sdf, sdf, by.x = "SubjectNumber", by.y = "CHILDid")
sum(duplicated(df_metabs_sdf_2$BarCode))
nrow(df_metabs_sdf_2) == nrow(df_metabs_sdf) # true good

table(df_metabs_sdf_2$Visit)
```

# Adding sample filters to be consistent w/t other analyses
Remove infants with no or inconsistent BF status data, and age at sample collection outliers, never breastfed (in that order to check n removed after each step by Visit)
```{r} 
sum(is.na(df_metabs_sdf_2$BF_status_3m_sampling2)) # rm 29 never BF
table(df_metabs_sdf_2$BF_status_3m_sampling2)

# doing first just to track numbers easier, after this filter how many rm by other filters?
df_metabs_sdf_2.1 <- subset(df_metabs_sdf_2, !is.na(BF_status_3m_sampling2)) 
table(df_metabs_sdf_2.1$Visit)

# Exclude Age at sample collection outliers
df_metabs_sdf_3 <- subset(df_metabs_sdf_2.1, (age_3m_months < 8 & age_3m_months > 1 & Visit == "3 Months") | (age_1y_months < 21 & age_1y_months > 8  & Visit == "1 Year")) 
nrow(df_metabs_sdf_3) #  remain - 
nrow(df_metabs_sdf_2.1) - nrow(df_metabs_sdf_3) # 13 samples removed
table(df_metabs_sdf_3$Visit)

df_metabs_sdf_4 <- subset(df_metabs_sdf_3, BF_status_3m_sampling2 != "Never breastfed") 
nrow(df_metabs_sdf_4) # was 1892 remain. Now 1879 removed 
nrow(df_metabs_sdf_3) - nrow(df_metabs_sdf_4) # 55 samples removed
table(df_metabs_sdf_4$Visit)

# New input df for data subset creation 
df_metabs_input <- df_metabs_sdf_4

### checking n's for specific analyses - scfa/nmr
table(df_metabs_input[!is.na(df_metabs_input$Propionate),]$Visit)

table(df_metabs_input[!is.na(df_metabs_input$Propionate),]$Visit, df_metabs_input[!is.na(df_metabs_input$Propionate),]$BF_status_3m_sampling2)
```

## Dataset types - split by comparison (slightly different set of individuals per comparison)
```{r}
vars_toKeep <- c("SubjectNumber", "Visit", "BF_status_3m_sampling", "Solids_3m_sampling", "asthma_3y") 
## only keeping selected vars from sdf to limit size of dfs!

## For plots & summary tables
df_metabs_sdf_m <- melt(df_metabs_input, measure.vars = c(ms_metabs, nmr_metabs), 
                        id.vars = vars_toKeep) %>%
                        merge(metab_annotation, ., by = "variable") ## adding metabolite annotations to melted df version


# Wide format Data Subsets For Tests, combine into list
### Split - separate tests for BF and Asthma 
df_metabs_sdf_BF <- subset(df_metabs_input, BF_status_3m_sampling != "Partial", 
                           select = c(vars_toKeep, ms_metabs, nmr_metabs))
df_metabs_sdf_BF$Subset <- "Breastfeeding"

## based on solids check, doing solids sensitivity analysis right away too
df_metabs_sdf_BF_noS <- subset(df_metabs_input, BF_status_3m_sampling != "Partial" & Solids_3m_sampling %in% c("No Solids", NA), 
                           select = c(vars_toKeep, ms_metabs, nmr_metabs)) ## just exluding those w/t known solids
df_metabs_sdf_BF_noS$Subset <- "Breastfeeding, no solids"

df_metabs_sdf_Ast <- subset(df_metabs_input, asthma_3y != "Possible", 
                           select = c(vars_toKeep, ms_metabs, nmr_metabs))
df_metabs_sdf_Ast$Subset <- "Asthma"


df_metabs_sdf_all <- rbind(df_metabs_sdf_BF, df_metabs_sdf_BF_noS, df_metabs_sdf_Ast)

## Dataset list for downstream test input = df_ls
df_ls <- split(df_metabs_sdf_all, interaction(df_metabs_sdf_all$Visit, df_metabs_sdf_all$Subset)) 
## main overall stratification should go first - Visit - strata_var1 for test later
names(df_ls)
```

## Data summaries (using long df format but same data as df_ls)
```{r}
metab_summaries_Asthma <- data_summary_V2(subset(df_metabs_sdf_m, asthma_3y != "Possible"), varname = "value", groupnames = c("variable", "class", "method", "Visit", "asthma_3y"))

metab_summaries_BF <- data_summary_V2(subset(df_metabs_sdf_m, BF_status_3m_sampling != "Partial"), varname = "value", groupnames = c("variable", "class", "method", "Visit", "BF_status_3m_sampling"))

# overall by visit - extra not saved
metab_summaries <- data_summary_V2(df_metabs_sdf_m, varname = "value", groupnames = c("variable", "class", "method", "Visit"))

# Extra - done for sensitivity analysis 
#metab_summaries_BF_noSolid <- data_summary_V2(subset(df_metabs_sdf_m, BF_status_3m_sampling != "Partial" & Solids_3m_sampling %in% c("No Solids", NA)), 
#                                   varname = "value", groupnames = c("variable", "class", "method", "Visit", "BF_status_3m_sampling"))

write.csv(metab_summaries_Asthma, paste0(path, "Tables/Res_Fig5C_Metab_Asthma_Summary.csv"))
write.csv(metab_summaries_BF, paste0(path, "Tables/Res_Fig5C_Metab_BF_Summary.csv"))
```

## Per-timepoint univariate tests 
```{r}
# checks
length(c("asthma_3y", "asthma_3y", rep("BF_status_3m_sampling", 4))) == length(df_ls) # good

# check - for each df, testing the right factor variable? Note that added BF comparisons are for a sensitivity analysis
names(df_ls)
c("asthma_3y", "asthma_3y", rep("BF_status_3m_sampling", 4))

# tests
test_df <- wilcox.unpaired.loops(df_ls = df_ls, 
                                 vars_names = metab_annotation$variable, # same continuous variables to test across each df/comparison
                                 factor_vars = c("asthma_3y", "asthma_3y", rep("BF_status_3m_sampling", 4)) # cat vars to test for each df 
                                 )

write.csv(test_df, file = paste0(path, "Tables/Res_Fig5CD_Wilcoxon_perTP_metabolites_Asthma_BF_withSensitivity.csv"))

## vars sig for anything
table(test_df$Subset, test_df$p.cat) 

# Sig table - use to annotate significance on plots later
length(unique(subset(test_df, p.cat != "")$variable)) # n to plot 
```

```{r}
# Edits for main plots in paper (for specific timepoint plots - decided on 3 month only, only specific metabolites, etc.)
## Plotting only some metabs: "Main SCFAs" , Other metabolites to plot:  Tryptophan, Tryptamine, Indole
plot_metabs <- c(metab_annotation[metab_annotation$class == "Main SCFAs",]$variable, "Tryptophan", "Tryptamine", "Indole")

## Decided to only move forward with main BF comparison at 3 months for metabolites 

### main long df subset
### want bf only apparently - also make matching main df to plot (for individual 3 month timepoint)
df_plots_sub_bf <- subset(df_metabs_sdf_m, BF_status_3m_sampling != "Partial" & Visit == "3 Months" & 
                          variable %in% plot_metabs[1:4]) 

### test res subset
test_df_sub_bf_0 <- subset(test_df, Subset == "3 Months.Breastfeeding" & 
                          variable %in% plot_metabs[1:4])

## summarized across all subsets - just for p-val positions in plots - NOT to be used for actual dta summaries
### just doing 3 month BF now, so just summarize that subset
summary_forplot <- data_summary_V2(df_plots_sub_bf, 
                                   varname = "value", groupnames = c("variable")) %>%
                      subset(., select = c("max", "variable"))

colnames(summary_forplot)[colnames(summary_forplot) == "max"] <- "y.position" ## relabel the summary stat want to use for pval pos.
summary_forplot$y.position <- ifelse(summary_forplot$y.position > 0,
                                     summary_forplot$y.position*0.8,
                                     summary_forplot$y.position*1.2) 
### depending if neg or pos, fraction used to get p-val position right will differ. 0 - 1*0.5

# add p-val annotation y.position to df for plotting
test_df_sub_bf <- merge(test_df_sub_bf_0, summary_forplot, by = "variable") 
nrow(test_df_sub_bf) == nrow(test_df_sub_bf_0) # great
```

## Generate paired datasets & organize in list
To compare 3M to 1Y values of same participant - stratified by BF and Asthma
- includes sensitivity analysis - exclude solids
```{r}
## One dataset per - asthma, BF
paired_asthma <- merge(df_ls$`3 Months.Asthma`, 
                       subset(df_ls$`1 Year.Asthma`, select = c("SubjectNumber", ms_metabs, nmr_metabs)), 
                       by = "SubjectNumber", suffixes = c("_3m", "_1y")) ## Just adding extra subset to make data cleaner looking
row.names(paired_asthma) <- paired_asthma$SubjectNumber ## good to keep rownames as subject ids to keep track of in traj dataset output later
dim(paired_asthma) # 467 samples matched

paired_bf <- merge(df_ls$`3 Months.Breastfeeding`, 
                   subset(df_ls$`1 Year.Breastfeeding`, select = c("SubjectNumber", ms_metabs, nmr_metabs)),
                   by = "SubjectNumber", suffixes = c("_3m", "_1y"))
row.names(paired_bf) <- paired_bf$SubjectNumber 


paired_bf_no_s <- merge(df_ls$`3 Months.Breastfeeding, no solids`, 
                        subset(df_ls$`1 Year.Breastfeeding, no solids`, select = c("SubjectNumber", ms_metabs, nmr_metabs)),
                        by = "SubjectNumber", suffixes = c("_3m", "_1y"))
row.names(paired_bf_no_s) <- paired_bf_no_s$SubjectNumber 


traj_df_ls <- list(paired_bf, paired_asthma, paired_bf_no_s)
names(traj_df_ls) <- c("BF", "Asthma", "NoSolids_BF")


# Stratified version for Paired wilcoxon tests comparing b/t timepoints
paired_df_ls <- c(split(paired_bf, paired_bf$BF_status_3m_sampling), split(paired_asthma, paired_asthma$asthma_3y), split(paired_bf_no_s, paired_bf_no_s$BF_status_3m_sampling))
names(paired_df_ls) <- paste(c("BF", "BF", "Asthma", "Asthma", "NoSolid_BF", "NoSolid_BF"), names(paired_df_ls), sep = "_") %>% gsub(" ", "_", .)
names(paired_df_ls) 
```

## Paired Wilcoxon tests - stratified by Asthma and breastfeeding; and trajectory analysis (using derived variable - change in concentration between 1 year and 3 months)
```{r}
test_df <- multi.wilcox.pair.loops(paired_df_ls = paired_df_ls,
                                  ## All varnames must have matching order, based on tests to do
                                   var_names_x = gsub("$", "_3m", c(ms_metabs, nmr_metabs)),
                                   var_names_y = gsub("$", "_1y", c(ms_metabs, nmr_metabs)),
                                   var_names = c(ms_metabs, nmr_metabs) ## extra, varnames matching x and y varnames but without the suffix
                                   )

# Extra - merge p-vals with Metabolite annotations: metab_annotation 
test_df_paired_strata <- merge(test_df_paired, metab_annotation, by = "variable")

### quick additional sanity check of sample sizes...(shouldn't really need to do this)
nrow(paired_df_ls$BF_Exclusive[!is.na(paired_df_ls$BF_Exclusive$Butyrate_1y) & !is.na(paired_df_ls$BF_Exclusive$Butyrate_3m),])
nrow(paired_df_ls$BF_Exclusive[!is.na(paired_df_ls$BF_No_BM$Butyrate_1y) & !is.na(paired_df_ls$BF_No_BM$Butyrate_3m),])
```

### Prep data & plot - Fig. 5C and 5D 
```{r, fig.height=7}
# datasets for plotting static timepoint res: plot_df_ls_4, test_df_sig_summ_sp_3
var_names_x <- gsub("$", "_3m", c(ms_metabs, nmr_metabs))
var_names_y <- gsub("$", "_1y", c(ms_metabs, nmr_metabs))

# for plotting stratified 3m vs 1y res: paired_df_ls, test_df (3m vs 1y p-vals )
# Just melting actually, will maybe re-split for plotting subsets separately later
plot_paired_df <- bind_rows(paired_df_ls, .id = "Subset") %>% melt(., measure.vars = c(var_names_x, var_names_y)) %>%
                      mutate(Visit = ifelse(grepl("_3m$", variable), "3 Months", "1 Year"), ## correct visit var for melted df
                             Metabolite = gsub("_3m$|_1y$", "", variable))  # correct variable
                    
## relabel the test var for each df - call it 'test_var'? Not needed for this because the comparison is always 'Visit'

# need a group1, group2 and y.position column in p-val df for this package - pubr
test_df$group1 <- unique(plot_paired_df$Visit)[1]
test_df$group2 <- unique(plot_paired_df$Visit)[2]
colnames(test_df)[colnames(test_df) == "variable"] <- "Metabolite" ## also relabel var id to match plotting df

# y.pos based on upper lim of data across subsets - just use for pval position NOT actual data summaries
summary_paired_df <- data_summary_V2(plot_paired_df, varname = "value", groupnames = c("Metabolite"))
summary_paired_sub <- subset(summary_paired_df, select = c("max", "Metabolite"))
colnames(summary_paired_sub)[colnames(summary_paired_sub) == "max"] <- "y.position" ## relabel the summary stat want to use for pval pos.
summary_paired_sub$y.position <- ifelse(summary_paired_sub$y.position > 0,
                                        summary_paired_sub$y.position*0.8,
                                        summary_paired_sub$y.position*1.2) 
# depending if neg or pos, fraction used to get p-val position right will differ. 0 - 1*0.5

test_df_2 <- merge(test_df, summary_paired_sub, by = c("Metabolite"))
nrow(test_df_2) == nrow(test_df) # great


table(plot_paired_df$Subset) ## also don't wana plot all subsets - exclude NoSolid_BF_Exclusive and NoSolid_BF_No_BM for main plot
plot_paired_sub <- subset(plot_paired_df, Metabolite %in% plot_metabs & Subset %ni% c("NoSolid_BF_Exclusive", "NoSolid_BF_No_BM"))
test_df_2_sub <- subset(test_df_2, Metabolite %in% plot_metabs & Subset %ni% c("NoSolid_BF_Exclusive", "NoSolid_BF_No_BM"))

table(plot_paired_sub$Subset)
## factor level order 
plot_paired_sub$Subset <- factor(plot_paired_sub$Subset, levels = c("BF_Exclusive" , "BF_No_BM", "Asthma_No", "Asthma_Yes"))
test_df_2_sub$Subset <- factor(test_df_2_sub$Subset, levels = c("BF_Exclusive" , "BF_No_BM", "Asthma_No", "Asthma_Yes"))


# decided to relabel levels: Exclusive BM; No BM & No Asthma; Asthma
plot_paired_sub$Subset <- revalue(plot_paired_sub$Subset, 
                                  c("BF_Exclusive" = "Exclusive BM", "BF_No_BM" = "No BM",
                                    "Asthma_No" = "No Asthma", "Asthma_Yes" = "Asthma"))
test_df_2_sub$Subset <- revalue(test_df_2_sub$Subset, 
                                  c("BF_Exclusive" = "Exclusive BM", "BF_No_BM" = "No BM",
                                    "Asthma_No" = "No Asthma", "Asthma_Yes" = "Asthma"))

plot_paired_sub$Visit <- factor(plot_paired_sub$Visit, levels = c("3 Months", "1 Year"))


vis_cols <- c("#f5877f", "#b02c23", "#31d462", "#017524", "#84aef5", "#3a609e", "#daa8ff", "#8552ab")

# Just so scales are easier to control - want diff scale for each metabolite 
### Decided - not to facet by metabolite but split by metab instead
plot_paired_sub_sp <- split(plot_paired_sub, plot_paired_sub$Metabolite)
test_df_2_sub_sp <- split(test_df_2_sub, test_df_2_sub$Metabolite)
identical(names(test_df_2_sub_sp), names(plot_paired_sub_sp)) # good, match order

## Also - want vis_cols aligned so that colors match - make list w/t matched order
# Originally: Red = Acetic Acid, Green = Butyrate, Blue = Prop., Purple = Valerate
names(plot_paired_sub_sp)
vis_cols_ls <- list(vis_cols[1:2], vis_cols[3:4], c("#f5ea51", "#807d57"),
                    vis_cols[5:6], c("#f5ab51", "#b87525"), c("#f5ab51", "#b87525"), 
                    vis_cols[7:8])
length(vis_cols_ls) == length(plot_paired_sub_sp)

test_df_2_sub_sp$Tryptamine$y.position <- -1 # not showing up b/c negative so not the same thing
test_df_2_sub_sp$Indole$y.position <- -3.5 # not showing up b/c negative so not the same thing
test_df_2_sub_sp$Tryptophan$y.position <- 1.3 # not showing up b/c negative so not the same thing

vis_plots <- list()
for(i in seq_along(plot_paired_sub_sp)) {
vis_plots[[i]] <- ggplot(plot_paired_sub_sp[[i]], aes(x = Visit, y = value)) + 
  facet_wrap(~Subset, ncol = 4) + 
  geom_violin(alpha=0.3) +
  geom_jitter(aes(color=Visit), alpha=0.70, size=1, width = 0.2) +
  stat_summary(fun=mean, geom="point", shape=23, size=1, fill="black") +
  scale_y_continuous(expand=c(0,0)) + 
  labs(x = "", y = "Normalized concentration") +
  theme(legend.position = "none", title = element_text(size = 10.5), axis.text.x = element_text(size = 11)) +
  scale_color_manual(values = vis_cols_ls[[i]]) +
  stat_pvalue_manual(test_df_2_sub_sp[[i]], label = "p.cat",  bracket.size = 0, color="black", size = 8) +
  ggtitle(names(plot_paired_sub_sp)[i])
}

# just selecting specific tryptophan related metabolites to plot like this - plotting SCFA metabolites differently (see below)
tiff(file = paste0(path, "Figures/Fig_5C_TryptophanMetabs_BFAsthma.tiff"), width = 1700, height = 2200, units = "px", res = 300)
plot_grid(vis_plots[[6]], vis_plots[[5]], vis_plots[[3]], ncol = 1)
dev.off()


# bf only 
plot_paired_sub_bf <- subset(plot_paired_sub, Subset %in% c("Exclusive BM", "No BM"))
test_df_2_sub_bf <- subset(test_df_2_sub, Subset %in% c("Exclusive BM", "No BM"))

# add static timepoint version to 3m vs 1 year comparison - just to get dimentions of fig right
# df_plots_sub_bf, test_df_sub_bf

test_df_sub_bf$Subset <- "3 Months"
names(test_df_sub_bf)[names(test_df_sub_bf) == "variable"] <- "Metabolite"

test_df_sub_bf_2 <- rbind(test_df_sub_bf, test_df_2_sub_bf)
test_df_sub_bf_2_sp <- split(test_df_sub_bf_2, test_df_sub_bf_2$Metabolite)


df_plots_sub_bf$Subset <- "3 Months"
names(df_plots_sub_bf)[names(df_plots_sub_bf) == "variable"] <- "Metabolite"

names(df_plots_sub_bf)[names(df_plots_sub_bf) == "BF_status_3m_sampling"] <- "test_var"
names(plot_paired_sub_bf)[names(plot_paired_sub_bf) == "Visit"] <- "test_var"

plot_paired_sub_2 <- rbind(select(plot_paired_sub_bf, test_var, Subset, value, Metabolite),
                           select(df_plots_sub_bf, test_var, Subset, value, Metabolite)
                           )

plot_paired_sub_2_sp <- split(plot_paired_sub_2, plot_paired_sub_2$Metabolite)

vis_cols_ls_2 <- list(c(vis_cols[1:2], "#f5877f", "#f5877f"), c(vis_cols[3:4], "#31d462", "#31d462"), 
                      c("#f5ea51", "#807d57", "#f5ea51", "#f5ea51"),
                    c(vis_cols[5:6], "#84aef5", "#84aef5"), c("#f5ab51", "#b87525", "#f5ab51", "#f5ab51"), 
                    c("#f5ab51", "#b87525", "#f5ab51", "#f5ab51"), 
                    c(vis_cols[7:8], "#daa8ff", "#daa8ff"))

test_df_sub_bf_2_sp$Valerate$y.position <- 1.8
test_df_sub_bf_2_sp$Butyrate$y.position <- 4.312

vis_plots_bf <- list()
for(i in seq_along(plot_paired_sub_2_sp)) {
vis_plots_bf[[i]] <- ggplot(plot_paired_sub_2_sp[[i]], aes(x = test_var, y = value)) + 
  facet_wrap(~Subset, scales = "free_x", ncol = 4) + 
  geom_violin(alpha=0.3) +
  geom_jitter(aes(color=test_var), alpha=0.70, size=1, width = 0.2) +
  stat_summary(fun=mean, geom="point", shape=23, size=1, fill="black") +
  scale_y_continuous(expand=c(0,0)) + 
  labs(x = "", y = "Normalized concentration") +
  theme(legend.position = "none", title = element_text(size = 10.5), axis.text.x = element_text(size = 11.5)) +
  scale_color_manual(values = vis_cols_ls_2[[i]]) +
  stat_pvalue_manual(test_df_sub_bf_2_sp[[i]], label = "p.cat",  bracket.size = 0, color="black", size = 8) +
  ggtitle(names(plot_paired_sub_2_sp)[i])
}

tiff(file = paste0(path, "Figures/Fig_5D_SCFA_Metabs_BFAsthma.tiff"), width = 1600, height = 2700, units = "px", res = 300)
plot_grid(vis_plots_bf[[1]], vis_plots_bf[[2]], vis_plots_bf[[4]], vis_plots_bf[[7]], ncol = 1)
dev.off()

break_seq <- seq(4, 7) ## need to do hack for acetic acid axis to get sizing right

tiff(file = paste0(path, "Figures/Fig_5D_SCFA_Metabs_BFAsthma.tiff"), width = 3500, height = 1450, units = "px", res = 300)
plot_grid(vis_plots_bf[[2]], vis_plots_bf[[7]],
          vis_plots_bf[[4]], 
          vis_plots_bf[[1]] + scale_y_continuous(breaks = break_seq, labels = paste0("     ", break_seq)), ncol = 2)
dev.off()
```
