---
title: "Figure 4. Timing of acquiring microbial functions in the first year of life is associated with breastfeeding and asthma."
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```
 
# Figure 4 - PreTCoA on stratified metagenomic functional data - using EC annotations 
EC and EggNog analyzed simultaneously here, but EggNog specific plots moved to Figure S7 (also see Fig_S7.Rmd, scripts are the same)

```{r}
## Setup 
R.version.string
library(phyloseq)
library(ggplot2)
library(tidyr)
library(reshape2)
library(dplyr)
library(cowplot)
library(ggpubr)
library(stringr)
library(here)

source(paste0(here::here(), "/Functions.R"))

theme_set(theme_bw() + theme(plot.title = element_text(hjust = 0.5,face = "bold"), strip.text.x = element_text(size = 11), strip.background = element_rect(colour="white", fill="white"), panel.grid.major = element_blank(), panel.grid.minor = element_blank(), text = element_text(size=12), axis.text = element_text(color="black", size=12), legend.key.size = unit(0.5,"cm")))
```

## Read in Metagenomic data 
```{r}
# Compiled stratified functional data
funcs_3M1Y_bin_s <- readRDS(file = paste0(path, "Data/Metagenome_metabolome/final_datasets/MainTaxaFunctionsBinaryTraj_sdf_above10EitherVis_v02.rds"))

names_bestPrevalence_s <- readRDS(file = paste0(path, "Data/Metagenome_metabolome/final_datasets/Ls_MainTaxaFuncs_best_Binary_above10_below90EitherVis_v02.rds"))
## some issues with number at beginning of strings
names_bestPrevalence_s$EC <- gsub("^", "X", names_bestPrevalence_s$EC) 
names_bestPrevalence_s$MetaCyc <- gsub("^\\d", "X1", names_bestPrevalence_s$MetaCyc)


# getting sample data
# taxonomy abudnances maybe used
Gut_3M1Y_all_filt <- readRDS(file = paste0(path, "Data/Metagenome_metabolome/final_datasets/03_Abund_metagenomic_Trajectory.rds"))
sdf <- sample_data(Gut_3M1Y_all_filt)
```

# Prevalence Trajectory Coordinate System & Permutation tests - Comparing Prevalence Trajectories Between Groups 
funcs_3M1Y_bin_s
## Gut - Prevalence Trajectory derived variables, creation of a prevalence trajectory feature table
To simplify making of prevalence trajectory scatter plots, first make the 'difference in prevalence' results for every pair of samples for each infant (1Y - 3M, e.g. 0 - 1 = -1; 1 - 0 = 1) - so a "-1" means it's present at 3 months but not 1 year, 1 means 1 year but not 3m and 0 means no change.
```{r} 
# Make dataset for Change in Prevalence from 3M to 1y (always TP2 - TP1) - Also append sdf to the end again - but NOTE that this is per-participant (just merging based on 1 year sampleid row.names but sample (rather than participant)-specific metadata is best to ignore)

## proof of subj # in same order in Visit 1 and 3 split - B/C df was ordered by subjectnumber first - below code shouldn't be used unless all are TRUE
### table(subset(funcs_3M1Y_bin_s$EggNog, Visit == "1 year")$subjectnumber == subset(funcs_3M1Y_bin_s$EggNog, Visit == "3 month")$subjectnumber)

# merge difference b/t timepoints - 1 year minus 3 months for same df - with sample data
funcs_3M1Y_bin_diff_s <- list()
for(i in seq(funcs_3M1Y_bin_s)) { 
  funcs_3M1Y_bin_diff_s[[i]] <- merge(subset(funcs_3M1Y_bin_s[[i]],
                                             Visit == "1 year", select =  
                                               intersect(names_bestPrevalence_s[[i]], colnames(funcs_3M1Y_bin_s[[i]]))
                                             ) - subset(funcs_3M1Y_bin_s[[i]], 
                                                        Visit == "3 month", 
                                                        select =  intersect(names_bestPrevalence_s[[i]], colnames(funcs_3M1Y_bin_s[[i]]))
                                                        ), 
                                      sdf, by = "row.names")
  funcs_3M1Y_bin_diff_s[[i]]$Row.names <- NULL
}
names(funcs_3M1Y_bin_diff_s) <- names(funcs_3M1Y_bin_s)
lapply(funcs_3M1Y_bin_diff_s, function(x) ncol(x) - ncol(sdf)) ## good, adds up


N_vars <- lapply(funcs_3M1Y_bin_diff_s, function(x) ncol(x) - ncol(sdf)) # number of taxa or vars, to use in downstream 
N_subj <- lapply(funcs_3M1Y_bin_diff_s, nrow) # number of subjects, to use in downstream 
```

## Asthma at 3 years
These scripts should also work if there are variable subject N per dataset
```{r} 
## input (could also do a loop across list of different variables groups to stratify by) - variable[[i]], Group_X[[i]], Group_Y[[i]], or mapply
data <- funcs_3M1Y_bin_diff_s # binary trajecotires
data_0 <-funcs_3M1Y_bin_s # original binary dataset
variable <- "asthma_3y"
Group_X <- "Yes" ## exactly as specified in variable
Group_Y <- "No" ## exactly as specified in variable
N <- N_vars # number of taxa in df, changing to N for generic "source" code below
```

```{r} 
# script
X <- list()
Y <- list()
df.pa <- list()
for(i in seq_along(data)) {
  X[[i]] <- data[[i]][data[[i]][[variable]] == Group_X,] %>% .[!is.na(.[[variable]]),]
  Y[[i]] <- data[[i]][data[[i]][[variable]] == Group_Y,] %>% .[!is.na(.[[variable]]),]
  df.pa[[i]] <- data.frame(X_P_Change=colSums(X[[i]][1:N[[i]]])*100/nrow(X[[i]]),
                           Y_P_Change=colSums(Y[[i]][1:N[[i]]])*100/nrow(Y[[i]]),
                           prevalence=colSums(data_0[[i]][1:N[[i]]])*100/(N_subj[[i]]*2)
                           ) %>% ## Can add taxa/variable annotations to data.frame as well. Excluding this for now for functions. Might add Taxa origin of function as annotation yet.
                 mutate(pa.change.diff = Y_P_Change - X_P_Change, vars = row.names(.)) %>% separate(vars, c("Function", "Taxa"), sep = "_")
}
names(df.pa) <- names(data)
  
# check sample sizes
lapply(X, nrow)
lapply(Y, nrow)
# check n variables
lapply(df.pa, nrow)
```

```{r} 
table(df.pa$EggNog$Taxa)

# Aesthetic things, not important -  make sure colors of taxa are always consistent no matter the number of taxa plotted 
cols <- c("#58b6ed", "#6aae9c","#9eae6a", "#b55174", "#6a7cae", 
          "#ae6a7c", "#ae976a", "#142647")

names(cols) <- unique(df.pa$EggNog$Taxa)
```

Plot Change in Prevalence 
```{r} 
# plot components
x_lab <- list()
y_lab <- list()
min_line <- list()
max_line <- list()

# plot
PreTCoA <- list()

for(i in seq_along(df.pa)) {
  
x_lab[[i]] <- paste(Group_X, " ", variable, " (n=", nrow(X[[i]]), ")\n1 Year - 3 Month Prevalence (%)", sep="")
y_lab[[i]] <- paste(Group_Y, " ", variable, " (n=", nrow(Y[[i]]), ")\n1 Year - 3 Month Prevalence (%)", sep="")

min_line[[i]] <- min(c(min(df.pa[[i]][["X_P_Change"]]), min(min(df.pa[[i]][["Y_P_Change"]]))))
max_line[[i]] <- max(c(max(df.pa[[i]][["X_P_Change"]]), max(max(df.pa[[i]][["Y_P_Change"]]))))

PreTCoA[[i]] <- ggplot() + 
  geom_point(data=df.pa[[i]], aes(x=X_P_Change, y=Y_P_Change, color = Taxa, size=prevalence), alpha=0.70) + 
  xlab(x_lab[[i]]) + ylab(y_lab[[i]]) + 
  scale_y_continuous(limits=c(min_line[[i]], max_line[[i]]), expand = c(0,0))+
  scale_x_continuous(limits=c(min_line[[i]], max_line[[i]]), expand = c(0,0))+
  annotate("segment", x = min_line[[i]], xend = max_line[[i]], y = min_line[[i]], yend = max_line[[i]], colour = "red", linewidth=0.25) + 
  ggtitle(names(df.pa)[[i]]) + scale_size(range = c(1,5)) + scale_color_manual(values = cols)
PreTCoA[[i]]
  
}

# saving data
df.pa_original_asthma <- df.pa ## saving original list for asthma first to not lose it
PreTCoA_asthma <- PreTCoA

PreTCoA_asthma
```


Median change in prevalence by asthma - overall
```{r} 
# Asthma
lapply(df.pa_original_asthma, function(x) round(summary(x[["X_P_Change"]]), 1))
```

```{r} 
# No Asthma
lapply(df.pa_original_asthma, function(x) round(summary(x[["Y_P_Change"]]), 1))
```

Median change in prevalence by asthma - stratified by Species
```{r} 
# Asthma
lapply(df.pa_original_asthma, function(x) data_summary_V2(x, varname = "X_P_Change", groupnames = "Taxa"))
```

```{r} 
# No Asthma
lapply(df.pa_original_asthma, function(x) data_summary_V2(x, varname = "Y_P_Change", groupnames = "Taxa"))
```

```{r} 
# diff in change
lapply(df.pa_original_asthma, function(x) data_summary_V2(x, varname = "pa.change.diff", groupnames = "Taxa"))
```

### Permutation test for Prevalence Trajectory analysis 

**Just assessing for EC - main results**
```{r} 
Res_ec_asthma <- perm_func_dfs_v2(X_df = df.pa_original_asthma$EC, 
                                  data = funcs_3M1Y_bin_diff_s$EC,
                                  variable = "asthma_3y", Group_X = "Yes", Group_Y = "No",
                                  N = N_vars$EC)

# global signfiicance?
original_wilcox_ec_asthma <- Res_ec_asthma$global_res
original_wilcox_ec_asthma

## Res_per_taxa  p-value assigned to each taxa; p_value = later in healthy; p_value_2 = earlier in healthy
Res_per_taxa_ec_asthma <- Res_ec_asthma$per_taxa_res
```

```{r}
## p_value threshold decision
p_thresh2 <- 0.01

# sig early
Res_per_taxa_ec_asthma_early <- subset(Res_per_taxa_ec_asthma, p_value_2 < p_thresh2)
Res_per_taxa_ec_asthma_early

# sig late
Res_per_taxa_ec_asthma_late2 <- subset(Res_per_taxa_ec_asthma, p_value < p_thresh2)
Res_per_taxa_ec_asthma_late2
```

## Breastfeeding 
These scripts should also work if there are variable subject N per dataset
```{r} 
## input (could also do a loop across list of different variables groups to stratify by) - variable[[i]], Group_X[[i]], Group_Y[[i]], or mapply
data <- funcs_3M1Y_bin_diff_s # binary trajecotires
data_0 <-funcs_3M1Y_bin_s # original binary dataset
variable <- "BF_status_3m_sampling"
Group_X <- "No BM" ## exactly as specified in variable
Group_Y <- "Exclusive" ## exactly as specified in variable
N <- N_vars # number of taxa in df, changing to N for generic "source" code below
```

```{r} 
# script
X <- list()
Y <- list()
df.pa <- list()
for(i in seq_along(data)) {
  X[[i]] <- data[[i]][data[[i]][[variable]] == Group_X,] %>% .[!is.na(.[[variable]]),]
  Y[[i]] <- data[[i]][data[[i]][[variable]] == Group_Y,] %>% .[!is.na(.[[variable]]),]
  df.pa[[i]] <- data.frame(X_P_Change=colSums(X[[i]][1:N[[i]]])*100/nrow(X[[i]]),
                           Y_P_Change=colSums(Y[[i]][1:N[[i]]])*100/nrow(Y[[i]]),
                           prevalence=colSums(data_0[[i]][1:N[[i]]])*100/(N_subj[[i]]*2)
                           ) %>% ## Can add taxa/variable annotations to data.frame as well. Excluding this for now for functions. Might add Taxa origin of function as annotation yet.
                 mutate(pa.change.diff = Y_P_Change - X_P_Change, vars = row.names(.)) %>% 
                 separate(vars, c("Function", "Taxa"), sep = "_") 
}
names(df.pa) <- names(data)
  
# check sample sizes
lapply(X, nrow)
lapply(Y, nrow)
# check n variables
lapply(df.pa, nrow)
```

Using same cols vector for colours as asthma plots

Plot Change in Prevalence 
```{r} 
# plot components
x_lab <- list()
y_lab <- list()
min_line <- list()
max_line <- list()

# plot
PreTCoA <- list()

for(i in seq_along(df.pa)) {
  
x_lab[[i]] <- paste(Group_X, " ", variable, " (n=", nrow(X[[i]]), ")\n1 Year - 3 Month Prevalence (%)", sep="")
y_lab[[i]] <- paste(Group_Y, " ", variable, " (n=", nrow(Y[[i]]), ")\n1 Year - 3 Month Prevalence (%)", sep="")

min_line[[i]] <- min(c(min(df.pa[[i]][["X_P_Change"]]), min(min(df.pa[[i]][["Y_P_Change"]]))))
max_line[[i]] <- max(c(max(df.pa[[i]][["X_P_Change"]]), max(max(df.pa[[i]][["Y_P_Change"]]))))

PreTCoA[[i]] <- ggplot() + 
  geom_point(data=df.pa[[i]], aes(x=X_P_Change, y=Y_P_Change, size=prevalence, color = Taxa), alpha=0.75) + 
  xlab(x_lab[[i]]) + ylab(y_lab[[i]]) + 
  scale_color_manual(values = cbPalette) +
  scale_y_continuous(limits=c(min_line[[i]], max_line[[i]]), expand = c(0,0))+
  scale_x_continuous(limits=c(min_line[[i]], max_line[[i]]), expand = c(0,0))+
  annotate("segment", x = min_line[[i]], xend = max_line[[i]], y = min_line[[i]], yend = max_line[[i]], colour = "red", linewidth=0.25) + 
  ggtitle(names(df.pa)[[i]]) + scale_size(range = c(1,5)) + scale_color_manual(values = cols)
PreTCoA[[i]]
  
}

# saving data
PreTCoA_bf <- PreTCoA
df.pa_original.BF <- df.pa ## saving original list for asthma first to not lose it

PreTCoA_bf
```
Probably the highly prevalent B. Longum functions in BF infants are from B. infantis - did a check of B. Longum and only B. infantis was associated with breastfeeding. 

**Median change in prevalence by BF**
```{r} 
# No BF
lapply(df.pa_original.BF, function(x) round(summary(x[["X_P_Change"]]), 1))
```

```{r} 
# BF
lapply(df.pa_original.BF, function(x) round(summary(x[["Y_P_Change"]]), 1))
```

**Median change in prevalence by BF - stratified by Species**
```{r} 
# No BM
lapply(df.pa_original.BF, function(x) data_summary_V2(x, varname = "X_P_Change", groupnames = "Taxa"))
```

```{r} 
# BM
lapply(df.pa_original.BF, function(x) data_summary_V2(x, varname = "Y_P_Change", groupnames = "Taxa"))
```

```{r} 
# diff in change
lapply(df.pa_original.BF, function(x) data_summary_V2(x, varname = "pa.change.diff", groupnames = "Taxa"))
```

### Permutation test for Prevalence Trajectory analysis 

**Just assessing for EC - main results**
```{r} 
Res_ec_bf <- perm_func_dfs_v2(X_df = df.pa_original.BF$EC, 
                                  data = funcs_3M1Y_bin_diff_s$EC,
                                  variable = "BF_status_3m_sampling", 
                                  Group_X = "No BM", Group_Y = "Exclusive",
                                  N = N_vars$EC)

# global signfiicance?
original_wilcox_ec_bf <- Res_ec_bf$global_res
original_wilcox_ec_bf

## Res_per_taxa  p-value assigned to each taxa; p_value = later in healthy; p_value_2 = earlier in healthy
Res_per_taxa_ec_bf <- Res_ec_bf$per_taxa_res
```

```{r}
# sig early
Res_per_taxa_ec_bf_early <- subset(Res_per_taxa_ec_bf, p_value_2 < p_thresh2)
Res_per_taxa_ec_bf_early

# sig late
Res_per_taxa_ec_bf_late2 <- subset(Res_per_taxa_ec_bf, p_value < p_thresh2)
Res_per_taxa_ec_bf_late2
```


## EC overlap b/t functions associated with BF and Asthma
0.01 - 8 shown
```{r} 
BF_asth3y_ec_01 <- intersect(row.names(Res_per_taxa_ec_bf_late2), row.names(Res_per_taxa_ec_asthma_late2))
length(BF_asth3y_ec_01) 

BF_asth3y_ec_01_2 <- data.frame(BF_asth3y_ec_01) %>% 
                    separate(BF_asth3y_ec_01, c("Function", "Taxa"), sep = "_")

table(BF_asth3y_ec_01_2$Taxa)
BF_asth3y_ec_01_2
```

# Generate Figure Panels  
## Figure 4A - PreTCoA 
Figure 4A - EC annotations (df # 3), Also plotting Figure S8 - EggNogg Annotations (df # 1) and MetaCyc (extra, df # 2)
```{r} 
# EC = first row, then EggNog, then MetaCyc
names(df.pa)
tiff(file = paste0(path, "Figures/Fig_4A_EC_Fig_S8_EggNog_MetaC_FuncStrataBFAsthma.tiff"),
     width = 2000, height = 3000, units = "px", res = 300)
plot_grid(PreTCoA_bf[[3]]+theme(legend.position = "none")+ggtitle("")+
          labs(x="No BM\n1 Year - 3 Month Prevalence (%)", y="Exclusive BM\n1 Year - 3 Month Prevalence (%)"), 
          PreTCoA_asthma[[3]]+theme(legend.position = "none")+ggtitle("")+
          labs(x="Asthma, 3 Years\n1 Year - 3 Month Prevalence (%)", y="No Asthma\n1 Year - 3 Month Prevalence (%)"),
          PreTCoA_bf[[1]]+theme(legend.position = "none")+ggtitle("")+
            labs(x="No BM\n1 Year - 3 Month Prevalence (%)", y="Exclusive BM\n1 Year - 3 Month Prevalence (%)"), 
          PreTCoA_asthma[[1]]+theme(legend.position = "none")+ggtitle("")+
            labs(x="Asthma, 3 Years\n1 Year - 3 Month Prevalence (%)", y="No Asthma\n1 Year - 3 Month Prevalence (%)"),
          PreTCoA_bf[[2]]+theme(legend.position = "none")+ggtitle("")+
            labs(x="No BM\n1 Year - 3 Month Prevalence (%)", y="Exclusive BM\n1 Year - 3 Month Prevalence (%)"), 
          PreTCoA_asthma[[2]]+theme(legend.position = "none")+ggtitle("")+
            labs(x="Asthma, 3 Years\n1 Year - 3 Month Prevalence (%)", y="No Asthma\n1 Year - 3 Month Prevalence (%)"), ncol = 2
          )
dev.off()

#tiff(file = paste0(path, "Figures/EC_forLegend.tiff"), width = 1500, height = 1500, units = "px", res = 300)
#PreTCoA_bf[[3]]
#dev.off()

## also save some plots to get legends (same for asthma and BF but differ b/t types)
#tiff(file = paste0(path, "Figures/EggNog_forLegend.tiff"), width = 1500, height = 1500, units = "px", res = 300)
#PreTCoA_bf[[1]]
#dev.off()
```

## Figure 4B - EC annotation violin plots - overall difference in change in prevalence stratified by species
Data frame of statistical test and summary statistics
```{r} 
# EC annotations - performing Wilcoxon signed rank test stratified by taxa - X_P_Change vs Y_P_Change
df_ls <- split(df.pa_original.BF$EC, df.pa_original.BF$EC$Taxa)
df_res_bf <- wilcox.pair.df.loops(df_ls, x_var = "X_P_Change", y_var =  "Y_P_Change",
                                  x_name = "No BM", y_name = "Exclusive", variable2 = "Breastfeeding\n3 Months")

df_ls <- split(df.pa_original_asthma$EC, df.pa_original_asthma$EC$Taxa)
df_res_ast <- wilcox.pair.df.loops(df_ls, x_var = "X_P_Change", y_var =  "Y_P_Change",
                                  x_name = "Yes", y_name = "No", variable2 = "Asthma\n3 Years")


## df for plot annotations - using pubr formatting
ec_res_for_plot <- rbind(df_res_bf, df_res_ast)
# need to match the long df version 
## so same as plot taxa names ## same group levels
ec_res_for_plot_2 <- ec_res_for_plot %>% 
                        mutate(Taxa =  gsub("\\.", " ", Taxa),
                               variable2 = factor(variable2, 
                                                  levels = c("Breastfeeding\n3 Months", "Asthma\n3 Years"))
                               )

# Extra for asthetics - want same y-positions across stratifications by taxa - also need to split by taxa anyway...
## make y position on plot a function of the overall max
ec_res_for_plot_ls <- split(ec_res_for_plot_2, ec_res_for_plot_2$Taxa)

# for pval annotation y-position will be the max of the max values across per-taxa stratified groups or close to this - *0.90 fraction 
ec_res_for_plot_ls <- lapply(ec_res_for_plot_ls, 
                              function(x) {x[["y.position"]] <- max(x[["max"]]*0.90); x})

# make second df for effect size annotation - closer to the minimum
# names(ec_res_for_plot_ls) == names(df.pa_asthma_bf_ec) # good!
ec_res_for_plot_ls_2 <- lapply(ec_res_for_plot_ls, function(x) {x[["y.position"]] <- min(x[["min"]]-1); x})
```

Full long format dataframes for plots 
```{r} 
 ## df for BF plot
df.pa_original.BF_L <- list()
for(i in seq(df.pa_original.BF)) {
  df.pa_original.BF_L[[i]] <- subset(melt(df.pa_original.BF[[i]], measure.vars = c("X_P_Change", "Y_P_Change"))) %>%
       # making variables to plot - consistent col names vs. asthma
                                mutate(variable2 = "Breastfeeding\n3 Months",
                                       Groups = ifelse(variable == "X_P_Change", "No BM", "Exclusive"),
                                       Taxa = str_to_sentence(gsub("\\.", " ", Taxa)) ## reformatting names - capitalizing first letter too
                                       )
}

## df for asthma plot
df.pa_original_asthma_L <- list()
for(i in seq(df.pa_original_asthma)) {
  df.pa_original_asthma_L[[i]] <- subset(melt(df.pa_original_asthma[[i]], measure.vars = c("X_P_Change", "Y_P_Change"))) %>%
       # making variables to plot - consistent col names vs. asthma
                                mutate(variable2 = "Asthma\n3 Years",
                                       Groups = ifelse(variable == "X_P_Change", "Yes", "No"),
                                       Taxa = str_to_sentence(gsub("\\.", " ", Taxa)) ## reformatting names - capitalizing first letter to0
                                       )
}


## Decided to Rbind to plot these together on second thought
df.pa_asthma_bf <- list()
for(i in seq(df.pa_original.BF_L)) {
  df.pa_asthma_bf[[i]] <- rbind(df.pa_original.BF_L[[i]],  df.pa_original_asthma_L[[i]]) %>%
                              mutate(variable2 = factor(variable2, levels = c("Breastfeeding\n3 Months", "Asthma\n3 Years")))
}
names(df.pa_asthma_bf) <- names(df.pa_original.BF)
# df.pa_asthma_bf = long format df for plotting


# keeping colours consistent as previous colours (despite slight diff in order)
unique(df.pa_asthma_bf[[1]]$Taxa)
cols # taxa in same order but modified names - space instead of dot
cols_2 <- cols
names(cols_2) <- unique(df.pa_original.BF_L[[1]]$Taxa)


# Issue with ggplot2 faceting - would want y-axis same for the same Taxa, but different ranges b/t taxa 
## Resolve - do separately and use loop instead of facet wrap to achieve this.
# Need different loop setup for this
df.pa_asthma_bf_ec <- split(df.pa_asthma_bf$EC, df.pa_asthma_bf$EC$Taxa)

## using this setup, will have to reorder colours based on these names instead
cols_3 <- cols_2[match(names(df.pa_asthma_bf_ec), names(cols_2))] # match - target, original names

# EC
box_plots_ec <- list()
for(i in seq_along(df.pa_asthma_bf_ec)) {
  box_plots_ec[[i]] <- ggplot(df.pa_asthma_bf_ec[[i]], aes(x=Groups, y=value, group=Groups)) + 
  facet_wrap(~variable2, scales = "free_x") + 
  geom_jitter(color = cols_3[[i]], shape=16, size=1.5, alpha=0.60, position = position_jitter(width = 0.2)) +
  geom_violin(trim = FALSE, alpha=0) + 
  stat_summary(fun=median, geom="point", shape=23, size=2, fill="black")  + 
  scale_y_continuous(expand=c(0,0)) + 
  theme(axis.text = element_text(size=10), legend.position = "none", strip.text.x = element_text(face="bold", size = 11),
        plot.title = element_text(face = "bold.italic", size = 12)) +
    labs(y = "", x = "") +
    ggtitle(names(df.pa_asthma_bf_ec)[[i]]) +
   stat_pvalue_manual(ec_res_for_plot_ls[[i]], label = "p.cat",  bracket.size = 0, color="black", size = 8) +
   stat_pvalue_manual(ec_res_for_plot_ls_2[[i]], label = "effect_lab",  bracket.size = 0, color="black", size = 3)
}

tiff(file = paste0(path, "Figures/Fig_4B_EC_FuncStrataBFAsthma.tiff"), width = 1875, height = 2775, units = "px", res = 300)
plot_grid(box_plots_ec[[7]], box_plots_ec[[2]],  ## re-ordered taxa based on author suggestion 
          box_plots_ec[[1]], box_plots_ec[[5]],
          box_plots_ec[[4]],  box_plots_ec[[3]],
          box_plots_ec[[6]],  box_plots_ec[[8]], ncol = 2)
dev.off()
```

## Save results associated with Figure 4
```{r} 
# saved per-taxa res - asthma and BF
write.csv(Res_per_taxa_ec_asthma, file = paste0(path, "Tables/Res_Fig4A_EC_StrataAsthmaPreTCo.csv"))
write.csv(Res_per_taxa_ec_bf, file = paste0(path, "Tables/Res_Fig4A_EC_StrataBFPreTCo.csv"))

# Also save associated result files - csvs with p-values
write.csv(ec_res_for_plot_2, file = paste0(path, "Tables/Res_Fig4B_EC_PairedWilcoxonStrata.csv"))

# Associated Detailed per-group summary stats
all_effect_size_perTaxa <- lapply(df.pa_asthma_bf, function(x) data_summary_V2(x, varname = "value", groupnames = c("Taxa", "variable2", "Groups")))
write.csv(all_effect_size_perTaxa$EC, file = paste0(path, "Tables/Res_Fig4B_EC_SummaryStatsStrata.csv"))
```

