---
title: "Delayed  colonization patterns of nasal and gut microbiota associated with exclusive breastfeeding and reduced asthma development (Figure 4)"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Setup 

```{r, warning=FALSE, message=FALSE}
R.version.string
library(phyloseq); packageVersion('phyloseq')
library(ggplot2); packageVersion('ggplot2')
library(cowplot); packageVersion('cowplot')
library(vegan); packageVersion('vegan')
library(tidyr); packageVersion('tidyr')
library(reshape2); packageVersion('reshape2')
library(dplyr); packageVersion('dplyr')  
library(lmPerm); packageVersion('lmPerm')
library(UpSetR); packageVersion('UpSetR')
library(ComplexHeatmap); packageVersion('ComplexHeatmap')
library(circlize); packageVersion('circlize')

source("Preprocessing/Functions.R")
```

```{r}
theme_set(theme_bw() + theme(plot.title = element_text(hjust = 0.5,face = "bold"), strip.text.x = element_text(size = 11), strip.background = element_rect(colour="white", fill="white"), panel.grid.major = element_blank(), panel.grid.minor = element_blank(), text = element_text(size=12), axis.text = element_text(color="black", size=12), legend.key.size = unit(0.5,"cm")))

cbPalette <- c("navy", "paleturquoise4",  "violetred3", "skyblue1", "goldenrod3")
```

### Microbiota Trajectory data subsets used
```{r}
Nasal_3M1Y_pa_filt <- readRDS("Data_subsets/Main_Nasal_Trajectory_Subset.rds") # Prevalence dataset

Gut_3M1Y_pa_filt <- readRDS("Data_subsets/Main_Gut_Trajectory_Subset.rds") # Prevalence dataset
```

```{r}
# splitting by timepoint for downstream analyses
N.pa.3m <- subset_samples(Nasal_3M1Y_pa_filt, Visit == "3 Months")
N.pa.1y <- subset_samples(Nasal_3M1Y_pa_filt, Visit == "1 Year")

G.pa.3m <- subset_samples(Gut_3M1Y_pa_filt, Visit == "3 Months")
G.pa.1y <- subset_samples(Gut_3M1Y_pa_filt, Visit == "1 Year")
```


## Nasal - McNemar test for overall prevalence trajectories

```{r}
# input
N_taxa <- ntaxa(Nasal_3M1Y_pa_filt)
N_subj <- nsamples(Nasal_3M1Y_pa_filt)/2
TP_1 <- N.pa.3m
TP_2 <- N.pa.1y

# setup
N_3m_otu <- as.data.frame(t(as(otu_table(TP_1), "matrix")))
N_1y_otu <- as.data.frame(t(as(otu_table(TP_2), "matrix")))

N_3m_otu_sdf <- data.frame(N_3m_otu, CHILDid=sample_data(TP_1)$CHILDid)
N_1y_otu_sdf <- data.frame(N_1y_otu, CHILDid=sample_data(TP_2)$CHILDid)

## don't like how data.frame function changed the column names, changing back
colnames(N_3m_otu_sdf)[1:N_taxa] <- colnames(N_3m_otu)
colnames(N_1y_otu_sdf)[1:N_taxa] <- colnames(N_1y_otu)

# Making sure 3m and 1y dataframe are in the same order
N_3m_otu_sdf <- N_3m_otu_sdf[order(N_3m_otu_sdf$CHILDid),]
N_1y_otu_sdf <- N_1y_otu_sdf[order(N_1y_otu_sdf$CHILDid),]

identical(N_3m_otu_sdf$CHILDid, N_1y_otu_sdf$CHILDid) # good

# are in the same order in both lists, remove ids for test
N_3m_otu_sdf$CHILDid <- NULL
N_1y_otu_sdf$CHILDid <- NULL

# Removing the persistently present taxa that cannot be tested (~100% prevalence at both timepoints)
N_3m_otu_sdf$`Streptococcus sp. 40` <- NULL 
N_1y_otu_sdf$`Streptococcus sp. 40`  <- NULL
```

```{r}
# McNemar test
## list of contingency tables
Nasal_Tab_3m1y <- mapply(FUN=table, as.list(N_3m_otu_sdf),  as.list(N_1y_otu_sdf) , SIMPLIFY = FALSE, USE.NAMES = TRUE) 

## RUN
Nasal_3m1y_Mcnem_p_df <- Nasal_Tab_3m1y %>% lapply(., FUN = mcnemar.test) %>% lapply(., function(x) {x["p.value"]}) %>% bind_rows(., .id = "OTU")

Nasal_3m1y_Mcnem_p_df$p.value.BH <- p.adjust(Nasal_3m1y_Mcnem_p_df$p.value, method="BH" , n = length(Nasal_3m1y_Mcnem_p_df$p.value))

# Calculate average proportion at 3 months and 1 year
prop_3m <- taxa_sums(TP_1)/N_subj
prop_1y <- taxa_sums(TP_2)/N_subj

props <- data.frame(Nasal_3M_P=prop_3m*100, Nasal_1Y_P=prop_1y*100, Nasal_3M_SE=sqrt(prop_3m*(1-prop_3m)/N_subj), Nasal_1Y_SE=sqrt(prop_1y*(1-prop_1y)/N_subj))
props$OTU <- row.names(props)

# strep is 100 and 0 SE so no test done, automatic 100% persistent (NA p values for this one)
Nasal_3m1y_Mcnem_p_df2 <- merge(Nasal_3m1y_Mcnem_p_df, props, by="OTU", all=TRUE) 

# Calc change in prevalence
Nasal_3m1y_Mcnem_p_df2$Nasal_changeP_1Y3M <- Nasal_3m1y_Mcnem_p_df2$Nasal_1Y_P - Nasal_3m1y_Mcnem_p_df2$Nasal_3M_P 
```

Using a more strict p-value threshold in addition to an effect size cutoff to define colonizers (checked volcano plots of nasal and gut before deciding on these thresholds)
```{r, warning=FALSE}
Nasal_3m1y_Mcnem_p_df2$Colonization_pattern_final <- ifelse(Nasal_3m1y_Mcnem_p_df2$p.value.BH < 0.001 & Nasal_3m1y_Mcnem_p_df2$Nasal_changeP_1Y3M < -7, "Early",
                                                     ifelse(Nasal_3m1y_Mcnem_p_df2$p.value.BH < 0.001 & Nasal_3m1y_Mcnem_p_df2$Nasal_changeP_1Y3M >7, "Late", "Persistent"))

### most taxa colonize later, and average change in prevalence for these late colonizers is 19.5%
data_summary_V2(Nasal_3m1y_Mcnem_p_df2, varname = "Nasal_changeP_1Y3M", groupnames = "Colonization_pattern_final") 
```


## Gut - McNemar test for overall prevalence trajectories

```{r}
# input
N_taxa_G <- ntaxa(Gut_3M1Y_pa_filt)
N_subj_G <- nsamples(Gut_3M1Y_pa_filt)/2
TP_1 <- G.pa.3m
TP_2 <- G.pa.1y

# setup

# using dataset with all 3 month and 1 year gut and nasal samples for now - same infants in all datasets to reduce variation 
G_3m_otu <- as.data.frame(t(as(otu_table(TP_1), "matrix")))
G_1y_otu <- as.data.frame(t(as(otu_table(TP_2), "matrix")))

G_3m_otu_sdf <- data.frame(G_3m_otu, CHILDid=sample_data(TP_1)$CHILDid)
G_1y_otu_sdf <- data.frame(G_1y_otu, CHILDid=sample_data(TP_2)$CHILDid)

## don't like how data.frame function changed the column names, changing back
colnames(G_3m_otu_sdf)[1:N_taxa_G] <- colnames(G_3m_otu)
colnames(G_1y_otu_sdf)[1:N_taxa_G] <- colnames(G_1y_otu)


G_3m_otu_sdf <- G_3m_otu_sdf[order(G_3m_otu_sdf$CHILDid),]
G_1y_otu_sdf <- G_1y_otu_sdf[order(G_1y_otu_sdf$CHILDid),]

identical(G_3m_otu_sdf$CHILDid, G_1y_otu_sdf$CHILDid) # good

# are in the same order in both lists
G_3m_otu_sdf$CHILDid <- NULL
G_1y_otu_sdf$CHILDid <- NULL
```

```{r}
# McNemar test

## list of contingency tables
Gut_Tab_3m1y <- mapply(FUN=table, as.list(G_3m_otu_sdf),  as.list(G_1y_otu_sdf), SIMPLIFY = FALSE, USE.NAMES = TRUE) 

## RUN
Gut_3m1y_Mcnem_p_df <- Gut_Tab_3m1y %>% lapply(., FUN = mcnemar.test) %>% lapply(., function(x) {x["p.value"]}) %>% bind_rows(., .id = "OTU")

Gut_3m1y_Mcnem_p_df$p.value.BH <- p.adjust(Gut_3m1y_Mcnem_p_df$p.value, method="BH" , n = length(Gut_3m1y_Mcnem_p_df$p.value))

# Calculate average proporition at 3 months and 1 year
prop_3m <- taxa_sums(TP_1)/N_subj_G
prop_1y <- taxa_sums(TP_2)/N_subj_G

props <- data.frame(Gut_3M_P=prop_3m*100, Gut_1Y_P=prop_1y*100, Gut_3M_SE=sqrt(prop_3m*(1-prop_3m)/N_subj_G), Gut_1Y_SE=sqrt(prop_1y*(1-prop_1y)/N_subj_G))
props$OTU <- row.names(props)

# strep is 100 and 0 SE so no test done, automatic 100% persistent (NA p values for this one)
Gut_3m1y_Mcnem_p_df2 <- merge(Gut_3m1y_Mcnem_p_df, props, by="OTU", all=TRUE) 

# Calc change in prevalence
Gut_3m1y_Mcnem_p_df2$Gut_changeP_1Y3M <- Gut_3m1y_Mcnem_p_df2$Gut_1Y_P - Gut_3m1y_Mcnem_p_df2$Gut_3M_P 
```

The Mcnemar test appears to be very sensitive. Use a more strict p-value threshold in addition to an effect size cutoff to define Late and Early colonizers as oppose to persistent (no change) colonizers.
```{r}
Gut_3m1y_Mcnem_p_df2$Colonization_pattern_final <- ifelse(Gut_3m1y_Mcnem_p_df2$p.value.BH < 0.001 & Gut_3m1y_Mcnem_p_df2$Gut_changeP_1Y3M < -7, "Early",
                                                          ifelse(Gut_3m1y_Mcnem_p_df2$p.value.BH < 0.001 & Gut_3m1y_Mcnem_p_df2$Gut_changeP_1Y3M >7, "Late", "Persistent"))

### most taxa colonize later (65), and average change in prevalence for these late colonizers is 15%
data_summary(Gut_3m1y_Mcnem_p_df2, varname = "Gut_changeP_1Y3M", groupnames = "Colonization_pattern_final") 
```

# Prevalence Trajectory Coordinate System & Permutation tests - Comparing Prevalence Trajectories Between Groups (Fig. 4A)

## Nasal - Prevalence Trajectory derived variables, creation of a prevalence trajectory feature table
To simplify making of prevalence trajectory scatter plots, first make the 'difference in prevalence' results for every pair of samples for each infant (1Y - 3M, e.g. 0 - 1 = -1; 1 - 0 = 1) - so a "-1" means it's present at 3 months but not 1 year, 1 means 1 year but not 3m and 0 means no change.
```{r} 
## extract otu table and use Participant IDs as row.names instead - will match b/t datasets
N_3m_otu <- data.frame(as.data.frame(t(as(otu_table(N.pa.3m), "matrix"))), row.names=sample_data(N.pa.3m)$CHILDid)
N_1y_otu <- data.frame(as.data.frame(t(as(otu_table(N.pa.1y), "matrix"))), row.names=sample_data(N.pa.1y)$CHILDid)

N_3m_otu <- N_3m_otu[order(row.names(N_3m_otu)),]
N_1y_otu <- N_1y_otu[order(row.names(N_1y_otu)),]

## Checking - colnames and Childid's are in the same order? Yes - good - can move forward
identical(colnames(N_3m_otu), colnames(N_1y_otu))
identical(row.names(N_1y_otu), row.names(N_3m_otu))

## 1 Year minus 3 Month Prevalence For each Taxa
change_3m1y_pa <- N_1y_otu - N_3m_otu # -1 = present at 3m not 1y, 1= present at 1y not 3m, 0=either present at both or neither timpoint
dim(change_3m1y_pa) # Same order as original physeq obj

# Sample data to merge with dataset
sdf_toAdd <- sample_data(N.pa.1y)[order(sample_data(N.pa.1y)$CHILDid),]
sdf_3m <- sample_data(N.pa.3m)[order(sample_data(N.pa.3m)$CHILDid),]
## check that it's in the same order as change_3m1y_pa
identical(row.names(change_3m1y_pa), sdf_toAdd$CHILDid) 
identical(row.names(change_3m1y_pa), sdf_3m$CHILDid) 

### same order so just merge data with sample metadata using data.frame
change_3m1y_pa_sdf <- data.frame(change_3m1y_pa, sdf_toAdd, RE_virus_3m = sdf_3m$RE_Virus) # Using 1 year sdf for variables (will be same for 3m and 1y for almost all) - but also adding RE virus at 3 months as "RE_virus_3m". Note that "RE_Virus" is the RE virus at 1 year.

N_taxa <- ncol(change_3m1y_pa) # number of taxa, to use in downstream code
N_subj <- nrow(change_3m1y_pa) # number of subjects, to use in downstream code
```
Objects used for Nasal trajectory plots: change_3m1y_pa_sdf, Nasal_3M1Y_pa_filt, N_taxa, N_subj 

## Gut - Prevalence Trajectory derived variables, creation of a prevalence trajectory feature table
To simplify making of prevalence trajectory scatter plots, first make the 'difference in prevalence' results for every pair of samples for each infant (1Y - 3M, e.g. 0 - 1 = -1; 1 - 0 = 1) - so a "-1" means it's present at 3 months but not 1 year, 1 means 1 year but not 3m and 0 means no change.
```{r} 
## extract otu table and use Participant IDs as row.names instead - will match b/t datasets
G_3m_otu <- data.frame(as.data.frame(t(as(otu_table(G.pa.3m), "matrix"))), row.names=sample_data(G.pa.3m)$CHILDid)
G_1y_otu <- data.frame(as.data.frame(t(as(otu_table(G.pa.1y), "matrix"))), row.names=sample_data(G.pa.1y)$CHILDid)

G_3m_otu <- G_3m_otu[order(row.names(G_3m_otu)),]
G_1y_otu <- G_1y_otu[order(row.names(G_1y_otu)),]

## Checking - colnames and Childid's are in the same order? Yes - good - can move forward
identical(colnames(G_3m_otu), colnames(G_1y_otu))
identical(row.names(G_3m_otu), row.names(G_1y_otu))

## 1 Year minus 3 Month Prevalence For each Taxa
change_3m1y_pa_G <- G_1y_otu - G_3m_otu # -1 = present at 3m not 1y, 1= present at 1y not 3m, 0=either present at both or neither timpoint
dim(change_3m1y_pa_G) # Same order as original physeq obj

sdf_toAdd <- sample_data(G.pa.1y)[order(sample_data(G.pa.1y)$CHILDid),]
sdf_3m <- sample_data(G.pa.3m)[order(sample_data(G.pa.3m)$CHILDid),]
## check that it's in the same order as change_3m1y_pa
identical(row.names(change_3m1y_pa_G), sdf_toAdd$CHILDid) # order matches
identical(row.names(change_3m1y_pa_G), sdf_3m$CHILDid) # order matches

### same order so just merge data with sample metadata using data.frame
change_3m1y_pa_G_sdf <- data.frame(change_3m1y_pa_G, sdf_toAdd, RE_virus_3m = sdf_3m$RE_Virus) # Using 1 year sdf for variables but also adding RE virus at 3 months as "RE_virus_3m". Note that "RE_Virus" is the RE virus at 1 year

N_taxa_G <- ncol(change_3m1y_pa_G) # number of taxa, to use in downstream code
N_subj_G <- nrow(change_3m1y_pa_G) # number of subjects, to use in downstream code
```
Objects used for gut trajectory plots: change_3m1y_pa_G_sdf, Gut_3M1Y_pa_filt, N_taxa_G, N_subj_G 

```{r}
## don't save with metadata - just save new derived data with CHILDid's as rownames: change_3m1y_pa, change_3m1y_pa_G; 
### merge before saving, but first differentiate nasal vs gut trajectory features by prepending N or G
colnames(change_3m1y_pa) <- gsub("^", "N_", colnames(change_3m1y_pa))
colnames(change_3m1y_pa_G) <- gsub("^", "G_", colnames(change_3m1y_pa_G))

change_3m1y_pa_all <- merge(change_3m1y_pa, change_3m1y_pa_G, by="row.names", all=TRUE)
row.names(change_3m1y_pa_all) <- change_3m1y_pa_all$Row.names
change_3m1y_pa_all$Row.names <- NULL
dim(change_3m1y_pa_all)
```

```{r} 
# setup for plots
## for plot annotations:
Main_phyla <- c('Proteobacteria','Firmicutes', "Actinobacteria", "Bacteroidetes")
```

## Nasal - Breastfeeding - Figure 4A
```{r} 
X <- subset(change_3m1y_pa_sdf, BF_status_3m_sampling == "No BM")
Y <- subset(change_3m1y_pa_sdf, BF_status_3m_sampling == "Exclusive")

# Make data.frame of prevalence like before BUT keeping as proprotions instead of Percentages for now
df.pa.nasal.BF <- data.frame(X_P_Change=colSums(X[1:N_taxa])*100/dim(X)[1],Y_P_Change=colSums(Y[1:N_taxa])*100/dim(Y)[1],
                             taxonomy=as(tax_table(Nasal_3M1Y_pa_filt), "matrix"), prevalence=taxa_sums(Nasal_3M1Y_pa_filt)*100/(N_subj*2)) # Taxonomy order actually maintained compared to physeq obj Nasal_3M1Y_pa_filt so can merge this way. prevalence column is the total prevalence so divided by total N across both timepoints (subject # x 2). 

# difference in change
df.pa.nasal.BF$pa.change.diff <- df.pa.nasal.BF$Y_P_Change - df.pa.nasal.BF$X_P_Change
```
Aesthetic things
```{r} 
# Plot by major phylum & annotate signif taxa
df.pa.nasal.BF$Phylum <- ifelse(df.pa.nasal.BF$taxonomy.Phylum %in% Main_phyla, df.pa.nasal.BF$taxonomy.Phylum, "Other")
#Reorder the phyla 
df.pa.nasal.BF$Phylum = factor(df.pa.nasal.BF$Phylum, levels=c(Main_phyla, "Other"))
```
Plot Change in Prevalence
```{r} 
x_lab <- paste("No BM (n=", nrow(X), ")\n1 Year - 3 Month Prevalence (%)", sep="")
y_lab <- paste("Exclusive BM (n=", nrow(Y), ")\n1 Year - 3 Month Prevalence (%)", sep="")

Fig_4A_N <- ggplot() + 
  geom_point(data=df.pa.nasal.BF, aes(x=X_P_Change, y=Y_P_Change, color=Phylum, size=prevalence), alpha=0.75) + 
  xlab(x_lab) + ylab(y_lab) + 
  scale_color_manual(values = cbPalette) +
  scale_y_continuous(limits=c(-30,60.6), expand = c(0,0))+
  scale_x_continuous(limits=c(-30,60.6), expand = c(0,0))+
  annotate("segment", x = -30, xend = 60, y = -30, yend = 60, colour = "red", size=0.5) + ggtitle("Nasal")
Fig_4A_N
```
Median change in prevalence of an OTU in BF and non-BF infants
```{r} 
# no BM
summary(df.pa.nasal.BF$X_P_Change)

# BM
summary(df.pa.nasal.BF$Y_P_Change)
```

**Permutation test for Prevalence Trajectory analysis - See Permutation_Tests_PrevTrajectories directory for source code**
```{r} 
source("Permutation_Tests_PrevTrajectories/03_BF_Nasal.R")

# Test output
# Res_per_taxa_BF_N - p-value assigned to each taxa; p_value = later in healthy; p_value_2 = earlier in healthy
original_wilcox_BF_N # Overall significance test
```

## Gut - Breastfeeding - Figure 4A
```{r} 
X <- subset(change_3m1y_pa_G_sdf, BF_status_3m_sampling == "No BM")
Y <- subset(change_3m1y_pa_G_sdf, BF_status_3m_sampling == "Exclusive")

# Make data.frame of prevalence like before BUT keeping as proprotions instead of Percentages for now
df.pa.gut.BF <- data.frame(X_P_Change=colSums(X[1:N_taxa_G])*100/dim(X)[1],Y_P_Change=colSums(Y[1:N_taxa_G])*100/dim(Y)[1], taxonomy=as(tax_table(Gut_3M1Y_pa_filt), "matrix"), prevalence=taxa_sums(Gut_3M1Y_pa_filt)*100/(N_subj_G*2)) # Taxonomy actually maintained compared to physeq obj, great

# difference in change
df.pa.gut.BF$pa.change.diff <- df.pa.gut.BF$Y_P_Change - df.pa.gut.BF$X_P_Change
```
Asthetic things, not important
```{r} 
## Plot by major phylum & annotate signif taxa
df.pa.gut.BF$Phylum <- ifelse(df.pa.gut.BF$taxonomy.Phylum %in% Main_phyla, df.pa.gut.BF$taxonomy.Phylum, "Other")
#Reorder the phyla 
df.pa.gut.BF$Phylum = factor(df.pa.gut.BF$Phylum, levels=c(Main_phyla, "Other"))
```
Plot Change in Prevalence 
```{r} 
x_lab <- paste("No BM (n=", nrow(X), ")\n1 Year - 3 Month Prevalence (%)", sep="")
y_lab <- paste("Exclusive BM (n=", nrow(Y), ")\n1 Year - 3 Month Prevalence (%)", sep="")

Fig_4A_G <- ggplot() + 
  geom_point(data=df.pa.gut.BF, aes(x=X_P_Change, y=Y_P_Change, color=Phylum, size=prevalence), alpha=0.75) + 
  xlab(x_lab) + ylab(y_lab) + 
  scale_color_manual(values = cbPalette) +
  scale_y_continuous(limits=c(-30,60), expand = c(0,0))+
  scale_x_continuous(limits=c(-30,60), expand = c(0,0))+
  annotate("segment", x = -30, xend = 60, y = -30, yend = 60, colour = "red", size=0.5) + ggtitle("Gut")
Fig_4A_G
```
Median change in prevalence of an OTU in BF and non-BF infants
```{r} 
# no BM
summary(df.pa.gut.BF$X_P_Change)

# BM
summary(df.pa.gut.BF$Y_P_Change)
```

**Permutation test for Prevalence Trajectory analysis - See Permutation_Tests_PrevTrajectories directory for source code**
```{r} 
source("Permutation_Tests_PrevTrajectories/04_BF_Gut.R")

# Test output
# Res_per_taxa_BF_G - p-value assigned to each taxa; p_value = later in healthy; p_value_2 = earlier in healthy
original_wilcox_BF_G # Overall significance test
```




```{r, include=FALSE}
# Save prevalence trajectory scatter plots, remove axis titles and re-add later


## BF exclusivity

Fig_4A_N_2 <- Fig_4A_N + theme(legend.position = "none", axis.title.x = element_blank(), axis.title.y = element_blank()) + ggtitle("")
tiff(file = "Figures/Fig_4A_Nasal_BF.tiff", width = 1275, height = 1275, units = "px", res = 300) 
Fig_4A_N_2
dev.off()

Fig_4A_G_2 <- Fig_4A_G + theme(legend.position = "none", axis.title.x = element_blank(), axis.title.y = element_blank()) + ggtitle("")
tiff(file = "Figures/Fig_4A_Gut_BF.tiff", width = 1275, height = 1275, units = "px", res = 300) 
Fig_4A_G_2 
dev.off()

```

# Plots of Individual taxa's Colonization patterns (Fig 4B & 4C)

### Select a few of the taxa with the lowest p-values for later colonization in healthy / BF groups vs Asthma / No BF
Only room for 4 nasal and 2 gut though so reduce p-val thresh
```{r}
p_vals_select <- merge(subset(Per_test_df_ls$BF_Nasal, select = c("p_value")), 
                       subset(Per_test_df_ls$Ast3y_Nasal, select = c("p_value")), 
                       by="row.names", suffixes = c("_BF", "_Ast"))

p_vals_select_G <- merge(subset(Per_test_df_ls$BF_Gut, select = c("p_value")), 
                         subset(Per_test_df_ls$Ast3y_Gut, select = c("p_value")), 
                         by="row.names", suffixes = c("_BF", "_Ast"))

##  Selected for Mediation if associated w/t both Asthma and BF
BF_asthma_Mostsig <- subset(p_vals_select, p_value_BF <0.01 & p_value_Ast <0.01)
  
BF_asthma_Mostsig_G <- subset(p_vals_select_G, p_value_BF <0.01 & p_value_Ast <0.01)

# But limit number selected to show in plot - picking the lowest cumulative p-values from BF and asthma
BF_asthma_Mostsig$p_val_sum <- BF_asthma_Mostsig$p_value_BF + BF_asthma_Mostsig$p_value_Ast
BF_asthma_Mostsig[order(BF_asthma_Mostsig$p_val_sum),] # select Pantoea.sp..4, Streptococcus.salivarius.5, Leptotrichia.sp..12; three are tied w/t 0.007 but picking Fusobacterium.sp..27 b/c effect size is larger for asthma.
To_select_N <- c("Streptococcus salivarius 5", "Pantoea sp. 4", "Leptotrichia sp. 12", "Fusobacterium sp. 27")

BF_asthma_Mostsig_G$p_val_sum <- BF_asthma_Mostsig_G$p_value_BF + BF_asthma_Mostsig_G$p_value_Ast
BF_asthma_Mostsig_G[order(BF_asthma_Mostsig_G$p_val_sum),] ## for gut also plot C. diff (also an interesting taxa) - but exclude from manuscript 
To_select_G <- c("Faecalibacterium sp. 13", "Flavonifractor sp.", "Monoglobus sp.", "Clostridioides difficile")
```

### Get Derived colonization pattern variables for selected taxa and merge with metadata used 
generated in 00_Processing_Derived_Trajectories.Rmd
```{r}
Traj_data_ls <- readRDS("Data_subsets/ALL_Trajectory_Derived_Data.rds") # Derived Trajectory variables 
```

```{r}
## identical(row.names(Traj_data_ls$Trajectories_Nasal_to181_Gut_at182), row.names(Traj_data_ls$Metadata))
Df_ColCount_P <- merge(subset(Traj_data_ls$Trajectories_Nasal_to181_Gut_at182, select = c(To_select_N, To_select_G)),
                            subset(Traj_data_ls$Metadata, select = c("BF_status_3m_sampling", "asthma_3y")), by="row.names")

DF_M_3m1y_ColCounts <- melt(Df_ColCount_P, variable.name = "Taxa") 
```

### Summarize percentages of dyads with these taxa as Early, Late, Persistent or Absent colonizers and plot
```{r, fig.width=10, fig.height=10}
# Need to Create a binary version for each of the colonization patterns, then calculate prevalence for each 
DF_M_3m1y_ColCounts$Early <- ifelse(DF_M_3m1y_ColCounts$value == 1, 1, 0) # for whether it's early or not
DF_M_3m1y_ColCounts$Persistent <- ifelse(DF_M_3m1y_ColCounts$value == 2, 1, 0) # for whether it's persistent or not
DF_M_3m1y_ColCounts$Late <- ifelse(DF_M_3m1y_ColCounts$value == 3, 1, 0) # for whether it's late or not
DF_M_3m1y_ColCounts$Absent <- ifelse(DF_M_3m1y_ColCounts$value == 0, 1, 0) # for whether it's absent or not

# Then melt using early/persistent/late/absent as measured vars to calculate prevalence for each pattern at once
DF_M_3m1y_ColCounts_m <- melt(DF_M_3m1y_ColCounts, measure.vars = c("Early", "Persistent", "Late", "Absent"), variable.name = "Pattern", value.name = "Occurance")

# Plot these for Asthma, BF and Colds
ColCounts_BF_summ <- prev_summary(subset(DF_M_3m1y_ColCounts_m, BF_status_3m_sampling != "Partial"), 
                     varname = "Occurance", groupnames = c("Taxa", "Pattern", "BF_status_3m_sampling")) 


ColCounts_Ast_summ <- prev_summary(subset(DF_M_3m1y_ColCounts_m, asthma_3y != "Possible"), 
                     varname = "Occurance", groupnames = c("Taxa", "Pattern", "asthma_3y"))


# Merge DFs
## First label datasets by variable for rbind & plot
ColCounts_Ast_summ$Variable <- "Asthma"
ColCounts_BF_summ$Variable <- "Breastfeeding status"

colnames(ColCounts_Ast_summ)[3] <- "Groups"
colnames(ColCounts_BF_summ)[3]<- "Groups"

ToPlot_Taxa_df <- rbind(ColCounts_Ast_summ, ColCounts_BF_summ)

# re-order groups 
ToPlot_Taxa_df$Groups <- factor(ToPlot_Taxa_df$Groups, levels = c("Exclusive", "No BM", "No", "Yes"))
ToPlot_Taxa_df$Pattern <- factor(ToPlot_Taxa_df$Pattern, levels = c("Absent", "Early", "Persistent", "Late"))
ToPlot_Taxa_df$Variable <- factor(ToPlot_Taxa_df$Variable, levels = c("Breastfeeding status", "Asthma"))

## Decided to not keep monoglobus & C. diff in plot for final figure version
ToPlot_Taxa_df_2 <- subset(ToPlot_Taxa_df, Taxa %ni% c("Monoglobus sp.", "Clostridioides difficile"))

Fig_4B_4C <- ggplot(ToPlot_Taxa_df_2, aes(x=Groups, y=Prevalence, fill=Pattern)) + 
  facet_wrap(~Taxa*Variable, scales = "free_x", ncol = 4) +
  geom_bar(stat="identity", position="stack") + 
  theme(axis.ticks.x = element_blank(), axis.text = element_text(size=10), strip.text.x = element_text(size = 8)) + 
  labs(y = "Percent of infants", title='', x="", fill = "Colonization/nPattern") + scale_fill_manual(values=c("grey", "#b58e8b", "#748f7a", "#619CFF")) + 
  scale_y_continuous(expand=c(0,0), limits = c(0,100))
Fig_4B_4C 
```

```{r, include=FALSE}
# Save figure 4B & C (as one fig) & output

# Individual taxa stacked bar plot - re-doing strip text later
tiff(file = "Figures/Fig_4B_4C.tiff", width = 2300, height = 2150, units = "px", res = 300) # warning only occurs after removing text but looks the same as version with text.
Fig_4B_4C
dev.off()

## making version without strip text too
Fig_4B_4C_V2 <- Fig_4B_4C + theme(strip.text.x = element_blank())
tiff(file = "Figures/Fig_4B_4C_blank.tiff", width = 2500, height = 1900, units = "px", res = 300) 
Fig_4B_4C_V2
dev.off()

write.csv(ToPlot_Taxa_df, file="Output/Fig4B_C_ColPats_SelectTaxa_BF_Ast_Res.csv")
```

