---
title: "Figure 1B & 1C. Nasal and gut microbiota composition in early-life in the CHILD cohort"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Setup 

```{r, message=FALSE, warning=FALSE}
R.version.string
library(phyloseq); packageVersion('phyloseq')
library(ggplot2); packageVersion('ggplot2')
library(tableone); packageVersion('tableone')
library(tidyr); packageVersion('tidyr')
library(reshape2); packageVersion('reshape2')
library(dplyr)
library(here)

source(paste0(here::here(), "/Functions.R"))

theme_set(theme_bw() + theme(plot.title = element_text(hjust = 0.5,face = "bold"), strip.text.x = element_text(size = 11), strip.background = element_rect(colour="white", fill="white"), panel.grid.major = element_blank(), panel.grid.minor = element_blank(), text = element_text(size=10), axis.text = element_text(color="black", size=12), legend.key.size = unit(0.5,"cm")))

mycols30 <- c("#e9fcff","#7435e2","#b9f2b7","#25007a","#ffed5c","#0153d2","#75a800","#93007a", "purple", "#ef0025","#fd6a00","#0191f9","#ffa32b","#004769","#c8ff97","#a3005a","#5d8300","#b5b2ff","#8e7a00","#ff77a1","#004b19","#ff7a5f","#00615b","#9f6000","#715333","#ffedb0","#540500","#00d488", "black","#ffbbd2","#01bfae", "#e82769", "grey")
```

**Data subsets**
```{r} 
data_NG2 <- readRDS(paste0(path, "Data/Data_subsets/Main_UnRarefied_NasalGut_Subset.rds"))

## total samples, not paired (used for Fig 1B & C stacked bars)
table(sample_data(data_NG2)$SampleType, sample_data(data_NG2)$Visit)
```

# Genus level relative abundances, overall microbiota composition (Figure 1B, 1C)
**Also see Table S1, uses the same scripts**

Aggregate to genus level and use relative data
```{r}
## aggregate to genus level 
data_NG2_G <- tax_glom(data_NG2, taxrank = "Genus")

NG_relative <- transform_sample_counts(data_NG2_G, function(x) x*100 / sum(x) )  
```

Split data by SampleType and time-point for taxa filter - want to plot genera present in 70% of samples at EITHER time-point
```{r}
genus_n_3m <- subset_samples(NG_relative, SampleType=="Nasal" & Visit == "3 Months")
genus_n_1y <- subset_samples(NG_relative, SampleType=="Nasal" & Visit == "1 Year")

genus_gut_3m <- subset_samples(NG_relative, SampleType=="Gut" & Visit == "3 Months")
genus_gut_1y <- subset_samples(NG_relative, SampleType=="Gut" & Visit == "1 Year")
```

Select taxa to plot
- Genera present in at least 70% of samples in at EITHER time-point will be colored in the plot with the remainder grouped into "Other"  
```{r}
data.p60_n_3m = filter_taxa(genus_n_3m, function(x) sum(x > 0) > (0.70*length(x)), TRUE) 
ntaxa(data.p60_n_3m) 
data.p60_n_1y = filter_taxa(genus_n_1y, function(x) sum(x > 0) > (0.70*length(x)), TRUE) 
ntaxa(data.p60_n_1y)

data.p60_gut_3m = filter_taxa(genus_gut_3m, function(x) sum(x > 0) > (0.70*length(x)), TRUE) 
ntaxa(data.p60_gut_3m)
data.p60_gut_1y = filter_taxa(genus_gut_1y, function(x) sum(x > 0) > (0.70*length(x)), TRUE) 
ntaxa(data.p60_gut_1y)
```

Nasal taxa to plot for 3-month & 1-year data
```{r}
Ntaxa_toPlot <- c(get_taxa_unique(data.p60_n_3m, taxonomic.rank = "Genus"), 
                 get_taxa_unique(data.p60_n_1y, taxonomic.rank = "Genus"))
Ntaxa_toPlot2 <- subset(Ntaxa_toPlot, !duplicated(Ntaxa_toPlot)) # remove dups found in more than 1 sample type
length(Ntaxa_toPlot2) # unique Genera 

# Use subset of all nasal samples
N_relative <- subset_samples(NG_relative, SampleType=="Nasal")

# Make overall nasal dataset of prevalent genera in either time-point 
N_relative_Sub <- subset_taxa(N_relative, Genus %in% Ntaxa_toPlot2)
N_relative_Sub
```

Gut taxa to plot
```{r}
Gtaxa_toPlot <- c(get_taxa_unique(data.p60_gut_3m, taxonomic.rank = "Genus"),
                 get_taxa_unique(data.p60_gut_1y, taxonomic.rank = "Genus"))
Gtaxa_toPlot2 <- subset(Gtaxa_toPlot, !duplicated(Gtaxa_toPlot))
length(Gtaxa_toPlot2) # unique Genera 

# Use subset of all nasal samples
G_relative <- subset_samples(NG_relative, SampleType=="Gut")

# Make overall nasal dataset of prevalent genera in either time-point 
G_relative_Sub <- subset_taxa(G_relative, Genus %in% Gtaxa_toPlot2)
G_relative_Sub # more than 13 b/c of contested [Eubacterium] with different Family level taxonomy - 4 different [Eubacterium] Genera - ok to group for plot.
```


(2) Make summary data for each subset that will be plotted 

Nasal dataset for stacked bars, stratify by Visit
```{r}
n_genus_melt <- psmelt(N_relative_Sub) 

#DATA Summary 
Genus_n_summary <- data_summary(n_genus_melt, varname = "Abundance", groupnames = c("Genus", "Visit"))

## Keep only these columns for plotting
Genus_n_summary2 <-  subset(Genus_n_summary, select = c("Mean", "Genus", "Visit"))

# use this alternative to melting the whole dataset to add 'other' group - to get 100% total, this is much quicker computationally vs melting all data
Genus_n_summary2_S <- split(Genus_n_summary2, Genus_n_summary2$Visit) 

Genus_n_summary3m <- rbind(Genus_n_summary2_S$`3 Months`, c(100-sum(Genus_n_summary2_S$`3 Months`$Mean), "Other", "3 Months")) # adding extra row for "Other" group, in same order: "Mean", "Genus", "Visit"

Genus_n_summary1y <- rbind(Genus_n_summary2_S$`1 Year`, c(100-sum(Genus_n_summary2_S$`1 Year`$Mean), "Other", "1 Year")) # adding extra row for "Other" group, in same order: "Mean", "Genus", "Visit"


Genus_N_summary3 <- rbind(Genus_n_summary3m, Genus_n_summary1y)
Genus_N_summary3$Mean <- as.numeric(Genus_N_summary3$Mean)
```

Gut dataset for stacked bars, stratify by Visit
```{r}
g_genus_melt <- psmelt(G_relative_Sub) 

#DATA Summary 
Genus_g_summary <- data_summary(g_genus_melt, varname = "Abundance", groupnames = c("Genus", "Visit"))

## Keep only these columns for plotting
Genus_g_summary2 <-  subset(Genus_g_summary, select = c("Mean", "Genus", "Visit"))


# use this alternative to melting the whole dataset to add 'other' group - to get 100% total, this is much quicker computationally vs melting all data
Genus_g_summary2_S <- split(Genus_g_summary2, Genus_g_summary2$Visit) 

Genus_g_summary3m <- rbind(Genus_g_summary2_S$`3 Months`, c(100-sum(Genus_g_summary2_S$`3 Months`$Mean), "Other", "3 Months")) # adding extra row for "Other" group, in same order: "Mean", "Genus", "Visit"

Genus_g_summary1y <- rbind(Genus_g_summary2_S$`1 Year`, c(100-sum(Genus_g_summary2_S$`1 Year`$Mean), "Other", "1 Year")) # adding extra row for "Other" group, in same order: "Mean", "Genus", "Visit"


Genus_g_summary3 <- rbind(Genus_g_summary3m, Genus_g_summary1y)
Genus_g_summary3$Mean <- as.numeric(Genus_g_summary3$Mean)
```


## Plot Figures 1B & 1C

**Genus factor levels for order of genera in plot**
Nasal
```{r}
# reordering factor levels in a way that will make sense visually

## 1. Nasal 3 months is the 1st bar, so the taxa found in nasal 3m should go first; then any remaining taxa found in Nasal 1Y; then "Other" taxa
Legend_1st <- sort(get_taxa_unique(data.p60_n_3m, taxonomic.rank = "Genus")) # sort will re-order alphabetically
Legend_2nd <- sort(setdiff(get_taxa_unique(data.p60_n_1y, taxonomic.rank = "Genus"), get_taxa_unique(data.p60_n_3m, taxonomic.rank = "Genus"))) # anything found in Nasal 1Y not in N 3M

Factor_order_n <- c(Legend_1st, Legend_2nd, "Other")

Genus_N_summary3$Genus <- factor(Genus_N_summary3$Genus, levels = Factor_order_n)
levels(Genus_N_summary3$Genus)
```

Gut 
```{r}
# reordering factor levels in a way that will make sense visually. 

## 1. Nasal 3 months is the 1st bar, so the taxa found in nasal 3m should go first; then any remaining taxa found in Nasal 1Y; then "Other" taxa
Legend_1st <- sort(get_taxa_unique(data.p60_gut_3m, taxonomic.rank = "Genus")) # sort will re-order alphabetically
Legend_2nd <- sort(setdiff(get_taxa_unique(data.p60_gut_1y, taxonomic.rank = "Genus"), get_taxa_unique(data.p60_gut_3m, taxonomic.rank = "Genus"))) 

Factor_order_g <- c(Legend_1st, Legend_2nd, "Other")

Genus_g_summary3$Genus <- factor(Genus_g_summary3$Genus, levels = Factor_order_g)
levels(Genus_g_summary3$Genus)
```

Make sure any taxa that are present in both nasal and gut are colored the same in both figures by using the names function to label colors
```{r}
unique_taxa <- c(setdiff(Factor_order_n, Factor_order_g), setdiff(Factor_order_g, Factor_order_n), intersect(Factor_order_n, Factor_order_g)) # taxa in nasal not in gut, gut not in nasal and taxa found in both nasal and gut in >70%

length(unique_taxa) # need 33 cols

 # label 33 cols
names(mycols30) <- unique_taxa
```

Coord flipped version with Nasal and Gut stacked bars split into two separate figures 
```{r, fig.width=10, fig.height=4}
# Nasal
## flip factor levels for Visit b/c coord_flip makes it backwards.
Genus_N_summary3$Visit <- factor(Genus_N_summary3$Visit, levels = rev(levels(Genus_N_summary3$Visit)))

Fig1B_Nasal <- ggplot(Genus_N_summary3, aes(x=Visit, y=Mean, fill=Genus)) + 
  geom_bar(stat="identity", position="stack") + scale_fill_manual(values = mycols30) + 
  scale_y_continuous(expand=c(0,0)) + labs( x="", y="Relative abundance (%)", fill="Genus") + coord_flip() +
  theme(axis.ticks.x = element_line(), axis.ticks.y = element_blank(), axis.title.x = element_text(size = 10), axis.text.x = element_text(size=9), axis.text.y = element_text(size=9),
        plot.margin = margin(1, 10, 1, 1)) 

# Gut
## flip factor levels for Visit b/c coord_flip makes it backwards.
Genus_g_summary3$Visit <- factor(Genus_g_summary3$Visit, levels = rev(levels(Genus_g_summary3$Visit)))
Fig1C_Gut <- ggplot(Genus_g_summary3, aes(x=Visit, y=Mean, fill=Genus)) + 
  geom_bar(stat="identity", position="stack") + scale_fill_manual(values = mycols30) + 
  scale_y_continuous(expand=c(0,0)) + labs( x="", y="Relative abundance (%)", fill="Genus") + coord_flip() +
  theme(axis.ticks.x = element_line(), axis.ticks.y = element_blank(), axis.title.x = element_text(size = 10), axis.text.x = element_text(size=9), axis.text.y = element_text(size=9),
        plot.margin = margin(1, 10, 1, 1)) 

Fig1B_Nasal + guides(fill = guide_legend(ncol=3)) 
Fig1C_Gut +  guides(fill = guide_legend(ncol=3)) 
```

```{r, include=FALSE}
tiff(file = paste0(path, "Figures/Fig_1B_nasal_composition.tiff"), width = 2800, height = 900, units = "px", res = 300)
Fig1B_Nasal + guides(fill = guide_legend(ncol=3)) 
dev.off()

tiff(file = paste0(path, "Figures/Fig_1C_gut_composition.tiff"), width = 2800, height = 900, units = "px", res = 300)
Fig1C_Gut +  guides(fill = guide_legend(ncol=3)) 
dev.off()
```


