---
title: "Figure 3. Nasal and gut microbiota colonization patterns and their associations with breastfeeding and asthma."
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Setup 

```{r, warning=FALSE, message=FALSE}
R.version.string
library(phyloseq); packageVersion('phyloseq')
library(ggplot2); packageVersion('ggplot2')
library(cowplot); packageVersion('cowplot')
library(vegan); packageVersion('vegan')
library(tidyr); packageVersion('tidyr')
library(reshape2); packageVersion('reshape2')
library(dplyr); packageVersion('dplyr')  
library(lmPerm); packageVersion('lmPerm')
library(UpSetR); packageVersion('UpSetR')
library(ComplexHeatmap); packageVersion('ComplexHeatmap')
library(circlize); packageVersion('circlize')
library(here)

source(paste0(here::here(), "/Functions.R"))

theme_set(theme_bw() + theme(plot.title = element_text(hjust = 0.5,face = "bold"), strip.text.x = element_text(size = 11), strip.background = element_rect(colour="white", fill="white"), panel.grid.major = element_blank(), panel.grid.minor = element_blank(), text = element_text(size=12), axis.text = element_text(color="black", size=12), legend.key.size = unit(0.5,"cm")))
```

### Microbiota Trajectory data subsets used
```{r}
Nasal_3M1Y_pa_filt <- readRDS(paste0(path, "Data/Data_subsets/Main_Nasal_Trajectory_Subset.rds")) # Prevalence dataset

Gut_3M1Y_pa_filt <- readRDS(paste0(path, "Data/Data_subsets/Main_Gut_Trajectory_Subset.rds")) # Prevalence dataset
```

```{r}
# splitting by timepoint for downstream analyses
N.pa.3m <- subset_samples(Nasal_3M1Y_pa_filt, Visit == "3 Months")
N.pa.1y <- subset_samples(Nasal_3M1Y_pa_filt, Visit == "1 Year")

G.pa.3m <- subset_samples(Gut_3M1Y_pa_filt, Visit == "3 Months")
G.pa.1y <- subset_samples(Gut_3M1Y_pa_filt, Visit == "1 Year")
```


# Volcano plots identifying general colonization patterns (Fig. 3A)

## Nasal - McNemar test for overall prevalence trajectories

```{r}
# input
N_taxa <- ntaxa(Nasal_3M1Y_pa_filt)
N_subj <- nsamples(Nasal_3M1Y_pa_filt)/2
TP_1 <- N.pa.3m
TP_2 <- N.pa.1y

# setup
N_3m_otu <- as.data.frame(t(as(otu_table(TP_1), "matrix")))
N_1y_otu <- as.data.frame(t(as(otu_table(TP_2), "matrix")))

N_3m_otu_sdf <- data.frame(N_3m_otu, CHILDid=sample_data(TP_1)$CHILDid)
N_1y_otu_sdf <- data.frame(N_1y_otu, CHILDid=sample_data(TP_2)$CHILDid)

## don't like how data.frame function changed the column names, changing back
colnames(N_3m_otu_sdf)[1:N_taxa] <- colnames(N_3m_otu)
colnames(N_1y_otu_sdf)[1:N_taxa] <- colnames(N_1y_otu)

# Making sure 3m and 1y dataframe are in the same order
N_3m_otu_sdf <- N_3m_otu_sdf[order(N_3m_otu_sdf$CHILDid),]
N_1y_otu_sdf <- N_1y_otu_sdf[order(N_1y_otu_sdf$CHILDid),]

identical(N_3m_otu_sdf$CHILDid, N_1y_otu_sdf$CHILDid) # good

# are in the same order in both lists, remove ids for test
N_3m_otu_sdf$CHILDid <- NULL
N_1y_otu_sdf$CHILDid <- NULL

# Removing the persistently present taxa that cannot be tested (~100% prevalence at both timepoints)
N_3m_otu_sdf$`Streptococcus sp. 40` <- NULL 
N_1y_otu_sdf$`Streptococcus sp. 40`  <- NULL
```

```{r}
# McNemar test
## list of contingency tables
Nasal_Tab_3m1y <- mapply(FUN=table, as.list(N_3m_otu_sdf),  as.list(N_1y_otu_sdf) , SIMPLIFY = FALSE, USE.NAMES = TRUE) 

## RUN
Nasal_3m1y_Mcnem_p_df <- Nasal_Tab_3m1y %>% lapply(., FUN = mcnemar.test) %>% 
                                            lapply(., function(x) {x["p.value"]}) %>% bind_rows(., .id = "OTU")

Nasal_3m1y_Mcnem_p_df$p.value.BH <- p.adjust(Nasal_3m1y_Mcnem_p_df$p.value, method="BH" , n = length(Nasal_3m1y_Mcnem_p_df$p.value))

# Calculate average proportion at 3 months and 1 year
prop_3m <- taxa_sums(TP_1)/N_subj
prop_1y <- taxa_sums(TP_2)/N_subj

props <- data.frame(Nasal_3M_P=prop_3m*100, Nasal_1Y_P=prop_1y*100, 
                    Nasal_3M_SE=sqrt(prop_3m*(1-prop_3m)/N_subj), 
                    Nasal_1Y_SE=sqrt(prop_1y*(1-prop_1y)/N_subj))
props$OTU <- row.names(props)

# strep is 100 and 0 SE so no test done, automatic 100% persistent (NA p values for this one)
Nasal_3m1y_Mcnem_p_df2 <- merge(Nasal_3m1y_Mcnem_p_df, props, by="OTU", all=TRUE) 

# Calc change in prevalence
Nasal_3m1y_Mcnem_p_df2$Nasal_changeP_1Y3M <- Nasal_3m1y_Mcnem_p_df2$Nasal_1Y_P - Nasal_3m1y_Mcnem_p_df2$Nasal_3M_P 
```

Volcano plots aesthetics
```{r}
# 1. Using a more strict p-value threshold in addition to an effect size cutoff to define colonizers (checked volcano plots of nasal and gut before deciding on these thresholds)
# 2. Adding additional annotation
# 3. labeling important/example early vs late taxa separately b/c they require different nudge_x's. 
Nasal_3m1y_Mcnem_p_df2 <- Nasal_3m1y_Mcnem_p_df2[order(-Nasal_3m1y_Mcnem_p_df2$Nasal_changeP_1Y3M),]
top_taxa <- Nasal_3m1y_Mcnem_p_df2$OTU[1:3]
bot_taxa <- Nasal_3m1y_Mcnem_p_df2$OTU[(nrow(Nasal_3m1y_Mcnem_p_df2)-2):nrow(Nasal_3m1y_Mcnem_p_df2)]

Nasal_3m1y_Mcnem_p_df3 <- Nasal_3m1y_Mcnem_p_df2 %>% mutate(
  Colonization_pattern_final = factor(ifelse(p.value.BH < 0.001 & Nasal_changeP_1Y3M < -7, "Early",
                                      ifelse(p.value.BH < 0.001 & Nasal_changeP_1Y3M >7, "Late", "Persistent")),
                                      levels = c("Early", "Persistent", "Late")),
  
  Label = ifelse(OTU %in% top_taxa, OTU, ""), Label2 = ifelse(OTU %in% bot_taxa, OTU, "")
)
  
# abbreviate long names and remove the "sp."
bot_taxa
top_taxa
Nasal_3m1y_Mcnem_p_df3$Label2 <- Nasal_3m1y_Mcnem_p_df3$Label2 %>% 
                                  gsub("Staphylococcus hominis 2", "S. hominis 2", .) %>% 
                                  gsub("sp\\.", "", .)
Nasal_3m1y_Mcnem_p_df3$Label <- gsub("sp\\.", "", Nasal_3m1y_Mcnem_p_df3$Label) %>%
                                  gsub("^", "italic('", .) %>%
                                  gsub("$", "')", .)
# To make everything italic... can change label within data.frame and then use parse = TRUE or use fontface=3

## Lab 2 needs to be split for separate nudges too...
```

```{r, warning=FALSE}
### most taxa colonize later, and average change in prevalence for these late colonizers is 19.5%
data_summary_V2(Nasal_3m1y_Mcnem_p_df3, varname = "Nasal_changeP_1Y3M", groupnames = "Colonization_pattern_final") 
```

```{r, warning=FALSE}
ann_size <- 2.8
axis_size <- 11
ax_title_s <- 11
bot_taxa <- unique(Nasal_3m1y_Mcnem_p_df3$Label2)
top_taxa <- unique(Nasal_3m1y_Mcnem_p_df3$Label)


Fig_3A_N <- ggplot(data=Nasal_3m1y_Mcnem_p_df3, aes(x=Nasal_changeP_1Y3M, y=-log10(p.value.BH), color=Colonization_pattern_final)) + 
  geom_point(size=2) + geom_hline(yintercept = -log10(0.001), color="black") + geom_vline(xintercept = 7, color="black")  + 
  geom_vline(xintercept = -7, color="black") + labs(color="Colonization", x="1 Year - 3 Month Prevalence (%)", y="-log10(FDR p-value)") + 
  ggtitle("Colonization Patterns Nasal (n=1545)") + scale_x_continuous(limits = c(-20, 52)) + 
  geom_text(aes(label=ifelse(Label == top_taxa[1], Label, "")), nudge_x = -9, nudge_y = 2, color="black", size=ann_size,  parse = TRUE) +
  geom_text(aes(label=ifelse(Label == top_taxa[2], Label, "")), nudge_x = -8, nudge_y = 3, color="black", size=ann_size,  parse = TRUE) +
  geom_text(aes(label=ifelse(Label == top_taxa[3], Label, "")), nudge_x = -11, color="black", size=ann_size,  parse = TRUE) +
  
  geom_text(aes(label=ifelse(Label2 == bot_taxa[2], Label2, "")), nudge_x = 7,nudge_y=-4, color="black", size=ann_size, fontface=3) +
  geom_text(aes(label=ifelse(Label2 == bot_taxa[3], Label2, "")), nudge_x = 13,nudge_y=0, color="black", size=ann_size, fontface=3) +
  geom_text(aes(label=ifelse(Label2 == bot_taxa[4], Label2, "")), nudge_x = 5,nudge_y=5.5, color="black", size=ann_size, fontface=3) +
  theme(axis.text = element_text(size=axis_size), axis.title = element_text(size= ax_title_s), title = element_text(size=11))
Fig_3A_N

tiff(file = paste0(path, "Figures/Fig_3A_Nasal_Volcano_Overall.tiff"), width = 1550, height = 1370, units = "px", res = 300) 
Fig_3A_N
dev.off()
```


## Gut - McNemar test for overall prevalence trajectories

```{r}
# input
N_taxa_G <- ntaxa(Gut_3M1Y_pa_filt)
N_subj_G <- nsamples(Gut_3M1Y_pa_filt)/2
TP_1 <- G.pa.3m
TP_2 <- G.pa.1y

# setup

# using dataset with all 3 month and 1 year gut and nasal samples for now - same infants in all datasets to reduce variation 
G_3m_otu <- as.data.frame(t(as(otu_table(TP_1), "matrix")))
G_1y_otu <- as.data.frame(t(as(otu_table(TP_2), "matrix")))

G_3m_otu_sdf <- data.frame(G_3m_otu, CHILDid=sample_data(TP_1)$CHILDid)
G_1y_otu_sdf <- data.frame(G_1y_otu, CHILDid=sample_data(TP_2)$CHILDid)

## don't like how data.frame function changed the column names, changing back
colnames(G_3m_otu_sdf)[1:N_taxa_G] <- colnames(G_3m_otu)
colnames(G_1y_otu_sdf)[1:N_taxa_G] <- colnames(G_1y_otu)


G_3m_otu_sdf <- G_3m_otu_sdf[order(G_3m_otu_sdf$CHILDid),]
G_1y_otu_sdf <- G_1y_otu_sdf[order(G_1y_otu_sdf$CHILDid),]

identical(G_3m_otu_sdf$CHILDid, G_1y_otu_sdf$CHILDid) # good

# are in the same order in both lists
G_3m_otu_sdf$CHILDid <- NULL
G_1y_otu_sdf$CHILDid <- NULL
```

```{r}
# McNemar test

## list of contingency tables
Gut_Tab_3m1y <- mapply(FUN=table, as.list(G_3m_otu_sdf),  as.list(G_1y_otu_sdf), SIMPLIFY = FALSE, USE.NAMES = TRUE) 

## RUN
Gut_3m1y_Mcnem_p_df <- Gut_Tab_3m1y %>% lapply(., FUN = mcnemar.test) %>% 
                                        lapply(., function(x) {x["p.value"]}) %>% bind_rows(., .id = "OTU")

Gut_3m1y_Mcnem_p_df$p.value.BH <- p.adjust(Gut_3m1y_Mcnem_p_df$p.value, method="BH" , n = length(Gut_3m1y_Mcnem_p_df$p.value))

# Calculate average proporition at 3 months and 1 year
prop_3m <- taxa_sums(TP_1)/N_subj_G
prop_1y <- taxa_sums(TP_2)/N_subj_G

props <- data.frame(Gut_3M_P=prop_3m*100, Gut_1Y_P=prop_1y*100, 
                    Gut_3M_SE=sqrt(prop_3m*(1-prop_3m)/N_subj_G), 
                    Gut_1Y_SE=sqrt(prop_1y*(1-prop_1y)/N_subj_G))
props$OTU <- row.names(props)

# strep is 100 and 0 SE so no test done, automatic 100% persistent (NA p values for this one)
Gut_3m1y_Mcnem_p_df2 <- merge(Gut_3m1y_Mcnem_p_df, props, by="OTU", all=TRUE) 

# Calc change in prevalence
Gut_3m1y_Mcnem_p_df2$Gut_changeP_1Y3M <- Gut_3m1y_Mcnem_p_df2$Gut_1Y_P - Gut_3m1y_Mcnem_p_df2$Gut_3M_P 
```

Volcano plots aesthetics 
```{r}
# 1. The Mcnemar test appears to be very sensitive. Use a more strict p-value threshold in addition to an effect size cutoff to define Late and Early colonizers as oppose to persistent (no change) colonizers.

# 2. Adding additional annotation
# 3. labeling important/example early vs late taxa separately b/c they require differnt nudge_x's. 
Gut_3m1y_Mcnem_p_df2 <- Gut_3m1y_Mcnem_p_df2[order(-Gut_3m1y_Mcnem_p_df2$Gut_changeP_1Y3M),]
top_taxa <- Gut_3m1y_Mcnem_p_df2$OTU[1:3]
bot_taxa <- Gut_3m1y_Mcnem_p_df2$OTU[(nrow(Gut_3m1y_Mcnem_p_df2)-2):nrow(Gut_3m1y_Mcnem_p_df2)]

Gut_3m1y_Mcnem_p_df3 <- Gut_3m1y_Mcnem_p_df2 %>% mutate(
  Colonization_pattern_final = factor(ifelse(p.value.BH < 0.001 & Gut_changeP_1Y3M < -7, "Early",
                                      ifelse(p.value.BH < 0.001 & Gut_changeP_1Y3M >7, "Late", "Persistent")),
                                      levels = c("Early", "Persistent", "Late")),
  
  Label = ifelse(OTU %in% top_taxa, OTU, ""), Label2 = ifelse(OTU %in% bot_taxa, OTU, "")
)

# abbreviate long names and remove the "sp."
Gut_3m1y_Mcnem_p_df3$Label2 <- Gut_3m1y_Mcnem_p_df3$Label2 %>% 
                                  gsub("Enterococcus faecalis 1", "E. faecalis 1", .) %>% 
                                  gsub(" sensu stricto|sp\\.", "", .)
Gut_3m1y_Mcnem_p_df3$Label <- gsub("sp\\.|\\[|\\]", "", Gut_3m1y_Mcnem_p_df3$Label)
Gut_3m1y_Mcnem_p_df3$Label <- gsub("Ruminococcus", "R.", Gut_3m1y_Mcnem_p_df3$Label)
```

```{r}
### most taxa colonize later (65), and average change in prevalence for these late colonizers is 15%
data_summary(Gut_3m1y_Mcnem_p_df3, varname = "Gut_changeP_1Y3M", groupnames = "Colonization_pattern_final") 
```


```{r, warning=FALSE}
#ann_size <- 2.8
#axis_size <- 9
#ax_title_s <- 10

ann_size <- 2.8
axis_size <- 11
ax_title_s <- 11
top_taxa <- unique(Gut_3m1y_Mcnem_p_df3$Label)

Fig_3A_G <- ggplot(data=Gut_3m1y_Mcnem_p_df3, aes(x=Gut_changeP_1Y3M, y=-log10(p.value.BH), color=Colonization_pattern_final)) + 
  geom_point(size=2) + geom_hline(yintercept = -log10(0.001), color="black") + geom_vline(xintercept = 7, color="black")  + 
  geom_vline(xintercept = -7, color="black") + labs(color="Colonization", x="1 Year - 3 Month Prevalence (%)", y="-log10(FDR p-value)") + 
  ggtitle("Colonization Patterns Gut (n=555)")+ scale_x_continuous(limits = c(-23.5, 52)) + 
  geom_text(aes(label=ifelse(Label == top_taxa[1], Label, "")), nudge_y=0.5, nudge_x=10, color="black", size=ann_size, fontface=3) +
  geom_text(aes(label=ifelse(Label == top_taxa[2], Label, "")), nudge_y=0, nudge_x=10, color="black", size=ann_size, fontface=3) + 
  geom_text(aes(label=ifelse(Label == top_taxa[3], Label, "")), nudge_y=0.5, nudge_x=-5, color="black", size=ann_size, fontface=3) + 
  geom_text(aes(label=Label2), nudge_x = 8,nudge_y=0.8, color="black", size=ann_size, fontface=3) +
  theme(axis.text = element_text(size=axis_size), axis.title =element_text(size=ax_title_s),  title = element_text(size=11))
Fig_3A_G

tiff(file = paste0(path, "Figures/Fig_3A_Gut_Volcano_Overall.tiff"), width = 1550, height = 1370,  units = "px", res = 300) 
Fig_3A_G
dev.off()
```


# Prevalence Trajectory Coordinate System & Permutation tests - Comparing Prevalence Trajectories Between Groups (Fig. 3B-D)

## Nasal - Prevalence Trajectory derived variables, creation of a prevalence trajectory feature table
To simplify making of prevalence trajectory scatter plots, first make the 'difference in prevalence' results for every pair of samples for each infant (1Y - 3M, e.g. 0 - 1 = -1; 1 - 0 = 1) - so a "-1" means it's present at 3 months but not 1 year, 1 means 1 year but not 3m and 0 means no change.
```{r} 
## extract otu table and use Participant IDs as row.names instead - will match b/t datasets
N_3m_otu <- data.frame(as.data.frame(t(as(otu_table(N.pa.3m), "matrix"))), row.names=sample_data(N.pa.3m)$CHILDid)
N_1y_otu <- data.frame(as.data.frame(t(as(otu_table(N.pa.1y), "matrix"))), row.names=sample_data(N.pa.1y)$CHILDid)

N_3m_otu <- N_3m_otu[order(row.names(N_3m_otu)),]
N_1y_otu <- N_1y_otu[order(row.names(N_1y_otu)),]

## Checking - colnames and Childid's are in the same order? Yes - good - can move forward
identical(colnames(N_3m_otu), colnames(N_1y_otu))
identical(row.names(N_1y_otu), row.names(N_3m_otu))

## 1 Year minus 3 Month Prevalence For each Taxa
change_3m1y_pa <- N_1y_otu - N_3m_otu # -1 = present at 3m not 1y, 1= present at 1y not 3m, 0=either present at both or neither timpoint
dim(change_3m1y_pa) # Same order as original physeq obj

# Sample data to merge with dataset
sdf_toAdd <- sample_data(N.pa.1y)[order(sample_data(N.pa.1y)$CHILDid),]
sdf_3m <- sample_data(N.pa.3m)[order(sample_data(N.pa.3m)$CHILDid),]
## check that it's in the same order as change_3m1y_pa
identical(row.names(change_3m1y_pa), sdf_toAdd$CHILDid) 
identical(row.names(change_3m1y_pa), sdf_3m$CHILDid) 

### same order so just merge data with sample metadata using data.frame
change_3m1y_pa_sdf <- data.frame(change_3m1y_pa, sdf_toAdd, RE_virus_3m = sdf_3m$RE_Virus) # Using 1 year sdf for variables (will be same for 3m and 1y for almost all) - but also adding RE virus at 3 months as "RE_virus_3m". Note that "RE_Virus" is the RE virus at 1 year.

N_taxa <- ncol(change_3m1y_pa) # number of taxa, to use in downstream code
N_subj <- nrow(change_3m1y_pa) # number of subjects, to use in downstream code
```
Objects used for Nasal trajectory plots: change_3m1y_pa_sdf, Nasal_3M1Y_pa_filt, N_taxa, N_subj 

## Gut - Prevalence Trajectory derived variables, creation of a prevalence trajectory feature table
To simplify making of prevalence trajectory scatter plots, first make the 'difference in prevalence' results for every pair of samples for each infant (1Y - 3M, e.g. 0 - 1 = -1; 1 - 0 = 1) - so a "-1" means it's present at 3 months but not 1 year, 1 means 1 year but not 3m and 0 means no change.
```{r} 
## extract otu table and use Participant IDs as row.names instead - will match b/t datasets
G_3m_otu <- data.frame(as.data.frame(t(as(otu_table(G.pa.3m), "matrix"))), row.names=sample_data(G.pa.3m)$CHILDid)
G_1y_otu <- data.frame(as.data.frame(t(as(otu_table(G.pa.1y), "matrix"))), row.names=sample_data(G.pa.1y)$CHILDid)

G_3m_otu <- G_3m_otu[order(row.names(G_3m_otu)),]
G_1y_otu <- G_1y_otu[order(row.names(G_1y_otu)),]

## Checking - colnames and Childid's are in the same order? Yes - good - can move forward
identical(colnames(G_3m_otu), colnames(G_1y_otu))
identical(row.names(G_3m_otu), row.names(G_1y_otu))

## 1 Year minus 3 Month Prevalence For each Taxa
change_3m1y_pa_G <- G_1y_otu - G_3m_otu # -1 = present at 3m not 1y, 1= present at 1y not 3m, 0=either present at both or neither timpoint
dim(change_3m1y_pa_G) # Same order as original physeq obj

sdf_toAdd <- sample_data(G.pa.1y)[order(sample_data(G.pa.1y)$CHILDid),]
sdf_3m <- sample_data(G.pa.3m)[order(sample_data(G.pa.3m)$CHILDid),]
## check that it's in the same order as change_3m1y_pa
identical(row.names(change_3m1y_pa_G), sdf_toAdd$CHILDid) # order matches
identical(row.names(change_3m1y_pa_G), sdf_3m$CHILDid) # order matches

### same order so just merge data with sample metadata using data.frame
change_3m1y_pa_G_sdf <- data.frame(change_3m1y_pa_G, sdf_toAdd, RE_virus_3m = sdf_3m$RE_Virus) # Using 1 year sdf for variables but also adding RE virus at 3 months as "RE_virus_3m". Note that "RE_Virus" is the RE virus at 1 year

N_taxa_G <- ncol(change_3m1y_pa_G) # number of taxa, to use in downstream code
N_subj_G <- nrow(change_3m1y_pa_G) # number of subjects, to use in downstream code
```
Objects used for gut trajectory plots: change_3m1y_pa_G_sdf, Gut_3M1Y_pa_filt, N_taxa_G, N_subj_G 

## Save prevalence trajectory feature tables for later use (in mediation analysis, Fig. 6)
```{r}
## don't save with metadata - just save new derived data with CHILDid's as rownames: change_3m1y_pa, change_3m1y_pa_G; 
### merge before saving, but first differentiate nasal vs gut trajectory features by prepending N or G
colnames(change_3m1y_pa) <- gsub("^", "N_", colnames(change_3m1y_pa))
colnames(change_3m1y_pa_G) <- gsub("^", "G_", colnames(change_3m1y_pa_G))

change_3m1y_pa_all <- merge(change_3m1y_pa, change_3m1y_pa_G, by="row.names", all=TRUE)
row.names(change_3m1y_pa_all) <- change_3m1y_pa_all$Row.names
change_3m1y_pa_all$Row.names <- NULL
dim(change_3m1y_pa_all)
write.csv(change_3m1y_pa_all, file=paste0(path, "Data/Data_subsets/Prevalence_Trajectory_Features_N171_G172.csv")) # The first 171 columns are nasal taxa, 172 to end of df is gut
```

```{r} 
# setup for plots
## for plot annotations:
Main_phyla <- c('Proteobacteria','Firmicutes', "Actinobacteria", "Bacteroidetes")
cbPalette <- c("navy", "paleturquoise4",  "violetred3", "skyblue1", "goldenrod3")
```

## Nasal - Breastfeeding - Figure 3C
```{r} 
X <- subset(change_3m1y_pa_sdf, BF_status_3m_sampling == "No BM")
Y <- subset(change_3m1y_pa_sdf, BF_status_3m_sampling == "Exclusive")

# Make data.frame of prevalence like before BUT keeping as proprotions instead of Percentages for now
df.pa.nasal.BF <- data.frame(X_P_Change=colSums(X[1:N_taxa])*100/dim(X)[1],Y_P_Change=colSums(Y[1:N_taxa])*100/dim(Y)[1],
                             taxonomy=as(tax_table(Nasal_3M1Y_pa_filt), "matrix"),
                             prevalence=taxa_sums(Nasal_3M1Y_pa_filt)*100/(N_subj*2)) # Taxonomy order actually maintained compared to physeq obj Nasal_3M1Y_pa_filt so can merge this way. prevalence column is the total prevalence so divided by total N across both timepoints (subject # x 2). 

# difference in change
df.pa.nasal.BF$pa.change.diff <- df.pa.nasal.BF$Y_P_Change - df.pa.nasal.BF$X_P_Change
```
Aesthetic things
```{r} 
# Plot by major phylum & annotate signif taxa
df.pa.nasal.BF$Phylum <- ifelse(df.pa.nasal.BF$taxonomy.Phylum %in% Main_phyla, df.pa.nasal.BF$taxonomy.Phylum, "Other")
#Reorder the phyla 
df.pa.nasal.BF$Phylum = factor(df.pa.nasal.BF$Phylum, levels=c(Main_phyla, "Other"))
```
Plot Change in Prevalence
```{r} 
x_lab <- paste("No BM (n=", nrow(X), ")\n1 Year - 3 Month Prevalence (%)", sep="")
y_lab <- paste("Exclusive BM (n=", nrow(Y), ")\n1 Year - 3 Month Prevalence (%)", sep="")

Fig_BF_N <- ggplot(data=df.pa.nasal.BF, aes(x=X_P_Change, y=Y_P_Change)) + 
  geom_point(aes(color=Phylum, size=prevalence), alpha=0.75) + 
  xlab(x_lab) + ylab(y_lab) + 
  scale_color_manual(values = cbPalette) +
  scale_y_continuous(limits=c(-30,60.6), expand = c(0,0))+
  scale_x_continuous(limits=c(-30,60.6), expand = c(0,0))+
  annotate("segment", x = -30, xend = 60, y = -30, yend = 60, colour = "red", linewidth=0.5) + ggtitle("PreTCO System - Nasal")
Fig_BF_N
```
Median change in prevalence of an OTU in BF and non-BF infants
```{r} 
# no BM
summary(df.pa.nasal.BF$X_P_Change)
# BM
summary(df.pa.nasal.BF$Y_P_Change)
```
**Permutation test for Prevalence Trajectory analysis - See 'Functions.R' for perm_func_dfs function**
```{r} 
Res_BF_N <- perm_func_dfs(df.pa.nasal.BF, change_3m1y_pa_sdf, 
                           variable = "BF_status_3m_sampling", Group_X = "No BM", Group_Y = "Exclusive", N_taxa = N_taxa)

# Test output
Res_BF_N$global_res # Overall significance test
# Res_per_taxa_BF_N - p-value assigned to each taxa; p_value = later in healthy; p_value_2 = earlier in healthy
Res_per_taxa_BF_N <- Res_BF_N$per_taxa_res
```

## Gut - Breastfeeding - Figure 3C
```{r} 
X <- subset(change_3m1y_pa_G_sdf, BF_status_3m_sampling == "No BM")
Y <- subset(change_3m1y_pa_G_sdf, BF_status_3m_sampling == "Exclusive")

# Make data.frame of prevalence like before BUT keeping as proprotions instead of Percentages for now
df.pa.gut.BF <- data.frame(X_P_Change=colSums(X[1:N_taxa_G])*100/dim(X)[1],Y_P_Change=colSums(Y[1:N_taxa_G])*100/dim(Y)[1], taxonomy=as(tax_table(Gut_3M1Y_pa_filt), "matrix"), prevalence=taxa_sums(Gut_3M1Y_pa_filt)*100/(N_subj_G*2)) # Taxonomy actually maintained compared to physeq obj, great

# difference in change
df.pa.gut.BF$pa.change.diff <- df.pa.gut.BF$Y_P_Change - df.pa.gut.BF$X_P_Change
```
Asthetic things, not important
```{r} 
## Plot by major phylum & annotate signif taxa
df.pa.gut.BF$Phylum <- ifelse(df.pa.gut.BF$taxonomy.Phylum %in% Main_phyla, df.pa.gut.BF$taxonomy.Phylum, "Other")
#Reorder the phyla 
df.pa.gut.BF$Phylum = factor(df.pa.gut.BF$Phylum, levels=c(Main_phyla, "Other"))
```
Plot Change in Prevalence 
```{r} 
x_lab <- paste("No BM (n=", nrow(X), ")\n1 Year - 3 Month Prevalence (%)", sep="")
y_lab <- paste("Exclusive BM (n=", nrow(Y), ")\n1 Year - 3 Month Prevalence (%)", sep="")
#df.pa.gut.BF$Label <- ifelse(row.names(df.pa.gut.BF) == "X.Ruminococcus..gnavus..sp.", "R. gnavus", "")

Fig_BF_G <- ggplot(data = df.pa.gut.BF, aes(x=X_P_Change, y=Y_P_Change)) + 
  geom_point(aes(color=Phylum, size=prevalence), alpha=0.75) + 
  xlab(x_lab) + ylab(y_lab) + 
  #geom_text(aes(label=Label), nudge_x = 0,nudge_y=0.8, color="black", size=5, fontface=3) +
  scale_color_manual(values = cbPalette) +
  scale_y_continuous(limits=c(-30,60), expand = c(0,0))+
  scale_x_continuous(limits=c(-30,60), expand = c(0,0))+
  annotate("segment", x = -30, xend = 60, y = -30, yend = 60, colour = "red", linewidth=0.5) + ggtitle("PreTCO System - Gut")
Fig_BF_G
```
Median change in prevalence of an OTU in BF and non-BF infants
```{r} 
# no BM
summary(df.pa.gut.BF$X_P_Change)

# BM
summary(df.pa.gut.BF$Y_P_Change)
```

**Permutation test for Prevalence Trajectory analysis**
```{r} 
Res_BF_G <- perm_func_dfs(df.pa.gut.BF, change_3m1y_pa_G_sdf, 
                           variable = "BF_status_3m_sampling", Group_X = "No BM", Group_Y = "Exclusive", N_taxa = N_taxa_G)

# Test output
Res_BF_G$global_res # Overall significance test
#  p-value assigned to each taxa; p_value = later in healthy; p_value_2 = earlier in healthy
Res_per_taxa_BF_G <- Res_BF_G$per_taxa_res
```

## Nasal - Asthma at 3 years - Figure 3C
```{r} 
X <- subset(change_3m1y_pa_sdf, asthma_3y == "Probable")
Y <- subset(change_3m1y_pa_sdf, asthma_3y == "No")

# Make data.frame of prevalence like before BUT keeping as proprotions instead of Percentages for now
df.pa.nasal.ast <- data.frame(X_P_Change=colSums(X[1:N_taxa])*100/dim(X)[1],Y_P_Change=colSums(Y[1:N_taxa])*100/dim(Y)[1], taxonomy=as(tax_table(Nasal_3M1Y_pa_filt), "matrix"), prevalence=taxa_sums(Nasal_3M1Y_pa_filt)*100/(N_subj*2)) 

# difference in change
df.pa.nasal.ast$pa.change.diff <- df.pa.nasal.ast$Y_P_Change - df.pa.nasal.ast$X_P_Change
```
Aesthetic things
```{r} 
# Plot by major phylum & annotate signif taxa
df.pa.nasal.ast$Phylum <- ifelse(df.pa.nasal.ast$taxonomy.Phylum %in% Main_phyla, df.pa.nasal.ast$taxonomy.Phylum, "Other")
#Reorder the phyla 
df.pa.nasal.ast$Phylum = factor(df.pa.nasal.ast$Phylum, levels=c(Main_phyla, "Other"))
```
Plot Change in Prevalence
```{r} 
x_lab <- paste("Asthma (n=", nrow(X), ")\n1 Year - 3 Month Prevalence (%)", sep="")
y_lab <- paste("No Asthma (n=", nrow(Y), ")\n1 Year - 3 Month Prevalence (%)", sep="")

Fig_ast_N <- ggplot(data=df.pa.nasal.ast, aes(x=X_P_Change, y=Y_P_Change)) + 
  geom_point(aes(color=Phylum, size=prevalence), alpha=0.75) + 
  xlab(x_lab) + ylab(y_lab) + 
  scale_color_manual(values = cbPalette) +
  scale_y_continuous(limits=c(-30,60), expand = c(0,0))+
  scale_x_continuous(limits=c(-30,60), expand = c(0,0))+
  annotate("segment", x = -30, xend = 60, y = -30, yend = 60, colour = "red", linewidth=0.5) + ggtitle("PreTCO System - Nasal")
Fig_ast_N
```
Median change in prevalence of an OTU in asthma and non-asthma infants
```{r} 
# Asthma
summary(df.pa.nasal.ast$X_P_Change)

# No Asthma
summary(df.pa.nasal.ast$Y_P_Change)
```

**Permutation test for Prevalence Trajectory analysis**
```{r} 
Res_ast_N <- perm_func_dfs(df.pa.nasal.ast, change_3m1y_pa_sdf, 
                           variable = "asthma_3y", Group_X = "Probable", Group_Y = "No", N_taxa = N_taxa)

# Overall significance test for nasal, asthma at 3 years test
Res_ast_N$global_res # sig
# Res_per_taxa_ast_N - p-value assigned to each taxa; p_value = later in healthy; p_value_2 = earlier in healthy
Res_per_taxa_ast_N <- Res_ast_N$per_taxa_res
```

## Gut - Asthma at 3 years - Figure 3C
```{r} 
X <- subset(change_3m1y_pa_G_sdf, asthma_3y == "Probable")
Y <- subset(change_3m1y_pa_G_sdf, asthma_3y == "No")

# Make data.frame of prevalence like before BUT keeping as proprotions instead of Percentages for now
df.pa.gut.ast <- data.frame(X_P_Change=colSums(X[1:N_taxa_G])*100/dim(X)[1],Y_P_Change=colSums(Y[1:N_taxa_G])*100/dim(Y)[1], taxonomy=as(tax_table(Gut_3M1Y_pa_filt), "matrix"), prevalence=taxa_sums(Gut_3M1Y_pa_filt)*100/(N_subj_G*2)) 

# difference in change
df.pa.gut.ast$pa.change.diff <- df.pa.gut.ast$Y_P_Change - df.pa.gut.ast$X_P_Change
```
Aesthetic things, not important
```{r} 
# Plot by major phylum & annotate signif taxa
df.pa.gut.ast$Phylum <- ifelse(df.pa.gut.ast$taxonomy.Phylum %in% Main_phyla, df.pa.gut.ast$taxonomy.Phylum, "Other")
#Reorder the phyla 
df.pa.gut.ast$Phylum = factor(df.pa.gut.ast$Phylum, levels=c(Main_phyla, "Other"))
```
Plot Change in Prevalence 
```{r} 
x_lab <- paste("Asthma (n=", nrow(X), ")\n1 Year - 3 Month Prevalence (%)", sep="")
y_lab <- paste("No Asthma (n=", nrow(Y), ")\n1 Year - 3 Month Prevalence (%)", sep="")
#df.pa.gut.ast$Label <- ifelse(row.names(df.pa.gut.ast) == "X.Ruminococcus..gnavus..sp.", "R. gnavus", "")

Fig_ast_G <- ggplot(data=df.pa.gut.ast, aes(x=X_P_Change, y=Y_P_Change, color=Phylum, size=prevalence)) + 
  geom_point(alpha=0.75) + 
  xlab(x_lab) + ylab(y_lab) + 
  #geom_text(aes(label=Label), nudge_x = 0,nudge_y=0.8, color="black", size=5, fontface=3) +
  scale_color_manual(values = cbPalette) +
  scale_y_continuous(limits=c(-30,60), expand = c(0,0))+
  scale_x_continuous(limits=c(-30,60), expand = c(0,0))+
  annotate("segment", x = -30, xend = 60, y = -30, yend = 60, colour = "red", linewidth=0.25) + ggtitle("PreTCO System - Gut")
Fig_ast_G
```
Median change in prevalence of an OTU in BF and non-BF infants
```{r} 
# Asthma
summary(df.pa.gut.ast$X_P_Change)

# No Asthma
summary(df.pa.gut.ast$Y_P_Change)
```

**Permutation test for Prevalence Trajectory analysis**
```{r} 
Res_ast_G <- perm_func_dfs(df.pa.gut.ast, change_3m1y_pa_G_sdf, 
                           variable = "asthma_3y", Group_X = "Probable", Group_Y = "No", N_taxa = N_taxa_G)

# Test output
Res_ast_G$global_res # Overall significance test
# p-value assigned to each taxa; p_value = later in healthy; p_value_2 = earlier in healthy
Res_per_taxa_ast_G <- Res_ast_G$per_taxa_res
```


## Extra Prevalence Trajectory analyses (Not shown but used for Figure 3D)

### Asthma at 3 years that persists to 5 years vs. no asthma at 3 and no asthma at 5 years
```{r} 
# want probable asthma at 3 years, and probable or possible at 5 years. Those missing 5 year or 3 year asthma data will be NA for this variable
change_3m1y_pa_sdf$asthma_3y_5y <- ifelse(change_3m1y_pa_sdf$asthma_3y == "Probable" & 
                                            change_3m1y_pa_sdf$asthma_5y %in% c("Probable", "Possible"), "Yes", 
                                          ifelse(change_3m1y_pa_sdf$asthma_3y == "No" & 
                                            change_3m1y_pa_sdf$asthma_5y == "No", "No", NA))
table(change_3m1y_pa_sdf$asthma_3y_5y)

# want probable asthma at 3 years, and probable or possible at 5 years. Those missing 5 year or 3 year asthma data will be NA for this variable  
change_3m1y_pa_G_sdf$asthma_3y_5y <- ifelse(change_3m1y_pa_G_sdf$asthma_3y == "Probable" & 
                                            change_3m1y_pa_G_sdf$asthma_5y  %in% c("Probable", "Possible"), "Yes", 
                                          ifelse(change_3m1y_pa_G_sdf$asthma_3y == "No" & 
                                            change_3m1y_pa_G_sdf$asthma_5y == "No", "No", NA))

table(change_3m1y_pa_G_sdf$asthma_3y_5y)
```

### Nasal - Asthma at 3 years that persists to 5 years
```{r} 
X <- subset(change_3m1y_pa_sdf, asthma_3y_5y == "Yes")
Y <- subset(change_3m1y_pa_sdf, asthma_3y_5y == "No")

# Make data.frame of prevalence like before BUT keeping as proprotions instead of Percentages for now
df.pa.nasal.ast.3y5y <- data.frame(X_P_Change=colSums(X[1:N_taxa])*100/dim(X)[1],Y_P_Change=colSums(Y[1:N_taxa])*100/dim(Y)[1], taxonomy=as(tax_table(Nasal_3M1Y_pa_filt), "matrix"), prevalence=taxa_sums(Nasal_3M1Y_pa_filt)*100/(N_subj*2)) 

# difference in change
df.pa.nasal.ast.3y5y$pa.change.diff <- df.pa.nasal.ast.3y5y$Y_P_Change - df.pa.nasal.ast.3y5y$X_P_Change
```
Aesthetic things
```{r} 
## Plot by major phylum & annotate signif taxa
df.pa.nasal.ast.3y5y$Phylum <- ifelse(df.pa.nasal.ast.3y5y$taxonomy.Phylum %in% Main_phyla, df.pa.nasal.ast.3y5y$taxonomy.Phylum, "Other")
#Reorder the phyla 
df.pa.nasal.ast.3y5y$Phylum = factor(df.pa.nasal.ast.3y5y$Phylum, levels=c(Main_phyla, "Other"))
```
Plot Change in Prevalence
```{r} 
x_lab <- paste("Asthma, 3 & 5 Years (n=", nrow(X), ")\n1 Year - 3 Month Prevalence (%)", sep="")
y_lab <- paste("No Asthma (n=", nrow(Y), ")\n1 Year - 3 Month Prevalence (%)", sep="")

Fig_3y5y_N <- ggplot(data=df.pa.nasal.ast.3y5y, aes(x=X_P_Change, y=Y_P_Change)) + 
  geom_point(aes(color=Phylum, size=prevalence), alpha=0.75) + 
  xlab(x_lab) + ylab(y_lab) + 
  scale_color_manual(values = cbPalette) +
  scale_y_continuous(limits=c(-30,60), expand = c(0,0))+
  scale_x_continuous(limits=c(-30,60), expand = c(0,0))+
  annotate("segment", x = -30, xend = 60, y = -30, yend = 60, colour = "red", linewidth=0.5) + ggtitle("PreTCO System - Nasal")
Fig_3y5y_N
```
Median change in prevalence of an OTU in asthma and non-asthma infants
```{r} 
# Asthma
summary(df.pa.nasal.ast.3y5y$X_P_Change)

# No Asthma
summary(df.pa.nasal.ast.3y5y$Y_P_Change)
```

**Permutation test for Prevalence Trajectory analysis**
```{r} 
Res_3y5yast_N <- perm_func_dfs(df.pa.nasal.ast.3y5y, change_3m1y_pa_sdf, 
                           variable = "asthma_3y_5y", Group_X = "Yes", Group_Y = "No", N_taxa = N_taxa)

# Test output
Res_3y5yast_N$global_res # Overall significance test
#  p-value assigned to each taxa; p_value = later in healthy; p_value_2 = earlier in healthy
Res_per_taxa_ast3y5y_N <- Res_3y5yast_N$per_taxa_res
```


### Gut - Asthma at 3 years that persists to 5 years
```{r} 
X <- subset(change_3m1y_pa_G_sdf, asthma_3y_5y == "Yes")
Y <- subset(change_3m1y_pa_G_sdf, asthma_3y_5y == "No")

# Make data.frame of prevalence like before BUT keeping as proprotions instead of Percentages for now
df.pa.gut.ast.3y5y <- data.frame(X_P_Change=colSums(X[1:N_taxa_G])*100/dim(X)[1],Y_P_Change=colSums(Y[1:N_taxa_G])*100/dim(Y)[1], taxonomy=as(tax_table(Gut_3M1Y_pa_filt), "matrix"), prevalence=taxa_sums(Gut_3M1Y_pa_filt)*100/(N_subj_G*2)) 

# difference in change
df.pa.gut.ast.3y5y$pa.change.diff <- df.pa.gut.ast.3y5y$Y_P_Change - df.pa.gut.ast.3y5y$X_P_Change
```
Aesthetic things, not important
```{r} 
df.pa.gut.ast.3y5y$Phylum <- ifelse(df.pa.gut.ast.3y5y$taxonomy.Phylum %in% c("Proteobacteria", "Actinobacteria", "Firmicutes", "Bacteroidetes"),
                                      df.pa.gut.ast.3y5y$taxonomy.Phylum, "Other")
#Reorder the phyla 
df.pa.gut.ast.3y5y$Phylum = factor(df.pa.gut.ast.3y5y$Phylum, levels=c('Proteobacteria','Firmicutes', "Actinobacteria", "Bacteroidetes", "Other"))
```
Plot Change in Prevalence 
```{r} 
x_lab <- paste("Asthma, 3 & 5 Years (n=", nrow(X), ")\n1 Year - 3 Month Prevalence (%)", sep="")
y_lab <- paste("No Asthma (n=", nrow(Y), ")\n1 Year - 3 Month Prevalence (%)", sep="")

Fig_3y5y_G <- ggplot(data=df.pa.gut.ast.3y5y, aes(x=X_P_Change, y=Y_P_Change)) + 
  geom_point(aes(color=Phylum, size=prevalence), alpha=0.75) + 
  xlab(x_lab) + ylab(y_lab) + 
  scale_color_manual(values = cbPalette) +
  scale_y_continuous(limits=c(-32,60), expand = c(0,0))+
  scale_x_continuous(limits=c(-32,60), expand = c(0,0))+
  annotate("segment", x = -32, xend = 60, y = -32, yend = 60, colour = "red", linewidth=0.25) + ggtitle("PreTCO System - Gut")
Fig_3y5y_G
```
Median change in prevalence of an OTU 
```{r} 
# Asthma
summary(df.pa.gut.ast.3y5y$X_P_Change)

# No Asthma
summary(df.pa.gut.ast.3y5y$Y_P_Change)
```

**Permutation test for Prevalence Trajectory analysis**
```{r} 
Res_3y5yast_G <- perm_func_dfs(df.pa.gut.ast.3y5y, change_3m1y_pa_G_sdf, 
                           variable = "asthma_3y_5y", Group_X = "Yes", Group_Y = "No", N_taxa = N_taxa_G)

# Test output
 # Overall significance test
Res_3y5yast_G$global_res
#  p-value assigned to each taxa; p_value = later in healthy; p_value_2 = earlier in healthy
Res_per_taxa_ast3y5y_G <- Res_3y5yast_G$per_taxa_res
```

```{r, include=FALSE}
# Save prevalence trajectory scatter plots, remove axis titles and re-add later

## saving a legend (all will be the same)
legend_PT <- plot_grid(get_legend(Fig_ast_N), ncol = 1)
tiff(file = paste0(path, "Figures/Fig_3C_Legend_PrevTraj.tiff"), width = 500, height = 1000, units = "px", res = 300) 
legend_PT
dev.off()


## BF exclusivity
Fig_BF_N + theme(legend.position = "none") 
Fig_4A_N_2 <- Fig_BF_N + theme(legend.position = "none", axis.title.x = element_blank(), axis.title.y = element_blank()) + ggtitle("")
tiff(file = paste0(path, "Figures/Fig_3C_Nasal_BF.tiff"), width = 1275, height = 1275, units = "px", res = 300) 
Fig_4A_N_2
dev.off()


Fig_4A_G_2 <- Fig_BF_G + theme(legend.position = "none", axis.title.x = element_blank(), axis.title.y = element_blank()) + ggtitle("")
tiff(file = paste0(path, "Figures/Fig_3C_Gut_BF.tiff"), width = 1275, height = 1275, units = "px", res = 300) 
Fig_4A_G_2 
dev.off()


## Asthma
Fig_3C_N_2 <- Fig_ast_N + theme(legend.position = "none", axis.title.x = element_blank(), axis.title.y = element_blank()) + ggtitle("")
tiff(file = paste0(path, "Figures/Fig_3D_Nasal_Asthma.tiff"), width = 1275, height = 1275, units = "px", res = 300) 
Fig_3C_N_2
dev.off()

Fig_3C_G_2 <- Fig_ast_G + theme(legend.position = "none", axis.title.x = element_blank(), axis.title.y = element_blank()) + ggtitle("")
tiff(file = paste0(path, "Figures/Fig_3D_Gut_Asthma.tiff"), width = 1275, height = 1275, units = "px", res = 300) 
Fig_3C_G_2
dev.off()


## Additional version - Make grid, control plot margins
t <- 1
r <- 0.5
b <- 0
l <- 0.2

tiff(file = paste0(path, "Figures/Fig_3C_D_merged.tiff"), width = 2200, height = 2200, units = "px", res = 300) 
plot_grid(Fig_ast_N + theme(legend.position = "none", plot.margin = unit(c(t, r, b, l), "cm")) + 
            labs(y = "Healthy\n1 Year - 3 Month Prevalence (%)", x =  "Asthma, 3 Years\n1 Year - 3 Month Prevalence (%)"),
          
          Fig_ast_G + theme(legend.position = "none", plot.margin = unit(c(t, r, b, l), "cm")) + labs(y = " \n ", x =  "Asthma, 3 Years\n1 Year - 3 Month Prevalence (%)"),
          
          Fig_BF_N + theme(legend.position = "none", plot.margin = unit(c(t, r, b, l), "cm")) + labs(y = "Exclusive BM\n1 Year - 3 Month Prevalence (%)", x =  "No BM\n1 Year - 3 Month Prevalence (%)"),
          Fig_BF_G + theme(legend.position = "none", plot.margin = unit(c(t, r, b, l), "cm")) + labs(y = " \n ", x =  "No BM\n1 Year - 3 Month Prevalence (%)"))
dev.off()

tiff(file = paste0(path, "Figures/Extra.tiff"), width = 2200, height = 2200, units = "px", res = 300) 
Fig_G_leg <- get_legend(Fig_ast_G + labs(y = " \n ", x =  "Asthma, 3 Years\n1 Year - 3 Month Prevalence (%)", size = "Prevalence") +
                          theme(legend.title = element_text(face = "bold")))
plot_grid(Fig_G_leg)
dev.off()
```


# Compile all prevalence trajectory results for heatmap and tables 
```{r}
# compile & save results for individ taxa

# saving everything to output
Per_test_df_ls <- list(Res_per_taxa_ast_N, Res_per_taxa_BF_N, 
                       Res_per_taxa_ast3y5y_N, 
                       Res_per_taxa_ast_G, Res_per_taxa_BF_G, Res_per_taxa_ast3y5y_G)
names(Per_test_df_ls) <-  c("Ast3y_Nasal", "BF_Nasal", "Ast3y_5y_Nasal", 
                            "Ast3y_Gut", "BF_Gut", "Ast3y_5y_Gut")

# Save as an overall rds
saveRDS(Per_test_df_ls, paste0(path, "Tables/PermTests/MAIN_Permutation_PrevTraj_Res.rds"))
##Per_test_df_ls <- readRDS(paste0(path, "Tables/PermTests/MAIN_Permutation_PrevTraj_Res.rds"))

# Save as individual csv files for others 
outpath_Pt1 <- paste0(path, "Tables/PermTests/Permutation_PrevTraj_Res")
outfiles <- paste(outpath_Pt1, names(Per_test_df_ls), ".csv", sep = "_")

mapply(function(x, y) write.csv(x, file=y), x=Per_test_df_ls, y=outfiles, SIMPLIFY = FALSE)
```

# Figure 3D - Heatmaps showing Taxa with Prevalence Trajectories that Differ by Asthma, compared to differences by other Phenotypes

### setups for downstream heatmaps
```{r}
## to make italic for just part of x axis, using extra packages...
library(stringr)
library(glue)
library(ggtext)

# comparisons - column names and label name matches in desired order 
comparisons <- c(Asthma3y5y = "No Asthma\nvs. Asthma\n[3 & 5 years]", 
                 BF = "Exclusive\nvs. No BM", 
                 Asthma = "No Asthma\nvs. Asthma\n[3 years]")
comparisons

# heatmap annotation plot.margin parameters - top, right, etc. 
t <- 0.2
r <- 0
b <- 4.4
l <- 0
```

## Nasal heatmap
Making dataframe showing p-values < 0.05 for Later colonizers as "1", Earlier colonizers as "-1"; and non-significant taxa as "0"
```{r}
# for nasal, all dataframes have the same order / generated in the same way - so can use data.frame to merge
## Binary dataframe for Late taxa < 0.05
p_vals_Binary_N <- data.frame(row.names = row.names(Per_test_df_ls$Ast3y_Nasal),
                           Asthma = ifelse(Per_test_df_ls$Ast3y_Nasal$p_value < 0.05, 1, 
                                           ifelse(Per_test_df_ls$Ast3y_Nasal$p_value_2 < 0.05, -1, 0)),
                           BF = ifelse(Per_test_df_ls$BF_Nasal$p_value < 0.05, 1, 
                                       ifelse(Per_test_df_ls$BF_Nasal$p_value_2 < 0.05, -1, 0)),
                           Asthma3y5y = ifelse(Per_test_df_ls$Ast3y_5y_Nasal$p_value < 0.05, 1, 
                                               ifelse(Per_test_df_ls$Ast3y_5y_Nasal$p_value_2 < 0.05, -1, 0))
                           )

## don't like how rownames were modified - change back using ref dataset - df.pa.nasal.ast
p_vals_Binary_N2 <- merge(p_vals_Binary_N, subset(df.pa.nasal.ast, select = c("taxonomy.ASV_taxonomy")), by="row.names")
p_vals_Binary_N2$Row.names <- NULL

# Also for ComplexHeatmap, can only have names as rownames
row.names(p_vals_Binary_N2) <- p_vals_Binary_N2$taxonomy.ASV_taxonomy
p_vals_Binary_N2$taxonomy.ASV_taxonomy <- NULL
```

```{r}
# Comparing all against those taxa significant for Asthma
p_vals_Binary_N_Ast <- subset(p_vals_Binary_N2, Asthma == 1 |  Asthma == -1) %>% 
                          ### Taxa as x axis, abbreviate names
                          mutate(Taxa = gsub("\\ sp\\.|Unclassified ", "", row.names(.)))

## reorder rows - first alphabetical, then main order by BF and Asthma 1
p_vals_Binary_N_Ast <- p_vals_Binary_N_Ast[order(-p_vals_Binary_N_Ast$Asthma, -p_vals_Binary_N_Ast$BF, -p_vals_Binary_N_Ast$Asthma3y5y, p_vals_Binary_N_Ast$Taxa),]
## want factor levels to match this order after sorting
p_vals_Binary_N_Ast$Taxa <- factor(p_vals_Binary_N_Ast$Taxa, levels = unique(p_vals_Binary_N_Ast$Taxa))
```

Heatmap showing the direction of association 
```{r, fig.width=10}
p_vals_Binary_N_Ast_m <- p_vals_Binary_N_Ast %>% 
  melt(., id.vars = "Taxa") %>%
  mutate(Comparison = revalue(factor(variable, levels = names(comparisons), 
               comparisons)),
         italic_x = gsub(" \\d{1,}$", "", Taxa),
         non_italic_x = ifelse(is.na(str_extract(Taxa, "\\d{1,}$") ), "", 
                               str_extract(Taxa, "\\d{1,}$") )
         ) %>%
         mutate(Taxa_2 = glue("<i>{italic_x}</i> {non_italic_x}"))

## factor levels were not maintained but...
unique(p_vals_Binary_N_Ast_m$Taxa) ## df already ordered how I want re. factor levels for taxa - but now want this for  Taxa_2 - should be easy
p_vals_Binary_N_Ast_m$Taxa_2 = factor(p_vals_Binary_N_Ast_m$Taxa_2, levels=unique(p_vals_Binary_N_Ast_m$Taxa_2))

main_h <- ggplot(p_vals_Binary_N_Ast_m, aes(Taxa_2, Comparison, fill = value)) +
                    geom_tile() + 
                    scale_fill_gradient2("",  low = "#ff8a90", mid = "#b3b3b3", high = "skyblue1",
                      midpoint = 0, limits=c(-1, 1), breaks=c(-1, 0, 1), labels = c("Earlier", "p > 0.05", "Later")) +
                    geom_hline(yintercept = 1.5, color = "#5e5d5d", linewidth = 0.5) +
                    geom_hline(yintercept = 2.5, color = "#5e5d5d", linewidth = 0.5) +
                    theme(axis.text.x= element_markdown(size=10, angle=90, vjust=0.5, hjust=1, color = "black"), 
                          axis.text.y = element_text(size = 10),
                          axis.title = element_blank(), legend.position = "none", axis.ticks = element_blank()
                          ) + scale_y_discrete(expand = c(0,0)) + scale_x_discrete(expand = c(0,0))

# Make side annotation: showing the percent taxa that are Late with no Asthma, that are also 'late' with other factors
## of the 41 that are sig for Asthma, 36 are Late w/t no asthma - 36 is the total
Direction_sig_Asthma_L <- subset(p_vals_Binary_N_Ast, Asthma == 1)

n_ast_tax <- nrow(Direction_sig_Asthma_L)
Pct_late <- data.frame(Asthma=100, 
                       BF=table(Direction_sig_Asthma_L$BF)[3]*100/n_ast_tax,
                       Asthma3y5y=table(Direction_sig_Asthma_L$Asthma3y5y)[2]*100/n_ast_tax) %>% 
                      t %>% as.data.frame %>% mutate(variable = factor(row.names(.), levels = names(comparisons)),
                                                     x_axis = "percent_late") 
Pct_late$pct <- Pct_late$'1' # Add annotations too: 36/36, 17/36, 12/36

annotation_h <- ggplot(Pct_late, aes(x_axis, variable, fill = pct)) +
                    geom_tile() + 
                    scale_fill_gradient2("",  low = "white", mid = "#f7b0bc", high = "red", midpoint = 50,limits = c(0, 100))+
                    theme(axis.text = element_blank(),
                          axis.title = element_blank(), legend.position = "none", axis.ticks = element_blank(),
                          plot.margin = unit(c(t, r, b, l), "cm")) + 
                    scale_y_discrete(expand = c(0,0)) + scale_x_discrete(expand = c(0,0))

Fig_3E_N <- plot_grid(main_h, annotation_h, ncol = 2, rel_widths = c(1, 0.02))
Fig_3E_N
```

## Gut Heatmap
Making dataframe showing p-values < 0.05 for Later colonizers as "1", Earlier colonizers as "-1"; and non-significant taxa as "0"
```{r}
# for nasal, all dataframes have the same order / generated in the same way - so can use data.frame to merge
## Binary dataframe for Late taxa < 0.05
p_vals_Binary_G <- data.frame(row.names = row.names(Per_test_df_ls$Ast3y_Gut),
                           Asthma = ifelse(Per_test_df_ls$Ast3y_Gut$p_value < 0.05, 1, 
                                           ifelse(Per_test_df_ls$Ast3y_Gut$p_value_2 < 0.05, -1, 0)),
                           BF = ifelse(Per_test_df_ls$BF_Gut$p_value < 0.05, 1, 
                                       ifelse(Per_test_df_ls$BF_Gut$p_value_2 < 0.05, -1, 0)),
                           Asthma3y5y = ifelse(Per_test_df_ls$Ast3y_5y_Gut$p_value < 0.05, 1,
                                               ifelse(Per_test_df_ls$Ast3y_5y_Gut$p_value_2 < 0.05, -1, 0))
                           )

## don't like how rownames were modified - change back using ref dataset - df.pa.gut.ast
p_vals_Binary_G2 <- merge(p_vals_Binary_G, subset(df.pa.gut.ast, select = c("taxonomy.ASV_taxonomy")), by="row.names")
p_vals_Binary_G2$Row.names <- NULL

# Also for ComplexHeatmap, can only have names as rownames
row.names(p_vals_Binary_G2) <- p_vals_Binary_G2$taxonomy.ASV_taxonomy
p_vals_Binary_G2$taxonomy.ASV_taxonomy <- NULL
```

```{r}
# Comparing all against those taxa significant for Asthma
p_vals_Binary_G_Ast <- subset(p_vals_Binary_G2, Asthma == 1 |  Asthma == -1) %>% 
                          ### Taxa as x axis, abbreviate names
  ### gsub UCG-008 too for gut so length of species names matches b/t nasal and gut makes it way easier to get figure sizes to match exactly
                          mutate(Taxa = gsub("\\ sp\\.|^\\[|\\]|Unclassified| UCG-008", "", row.names(.)))

## reorder rows - first alphabetical, then main order by BF and Asthma 1
p_vals_Binary_G_Ast <- p_vals_Binary_G_Ast[order(-p_vals_Binary_G_Ast$Asthma, -p_vals_Binary_G_Ast$BF, -p_vals_Binary_G_Ast$Asthma3y5y, p_vals_Binary_G_Ast$Taxa),]
## want factor levels to match this order after sorting
p_vals_Binary_G_Ast$Taxa <- factor(p_vals_Binary_G_Ast$Taxa, levels = unique(p_vals_Binary_G_Ast$Taxa))
```

Heatmap showing the direction of association 
```{r, fig.width=10}
p_vals_binary_m <- p_vals_Binary_G_Ast %>% 
  melt(., id.vars = "Taxa") %>%
  mutate(Comparison = revalue(factor(variable, levels = names(comparisons), 
               comparisons)),
         italic_x = gsub(" \\d{1,}$", "", Taxa),
         non_italic_x = ifelse(is.na(str_extract(Taxa, "\\d{1,}$") ), "", 
                               str_extract(Taxa, "\\d{1,}$") )
         ) %>%
         mutate(Taxa_2 = glue("<i>{italic_x}</i> {non_italic_x}"))

## factor levels were not maintained but...
unique(p_vals_binary_m$Taxa) ## df already ordered how I want re. factor levels for taxa - but now want this for  Taxa_2 - should be easy
p_vals_binary_m$Taxa_2 = factor(p_vals_binary_m$Taxa_2, levels=unique(p_vals_binary_m$Taxa_2))

main_h <- ggplot(p_vals_binary_m, aes(Taxa_2, Comparison, fill = value)) +
                    geom_tile() + 
                    scale_fill_gradient2("",  low = "#ff8a90", mid = "#b3b3b3", high = "skyblue1",
                      midpoint = 0, limits=c(-1, 1), breaks=c(-1, 0, 1), labels = c("Earlier", "p > 0.05", "Later")) +
                    geom_hline(yintercept = 1.5, color = "#5e5d5d", linewidth = 0.5) +
                    geom_hline(yintercept = 2.5, color = "#5e5d5d", linewidth = 0.5) +
                    theme(axis.text.x= element_markdown(size=10, angle=90, vjust=0.5, hjust=1, color = "black"), 
                          axis.text.y = element_text(size = 10),
                          axis.title = element_blank(), legend.position = "none", axis.ticks = element_blank()
                          ) + scale_y_discrete(expand = c(0,0)) + scale_x_discrete(expand = c(0,0))

# Make side annotation: showing the percent taxa that are Late with no Asthma, that are also 'late' with other factors
Direction_sig_Asthma_L <- subset(p_vals_Binary_G_Ast, Asthma == 1)

n_ast_tax <- nrow(Direction_sig_Asthma_L)
Pct_late <- data.frame(Asthma=100, 
                       BF=table(Direction_sig_Asthma_L$BF)[2]*100/n_ast_tax,
                       Asthma3y5y=table(Direction_sig_Asthma_L$Asthma3y5y)[2]*100/n_ast_tax) %>% 
                      t %>% as.data.frame %>% mutate(variable = factor(row.names(.), levels = names(comparisons)),
                                                     x_axis = "percent_late") 
Pct_late$pct <- Pct_late$'1' # Add annotations too: 36/36, 17/36, 12/36

annotation_h <- ggplot(Pct_late, aes(x_axis, variable, fill = pct)) +
                    geom_tile() + 
                    scale_fill_gradient2("",  low = "white", mid = "#f7b0bc", high = "red", midpoint = 50,limits = c(0, 100))+
                    theme(axis.text = element_blank(),
                          axis.title = element_blank(), legend.position = "none", axis.ticks = element_blank(),
                          plot.margin = unit(c(t, r, b, l), "cm")) + 
                    scale_y_discrete(expand = c(0,0)) + scale_x_discrete(expand = c(0,0))

Fig_3E_G <- plot_grid(main_h, annotation_h, ncol = 2, rel_widths = c(1, 0.02))
Fig_3E_G
```


```{r, include=FALSE}
## save
tiff(file = paste0(path, "Figures/Fig_3D_Nasal_heatmap.tiff"), width = 1920, height = 1000, units = "px", res = 300) 
Fig_3E_N
dev.off()

tiff(file = paste0(path, "Figures/Fig_3D_Gut_heatmap.tiff"), width = 1920, height = 1000, units = "px", res = 300) 
Fig_3E_G
dev.off()
```

