---
title: "Table S1. Relative abundance (%) of Genera present in 70% of samples at either 3-months or 1-year for both nasal and gut niches."
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Setup 

```{r, message=FALSE, warning=FALSE}
R.version.string
library(phyloseq); packageVersion('phyloseq')
library(tableone); packageVersion('tableone')
library(tidyr); packageVersion('tidyr')
library(reshape2); packageVersion('reshape2')
library(knitr); packageVersion('knitr')
library(here)

source(paste0(here::here(), "/Functions.R"))
```

**Data used**
```{r} 
data_NG2 <- readRDS(paste0(path, "Data/Data_subsets/Main_UnRarefied_NasalGut_Subset.rds"))
```

# Genus level relative abundances, overall microbiota composition (Table S1)
**Also see Figure 1B, uses the same scripts**

Aggregate to genus level and use relative data
```{r}
## aggregate to genus level 
data_NG2_G <- tax_glom(data_NG2, taxrank = "Genus")

NG_relative <- transform_sample_counts(data_NG2_G, function(x) x*100 / sum(x) )  
```

Split data by SampleType and time-point for taxa filter - want to plot genera present in 70% of samples at EITHER time-point
```{r}
genus_n_3m <- subset_samples(NG_relative, SampleType=="Nasal" & Visit == "3 Months")
genus_n_1y <- subset_samples(NG_relative, SampleType=="Nasal" & Visit == "1 Year")

genus_gut_3m <- subset_samples(NG_relative, SampleType=="Gut" & Visit == "3 Months")
genus_gut_1y <- subset_samples(NG_relative, SampleType=="Gut" & Visit == "1 Year")
```

Select taxa to plot
- Genera present in at least 70% of samples in at EITHER time-point will be colored in the plot with the remainder grouped into "Other"  
```{r}
data.p60_n_3m = filter_taxa(genus_n_3m, function(x) sum(x > 0) > (0.70*length(x)), TRUE) 
ntaxa(data.p60_n_3m) 
data.p60_n_1y = filter_taxa(genus_n_1y, function(x) sum(x > 0) > (0.70*length(x)), TRUE) 
ntaxa(data.p60_n_1y)

data.p60_gut_3m = filter_taxa(genus_gut_3m, function(x) sum(x > 0) > (0.70*length(x)), TRUE) 
ntaxa(data.p60_gut_3m)
data.p60_gut_1y = filter_taxa(genus_gut_1y, function(x) sum(x > 0) > (0.70*length(x)), TRUE) 
ntaxa(data.p60_gut_1y)
```

Nasal taxa to plot for 3-month & 1-year data
```{r}
Ntaxa_toPlot <- c(get_taxa_unique(data.p60_n_3m, taxonomic.rank = "Genus"), 
                 get_taxa_unique(data.p60_n_1y, taxonomic.rank = "Genus"))
Ntaxa_toPlot2 <- subset(Ntaxa_toPlot, !duplicated(Ntaxa_toPlot)) # remove dups found in more than 1 sample type
length(Ntaxa_toPlot2) # unique Genera 

# Use subset of all nasal samples
N_relative <- subset_samples(NG_relative, SampleType=="Nasal")

# Make overall nasal dataset of prevalent genera in either time-point 
N_relative_Sub <- subset_taxa(N_relative, Genus %in% Ntaxa_toPlot2)
N_relative_Sub
```

Gut taxa to plot
```{r}
Gtaxa_toPlot <- c(get_taxa_unique(data.p60_gut_3m, taxonomic.rank = "Genus"),
                 get_taxa_unique(data.p60_gut_1y, taxonomic.rank = "Genus"))
Gtaxa_toPlot2 <- subset(Gtaxa_toPlot, !duplicated(Gtaxa_toPlot))
length(Gtaxa_toPlot2) # unique Genera 

# Use subset of all nasal samples
G_relative <- subset_samples(NG_relative, SampleType=="Gut")

# Make overall nasal dataset of prevalent genera in either time-point 
G_relative_Sub <- subset_taxa(G_relative, Genus %in% Gtaxa_toPlot2)
G_relative_Sub # more than 13 b/c of contested [Eubacterium] with different Family level taxonomy - 4 different [Eubacterium] Genera - ok to group for plot.
```


(2) Make summary data for each subset that will be plotted 

Nasal dataset for stacked bars, stratify by Visit
```{r}
n_genus_melt <- psmelt(N_relative_Sub) 

#DATA Summary 
Genus_n_summary <- data_summary(n_genus_melt, varname = "Abundance", groupnames = c("Genus", "Visit"))

## Keep only these columns for plotting
Genus_n_summary2 <-  subset(Genus_n_summary, select = c("Mean", "Genus", "Visit"))

# use this alternative to melting the whole dataset to add 'other' group - to get 100% total, this is much quicker computationally vs melting all data
Genus_n_summary2_S <- split(Genus_n_summary2, Genus_n_summary2$Visit) 

Genus_n_summary3m <- rbind(Genus_n_summary2_S$`3 Months`, c(100-sum(Genus_n_summary2_S$`3 Months`$Mean), "Other", "3 Months")) # adding extra row for "Other" group, in same order: "Mean", "Genus", "Visit"

Genus_n_summary1y <- rbind(Genus_n_summary2_S$`1 Year`, c(100-sum(Genus_n_summary2_S$`1 Year`$Mean), "Other", "1 Year")) # adding extra row for "Other" group, in same order: "Mean", "Genus", "Visit"


Genus_N_summary3 <- rbind(Genus_n_summary3m, Genus_n_summary1y)
Genus_N_summary3$Mean <- as.numeric(Genus_N_summary3$Mean)
```

Gut dataset for stacked bars, stratify by Visit
```{r}
g_genus_melt <- psmelt(G_relative_Sub) 

#DATA Summary 
Genus_g_summary <- data_summary(g_genus_melt, varname = "Abundance", groupnames = c("Genus", "Visit"))

## Keep only these columns for plotting
Genus_g_summary2 <-  subset(Genus_g_summary, select = c("Mean", "Genus", "Visit"))


# use this alternative to melting the whole dataset to add 'other' group - to get 100% total, this is much quicker computationally vs melting all data
Genus_g_summary2_S <- split(Genus_g_summary2, Genus_g_summary2$Visit) 

Genus_g_summary3m <- rbind(Genus_g_summary2_S$`3 Months`, c(100-sum(Genus_g_summary2_S$`3 Months`$Mean), "Other", "3 Months")) # adding extra row for "Other" group, in same order: "Mean", "Genus", "Visit"

Genus_g_summary1y <- rbind(Genus_g_summary2_S$`1 Year`, c(100-sum(Genus_g_summary2_S$`1 Year`$Mean), "Other", "1 Year")) # adding extra row for "Other" group, in same order: "Mean", "Genus", "Visit"


Genus_g_summary3 <- rbind(Genus_g_summary3m, Genus_g_summary1y)
Genus_g_summary3$Mean <- as.numeric(Genus_g_summary3$Mean)
```

## Generate Table
```{r}
Genus_n_summary$Type <- "Nasal"
Genus_g_summary$Type <- "Gut"

Genus_summary <- rbind(Genus_n_summary, Genus_g_summary)

write.csv(Genus_summary, file=paste0(path, "Tables/TableS1_Genus_level_relative_abundance.csv"))


Genus_summary_split <- split(Genus_summary, Genus_summary$Type)

Genus_summary_nasal <- Genus_summary_split$Nasal[order(Genus_summary_split$Nasal$Visit, -Genus_summary_split$Nasal$Median),]
Genus_summary_gut <- Genus_summary_split$Gut[order(Genus_summary_split$Gut$Visit, -Genus_summary_split$Gut$Median),]

kable(data.frame(Type  = Genus_summary_nasal$Type, Genus_summary_nasal[1:2], round(Genus_summary_nasal[3:11], 2)))

kable(data.frame(Type  = Genus_summary_gut$Type, Genus_summary_gut[1:2], round(Genus_summary_gut[3:11], 2)))
```



