---
title: "Table S4. Permutation test results for the difference in the change in prevalence (1 year – 3 months) between infants exclusively breastfed and those no longer breastfed, and those with and without asthma at 3 years."
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Setup 

```{r, warning=FALSE, message=FALSE}
R.version.string
library(phyloseq); packageVersion('phyloseq')
library(dplyr); packageVersion('dplyr')  
library(here)

source(paste0(here::here(), "/Functions.R"))
```

### Microbiota Trajectory data subsets used
```{r}
Nasal_3M1Y_pa_filt <- readRDS(paste0(path, "Data/Data_subsets/Main_Nasal_Trajectory_Subset.rds")) # Prevalence dataset

Gut_3M1Y_pa_filt <- readRDS(paste0(path, "Data/Data_subsets/Main_Gut_Trajectory_Subset.rds")) # Prevalence dataset
```

```{r}
# splitting by timepoint for downstream analyses
N.pa.3m <- subset_samples(Nasal_3M1Y_pa_filt, Visit == "3 Months")
N.pa.1y <- subset_samples(Nasal_3M1Y_pa_filt, Visit == "1 Year")

G.pa.3m <- subset_samples(Gut_3M1Y_pa_filt, Visit == "3 Months")
G.pa.1y <- subset_samples(Gut_3M1Y_pa_filt, Visit == "1 Year")
```


# Table S4 permutation test results for breastfeeding and asthma are generated for Figure 3
Same scripts as Fig_3.Rmd for these results

## Nasal - Prevalence Trajectory derived variables, creation of a prevalence trajectory feature table
```{r} 
## extract otu table and use Participant IDs as row.names instead - will match b/t datasets
N_3m_otu <- data.frame(as.data.frame(t(as(otu_table(N.pa.3m), "matrix"))), row.names=sample_data(N.pa.3m)$CHILDid)
N_1y_otu <- data.frame(as.data.frame(t(as(otu_table(N.pa.1y), "matrix"))), row.names=sample_data(N.pa.1y)$CHILDid)

N_3m_otu <- N_3m_otu[order(row.names(N_3m_otu)),]
N_1y_otu <- N_1y_otu[order(row.names(N_1y_otu)),]

## Checking - colnames and Childid's are in the same order? Yes - good - can move forward
identical(colnames(N_3m_otu), colnames(N_1y_otu))
identical(row.names(N_1y_otu), row.names(N_3m_otu))

## 1 Year minus 3 Month Prevalence For each Taxa
change_3m1y_pa <- N_1y_otu - N_3m_otu # -1 = present at 3m not 1y, 1= present at 1y not 3m, 0=either present at both or neither timpoint
dim(change_3m1y_pa) # Same order as original physeq obj

# Sample data to merge with dataset
sdf_toAdd <- sample_data(N.pa.1y)[order(sample_data(N.pa.1y)$CHILDid),]
sdf_3m <- sample_data(N.pa.3m)[order(sample_data(N.pa.3m)$CHILDid),]
## check that it's in the same order as change_3m1y_pa
identical(row.names(change_3m1y_pa), sdf_toAdd$CHILDid) 
identical(row.names(change_3m1y_pa), sdf_3m$CHILDid) 

### same order so just merge data with sample metadata using data.frame
change_3m1y_pa_sdf <- data.frame(change_3m1y_pa, sdf_toAdd, RE_virus_3m = sdf_3m$RE_Virus) # Using 1 year sdf for variables (will be same for 3m and 1y for almost all) - but also adding RE virus at 3 months as "RE_virus_3m". Note that "RE_Virus" is the RE virus at 1 year.

N_taxa <- ncol(change_3m1y_pa) # number of taxa, to use in downstream code
N_subj <- nrow(change_3m1y_pa) # number of subjects, to use in downstream code
```

## Gut - Prevalence Trajectory derived variables, creation of a prevalence trajectory feature table
```{r} 
## extract otu table and use Participant IDs as row.names instead - will match b/t datasets
G_3m_otu <- data.frame(as.data.frame(t(as(otu_table(G.pa.3m), "matrix"))), row.names=sample_data(G.pa.3m)$CHILDid)
G_1y_otu <- data.frame(as.data.frame(t(as(otu_table(G.pa.1y), "matrix"))), row.names=sample_data(G.pa.1y)$CHILDid)

G_3m_otu <- G_3m_otu[order(row.names(G_3m_otu)),]
G_1y_otu <- G_1y_otu[order(row.names(G_1y_otu)),]

## Checking - colnames and Childid's are in the same order? Yes - good - can move forward
identical(colnames(G_3m_otu), colnames(G_1y_otu))
identical(row.names(G_3m_otu), row.names(G_1y_otu))

## 1 Year minus 3 Month Prevalence For each Taxa
change_3m1y_pa_G <- G_1y_otu - G_3m_otu # -1 = present at 3m not 1y, 1= present at 1y not 3m, 0=either present at both or neither timpoint
dim(change_3m1y_pa_G) # Same order as original physeq obj

sdf_toAdd <- sample_data(G.pa.1y)[order(sample_data(G.pa.1y)$CHILDid),]
sdf_3m <- sample_data(G.pa.3m)[order(sample_data(G.pa.3m)$CHILDid),]
## check that it's in the same order as change_3m1y_pa
identical(row.names(change_3m1y_pa_G), sdf_toAdd$CHILDid) # order matches
identical(row.names(change_3m1y_pa_G), sdf_3m$CHILDid) # order matches

### same order so just merge data with sample metadata using data.frame
change_3m1y_pa_G_sdf <- data.frame(change_3m1y_pa_G, sdf_toAdd, RE_virus_3m = sdf_3m$RE_Virus) # Using 1 year sdf for variables but also adding RE virus at 3 months as "RE_virus_3m". Note that "RE_Virus" is the RE virus at 1 year

N_taxa_G <- ncol(change_3m1y_pa_G) # number of taxa, to use in downstream code
N_subj_G <- nrow(change_3m1y_pa_G) # number of subjects, to use in downstream code
```

## Nasal - Breastfeeding 
```{r} 
X <- subset(change_3m1y_pa_sdf, BF_status_3m_sampling == "No BM")
Y <- subset(change_3m1y_pa_sdf, BF_status_3m_sampling == "Exclusive")

# Make data.frame of prevalence like before BUT keeping as proprotions instead of Percentages for now
df.pa.nasal.BF <- data.frame(X_P_Change=colSums(X[1:N_taxa])*100/dim(X)[1],Y_P_Change=colSums(Y[1:N_taxa])*100/dim(Y)[1],
                             taxonomy=as(tax_table(Nasal_3M1Y_pa_filt), "matrix"),
                             prevalence=taxa_sums(Nasal_3M1Y_pa_filt)*100/(N_subj*2)) 

df.pa.nasal.BF$pa.change.diff <- df.pa.nasal.BF$Y_P_Change - df.pa.nasal.BF$X_P_Change
```
## Gut - Breastfeeding 
```{r} 
X <- subset(change_3m1y_pa_G_sdf, BF_status_3m_sampling == "No BM")
Y <- subset(change_3m1y_pa_G_sdf, BF_status_3m_sampling == "Exclusive")

# Make data.frame of prevalence like before BUT keeping as proprotions instead of Percentages for now
df.pa.gut.BF <- data.frame(X_P_Change=colSums(X[1:N_taxa_G])*100/dim(X)[1],Y_P_Change=colSums(Y[1:N_taxa_G])*100/dim(Y)[1], taxonomy=as(tax_table(Gut_3M1Y_pa_filt), "matrix"), prevalence=taxa_sums(Gut_3M1Y_pa_filt)*100/(N_subj_G*2)) 

# difference in change
df.pa.gut.BF$pa.change.diff <- df.pa.gut.BF$Y_P_Change - df.pa.gut.BF$X_P_Change
```
## Nasal - Asthma at 3 years 
```{r} 
X <- subset(change_3m1y_pa_sdf, asthma_3y == "Probable")
Y <- subset(change_3m1y_pa_sdf, asthma_3y == "No")

# Make data.frame of prevalence like before BUT keeping as proprotions instead of Percentages for now
df.pa.nasal.ast <- data.frame(X_P_Change=colSums(X[1:N_taxa])*100/dim(X)[1],Y_P_Change=colSums(Y[1:N_taxa])*100/dim(Y)[1], taxonomy=as(tax_table(Nasal_3M1Y_pa_filt), "matrix"), prevalence=taxa_sums(Nasal_3M1Y_pa_filt)*100/(N_subj*2)) 

# difference in change
df.pa.nasal.ast$pa.change.diff <- df.pa.nasal.ast$Y_P_Change - df.pa.nasal.ast$X_P_Change
```
## Gut - Asthma at 3 years
```{r} 
X <- subset(change_3m1y_pa_G_sdf, asthma_3y == "Probable")
Y <- subset(change_3m1y_pa_G_sdf, asthma_3y == "No")

# Make data.frame of prevalence like before BUT keeping as proprotions instead of Percentages for now
df.pa.gut.ast <- data.frame(X_P_Change=colSums(X[1:N_taxa_G])*100/dim(X)[1],Y_P_Change=colSums(Y[1:N_taxa_G])*100/dim(Y)[1], taxonomy=as(tax_table(Gut_3M1Y_pa_filt), "matrix"), prevalence=taxa_sums(Gut_3M1Y_pa_filt)*100/(N_subj_G*2)) 

# difference in change
df.pa.gut.ast$pa.change.diff <- df.pa.gut.ast$Y_P_Change - df.pa.gut.ast$X_P_Change
```

# Additional permutation results from Fig_3.Rmd used(saved in chunk 52) 
```{r}
Per_test_df_ls <- readRDS(paste0(path, "Tables/PermTests/MAIN_Permutation_PrevTraj_Res.rds"))
# NOTE - effect size sign is flipped in these results, but results are the same - in this case negative effect size = Later in Exclusive BF, later in 'healthy' vs asthma, etc. 
```

```{r}
Table_S4_N_BF <- merge(subset(Per_test_df_ls$BF_Nasal, select = c("p_value", "p_value_2")), 
                    subset(df.pa.nasal.BF, select = c("X_P_Change", "Y_P_Change", "pa.change.diff")), 
                    by="row.names")

Table_S4_N_Asthma <- merge(subset(Per_test_df_ls$Ast3y_Nasal, select = c("p_value", "p_value_2")), 
                    subset(df.pa.nasal.ast, select = c("X_P_Change", "Y_P_Change", "pa.change.diff", 
                                                       "prevalence", "taxonomy.Phylum", "taxonomy.Family" , "taxonomy.Genus", "taxonomy.ASV")), 
                    by="row.names")

Table_S4_N <- merge(Table_S4_N_BF, Table_S4_N_Asthma, by = "Row.names", suffixes = c("_BF", "_Asthma"))


Table_S4_G_BF <- merge(subset(Per_test_df_ls$BF_Gut, select = c("p_value", "p_value_2")), 
                    subset(df.pa.gut.BF, select = c("X_P_Change", "Y_P_Change", "pa.change.diff")), 
                    by="row.names")

Table_S4_G_Asthma <- merge(subset(Per_test_df_ls$Ast3y_Gut, select = c("p_value", "p_value_2")), 
                    subset(df.pa.gut.ast, select = c("X_P_Change", "Y_P_Change", "pa.change.diff", 
                                                       "prevalence", "taxonomy.Phylum", "taxonomy.Family" , "taxonomy.Genus", "taxonomy.ASV")), 
                    by="row.names")

Table_S4_G <- merge(Table_S4_G_BF, Table_S4_G_Asthma, by = "Row.names", suffixes = c("_BF", "_Asthma"))

# change some colnames
colnames(Table_S4_N) <- gsub("taxonomy\\.", "", colnames(Table_S4_N))
colnames(Table_S4_N)[names(Table_S4_N) == "Row.names"] <- "OTU_Taxonomy"
colnames(Table_S4_N)[4:6]
colnames(Table_S4_N)[4:6] <- c("Delta_P_NoBM", "Delta_P_Exclusive", "Delta_P_Diff_BM")
colnames(Table_S4_N)[9:11]
colnames(Table_S4_N)[9:11] <- c("Delta_P_Asthma3Y", "Delta_P_NoAsthma", "Delta_P_Diff_Asthma3Y")
colnames(Table_S4_N)

colnames(Table_S4_G) <- gsub("taxonomy\\.", "", colnames(Table_S4_G))
colnames(Table_S4_G)[names(Table_S4_G) == "Row.names"] <- "OTU_Taxonomy"
colnames(Table_S4_G)[4:6]
colnames(Table_S4_G)[4:6] <- c("Delta_P_NoBM", "Delta_P_Exclusive", "Delta_P_Diff_BM")
colnames(Table_S4_G)[9:11]
colnames(Table_S4_G)[9:11] <- c("Delta_P_Asthma3Y", "Delta_P_NoAsthma", "Delta_P_Diff_Asthma3Y")
colnames(Table_S4_G)

write.csv(Table_S4_N, file=paste0(path, "Tables/TableS4_Permutation_test_res_Nasal.csv"))
write.csv(Table_S4_G, file=paste0(path, "Tables/TableS4_Permutation_test_res_Gut.csv"))
```

