---
title: "Table S4"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Setup 

```{r, warning=FALSE, message=FALSE}
R.version.string
library(phyloseq); packageVersion('phyloseq')
library(ggplot2); packageVersion('ggplot2')
library(cowplot); packageVersion('cowplot')
library(vegan); packageVersion('vegan')
library(tidyr); packageVersion('tidyr')
library(reshape2); packageVersion('reshape2')
library(dplyr); packageVersion('dplyr')  
library(lmPerm); packageVersion('lmPerm')
library(UpSetR); packageVersion('UpSetR')
library(ComplexHeatmap); packageVersion('ComplexHeatmap')
library(circlize); packageVersion('circlize')

source("Preprocessing/Functions.R")
```

```{r}
theme_set(theme_bw() + theme(plot.title = element_text(hjust = 0.5,face = "bold"), strip.text.x = element_text(size = 11), strip.background = element_rect(colour="white", fill="white"), panel.grid.major = element_blank(), panel.grid.minor = element_blank(), text = element_text(size=12), axis.text = element_text(color="black", size=12), legend.key.size = unit(0.5,"cm")))

cbPalette <- c("navy", "paleturquoise4",  "violetred3", "skyblue1", "goldenrod3")
```

### Microbiota Trajectory data subsets used
```{r}
Nasal_3M1Y_pa_filt <- readRDS("Data_subsets/Main_Nasal_Trajectory_Subset.rds") # Prevalence dataset

Gut_3M1Y_pa_filt <- readRDS("Data_subsets/Main_Gut_Trajectory_Subset.rds") # Prevalence dataset
```

```{r}
# splitting by timepoint for downstream analyses
N.pa.3m <- subset_samples(Nasal_3M1Y_pa_filt, Visit == "3 Months")
N.pa.1y <- subset_samples(Nasal_3M1Y_pa_filt, Visit == "1 Year")

G.pa.3m <- subset_samples(Gut_3M1Y_pa_filt, Visit == "3 Months")
G.pa.1y <- subset_samples(Gut_3M1Y_pa_filt, Visit == "1 Year")
```

## Nasal - McNemar test for overall prevalence trajectories

```{r}
# input
N_taxa <- ntaxa(Nasal_3M1Y_pa_filt)
N_subj <- nsamples(Nasal_3M1Y_pa_filt)/2
TP_1 <- N.pa.3m
TP_2 <- N.pa.1y

# setup
N_3m_otu <- as.data.frame(t(as(otu_table(TP_1), "matrix")))
N_1y_otu <- as.data.frame(t(as(otu_table(TP_2), "matrix")))

N_3m_otu_sdf <- data.frame(N_3m_otu, CHILDid=sample_data(TP_1)$CHILDid)
N_1y_otu_sdf <- data.frame(N_1y_otu, CHILDid=sample_data(TP_2)$CHILDid)

## don't like how data.frame function changed the column names, changing back
colnames(N_3m_otu_sdf)[1:N_taxa] <- colnames(N_3m_otu)
colnames(N_1y_otu_sdf)[1:N_taxa] <- colnames(N_1y_otu)

# Making sure 3m and 1y dataframe are in the same order
N_3m_otu_sdf <- N_3m_otu_sdf[order(N_3m_otu_sdf$CHILDid),]
N_1y_otu_sdf <- N_1y_otu_sdf[order(N_1y_otu_sdf$CHILDid),]

identical(N_3m_otu_sdf$CHILDid, N_1y_otu_sdf$CHILDid) # good

# are in the same order in both lists, remove ids for test
N_3m_otu_sdf$CHILDid <- NULL
N_1y_otu_sdf$CHILDid <- NULL

# Removing the persistently present taxa that cannot be tested (~100% prevalence at both timepoints)
N_3m_otu_sdf$`Streptococcus sp. 40` <- NULL 
N_1y_otu_sdf$`Streptococcus sp. 40`  <- NULL
```

```{r}
# McNemar test
## list of contingency tables
Nasal_Tab_3m1y <- mapply(FUN=table, as.list(N_3m_otu_sdf),  as.list(N_1y_otu_sdf) , SIMPLIFY = FALSE, USE.NAMES = TRUE) 

## RUN
Nasal_3m1y_Mcnem_p_df <- Nasal_Tab_3m1y %>% lapply(., FUN = mcnemar.test) %>% lapply(., function(x) {x["p.value"]}) %>% bind_rows(., .id = "OTU")

Nasal_3m1y_Mcnem_p_df$p.value.BH <- p.adjust(Nasal_3m1y_Mcnem_p_df$p.value, method="BH" , n = length(Nasal_3m1y_Mcnem_p_df$p.value))

# Calculate average proportion at 3 months and 1 year
prop_3m <- taxa_sums(TP_1)/N_subj
prop_1y <- taxa_sums(TP_2)/N_subj

props <- data.frame(Nasal_3M_P=prop_3m*100, Nasal_1Y_P=prop_1y*100, Nasal_3M_SE=sqrt(prop_3m*(1-prop_3m)/N_subj), Nasal_1Y_SE=sqrt(prop_1y*(1-prop_1y)/N_subj))
props$OTU <- row.names(props)

# strep is 100 and 0 SE so no test done, automatic 100% persistent (NA p values for this one)
Nasal_3m1y_Mcnem_p_df2 <- merge(Nasal_3m1y_Mcnem_p_df, props, by="OTU", all=TRUE) 

# Calc change in prevalence
Nasal_3m1y_Mcnem_p_df2$Nasal_changeP_1Y3M <- Nasal_3m1y_Mcnem_p_df2$Nasal_1Y_P - Nasal_3m1y_Mcnem_p_df2$Nasal_3M_P 
```

Using a more strict p-value threshold in addition to an effect size cutoff to define colonizers (checked volcano plots of nasal and gut before deciding on these thresholds)
```{r, warning=FALSE}
Nasal_3m1y_Mcnem_p_df2$Colonization_pattern_final <- ifelse(Nasal_3m1y_Mcnem_p_df2$p.value.BH < 0.001 & Nasal_3m1y_Mcnem_p_df2$Nasal_changeP_1Y3M < -7, "Early",
                                                     ifelse(Nasal_3m1y_Mcnem_p_df2$p.value.BH < 0.001 & Nasal_3m1y_Mcnem_p_df2$Nasal_changeP_1Y3M >7, "Late", "Persistent"))

### most taxa colonize later, and average change in prevalence for these late colonizers is 19.5%
data_summary_V2(Nasal_3m1y_Mcnem_p_df2, varname = "Nasal_changeP_1Y3M", groupnames = "Colonization_pattern_final") 
```

## Gut - McNemar test for overall prevalence trajectories

```{r}
# input
N_taxa_G <- ntaxa(Gut_3M1Y_pa_filt)
N_subj_G <- nsamples(Gut_3M1Y_pa_filt)/2
TP_1 <- G.pa.3m
TP_2 <- G.pa.1y

# setup

# using dataset with all 3 month and 1 year gut and nasal samples for now - same infants in all datasets to reduce variation 
G_3m_otu <- as.data.frame(t(as(otu_table(TP_1), "matrix")))
G_1y_otu <- as.data.frame(t(as(otu_table(TP_2), "matrix")))

G_3m_otu_sdf <- data.frame(G_3m_otu, CHILDid=sample_data(TP_1)$CHILDid)
G_1y_otu_sdf <- data.frame(G_1y_otu, CHILDid=sample_data(TP_2)$CHILDid)

## don't like how data.frame function changed the column names, changing back
colnames(G_3m_otu_sdf)[1:N_taxa_G] <- colnames(G_3m_otu)
colnames(G_1y_otu_sdf)[1:N_taxa_G] <- colnames(G_1y_otu)


G_3m_otu_sdf <- G_3m_otu_sdf[order(G_3m_otu_sdf$CHILDid),]
G_1y_otu_sdf <- G_1y_otu_sdf[order(G_1y_otu_sdf$CHILDid),]

identical(G_3m_otu_sdf$CHILDid, G_1y_otu_sdf$CHILDid) # good

# are in the same order in both lists
G_3m_otu_sdf$CHILDid <- NULL
G_1y_otu_sdf$CHILDid <- NULL
```

```{r}
# McNemar test

## list of contingency tables
Gut_Tab_3m1y <- mapply(FUN=table, as.list(G_3m_otu_sdf),  as.list(G_1y_otu_sdf), SIMPLIFY = FALSE, USE.NAMES = TRUE) 

## RUN
Gut_3m1y_Mcnem_p_df <- Gut_Tab_3m1y %>% lapply(., FUN = mcnemar.test) %>% lapply(., function(x) {x["p.value"]}) %>% bind_rows(., .id = "OTU")

Gut_3m1y_Mcnem_p_df$p.value.BH <- p.adjust(Gut_3m1y_Mcnem_p_df$p.value, method="BH" , n = length(Gut_3m1y_Mcnem_p_df$p.value))

# Calculate average proporition at 3 months and 1 year
prop_3m <- taxa_sums(TP_1)/N_subj_G
prop_1y <- taxa_sums(TP_2)/N_subj_G

props <- data.frame(Gut_3M_P=prop_3m*100, Gut_1Y_P=prop_1y*100, Gut_3M_SE=sqrt(prop_3m*(1-prop_3m)/N_subj_G), Gut_1Y_SE=sqrt(prop_1y*(1-prop_1y)/N_subj_G))
props$OTU <- row.names(props)

# strep is 100 and 0 SE so no test done, automatic 100% persistent (NA p values for this one)
Gut_3m1y_Mcnem_p_df2 <- merge(Gut_3m1y_Mcnem_p_df, props, by="OTU", all=TRUE) 

# Calc change in prevalence
Gut_3m1y_Mcnem_p_df2$Gut_changeP_1Y3M <- Gut_3m1y_Mcnem_p_df2$Gut_1Y_P - Gut_3m1y_Mcnem_p_df2$Gut_3M_P 
```

The Mcnemar test appears to be very sensitive. Use a more strict p-value threshold in addition to an effect size cutoff to define Late and Early colonizers as oppose to persistent (no change) colonizers.
```{r}
Gut_3m1y_Mcnem_p_df2$Colonization_pattern_final <- ifelse(Gut_3m1y_Mcnem_p_df2$p.value.BH < 0.001 & Gut_3m1y_Mcnem_p_df2$Gut_changeP_1Y3M < -7, "Early",
                                                          ifelse(Gut_3m1y_Mcnem_p_df2$p.value.BH < 0.001 & Gut_3m1y_Mcnem_p_df2$Gut_changeP_1Y3M >7, "Late", "Persistent"))

### most taxa colonize later (65), and average change in prevalence for these late colonizers is 15%
data_summary(Gut_3m1y_Mcnem_p_df2, varname = "Gut_changeP_1Y3M", groupnames = "Colonization_pattern_final") 
```


### Compile & save all prevalence trajectory results
```{r}
# compile & save results for individ taxa

# saving everything to output
Per_test_df_ls <- list(Res_per_taxa_ast_N, Res_per_taxa_BF_N, Res_per_taxa_ast5y_N, 
                       Res_per_taxa_ast3y5y_N, Res_per_taxa_wheeze_N, Res_per_taxa_atopy_N, Res_per_taxa_Cold_N, 
                       Res_per_taxa_ast_G, Res_per_taxa_BF_G, Res_per_taxa_ast5y_G, Res_per_taxa_ast3y5y_G, 
                       Res_per_taxa_wheeze_G, Res_per_taxa_atopy_G, Res_per_taxa_Cold_G)
names(Per_test_df_ls) <-  c("Ast3y_Nasal", "BF_Nasal", "Ast5y_Nasal", "Ast3y_5y_Nasal", "Wheeze_Nasal", "Atopy_Nasal", "Cold_Nasal", 
                            "Ast3y_Gut", "BF_Gut", "Ast5y_Gut", "Ast3y_5y_Gut", "Wheeze_Gut", "Atopy_Gut", "Cold_Gut")

# Save as an overall rds file in Output/PermTests
saveRDS(Per_test_df_ls, "Output/PermTests/ALL_Permutation_PrevTraj_Res.rds")

# Save as individual csv files for others 
outpath_Pt1 <- "Output/PermTests/Permutation_PrevTraj_Res"
outfiles <- paste(outpath_Pt1, names(Per_test_df_ls), ".csv", sep = "_")

mapply(function(x, y) write.csv(x, file=y, row.names = FALSE, quote = FALSE), x=Per_test_df_ls, y=outfiles, SIMPLIFY = FALSE)
```


```{r, include=FALSE}
## Merge & save Main Asthma and BF results for Table S4
Table_S4_N <- merge(Res_per_taxa_ast_N, Res_per_taxa_BF_N, by="row.names", suffixes = c("_Asthma", "_BF"))
Table_S4_N$original_effect_diff_Asthma <- NULL # using effect size from other dataframe, the direction should ideally be flipped though it doesn't really matter
Table_S4_N$original_effect_diff_BF <- NULL

Table_S4_G <- merge(Res_per_taxa_ast_G, Res_per_taxa_BF_G, by="row.names", suffixes = c("_Asthma", "_BF"))
Table_S4_G$original_effect_diff_Asthma<- NULL 
Table_S4_G$original_effect_diff_BF <- NULL

# Nasal effect sizes and taxonomy
To_rm <- intersect(colnames(df.pa.nasal.ast), colnames(df.pa.nasal.BF))[3:13] ## remove from 1 so not replicated
Table_S4_N_est <- merge(subset(df.pa.nasal.ast, select =setdiff(colnames(df.pa.nasal.ast), To_rm)), 
                        df.pa.nasal.BF, by="row.names", suffixes = c("_Asthma", "_BF"))
Table_S4_N_est$Phylum <- NULL
Table_S4_N_est$taxonomy.Kingdom <- NULL
Table_S4_N_est$taxonomy.LefSe_Taxonomy_toGenus <- NULL

# Gut effect sizes and taxonomy
To_rm <- intersect(colnames(df.pa.gut.ast), colnames(df.pa.gut.BF))[3:13] ## remove from 1 so not replicated
Table_S4_G_est <- merge(subset(df.pa.gut.ast, select =setdiff(colnames(df.pa.gut.ast), To_rm)), 
                        df.pa.gut.BF, by="row.names", suffixes = c("_Asthma", "_BF"))
Table_S4_G_est$Phylum <- NULL
Table_S4_G_est$taxonomy.Kingdom <- NULL
Table_S4_G_est$taxonomy.LefSe_Taxonomy_toGenus <- NULL

## merge
Table_S4_N2 <- merge(Table_S4_N, Table_S4_N_est, by="Row.names")
Table_S4_G2 <- merge(Table_S4_G, Table_S4_G_est, by="Row.names")

colnames(Table_S4_N2)[names(Table_S4_N2) == "Row.names"] <- "OTU_Taxonomy"
colnames(Table_S4_G2)[names(Table_S4_G2) == "Row.names"] <- "OTU_Taxonomy"

write.csv(Table_S4_N, file="Output/TableS4_Permutation_test_res_Nasal.csv")
write.csv(Table_S4_G, file="Output/TableS4_Permutation_test_res_Gut.csv")
```

