---
title: "Overlap in associations between microbiota prevalence trajectories and breastfeeding and respiratory phenotypes (Figure S5)"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Setup 

```{r, warning=FALSE, message=FALSE}
R.version.string
library(phyloseq); packageVersion('phyloseq')
library(ggplot2); packageVersion('ggplot2')
library(cowplot); packageVersion('cowplot')
library(vegan); packageVersion('vegan')
library(tidyr); packageVersion('tidyr')
library(reshape2); packageVersion('reshape2')
library(dplyr); packageVersion('dplyr')  
library(lmPerm); packageVersion('lmPerm')
library(UpSetR); packageVersion('UpSetR')
library(ComplexHeatmap); packageVersion('ComplexHeatmap')
library(circlize); packageVersion('circlize')

source("Preprocessing/Functions.R")
```

```{r}
theme_set(theme_bw() + theme(plot.title = element_text(hjust = 0.5,face = "bold"), strip.text.x = element_text(size = 11), strip.background = element_rect(colour="white", fill="white"), panel.grid.major = element_blank(), panel.grid.minor = element_blank(), text = element_text(size=12), axis.text = element_text(color="black", size=12), legend.key.size = unit(0.5,"cm")))

cbPalette <- c("navy", "paleturquoise4",  "violetred3", "skyblue1", "goldenrod3")
```

### Microbiota Trajectory data subsets used
```{r}
Nasal_3M1Y_pa_filt <- readRDS("Data_subsets/Main_Nasal_Trajectory_Subset.rds") # Prevalence dataset

Gut_3M1Y_pa_filt <- readRDS("Data_subsets/Main_Gut_Trajectory_Subset.rds") # Prevalence dataset
```

```{r}
# splitting by timepoint for downstream analyses
N.pa.3m <- subset_samples(Nasal_3M1Y_pa_filt, Visit == "3 Months")
N.pa.1y <- subset_samples(Nasal_3M1Y_pa_filt, Visit == "1 Year")

G.pa.3m <- subset_samples(Gut_3M1Y_pa_filt, Visit == "3 Months")
G.pa.1y <- subset_samples(Gut_3M1Y_pa_filt, Visit == "1 Year")
```


## Nasal - McNemar test for overall prevalence trajectories

```{r}
# input
N_taxa <- ntaxa(Nasal_3M1Y_pa_filt)
N_subj <- nsamples(Nasal_3M1Y_pa_filt)/2
TP_1 <- N.pa.3m
TP_2 <- N.pa.1y

# setup
N_3m_otu <- as.data.frame(t(as(otu_table(TP_1), "matrix")))
N_1y_otu <- as.data.frame(t(as(otu_table(TP_2), "matrix")))

N_3m_otu_sdf <- data.frame(N_3m_otu, CHILDid=sample_data(TP_1)$CHILDid)
N_1y_otu_sdf <- data.frame(N_1y_otu, CHILDid=sample_data(TP_2)$CHILDid)

## don't like how data.frame function changed the column names, changing back
colnames(N_3m_otu_sdf)[1:N_taxa] <- colnames(N_3m_otu)
colnames(N_1y_otu_sdf)[1:N_taxa] <- colnames(N_1y_otu)

# Making sure 3m and 1y dataframe are in the same order
N_3m_otu_sdf <- N_3m_otu_sdf[order(N_3m_otu_sdf$CHILDid),]
N_1y_otu_sdf <- N_1y_otu_sdf[order(N_1y_otu_sdf$CHILDid),]

identical(N_3m_otu_sdf$CHILDid, N_1y_otu_sdf$CHILDid) # good

# are in the same order in both lists, remove ids for test
N_3m_otu_sdf$CHILDid <- NULL
N_1y_otu_sdf$CHILDid <- NULL

# Removing the persistently present taxa that cannot be tested (~100% prevalence at both timepoints)
N_3m_otu_sdf$`Streptococcus sp. 40` <- NULL 
N_1y_otu_sdf$`Streptococcus sp. 40`  <- NULL
```

```{r}
# McNemar test
## list of contingency tables
Nasal_Tab_3m1y <- mapply(FUN=table, as.list(N_3m_otu_sdf),  as.list(N_1y_otu_sdf) , SIMPLIFY = FALSE, USE.NAMES = TRUE) 

## RUN
Nasal_3m1y_Mcnem_p_df <- Nasal_Tab_3m1y %>% lapply(., FUN = mcnemar.test) %>% lapply(., function(x) {x["p.value"]}) %>% bind_rows(., .id = "OTU")

Nasal_3m1y_Mcnem_p_df$p.value.BH <- p.adjust(Nasal_3m1y_Mcnem_p_df$p.value, method="BH" , n = length(Nasal_3m1y_Mcnem_p_df$p.value))

# Calculate average proportion at 3 months and 1 year
prop_3m <- taxa_sums(TP_1)/N_subj
prop_1y <- taxa_sums(TP_2)/N_subj

props <- data.frame(Nasal_3M_P=prop_3m*100, Nasal_1Y_P=prop_1y*100, Nasal_3M_SE=sqrt(prop_3m*(1-prop_3m)/N_subj), Nasal_1Y_SE=sqrt(prop_1y*(1-prop_1y)/N_subj))
props$OTU <- row.names(props)

# strep is 100 and 0 SE so no test done, automatic 100% persistent (NA p values for this one)
Nasal_3m1y_Mcnem_p_df2 <- merge(Nasal_3m1y_Mcnem_p_df, props, by="OTU", all=TRUE) 

# Calc change in prevalence
Nasal_3m1y_Mcnem_p_df2$Nasal_changeP_1Y3M <- Nasal_3m1y_Mcnem_p_df2$Nasal_1Y_P - Nasal_3m1y_Mcnem_p_df2$Nasal_3M_P 
```

Using a more strict p-value threshold in addition to an effect size cutoff to define colonizers (checked volcano plots of nasal and gut before deciding on these thresholds)
```{r, warning=FALSE}
Nasal_3m1y_Mcnem_p_df2$Colonization_pattern_final <- ifelse(Nasal_3m1y_Mcnem_p_df2$p.value.BH < 0.001 & Nasal_3m1y_Mcnem_p_df2$Nasal_changeP_1Y3M < -7, "Early",
                                                     ifelse(Nasal_3m1y_Mcnem_p_df2$p.value.BH < 0.001 & Nasal_3m1y_Mcnem_p_df2$Nasal_changeP_1Y3M >7, "Late", "Persistent"))

### most taxa colonize later, and average change in prevalence for these late colonizers is 19.5%
data_summary_V2(Nasal_3m1y_Mcnem_p_df2, varname = "Nasal_changeP_1Y3M", groupnames = "Colonization_pattern_final") 
```


## Gut - McNemar test for overall prevalence trajectories

```{r}
# input
N_taxa_G <- ntaxa(Gut_3M1Y_pa_filt)
N_subj_G <- nsamples(Gut_3M1Y_pa_filt)/2
TP_1 <- G.pa.3m
TP_2 <- G.pa.1y

# setup

# using dataset with all 3 month and 1 year gut and nasal samples for now - same infants in all datasets to reduce variation 
G_3m_otu <- as.data.frame(t(as(otu_table(TP_1), "matrix")))
G_1y_otu <- as.data.frame(t(as(otu_table(TP_2), "matrix")))

G_3m_otu_sdf <- data.frame(G_3m_otu, CHILDid=sample_data(TP_1)$CHILDid)
G_1y_otu_sdf <- data.frame(G_1y_otu, CHILDid=sample_data(TP_2)$CHILDid)

## don't like how data.frame function changed the column names, changing back
colnames(G_3m_otu_sdf)[1:N_taxa_G] <- colnames(G_3m_otu)
colnames(G_1y_otu_sdf)[1:N_taxa_G] <- colnames(G_1y_otu)


G_3m_otu_sdf <- G_3m_otu_sdf[order(G_3m_otu_sdf$CHILDid),]
G_1y_otu_sdf <- G_1y_otu_sdf[order(G_1y_otu_sdf$CHILDid),]

identical(G_3m_otu_sdf$CHILDid, G_1y_otu_sdf$CHILDid) # good

# are in the same order in both lists
G_3m_otu_sdf$CHILDid <- NULL
G_1y_otu_sdf$CHILDid <- NULL
```

```{r}
# McNemar test

## list of contingency tables
Gut_Tab_3m1y <- mapply(FUN=table, as.list(G_3m_otu_sdf),  as.list(G_1y_otu_sdf), SIMPLIFY = FALSE, USE.NAMES = TRUE) 

## RUN
Gut_3m1y_Mcnem_p_df <- Gut_Tab_3m1y %>% lapply(., FUN = mcnemar.test) %>% lapply(., function(x) {x["p.value"]}) %>% bind_rows(., .id = "OTU")

Gut_3m1y_Mcnem_p_df$p.value.BH <- p.adjust(Gut_3m1y_Mcnem_p_df$p.value, method="BH" , n = length(Gut_3m1y_Mcnem_p_df$p.value))

# Calculate average proporition at 3 months and 1 year
prop_3m <- taxa_sums(TP_1)/N_subj_G
prop_1y <- taxa_sums(TP_2)/N_subj_G

props <- data.frame(Gut_3M_P=prop_3m*100, Gut_1Y_P=prop_1y*100, Gut_3M_SE=sqrt(prop_3m*(1-prop_3m)/N_subj_G), Gut_1Y_SE=sqrt(prop_1y*(1-prop_1y)/N_subj_G))
props$OTU <- row.names(props)

# strep is 100 and 0 SE so no test done, automatic 100% persistent (NA p values for this one)
Gut_3m1y_Mcnem_p_df2 <- merge(Gut_3m1y_Mcnem_p_df, props, by="OTU", all=TRUE) 

# Calc change in prevalence
Gut_3m1y_Mcnem_p_df2$Gut_changeP_1Y3M <- Gut_3m1y_Mcnem_p_df2$Gut_1Y_P - Gut_3m1y_Mcnem_p_df2$Gut_3M_P 
```

The Mcnemar test appears to be very sensitive. Use a more strict p-value threshold in addition to an effect size cutoff to define Late and Early colonizers as oppose to persistent (no change) colonizers.
```{r}
Gut_3m1y_Mcnem_p_df2$Colonization_pattern_final <- ifelse(Gut_3m1y_Mcnem_p_df2$p.value.BH < 0.001 & Gut_3m1y_Mcnem_p_df2$Gut_changeP_1Y3M < -7, "Early",
                                                          ifelse(Gut_3m1y_Mcnem_p_df2$p.value.BH < 0.001 & Gut_3m1y_Mcnem_p_df2$Gut_changeP_1Y3M >7, "Late", "Persistent"))

### most taxa colonize later (65), and average change in prevalence for these late colonizers is 15%
data_summary(Gut_3m1y_Mcnem_p_df2, varname = "Gut_changeP_1Y3M", groupnames = "Colonization_pattern_final") 
```


# Prevalence Trajectory Coordinate System & Permutation tests - Comparing Prevalence Trajectories Between Groups (Fig. S5)

## Nasal - Prevalence Trajectory derived variables, creation of a prevalence trajectory feature table
To simplify making of prevalence trajectory scatter plots, first make the 'difference in prevalence' results for every pair of samples for each infant (1Y - 3M, e.g. 0 - 1 = -1; 1 - 0 = 1) - so a "-1" means it's present at 3 months but not 1 year, 1 means 1 year but not 3m and 0 means no change.
```{r} 
## extract otu table and use Participant IDs as row.names instead - will match b/t datasets
N_3m_otu <- data.frame(as.data.frame(t(as(otu_table(N.pa.3m), "matrix"))), row.names=sample_data(N.pa.3m)$CHILDid)
N_1y_otu <- data.frame(as.data.frame(t(as(otu_table(N.pa.1y), "matrix"))), row.names=sample_data(N.pa.1y)$CHILDid)

N_3m_otu <- N_3m_otu[order(row.names(N_3m_otu)),]
N_1y_otu <- N_1y_otu[order(row.names(N_1y_otu)),]

## Checking - colnames and Childid's are in the same order? Yes - good - can move forward
identical(colnames(N_3m_otu), colnames(N_1y_otu))
identical(row.names(N_1y_otu), row.names(N_3m_otu))

## 1 Year minus 3 Month Prevalence For each Taxa
change_3m1y_pa <- N_1y_otu - N_3m_otu # -1 = present at 3m not 1y, 1= present at 1y not 3m, 0=either present at both or neither timpoint
dim(change_3m1y_pa) # Same order as original physeq obj

# Sample data to merge with dataset
sdf_toAdd <- sample_data(N.pa.1y)[order(sample_data(N.pa.1y)$CHILDid),]
sdf_3m <- sample_data(N.pa.3m)[order(sample_data(N.pa.3m)$CHILDid),]
## check that it's in the same order as change_3m1y_pa
identical(row.names(change_3m1y_pa), sdf_toAdd$CHILDid) 
identical(row.names(change_3m1y_pa), sdf_3m$CHILDid) 

### same order so just merge data with sample metadata using data.frame
change_3m1y_pa_sdf <- data.frame(change_3m1y_pa, sdf_toAdd, RE_virus_3m = sdf_3m$RE_Virus) # Using 1 year sdf for variables (will be same for 3m and 1y for almost all) - but also adding RE virus at 3 months as "RE_virus_3m". Note that "RE_Virus" is the RE virus at 1 year.

N_taxa <- ncol(change_3m1y_pa) # number of taxa, to use in downstream code
N_subj <- nrow(change_3m1y_pa) # number of subjects, to use in downstream code
```
Objects used for Nasal trajectory plots: change_3m1y_pa_sdf, Nasal_3M1Y_pa_filt, N_taxa, N_subj 

## Gut - Prevalence Trajectory derived variables, creation of a prevalence trajectory feature table
To simplify making of prevalence trajectory scatter plots, first make the 'difference in prevalence' results for every pair of samples for each infant (1Y - 3M, e.g. 0 - 1 = -1; 1 - 0 = 1) - so a "-1" means it's present at 3 months but not 1 year, 1 means 1 year but not 3m and 0 means no change.
```{r} 
## extract otu table and use Participant IDs as row.names instead - will match b/t datasets
G_3m_otu <- data.frame(as.data.frame(t(as(otu_table(G.pa.3m), "matrix"))), row.names=sample_data(G.pa.3m)$CHILDid)
G_1y_otu <- data.frame(as.data.frame(t(as(otu_table(G.pa.1y), "matrix"))), row.names=sample_data(G.pa.1y)$CHILDid)

G_3m_otu <- G_3m_otu[order(row.names(G_3m_otu)),]
G_1y_otu <- G_1y_otu[order(row.names(G_1y_otu)),]

## Checking - colnames and Childid's are in the same order? Yes - good - can move forward
identical(colnames(G_3m_otu), colnames(G_1y_otu))
identical(row.names(G_3m_otu), row.names(G_1y_otu))

## 1 Year minus 3 Month Prevalence For each Taxa
change_3m1y_pa_G <- G_1y_otu - G_3m_otu # -1 = present at 3m not 1y, 1= present at 1y not 3m, 0=either present at both or neither timpoint
dim(change_3m1y_pa_G) # Same order as original physeq obj

sdf_toAdd <- sample_data(G.pa.1y)[order(sample_data(G.pa.1y)$CHILDid),]
sdf_3m <- sample_data(G.pa.3m)[order(sample_data(G.pa.3m)$CHILDid),]
## check that it's in the same order as change_3m1y_pa
identical(row.names(change_3m1y_pa_G), sdf_toAdd$CHILDid) # order matches
identical(row.names(change_3m1y_pa_G), sdf_3m$CHILDid) # order matches

### same order so just merge data with sample metadata using data.frame
change_3m1y_pa_G_sdf <- data.frame(change_3m1y_pa_G, sdf_toAdd, RE_virus_3m = sdf_3m$RE_Virus) # Using 1 year sdf for variables but also adding RE virus at 3 months as "RE_virus_3m". Note that "RE_Virus" is the RE virus at 1 year

N_taxa_G <- ncol(change_3m1y_pa_G) # number of taxa, to use in downstream code
N_subj_G <- nrow(change_3m1y_pa_G) # number of subjects, to use in downstream code
```
Objects used for gut trajectory plots: change_3m1y_pa_G_sdf, Gut_3M1Y_pa_filt, N_taxa_G, N_subj_G 

## Save prevalence trajectory feature tables for later use (in mediation analysis, Fig. 6)
```{r}
## don't save with metadata - just save new derived data with CHILDid's as rownames: change_3m1y_pa, change_3m1y_pa_G; 
### merge before saving, but first differentiate nasal vs gut trajectory features by prepending N or G
colnames(change_3m1y_pa) <- gsub("^", "N_", colnames(change_3m1y_pa))
colnames(change_3m1y_pa_G) <- gsub("^", "G_", colnames(change_3m1y_pa_G))

change_3m1y_pa_all <- merge(change_3m1y_pa, change_3m1y_pa_G, by="row.names", all=TRUE)
row.names(change_3m1y_pa_all) <- change_3m1y_pa_all$Row.names
change_3m1y_pa_all$Row.names <- NULL
dim(change_3m1y_pa_all)
write.csv(change_3m1y_pa_all, file="Output/Prevalence_Trajectory_Features_N171_G172.csv") # The first 171 columns are nasal taxa, 172 to end of df is gut
```

```{r} 
# setup for plots
## for plot annotations:
Main_phyla <- c('Proteobacteria','Firmicutes', "Actinobacteria", "Bacteroidetes")
```


### Extra Prevalence Trajectory analyses (Not shown but used for Figure 3D & S5)
```{r} 
# creating wheeze trajectory /atopy at 3 years combination variable

## Nasal
change_3m1y_pa_sdf$Wheeze_atopy_Traj <- ifelse(!is.na(change_3m1y_pa_sdf$WheezeTraj) & !is.na(change_3m1y_pa_sdf$atopy3y), paste(change_3m1y_pa_sdf$WheezeTraj, change_3m1y_pa_sdf$atopy3y, sep = ", Atopy "), NA) 

## Gut
change_3m1y_pa_G_sdf$Wheeze_atopy_Traj <- ifelse(!is.na(change_3m1y_pa_G_sdf$WheezeTraj) & !is.na(change_3m1y_pa_G_sdf$atopy3y), paste(change_3m1y_pa_G_sdf$WheezeTraj, change_3m1y_pa_G_sdf$atopy3y, sep = ", Atopy "), NA) # Using Atopy at 3Y b/c had better luck with that & b/c asthma_3y is the main asthma variable used. 

# Focusing on:  "Persistent, Atopy No"; "Never/infrequent, Atopy Yes"; "Never/infrequent, Atopy No"
```

#### Nasal, Persistent Wheeze without Atopy at 3 years vs. no Wheeze or Atopy 
```{r} 
X <- subset(change_3m1y_pa_sdf, Wheeze_atopy_Traj == "Persistent, Atopy No")
Y <- subset(change_3m1y_pa_sdf, Wheeze_atopy_Traj == "Never/infrequent, Atopy No")

# Make data.frame of prevalence like before BUT keeping as proprotions instead of Percentages for now
df.pa.nasal.Whz.atopy <- data.frame(X_P_Change=colSums(X[1:N_taxa])*100/dim(X)[1],Y_P_Change=colSums(Y[1:N_taxa])*100/dim(Y)[1], taxonomy=as(tax_table(Nasal_3M1Y_pa_filt), "matrix"), prevalence=taxa_sums(Nasal_3M1Y_pa_filt)*100/(N_subj*2)) 

# difference in change
df.pa.nasal.Whz.atopy$pa.change.diff <- df.pa.nasal.Whz.atopy$Y_P_Change - df.pa.nasal.Whz.atopy$X_P_Change
```
- Aesthetic things
```{r} 
## Plot by major phylum & annotate signif taxa
df.pa.nasal.Whz.atopy$Phylum <- ifelse(df.pa.nasal.Whz.atopy$taxonomy.Phylum %in% Main_phyla, df.pa.nasal.Whz.atopy$taxonomy.Phylum, "Other")
#Reorder the phyla 
df.pa.nasal.Whz.atopy$Phylum = factor(df.pa.nasal.Whz.atopy$Phylum, levels=c(Main_phyla, "Other"))
```
- Plot Change in Prevalence 
```{r} 
x_lab <- paste("Wheeze, no Atopy (n=", nrow(X), ")\n1 Year - 3 Month Prevalence (%)", sep="")
y_lab <- paste("No Wheeze, no Atopy (n=", nrow(Y), ")\n1 Year - 3 Month Prevalence (%)", sep="")

Fig_Ext_N_Wheeze <- ggplot() + 
  geom_point(data=df.pa.nasal.Whz.atopy, aes(x=X_P_Change, y=Y_P_Change, color=Phylum, size=prevalence), alpha=0.75) + 
  xlab(x_lab) + ylab(y_lab) + 
  scale_color_manual(values = cbPalette) +
  scale_y_continuous(limits=c(-37,60.1), expand = c(0,0))+
  scale_x_continuous(limits=c(-37,60.1), expand = c(0,0))+
  annotate("segment", x = -37, xend = 60, y = -30, yend = 60, colour = "red", size=0.5) + ggtitle("Nasal")
Fig_Ext_N_Wheeze
```

**Permutation test for Prevalence Trajectory analysis - See Permutation_Tests_PrevTrajectories directory for source code**
```{r} 
source("Permutation_Tests_PrevTrajectories/09_Wheeze_Nasal.R")

# Test output
# Res_per_taxa_wheeze_N - p-value assigned to each taxa; p_value = later in healthy; p_value_2 = earlier in healthy
original_wilcox_wheeze_N # Overall significance test
```


#### Gut, Persistent Wheeze without Atopy at 3 years vs. no Wheeze or Atopy (Not shown but used for Figure 3D & S5)
```{r} 
X <- subset(change_3m1y_pa_G_sdf, Wheeze_atopy_Traj == "Persistent, Atopy No")
Y <- subset(change_3m1y_pa_G_sdf, Wheeze_atopy_Traj == "Never/infrequent, Atopy No")

# Make data.frame of prevalence like before BUT keeping as proprotions instead of Percentages for now
df.pa.gut.Whz.atopy <- data.frame(X_P_Change=colSums(X[1:N_taxa_G])*100/dim(X)[1],Y_P_Change=colSums(Y[1:N_taxa_G])*100/dim(Y)[1], taxonomy=as(tax_table(Gut_3M1Y_pa_filt), "matrix"), prevalence=taxa_sums(Gut_3M1Y_pa_filt)*100/(N_subj_G*2)) 

# difference in change
df.pa.gut.Whz.atopy$pa.change.diff <- df.pa.gut.Whz.atopy$Y_P_Change - df.pa.gut.Whz.atopy$X_P_Change
```
- Aesthetic things
```{r} 
## Plot by major phylum & annotate signif taxa
df.pa.gut.Whz.atopy$Phylum <- ifelse(df.pa.gut.Whz.atopy$taxonomy.Phylum %in% Main_phyla, df.pa.gut.Whz.atopy$taxonomy.Phylum, "Other")
#Reorder the phyla 
df.pa.gut.Whz.atopy$Phylum = factor(df.pa.gut.Whz.atopy$Phylum, levels=c(Main_phyla, "Other"))
```
- Plot Change in Prevalence 
```{r} 
x_lab <- paste("Wheeze, no Atopy (n=", nrow(X), ")\n1 Year - 3 Month Prevalence (%)", sep="")
y_lab <- paste("No Wheeze, no Atopy (n=", nrow(Y), ")\n1 Year - 3 Month Prevalence (%)", sep="")

Fig_Ext_G_Wheeze <- ggplot() + 
  geom_point(data=df.pa.gut.Whz.atopy, aes(x=X_P_Change, y=Y_P_Change, color=Phylum, size=prevalence), alpha=0.75) + 
  xlab(x_lab) + ylab(y_lab) + 
  scale_color_manual(values = cbPalette) +
  scale_y_continuous(limits=c(-37,60.1), expand = c(0,0))+
  scale_x_continuous(limits=c(-37,60.1), expand = c(0,0))+
  annotate("segment", x = -37, xend = 60, y = -30, yend = 60, colour = "red", size=0.5) + ggtitle("Gut")
Fig_Ext_G_Wheeze
```

**Permutation test for Prevalence Trajectory analysis - See Permutation_Tests_PrevTrajectories directory for source code**
```{r} 
source("Permutation_Tests_PrevTrajectories/10_Wheeze_Gut.R")

# Test output
# Res_per_taxa_wheeze_G - p-value assigned to each taxa; p_value = later in healthy; p_value_2 = earlier in healthy
original_wilcox_wheeze_G # Overall significance test
```

#### Nasal, Atopy at 3 years without Persistent Wheeze vs. no Atopy or Wheeze (Not shown but used for Figure 3D & S5)
```{r} 
X <- subset(change_3m1y_pa_sdf, Wheeze_atopy_Traj == "Never/infrequent, Atopy Yes")
Y <- subset(change_3m1y_pa_sdf, Wheeze_atopy_Traj == "Never/infrequent, Atopy No")

# Make data.frame of prevalence like before BUT keeping as proprotions instead of Percentages for now
df.pa.nasal.NoWhz.atopy <- data.frame(X_P_Change=colSums(X[1:N_taxa])*100/dim(X)[1],Y_P_Change=colSums(Y[1:N_taxa])*100/dim(Y)[1], taxonomy=as(tax_table(Nasal_3M1Y_pa_filt), "matrix"), prevalence=taxa_sums(Nasal_3M1Y_pa_filt)*100/(N_subj*2)) 

# difference in change
df.pa.nasal.NoWhz.atopy$pa.change.diff <- df.pa.nasal.NoWhz.atopy$Y_P_Change - df.pa.nasal.NoWhz.atopy$X_P_Change
```
- Aesthetic things
```{r} 
## Plot by major phylum & annotate signif taxa
df.pa.nasal.NoWhz.atopy$Phylum <- ifelse(df.pa.nasal.NoWhz.atopy$taxonomy.Phylum %in% Main_phyla, df.pa.nasal.NoWhz.atopy$taxonomy.Phylum, "Other")
#Reorder the phyla 
df.pa.nasal.NoWhz.atopy$Phylum = factor(df.pa.nasal.NoWhz.atopy$Phylum, levels=c(Main_phyla, "Other"))
```
- Plot Change in Prevalence 
```{r} 
x_lab <- paste("Atopy, no Wheeze (n=", nrow(X), ")\n1 Year - 3 Month Prevalence (%)", sep="")
y_lab <- paste("No Atopy, no Wheeze (n=", nrow(Y), ")\n1 Year - 3 Month Prevalence (%)", sep="")

Fig_Ext_N_Atopy <- ggplot() + 
  geom_point(data=df.pa.nasal.NoWhz.atopy, aes(x=X_P_Change, y=Y_P_Change, color=Phylum, size=prevalence), alpha=0.75) + 
  xlab(x_lab) + 
  ylab(y_lab) + 
  scale_color_manual(values = cbPalette) +
  scale_y_continuous(limits=c(-37,60.1), expand = c(0,0))+
  scale_x_continuous(limits=c(-37,60.1), expand = c(0,0))+
  annotate("segment", x = -37, xend = 60, y = -37, yend = 60, colour = "red", size=0.5) + ggtitle("Nasal")
Fig_Ext_N_Atopy
```

**Permutation test for Prevalence Trajectory analysis - See Permutation_Tests_PrevTrajectories directory for source code**
```{r} 
source("Permutation_Tests_PrevTrajectories/11_Atopy_Nasal.R")

# Test output
# Res_per_taxa_atopy_N - p-value assigned to each taxa; p_value = later in healthy; p_value_2 = earlier in healthy
original_wilcox_atopy_N # Overall significance test
```


#### Gut, Atopy at 3 years without Persistent Wheeze vs. no Atopy or Wheeze (Not shown but used for Figure 3D & S5)
```{r} 
X <- subset(change_3m1y_pa_G_sdf, Wheeze_atopy_Traj == "Never/infrequent, Atopy Yes")
Y <- subset(change_3m1y_pa_G_sdf, Wheeze_atopy_Traj == "Never/infrequent, Atopy No")

# Make data.frame of prevalence like before BUT keeping as proprotions instead of Percentages for now
df.pa.gut.NoWhz.atopy <- data.frame(X_P_Change=colSums(X[1:N_taxa_G])*100/dim(X)[1],Y_P_Change=colSums(Y[1:N_taxa_G])*100/dim(Y)[1], taxonomy=as(tax_table(Gut_3M1Y_pa_filt), "matrix"), prevalence=taxa_sums(Gut_3M1Y_pa_filt)*100/(N_subj_G*2)) 

# difference in change
df.pa.gut.NoWhz.atopy$pa.change.diff <- df.pa.gut.NoWhz.atopy$Y_P_Change - df.pa.gut.NoWhz.atopy$X_P_Change
```
- Aesthetic things
```{r} 
## Plot by major phylum & annotate signif taxa
df.pa.gut.NoWhz.atopy$Phylum <- ifelse(df.pa.gut.NoWhz.atopy$taxonomy.Phylum %in% Main_phyla, df.pa.gut.NoWhz.atopy$taxonomy.Phylum, "Other")
#Reorder the phyla 
df.pa.gut.NoWhz.atopy$Phylum = factor(df.pa.gut.NoWhz.atopy$Phylum, levels=c(Main_phyla, "Other"))
```
- Plot Change in Prevalence 
```{r} 
x_lab <- paste("Atopy, no Wheeze (n=", nrow(X), ")\n1 Year - 3 Month Prevalence (%)", sep="")
y_lab <- paste("No Atopy, no Wheeze (n=", nrow(Y), ")\n1 Year - 3 Month Prevalence (%)", sep="")

Fig_Ext_G_Atopy <- ggplot() + 
  geom_point(data=df.pa.gut.NoWhz.atopy, aes(x=X_P_Change, y=Y_P_Change, color=Phylum, size=prevalence), alpha=0.75) + 
  xlab(x_lab) + ylab(y_lab) + 
  scale_color_manual(values = cbPalette) +
  scale_y_continuous(limits=c(-37,60.1), expand = c(0,0))+
  scale_x_continuous(limits=c(-37,60.1), expand = c(0,0))+
  annotate("segment", x = -37, xend = 60, y = -30, yend = 60, colour = "red", size=0.5) + ggtitle("Gut")
Fig_Ext_G_Atopy
```

**Permutation test for Prevalence Trajectory analysis - See Permutation_Tests_PrevTrajectories directory for source code**
```{r} 
source("Permutation_Tests_PrevTrajectories/12_Atopy_Gut.R")

# Test output
# Res_per_taxa_atopy_G - p-value assigned to each taxa; p_value = later in healthy; p_value_2 = earlier in healthy
original_wilcox_atopy_G # Overall significance test
```


#### Nasal, Colds 0-3 months (Not shown but used for Figure 3D & S5)
```{r} 
X <- subset(change_3m1y_pa_sdf, colds_0to3m == "Yes")
Y <- subset(change_3m1y_pa_sdf, colds_0to3m == "No")

# Make data.frame of prevalence like before BUT keeping as proprotions instead of Percentages for now
df.pa.nasal.C1 <- data.frame(X_P_Change=colSums(X[1:N_taxa])*100/dim(X)[1],Y_P_Change=colSums(Y[1:N_taxa])*100/dim(Y)[1], taxonomy=as(tax_table(Nasal_3M1Y_pa_filt), "matrix"), prevalence=taxa_sums(Nasal_3M1Y_pa_filt)*100/(N_subj*2)) 

# difference in change
df.pa.nasal.C1$pa.change.diff <- df.pa.nasal.C1$Y_P_Change - df.pa.nasal.C1$X_P_Change
```
Aesthetic things
```{r} 
## Plot by major phylum & annotate signif taxa
df.pa.nasal.C1$Phylum <- ifelse(df.pa.nasal.C1$taxonomy.Phylum %in% Main_phyla, df.pa.nasal.C1$taxonomy.Phylum, "Other")
#Reorder the phyla 
df.pa.nasal.C1$Phylum = factor(df.pa.nasal.C1$Phylum, levels=c(Main_phyla, "Other"))
```
Plot Change in Prevalence 
```{r} 
x_lab <- paste("Colds (n=", nrow(X), ")\n1 Year - 3 Month Prevalence (%)", sep="")
y_lab <- paste("No Colds (n=", nrow(Y), ")\n1 Year - 3 Month Prevalence (%)", sep="")

Fig_Ext_N_Cold <- ggplot() + 
  geom_point(data=df.pa.nasal.C1, aes(x=X_P_Change, y=Y_P_Change, color=Phylum, size=prevalence), alpha=0.75) + 
  xlab(x_lab) + ylab(y_lab) + 
  scale_color_manual(values = cbPalette) +
  scale_y_continuous(limits=c(-37,60.1), expand = c(0,0))+
  scale_x_continuous(limits=c(-37,60.1), expand = c(0,0))+
  annotate("segment", x = -37, xend = 60, y = -37, yend = 60, colour = "red", size=0.5) + ggtitle("Nasal")
Fig_Ext_N_Cold
```

**Permutation test for Prevalence Trajectory analysis - See Permutation_Tests_PrevTrajectories directory for source code**
```{r} 
source("Permutation_Tests_PrevTrajectories/13_Colds_Nasal.R")

# Test output
# Res_per_taxa_Cold_N - p-value assigned to each taxa; p_value = later in healthy; p_value_2 = earlier in healthy
original_wilcox_Cold_N # Overall significance test
```


#### Gut, Colds 0-3 months (Not shown but used for Figure 3D & S5)
```{r} 
X <- subset(change_3m1y_pa_G_sdf, colds_0to3m == "Yes")
Y <- subset(change_3m1y_pa_G_sdf, colds_0to3m == "No")

# Make data.frame of prevalence like before BUT keeping as proprotions instead of Percentages for now
df.pa.gut.C <- data.frame(X_P_Change=colSums(X[1:N_taxa_G])*100/dim(X)[1],Y_P_Change=colSums(Y[1:N_taxa_G])*100/dim(Y)[1], taxonomy=as(tax_table(Gut_3M1Y_pa_filt), "matrix"), prevalence=taxa_sums(Gut_3M1Y_pa_filt)*100/(N_subj_G*2))

# difference in change
df.pa.gut.C$pa.change.diff <- df.pa.gut.C$Y_P_Change - df.pa.gut.C$X_P_Change
```
Asthetic things, not important
```{r} 
## Plot by major phylum & annotate signif taxa
df.pa.gut.C$Phylum <- ifelse(df.pa.gut.C$taxonomy.Phylum %in% Main_phyla, df.pa.gut.C$taxonomy.Phylum, "Other")
#Reorder the phyla 
df.pa.gut.C$Phylum = factor(df.pa.gut.C$Phylum, levels=c(Main_phyla, "Other"))
```
Plot Change in Prevalence 
```{r} 
x_lab <- paste("Colds (n=", nrow(X), ")\n1 Year - 3 Month Prevalence (%)", sep="")
y_lab <- paste("No Colds (n=", nrow(Y), ")\n1 Year - 3 Month Prevalence (%)", sep="")

Fig_Ext_G_Cold <- ggplot() + 
  geom_point(data=df.pa.gut.C, aes(x=X_P_Change, y=Y_P_Change, color=Phylum, size=prevalence), alpha=0.75) + 
  xlab(x_lab) + ylab(y_lab) + 
  scale_color_manual(values = cbPalette) +
  scale_y_continuous(limits=c(-37,60), expand = c(0,0))+
  scale_x_continuous(limits=c(-37,60), expand = c(0,0))+
  annotate("segment", x = -37, xend = 60, y = -37, yend = 60, colour = "red", size=0.5) + ggtitle("Gut")
Fig_Ext_G_Cold
```

**Permutation test for Prevalence Trajectory analysis - See Permutation_Tests_PrevTrajectories directory for source code**
```{r} 
source("Permutation_Tests_PrevTrajectories/14_Colds_Gut.R")

# Test output
# Res_per_taxa_Cold_G - p-value assigned to each taxa; p_value = later in healthy; p_value_2 = earlier in healthy
original_wilcox_Cold_G # Overall significance test
```

### Compile & save all prevalence trajectory results
```{r}
# compile & save results for individ taxa

# saving everything to output
Per_test_df_ls <- list(Res_per_taxa_ast_N, Res_per_taxa_BF_N, Res_per_taxa_ast5y_N, 
                       Res_per_taxa_ast3y5y_N, Res_per_taxa_wheeze_N, Res_per_taxa_atopy_N, Res_per_taxa_Cold_N, 
                       Res_per_taxa_ast_G, Res_per_taxa_BF_G, Res_per_taxa_ast5y_G, Res_per_taxa_ast3y5y_G, 
                       Res_per_taxa_wheeze_G, Res_per_taxa_atopy_G, Res_per_taxa_Cold_G)
names(Per_test_df_ls) <-  c("Ast3y_Nasal", "BF_Nasal", "Ast5y_Nasal", "Ast3y_5y_Nasal", "Wheeze_Nasal", "Atopy_Nasal", "Cold_Nasal", 
                            "Ast3y_Gut", "BF_Gut", "Ast5y_Gut", "Ast3y_5y_Gut", "Wheeze_Gut", "Atopy_Gut", "Cold_Gut")

# Save as an overall rds file in Output/PermTests
saveRDS(Per_test_df_ls, "Output/PermTests/ALL_Permutation_PrevTraj_Res.rds")

# Save as individual csv files for others 
outpath_Pt1 <- "Output/PermTests/Permutation_PrevTraj_Res"
outfiles <- paste(outpath_Pt1, names(Per_test_df_ls), ".csv", sep = "_")

mapply(function(x, y) write.csv(x, file=y, row.names = FALSE, quote = FALSE), x=Per_test_df_ls, y=outfiles, SIMPLIFY = FALSE)
```


# Figure S5 - Upset plots for overlap among taxa identified as significantly later colonizers in the 'healthy' group
```{r}
Per_test_df_ls_Sig <- lapply(Per_test_df_ls, function(x) {subset(x, p_value < 0.05)})
lapply(Per_test_df_ls_Sig, function(x) {nrow(x)}) ## nrow of each dataframe is the number of significant late taxa with p<0.05 - of the 170 testable Nasal taxa and 115 gut taxa (not including the Strep.40 present in 100% of samples)

Per_test_df_ls_Sig_0.01 <- lapply(Per_test_df_ls, function(x) {subset(x, p_value < 0.01)})
lapply(Per_test_df_ls_Sig_0.01, function(x) {nrow(x)}) ## nrow of each dataframe is the number of significant late taxa with p<0.05 - of the 170 testable taxa (not including the Strep.40 present in 100% of samples)
```

## Nasal, Overlap in significant late colonizers between groups (Fig S5B & S5D)
- p value < 0.05 
```{r} 
Per_test_df_ls_Sig_N <- Per_test_df_ls_Sig[1:7] # nasal dataframes

listInput <- list(Asthma = row.names(Per_test_df_ls_Sig_N$Ast3y_Nasal), Colds = row.names(Per_test_df_ls_Sig_N$Cold_Nasal), 
                  Wheeze = row.names(Per_test_df_ls_Sig_N$Wheeze_Nasal), Atopy = row.names(Per_test_df_ls_Sig_N$Atopy_Nasal), 
                  BF = row.names(Per_test_df_ls_Sig_N$BF_Nasal))

tiff(file = "Figures/Fig_S5B_N_p0.05_upset.tiff", width = 1200, height = 1200, units = "px", res = 300) 
upset(fromList(listInput), keep.order = T, nsets = 6, sets = rev(c("BF", "Asthma", "Wheeze", "Colds", "Atopy"))) 
dev.off()

df_listInput_1 <- fromList2(listInput)
# checking total overlap with Asthma for each group
sum(df_listInput_1$BF == 1 & df_listInput_1$Asthma == 1) # 17
sum(df_listInput_1$Wheeze == 1 & df_listInput_1$Asthma == 1) # 17
sum(df_listInput_1$Colds == 1 & df_listInput_1$Asthma == 1) # 12
sum(df_listInput_1$Atopy == 1 & df_listInput_1$Asthma == 1) # 3 
```
- p value < 0.01 
```{r} 
Per_test_df_ls_Sig_0.01N <- Per_test_df_ls_Sig_0.01[1:7] # nasal dataframes

listInput2 <- list(Asthma = row.names(Per_test_df_ls_Sig_0.01N$Ast3y_Nasal), Colds = row.names(Per_test_df_ls_Sig_0.01N$Cold_Nasal), 
                  Wheeze = row.names(Per_test_df_ls_Sig_0.01N$Wheeze_Nasal), Atopy = row.names(Per_test_df_ls_Sig_0.01N$Atopy_Nasal), 
                  BF = row.names(Per_test_df_ls_Sig_0.01N$BF_Nasal))


tiff(file = "Figures/Fig_S5D_N_p0.01_upset.tiff", width = 1200, height = 1200, units = "px", res = 300) 
upset(fromList(listInput2), keep.order = T, sets = rev(c("BF", "Asthma", "Wheeze", "Colds", "Atopy")), nsets = 6)
dev.off()

df_listInput2 <- fromList2(listInput2) # IDs of overlap taxa
```

## Gut, Overlap in significant taxa between groups (Fig S5C & S5E)
- p value < 0.05
```{r} 
Per_test_df_ls_Sig_G <- Per_test_df_ls_Sig[8:14] # gut dataframes

listInputG <- list(Asthma = row.names(Per_test_df_ls_Sig_G$Ast3y_Gut), Colds = row.names(Per_test_df_ls_Sig_G$Cold_Gut), 
                  Wheeze = row.names(Per_test_df_ls_Sig_G$Wheeze_Gut), Atopy = row.names(Per_test_df_ls_Sig_G$Atopy_Gut), 
                  BF = row.names(Per_test_df_ls_Sig_G$BF_Gut))

tiff(file = "Figures/Fig_S5C_G_p0.05_upset.tiff", width = 1200, height = 1200, units = "px", res = 300) 
upset(fromList(listInputG), keep.order = T, nsets = 6, sets = rev(c("BF", "Asthma", "Wheeze", "Colds", "Atopy"))) 
dev.off()

df_listInputG_1 <- fromList2(listInputG)
# checking total overlap with Asthma for each group
sum(df_listInputG_1$BF == 1 & df_listInputG_1$Asthma == 1) # 10
sum(df_listInputG_1$Wheeze == 1 & df_listInputG_1$Asthma == 1) # 1
sum(df_listInputG_1$Colds == 1 & df_listInputG_1$Asthma == 1) # 3
sum(df_listInputG_1$Atopy == 1 & df_listInputG_1$Asthma == 1) # 0
```
- p value < 0.01 
```{r} 
Per_test_df_ls_Sig0.01G <- Per_test_df_ls_Sig_0.01[8:14] # gut dataframes

listInput2G <- list(Asthma = row.names(Per_test_df_ls_Sig0.01G$Ast3y_Gut), Colds = row.names(Per_test_df_ls_Sig0.01G$Cold_Gut), 
                  Wheeze = row.names(Per_test_df_ls_Sig0.01G$Wheeze_Gut), Atopy = row.names(Per_test_df_ls_Sig0.01G$Atopy_Gut), 
                  BF = row.names(Per_test_df_ls_Sig0.01G$BF_Gut))

tiff(file = "Figures/Fig_S5C_G_p0.01_upset.tiff", width = 1200, height = 1200, units = "px", res = 300) 
upset(fromList(listInput2G), keep.order = T, sets = rev(c("BF", "Asthma", "Wheeze", "Colds", "Atopy")), nsets = 6)
dev.off()

df_listInput2G <- fromList2(listInput2G) # IDs of overlap taxa
```


