---
title: "Figure S6. Overlap in associations between microbiota prevalence trajectories and breastfeeding and respiratory phenotypes."
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Setup 

```{r, warning=FALSE, message=FALSE}
R.version.string
library(phyloseq); packageVersion('phyloseq')
library(tidyr); packageVersion('tidyr')
library(reshape2); packageVersion('reshape2')
library(dplyr); packageVersion('dplyr')  
library(UpSetR); packageVersion('UpSetR')
library(here)

source(paste0(here::here(), "/Functions.R"))
```

### Microbiota Trajectory data subsets used
```{r}
Nasal_3M1Y_pa_filt <- readRDS(paste0(path, "Data/Data_subsets/Main_Nasal_Trajectory_Subset.rds")) # Prevalence dataset

Gut_3M1Y_pa_filt <- readRDS(paste0(path, "Data/Data_subsets/Main_Gut_Trajectory_Subset.rds")) # Prevalence dataset
```

```{r}
# splitting by timepoint for downstream analyses
N.pa.3m <- subset_samples(Nasal_3M1Y_pa_filt, Visit == "3 Months")
N.pa.1y <- subset_samples(Nasal_3M1Y_pa_filt, Visit == "1 Year")

G.pa.3m <- subset_samples(Gut_3M1Y_pa_filt, Visit == "3 Months")
G.pa.1y <- subset_samples(Gut_3M1Y_pa_filt, Visit == "1 Year")
```


# Prevalence Trajectory Coordinate System & Permutation tests for Figure S6 
**similar scripts compared to main Figure 3**

## Nasal - Prevalence Trajectory derived variables, creation of a prevalence trajectory feature table
To simplify making of prevalence trajectory scatter plots, first make the 'difference in prevalence' results for every pair of samples for each infant (1Y - 3M, e.g. 0 - 1 = -1; 1 - 0 = 1) - so a "-1" means it's present at 3 months but not 1 year, 1 means 1 year but not 3m and 0 means no change.
```{r} 
## extract otu table and use Participant IDs as row.names instead - will match b/t datasets
N_3m_otu <- data.frame(as.data.frame(t(as(otu_table(N.pa.3m), "matrix"))), row.names=sample_data(N.pa.3m)$CHILDid)
N_1y_otu <- data.frame(as.data.frame(t(as(otu_table(N.pa.1y), "matrix"))), row.names=sample_data(N.pa.1y)$CHILDid)

N_3m_otu <- N_3m_otu[order(row.names(N_3m_otu)),]
N_1y_otu <- N_1y_otu[order(row.names(N_1y_otu)),]

## Checking - colnames and Childid's are in the same order? Yes - good - can move forward
identical(colnames(N_3m_otu), colnames(N_1y_otu))
identical(row.names(N_1y_otu), row.names(N_3m_otu))

## 1 Year minus 3 Month Prevalence For each Taxa
change_3m1y_pa <- N_1y_otu - N_3m_otu # -1 = present at 3m not 1y, 1= present at 1y not 3m, 0=either present at both or neither timpoint
dim(change_3m1y_pa) # Same order as original physeq obj

# Sample data to merge with dataset
sdf_toAdd <- sample_data(N.pa.1y)[order(sample_data(N.pa.1y)$CHILDid),]
sdf_3m <- sample_data(N.pa.3m)[order(sample_data(N.pa.3m)$CHILDid),]
## check that it's in the same order as change_3m1y_pa
identical(row.names(change_3m1y_pa), sdf_toAdd$CHILDid) 
identical(row.names(change_3m1y_pa), sdf_3m$CHILDid) 

### same order so just merge data with sample metadata using data.frame
change_3m1y_pa_sdf <- data.frame(change_3m1y_pa, sdf_toAdd, RE_virus_3m = sdf_3m$RE_Virus) # Using 1 year sdf for variables (will be same for 3m and 1y for almost all) - but also adding RE virus at 3 months as "RE_virus_3m". Note that "RE_Virus" is the RE virus at 1 year.

N_taxa <- ncol(change_3m1y_pa) # number of taxa, to use in downstream code
N_subj <- nrow(change_3m1y_pa) # number of subjects, to use in downstream code
```
Objects used for Nasal trajectory plots: change_3m1y_pa_sdf, Nasal_3M1Y_pa_filt, N_taxa, N_subj 

## Gut - Prevalence Trajectory derived variables, creation of a prevalence trajectory feature table
To simplify making of prevalence trajectory scatter plots, first make the 'difference in prevalence' results for every pair of samples for each infant (1Y - 3M, e.g. 0 - 1 = -1; 1 - 0 = 1) - so a "-1" means it's present at 3 months but not 1 year, 1 means 1 year but not 3m and 0 means no change.
```{r} 
## extract otu table and use Participant IDs as row.names instead - will match b/t datasets
G_3m_otu <- data.frame(as.data.frame(t(as(otu_table(G.pa.3m), "matrix"))), row.names=sample_data(G.pa.3m)$CHILDid)
G_1y_otu <- data.frame(as.data.frame(t(as(otu_table(G.pa.1y), "matrix"))), row.names=sample_data(G.pa.1y)$CHILDid)

G_3m_otu <- G_3m_otu[order(row.names(G_3m_otu)),]
G_1y_otu <- G_1y_otu[order(row.names(G_1y_otu)),]

## Checking - colnames and Childid's are in the same order? Yes - good - can move forward
identical(colnames(G_3m_otu), colnames(G_1y_otu))
identical(row.names(G_3m_otu), row.names(G_1y_otu))

## 1 Year minus 3 Month Prevalence For each Taxa
change_3m1y_pa_G <- G_1y_otu - G_3m_otu # -1 = present at 3m not 1y, 1= present at 1y not 3m, 0=either present at both or neither timpoint
dim(change_3m1y_pa_G) # Same order as original physeq obj

sdf_toAdd <- sample_data(G.pa.1y)[order(sample_data(G.pa.1y)$CHILDid),]
sdf_3m <- sample_data(G.pa.3m)[order(sample_data(G.pa.3m)$CHILDid),]
## check that it's in the same order as change_3m1y_pa
identical(row.names(change_3m1y_pa_G), sdf_toAdd$CHILDid) # order matches
identical(row.names(change_3m1y_pa_G), sdf_3m$CHILDid) # order matches

### same order so just merge data with sample metadata using data.frame
change_3m1y_pa_G_sdf <- data.frame(change_3m1y_pa_G, sdf_toAdd, RE_virus_3m = sdf_3m$RE_Virus) # Using 1 year sdf for variables but also adding RE virus at 3 months as "RE_virus_3m". Note that "RE_Virus" is the RE virus at 1 year

N_taxa_G <- ncol(change_3m1y_pa_G) # number of taxa, to use in downstream code
N_subj_G <- nrow(change_3m1y_pa_G) # number of subjects, to use in downstream code
```
Objects used for gut trajectory plots: change_3m1y_pa_G_sdf, Gut_3M1Y_pa_filt, N_taxa_G, N_subj_G 


#### Extra variables - Asthma at 3 years that persists to 5 years vs. no asthma at 3 and no asthma at 5 years and atopy without wheeze (just used in Figure S6B and S6C)
```{r} 
# want probable asthma at 3 years, and probable or possible at 5 years. Those missing 5 year or 3 year asthma data will be NA for this variable
change_3m1y_pa_sdf$asthma_3y_5y <- ifelse(change_3m1y_pa_sdf$asthma_3y == "Probable" & 
                                            change_3m1y_pa_sdf$asthma_5y %in% c("Probable", "Possible"), "Yes", 
                                          ifelse(change_3m1y_pa_sdf$asthma_3y == "No" & 
                                            change_3m1y_pa_sdf$asthma_5y == "No", "No", NA))
table(change_3m1y_pa_sdf$asthma_3y_5y)

# want probable asthma at 3 years, and probable or possible at 5 years. Those missing 5 year or 3 year asthma data will be NA for this variable  
change_3m1y_pa_G_sdf$asthma_3y_5y <- ifelse(change_3m1y_pa_G_sdf$asthma_3y == "Probable" & 
                                            change_3m1y_pa_G_sdf$asthma_5y  %in% c("Probable", "Possible"), "Yes", 
                                          ifelse(change_3m1y_pa_G_sdf$asthma_3y == "No" & 
                                            change_3m1y_pa_G_sdf$asthma_5y == "No", "No", NA))

table(change_3m1y_pa_G_sdf$asthma_3y_5y)

# Atopy without wheeze
## Nasal
change_3m1y_pa_sdf$Wheeze_atopy_Traj <- ifelse(!is.na(change_3m1y_pa_sdf$WheezeTraj) & !is.na(change_3m1y_pa_sdf$atopy3y), paste(change_3m1y_pa_sdf$WheezeTraj, change_3m1y_pa_sdf$atopy3y, sep = ", Atopy "), NA) 

## Gut
change_3m1y_pa_G_sdf$Wheeze_atopy_Traj <- ifelse(!is.na(change_3m1y_pa_G_sdf$WheezeTraj) & !is.na(change_3m1y_pa_G_sdf$atopy3y), paste(change_3m1y_pa_G_sdf$WheezeTraj, change_3m1y_pa_G_sdf$atopy3y, sep = ", Atopy "), NA) 
# Focusing on:  "Persistent, Atopy No"; "Never/infrequent, Atopy Yes"; "Never/infrequent, Atopy No"
```


#### Nasal - Breastfeeding
```{r} 
X <- subset(change_3m1y_pa_sdf, BF_status_3m_sampling == "No BM")
Y <- subset(change_3m1y_pa_sdf, BF_status_3m_sampling == "Exclusive")

# Make data.frame of prevalence like before BUT keeping as proprotions instead of Percentages for now
df.pa.nasal.BF <- data.frame(X_P_Change=colSums(X[1:N_taxa])*100/dim(X)[1],Y_P_Change=colSums(Y[1:N_taxa])*100/dim(Y)[1],
                             taxonomy=as(tax_table(Nasal_3M1Y_pa_filt), "matrix"), prevalence=taxa_sums(Nasal_3M1Y_pa_filt)*100/(N_subj*2)) # Taxonomy order actually maintained compared to physeq obj Nasal_3M1Y_pa_filt so can merge this way. prevalence column is the total prevalence so divided by total N across both timepoints (subject # x 2). 

# difference in change
df.pa.nasal.BF$pa.change.diff <- df.pa.nasal.BF$Y_P_Change - df.pa.nasal.BF$X_P_Change
```
**Permutation test for Prevalence Trajectory analysis - See 'Functions.R' for perm_func_dfs function**
```{r} 
Res_BF_N <- perm_func_dfs(df.pa.nasal.BF, change_3m1y_pa_sdf, 
                           variable = "BF_status_3m_sampling", Group_X = "No BM", Group_Y = "Exclusive", N_taxa = N_taxa)
# Test output
Res_BF_N$global_res # Overall significance test
# Res_per_taxa_BF_N - p-value assigned to each taxa; p_value = later in healthy; p_value_2 = earlier in healthy
Res_per_taxa_BF_N <- Res_BF_N$per_taxa_res
```

#### Gut - Breastfeeding
```{r} 
X <- subset(change_3m1y_pa_G_sdf, BF_status_3m_sampling == "No BM")
Y <- subset(change_3m1y_pa_G_sdf, BF_status_3m_sampling == "Exclusive")

# Make data.frame of prevalence like before BUT keeping as proprotions instead of Percentages for now
df.pa.gut.BF <- data.frame(X_P_Change=colSums(X[1:N_taxa_G])*100/dim(X)[1],Y_P_Change=colSums(Y[1:N_taxa_G])*100/dim(Y)[1], taxonomy=as(tax_table(Gut_3M1Y_pa_filt), "matrix"), prevalence=taxa_sums(Gut_3M1Y_pa_filt)*100/(N_subj_G*2)) # Taxonomy actually maintained compared to physeq obj, great

# difference in change
df.pa.gut.BF$pa.change.diff <- df.pa.gut.BF$Y_P_Change - df.pa.gut.BF$X_P_Change
```
**Permutation test for Prevalence Trajectory analysis**
```{r} 
Res_BF_G <- perm_func_dfs(df.pa.gut.BF, change_3m1y_pa_G_sdf, 
                           variable = "BF_status_3m_sampling", Group_X = "No BM", Group_Y = "Exclusive", N_taxa = N_taxa_G)

# Test output
Res_BF_G$global_res # Overall significance test
#  p-value assigned to each taxa; p_value = later in healthy; p_value_2 = earlier in healthy
Res_per_taxa_BF_G <- Res_BF_G$per_taxa_res
```

#### Nasal - Asthma at 3 years
```{r} 
X <- subset(change_3m1y_pa_sdf, asthma_3y == "Probable")
Y <- subset(change_3m1y_pa_sdf, asthma_3y == "No")

# Make data.frame of prevalence like before BUT keeping as proprotions instead of Percentages for now
df.pa.nasal.ast <- data.frame(X_P_Change=colSums(X[1:N_taxa])*100/dim(X)[1],Y_P_Change=colSums(Y[1:N_taxa])*100/dim(Y)[1], taxonomy=as(tax_table(Nasal_3M1Y_pa_filt), "matrix"), prevalence=taxa_sums(Nasal_3M1Y_pa_filt)*100/(N_subj*2)) 

# difference in change
df.pa.nasal.ast$pa.change.diff <- df.pa.nasal.ast$Y_P_Change - df.pa.nasal.ast$X_P_Change
```
**Permutation test for Prevalence Trajectory analysis**
```{r} 
Res_ast_N <- perm_func_dfs(df.pa.nasal.ast, change_3m1y_pa_sdf, 
                           variable = "asthma_3y", Group_X = "Probable", Group_Y = "No", N_taxa = N_taxa)

# Overall significance test for nasal, asthma at 3 years test
Res_ast_N$global_res # sig
# Res_per_taxa_ast_N - p-value assigned to each taxa; p_value = later in healthy; p_value_2 = earlier in healthy
Res_per_taxa_ast_N <- Res_ast_N$per_taxa_res
```

#### Gut - Asthma at 3 years
```{r} 
X <- subset(change_3m1y_pa_G_sdf, asthma_3y == "Probable")
Y <- subset(change_3m1y_pa_G_sdf, asthma_3y == "No")

# Make data.frame of prevalence like before BUT keeping as proprotions instead of Percentages for now
df.pa.gut.ast <- data.frame(X_P_Change=colSums(X[1:N_taxa_G])*100/dim(X)[1],Y_P_Change=colSums(Y[1:N_taxa_G])*100/dim(Y)[1], taxonomy=as(tax_table(Gut_3M1Y_pa_filt), "matrix"), prevalence=taxa_sums(Gut_3M1Y_pa_filt)*100/(N_subj_G*2)) 

# difference in change
df.pa.gut.ast$pa.change.diff <- df.pa.gut.ast$Y_P_Change - df.pa.gut.ast$X_P_Change
```
**Permutation test for Prevalence Trajectory analysis**
```{r} 
Res_ast_G <- perm_func_dfs(df.pa.gut.ast, change_3m1y_pa_G_sdf, 
                           variable = "asthma_3y", Group_X = "Probable", Group_Y = "No", N_taxa = N_taxa_G)
# Test output
Res_ast_G$global_res # Overall significance test
# p-value assigned to each taxa; p_value = later in healthy; p_value_2 = earlier in healthy
Res_per_taxa_ast_G <- Res_ast_G$per_taxa_res
```

#### Nasal - Asthma at 5 years 
```{r} 
X <- subset(change_3m1y_pa_sdf, asthma_5y == "Probable")
Y <- subset(change_3m1y_pa_sdf, asthma_5y == "No")

# Make data.frame of prevalence like before BUT keeping as proprotions instead of Percentages for now
df.pa.nasal.ast5 <- data.frame(X_P_Change=colSums(X[1:N_taxa])*100/dim(X)[1],Y_P_Change=colSums(Y[1:N_taxa])*100/dim(Y)[1], taxonomy=as(tax_table(Nasal_3M1Y_pa_filt), "matrix"), prevalence=taxa_sums(Nasal_3M1Y_pa_filt)*100/(N_subj*2)) 

# difference in change
df.pa.nasal.ast5$pa.change.diff <- df.pa.nasal.ast5$Y_P_Change - df.pa.nasal.ast5$X_P_Change
```
**Permutation test for Prevalence Trajectory analysis**
```{r} 
Res_ast5y_N <- perm_func_dfs(df.pa.nasal.ast5, change_3m1y_pa_sdf, 
                           variable = "asthma_5y", Group_X = "Probable", Group_Y = "No", N_taxa = N_taxa)
# Test output
Res_ast5y_N$global_res # Overall significance test; p_value = later in healthy
# Res_per_taxa_ast5y_N - p-value assigned to each taxa
Res_per_taxa_ast5y_N <- Res_ast5y_N$per_taxa_res
```

#### Gut - Asthma at 5 years
```{r} 
X <- subset(change_3m1y_pa_G_sdf, asthma_5y == "Probable")
Y <- subset(change_3m1y_pa_G_sdf, asthma_5y == "No")

# Make data.frame of prevalence like before BUT keeping as proprotions instead of Percentages for now
df.pa.gut.ast5 <- data.frame(X_P_Change=colSums(X[1:N_taxa_G])*100/dim(X)[1],Y_P_Change=colSums(Y[1:N_taxa_G])*100/dim(Y)[1], taxonomy=as(tax_table(Gut_3M1Y_pa_filt), "matrix"), prevalence=taxa_sums(Gut_3M1Y_pa_filt)*100/(N_subj_G*2)) 

# difference in change
df.pa.gut.ast5$pa.change.diff <- df.pa.gut.ast5$Y_P_Change - df.pa.gut.ast5$X_P_Change
```
**Permutation test for Prevalence Trajectory analysis**
```{r} 
Res_ast5y_G <- perm_func_dfs(df.pa.gut.ast5, change_3m1y_pa_G_sdf, 
                           variable = "asthma_5y", Group_X = "Probable", Group_Y = "No", N_taxa = N_taxa_G)
# Test output
Res_ast5y_G$global_res # Overall significance test; p_value = later in healthy
# Res_per_taxa_ast5y_G - p-value assigned to each taxa
Res_per_taxa_ast5y_G <- Res_ast5y_G$per_taxa_res
```

#### Nasal - Asthma at 3 years that persists to 5 years
```{r} 
X <- subset(change_3m1y_pa_sdf, asthma_3y_5y == "Yes")
Y <- subset(change_3m1y_pa_sdf, asthma_3y_5y == "No")

# Make data.frame of prevalence like before BUT keeping as proprotions instead of Percentages for now
df.pa.nasal.ast.3y5y <- data.frame(X_P_Change=colSums(X[1:N_taxa])*100/dim(X)[1],Y_P_Change=colSums(Y[1:N_taxa])*100/dim(Y)[1], taxonomy=as(tax_table(Nasal_3M1Y_pa_filt), "matrix"), prevalence=taxa_sums(Nasal_3M1Y_pa_filt)*100/(N_subj*2)) 

# difference in change
df.pa.nasal.ast.3y5y$pa.change.diff <- df.pa.nasal.ast.3y5y$Y_P_Change - df.pa.nasal.ast.3y5y$X_P_Change
```
**Permutation test for Prevalence Trajectory analysis**
```{r} 
Res_3y5yast_N <- perm_func_dfs(df.pa.nasal.ast.3y5y, change_3m1y_pa_sdf, 
                           variable = "asthma_3y_5y", Group_X = "Yes", Group_Y = "No", N_taxa = N_taxa)

# Test output
Res_3y5yast_N$global_res # Overall significance test
#  p-value assigned to each taxa; p_value = later in healthy; p_value_2 = earlier in healthy
Res_per_taxa_ast3y5y_N <- Res_3y5yast_N$per_taxa_res
```

### Gut - Asthma at 3 years that persists to 5 years
```{r} 
X <- subset(change_3m1y_pa_G_sdf, asthma_3y_5y == "Yes")
Y <- subset(change_3m1y_pa_G_sdf, asthma_3y_5y == "No")

# Make data.frame of prevalence like before BUT keeping as proprotions instead of Percentages for now
df.pa.gut.ast.3y5y <- data.frame(X_P_Change=colSums(X[1:N_taxa_G])*100/dim(X)[1],Y_P_Change=colSums(Y[1:N_taxa_G])*100/dim(Y)[1], taxonomy=as(tax_table(Gut_3M1Y_pa_filt), "matrix"), prevalence=taxa_sums(Gut_3M1Y_pa_filt)*100/(N_subj_G*2)) 

# difference in change
df.pa.gut.ast.3y5y$pa.change.diff <- df.pa.gut.ast.3y5y$Y_P_Change - df.pa.gut.ast.3y5y$X_P_Change
```
**Permutation test for Prevalence Trajectory analysis**
```{r} 
Res_3y5yast_G <- perm_func_dfs(df.pa.gut.ast.3y5y, change_3m1y_pa_G_sdf, 
                           variable = "asthma_3y_5y", Group_X = "Yes", Group_Y = "No", N_taxa = N_taxa_G)

# Test output
 # Overall significance test
Res_3y5yast_G$global_res
#  p-value assigned to each taxa; p_value = later in healthy; p_value_2 = earlier in healthy
Res_per_taxa_ast3y5y_G <- Res_3y5yast_G$per_taxa_res
```

#### Nasal, Atopy at 3 years without Persistent Wheeze vs. no Atopy or Wheeze 
```{r} 
X <- subset(change_3m1y_pa_sdf, Wheeze_atopy_Traj == "Never/infrequent, Atopy Yes")
Y <- subset(change_3m1y_pa_sdf, Wheeze_atopy_Traj == "Never/infrequent, Atopy No")

# Make data.frame of prevalence like before BUT keeping as proprotions instead of Percentages for now
df.pa.nasal.NoWhz.atopy <- data.frame(X_P_Change=colSums(X[1:N_taxa])*100/dim(X)[1],Y_P_Change=colSums(Y[1:N_taxa])*100/dim(Y)[1], taxonomy=as(tax_table(Nasal_3M1Y_pa_filt), "matrix"), prevalence=taxa_sums(Nasal_3M1Y_pa_filt)*100/(N_subj*2)) 

# difference in change
df.pa.nasal.NoWhz.atopy$pa.change.diff <- df.pa.nasal.NoWhz.atopy$Y_P_Change - df.pa.nasal.NoWhz.atopy$X_P_Change
```
**Permutation test for Prevalence Trajectory analysis**
```{r} 
Res_atopy_N <- perm_func_dfs(df.pa.nasal.NoWhz.atopy, change_3m1y_pa_sdf, 
                           variable = "Wheeze_atopy_Traj", 
                           Group_X = "Never/infrequent, Atopy Yes", Group_Y = "Never/infrequent, Atopy No", N_taxa = N_taxa)
 # Overall significance test
Res_atopy_N$global_res
#  p-value assigned to each taxa; p_value = later in healthy; p_value_2 = earlier in healthy
Res_per_taxa_atopy_N <- Res_atopy_N$per_taxa_res
```

#### Gut, Atopy at 3 years without Persistent Wheeze vs. no Atopy or Wheeze
```{r} 
X <- subset(change_3m1y_pa_G_sdf, Wheeze_atopy_Traj == "Never/infrequent, Atopy Yes")
Y <- subset(change_3m1y_pa_G_sdf, Wheeze_atopy_Traj == "Never/infrequent, Atopy No")

# Make data.frame of prevalence like before BUT keeping as proprotions instead of Percentages for now
df.pa.gut.NoWhz.atopy <- data.frame(X_P_Change=colSums(X[1:N_taxa_G])*100/dim(X)[1],Y_P_Change=colSums(Y[1:N_taxa_G])*100/dim(Y)[1], taxonomy=as(tax_table(Gut_3M1Y_pa_filt), "matrix"), prevalence=taxa_sums(Gut_3M1Y_pa_filt)*100/(N_subj_G*2)) 

# difference in change
df.pa.gut.NoWhz.atopy$pa.change.diff <- df.pa.gut.NoWhz.atopy$Y_P_Change - df.pa.gut.NoWhz.atopy$X_P_Change
```
**Permutation test for Prevalence Trajectory analysis**
```{r} 
Res_atopy_G <- perm_func_dfs(df.pa.gut.NoWhz.atopy, change_3m1y_pa_G_sdf, 
                           variable = "Wheeze_atopy_Traj", 
                           Group_X = "Never/infrequent, Atopy Yes", Group_Y = "Never/infrequent, Atopy No", N_taxa = N_taxa_G)
 # Overall significance test
Res_atopy_G$global_res
#  p-value assigned to each taxa; p_value = later in healthy; p_value_2 = earlier in healthy
Res_per_taxa_atopy_G <- Res_atopy_G$per_taxa_res
```

#### Nasal, Colds 0-3 months 
```{r} 
X <- subset(change_3m1y_pa_sdf, colds_0to3m == "Yes")
Y <- subset(change_3m1y_pa_sdf, colds_0to3m == "No")

# Make data.frame of prevalence like before BUT keeping as proprotions instead of Percentages for now
df.pa.nasal.C1 <- data.frame(X_P_Change=colSums(X[1:N_taxa])*100/dim(X)[1],Y_P_Change=colSums(Y[1:N_taxa])*100/dim(Y)[1], taxonomy=as(tax_table(Nasal_3M1Y_pa_filt), "matrix"), prevalence=taxa_sums(Nasal_3M1Y_pa_filt)*100/(N_subj*2)) 

# difference in change
df.pa.nasal.C1$pa.change.diff <- df.pa.nasal.C1$Y_P_Change - df.pa.nasal.C1$X_P_Change
```
**Permutation test for Prevalence Trajectory analysis**
```{r} 
Res_Cold_N <- perm_func_dfs(df.pa.nasal.C1, change_3m1y_pa_sdf, 
                           variable = "colds_0to3m", Group_X = "Yes", Group_Y = "No", N_taxa = N_taxa)
 # Overall significance test
Res_Cold_N$global_res
#  p-value assigned to each taxa; p_value = later in healthy; p_value_2 = earlier in healthy
Res_per_taxa_Cold_N <- Res_Cold_N$per_taxa_res
```

#### Gut, Colds 0-3 months
```{r} 
X <- subset(change_3m1y_pa_G_sdf, colds_0to3m == "Yes")
Y <- subset(change_3m1y_pa_G_sdf, colds_0to3m == "No")

# Make data.frame of prevalence like before BUT keeping as proprotions instead of Percentages for now
df.pa.gut.C <- data.frame(X_P_Change=colSums(X[1:N_taxa_G])*100/dim(X)[1],Y_P_Change=colSums(Y[1:N_taxa_G])*100/dim(Y)[1], taxonomy=as(tax_table(Gut_3M1Y_pa_filt), "matrix"), prevalence=taxa_sums(Gut_3M1Y_pa_filt)*100/(N_subj_G*2))

# difference in change
df.pa.gut.C$pa.change.diff <- df.pa.gut.C$Y_P_Change - df.pa.gut.C$X_P_Change
```
**Permutation test for Prevalence Trajectory analysis**
```{r} 
Res_Cold_G <- perm_func_dfs(df.pa.gut.C, change_3m1y_pa_G_sdf, 
                           variable = "colds_0to3m", Group_X = "Yes", Group_Y = "No", N_taxa = N_taxa_G)
 # Overall significance test
Res_Cold_G$global_res
#  p-value assigned to each taxa; p_value = later in healthy; p_value_2 = earlier in healthy
Res_per_taxa_Cold_G <- Res_Cold_G$per_taxa_res
```

# Compile & save all prevalence trajectory results
```{r}
# saving everything to output
Per_test_df_ls <- list(Res_per_taxa_ast_N, Res_per_taxa_BF_N, Res_per_taxa_ast5y_N, 
                       Res_per_taxa_ast3y5y_N, Res_per_taxa_atopy_N, Res_per_taxa_Cold_N, 
                       Res_per_taxa_ast_G, Res_per_taxa_BF_G, Res_per_taxa_ast5y_G, Res_per_taxa_ast3y5y_G, 
                       Res_per_taxa_atopy_G, Res_per_taxa_Cold_G)
names(Per_test_df_ls) <-  c("Ast3y_Nasal", "BF_Nasal", "Ast5y_Nasal", "Ast3y_5y_Nasal", "Atopy_Nasal", "Cold_Nasal", 
                            "Ast3y_Gut", "BF_Gut", "Ast5y_Gut", "Ast3y_5y_Gut", "Atopy_Gut", "Cold_Gut")

# Save as an overall rds file in Tables/PermTests/
saveRDS(Per_test_df_ls, paste0(path, "Tables/PermTests/ALL_Permutation_PrevTraj_Res.rds"))
```

# Figure S6 - Upset plots for overlap among taxa identified as significantly later colonizers in the 'healthy' group
```{r}
# p-value = for later in healthy group
Per_test_df_ls_Sig <- lapply(Per_test_df_ls, function(x) {subset(x, p_value < 0.05)})
lapply(Per_test_df_ls_Sig, function(x) {nrow(x)}) ## nrow of each dataframe is the number of significant late taxa with p<0.05 - of the 170 testable Nasal taxa and 115 gut taxa (not including the Strep.40 present in 100% of samples)
```

## Nasal, Overlap in significant late colonizers between groups (Figure S6B)
- p value < 0.05 
```{r} 
groups <- rev(c("BF", "Asthma, 3 Years", "Asthma, 3 & 5 Years", "Colds", "Atopy"))

Per_test_df_ls_Sig_N <- Per_test_df_ls_Sig[1:6] # nasal dataframes

listInput <- list("Asthma, 3 Years" = row.names(Per_test_df_ls_Sig_N$Ast3y_Nasal), 
                  Colds = row.names(Per_test_df_ls_Sig_N$Cold_Nasal), 
                  "Asthma, 3 & 5 Years" = row.names(Per_test_df_ls_Sig_N$Ast3y_5y_Nasal), 
                  Atopy = row.names(Per_test_df_ls_Sig_N$Atopy_Nasal), 
                  BF = row.names(Per_test_df_ls_Sig_N$BF_Nasal))

tiff(file = paste0(path, "Figures/Fig_S5B_N_p0.05_upset.tiff"), width = 1300, height = 1200, units = "px", res = 300) 
upset(fromList(listInput), keep.order = T, nsets = 6, sets = groups) 
dev.off()

df_listInput_1 <- fromList2(listInput)

# checking total overlap with Asthma for each group
sum(df_listInput_1$BF == 1 & df_listInput_1$`Asthma, 3 Years` == 1) # 17
sum(df_listInput_1$`Asthma, 3 & 5 Years` == 1 & df_listInput_1$`Asthma, 3 Years` == 1) # 21
sum(df_listInput_1$Colds == 1 & df_listInput_1$`Asthma, 3 Years` == 1) # 12
sum(df_listInput_1$Atopy == 1 & df_listInput_1$`Asthma, 3 Years` == 1) # 3 
```

## Gut, Overlap in significant taxa between groups (Figure S6C)
- p value < 0.05
```{r} 
Per_test_df_ls_Sig_G <- Per_test_df_ls_Sig[7:12] # gut dataframes

listInputG <- list("Asthma, 3 Years" = row.names(Per_test_df_ls_Sig_G$Ast3y_Gut), 
                  Colds = row.names(Per_test_df_ls_Sig_G$Cold_Gut), 
                  "Asthma, 3 & 5 Years" = row.names(Per_test_df_ls_Sig_G$Ast3y_5y_Gut), 
                  Atopy = row.names(Per_test_df_ls_Sig_G$Atopy_Gut), 
                  BF = row.names(Per_test_df_ls_Sig_G$BF_Gut))


tiff(file = paste0(path, "Figures/Fig_S5C_G_p0.05_upset.tiff"), width = 1300, height = 1200, units = "px", res = 300) 
upset(fromList(listInputG), keep.order = T, nsets = 6, sets = groups) 
dev.off()

df_listInputG_1 <- fromList2(listInputG)
# checking total overlap with Asthma for each group
sum(df_listInputG_1$BF == 1 & df_listInputG_1$`Asthma, 3 Years` == 1) # 10
sum(df_listInputG_1$`Asthma, 3 & 5 Years` == 1 & df_listInputG_1$`Asthma, 3 Years` == 1) # 13
sum(df_listInputG_1$Colds == 1 & df_listInputG_1$`Asthma, 3 Years` == 1) # 3
sum(df_listInputG_1$Atopy == 1 & df_listInputG_1$`Asthma, 3 Years` == 1) # 0
```

