---
title: "Figure S5. Microbiota richness and colonization patterns between infants with and without asthma at 5 years, and with and without asthma at both 3 and 5 years."
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Setup

```{r, warning=FALSE, message=FALSE}
R.version.string
library(phyloseq); packageVersion('phyloseq')
library(ggplot2); packageVersion('ggplot2')
library(dplyr); packageVersion('dplyr')   
library(tidyr); packageVersion('tidyr')
library(reshape2); packageVersion('reshape2')
library(ggsignif); packageVersion('ggsignif')
library(cowplot); packageVersion('cowplot')
library(here)

source(paste0(here::here(), "/Functions.R"))

theme_set(theme_bw() + theme(plot.title = element_text(hjust = 0.5,face = "bold"), strip.text.x = element_text(size = 11), strip.background = element_rect(colour="white", fill="white"), panel.grid.major = element_blank(), panel.grid.minor = element_blank(), text = element_text(size=12), axis.text = element_text(color="black", size=12), legend.key.size = unit(0.5,"cm")))

Trj_cols <- c("#f2aeb0", "#f08a0c")
```

**Data subsets used**
Microbiota Trajectory data subsets used
```{r}
Nasal_3M1Y_pa_filt <- readRDS(paste0(path, "Data/Data_subsets/Main_Nasal_Trajectory_Subset.rds")) # Prevalence dataset

Gut_3M1Y_pa_filt <- readRDS(paste0(path, "Data/Data_subsets/Main_Gut_Trajectory_Subset.rds")) # Prevalence dataset

# splitting by timepoint for downstream analyses
N.pa.3m <- subset_samples(Nasal_3M1Y_pa_filt, Visit == "3 Months")
N.pa.1y <- subset_samples(Nasal_3M1Y_pa_filt, Visit == "1 Year")

G.pa.3m <- subset_samples(Gut_3M1Y_pa_filt, Visit == "3 Months")
G.pa.1y <- subset_samples(Gut_3M1Y_pa_filt, Visit == "1 Year")

# Derived Trajectory variables for Fig S5B & S5C
Traj_data_ls <- readRDS(paste0(path, "Data/Data_subsets/ALL_Trajectory_Derived_Data.rds"))
```


# Figure S5A & S5C - Prevalence Trajectory Coordinate plots for Asthma at 5 years and Asthma at 3 years that persists to 5 years 

### Nasal - Prevalence Trajectory derived variables, creation of a prevalence trajectory feature table
To simplify making of prevalence trajectory scatter plots, first make the 'difference in prevalence' results for every pair of samples for each infant (1Y - 3M, e.g. 0 - 1 = -1; 1 - 0 = 1) - so a "-1" means it's present at 3 months but not 1 year, 1 means 1 year but not 3m and 0 means no change.
```{r} 
## extract otu table and use Participant IDs as row.names instead - will match b/t datasets
N_3m_otu <- data.frame(as.data.frame(t(as(otu_table(N.pa.3m), "matrix"))), row.names=sample_data(N.pa.3m)$CHILDid)
N_1y_otu <- data.frame(as.data.frame(t(as(otu_table(N.pa.1y), "matrix"))), row.names=sample_data(N.pa.1y)$CHILDid)

N_3m_otu <- N_3m_otu[order(row.names(N_3m_otu)),]
N_1y_otu <- N_1y_otu[order(row.names(N_1y_otu)),]

## Checking - colnames and Childid's are in the same order? Yes - good - can move forward
identical(colnames(N_3m_otu), colnames(N_1y_otu))
identical(row.names(N_1y_otu), row.names(N_3m_otu))

## 1 Year minus 3 Month Prevalence For each Taxa
change_3m1y_pa <- N_1y_otu - N_3m_otu # -1 = present at 3m not 1y, 1= present at 1y not 3m, 0=either present at both or neither timpoint
dim(change_3m1y_pa) # Same order as original physeq obj

# Sample data to merge with dataset
sdf_toAdd <- sample_data(N.pa.1y)[order(sample_data(N.pa.1y)$CHILDid),]
sdf_3m <- sample_data(N.pa.3m)[order(sample_data(N.pa.3m)$CHILDid),]
## check that it's in the same order as change_3m1y_pa
identical(row.names(change_3m1y_pa), sdf_toAdd$CHILDid) 
identical(row.names(change_3m1y_pa), sdf_3m$CHILDid) 

### same order so just merge data with sample metadata using data.frame
change_3m1y_pa_sdf <- data.frame(change_3m1y_pa, sdf_toAdd, RE_virus_3m = sdf_3m$RE_Virus) # Using 1 year sdf for variables (will be same for 3m and 1y for almost all) - but also adding RE virus at 3 months as "RE_virus_3m". Note that "RE_Virus" is the RE virus at 1 year.

N_taxa <- ncol(change_3m1y_pa) # number of taxa, to use in downstream code
N_subj <- nrow(change_3m1y_pa) # number of subjects, to use in downstream code
```

```{r} 
# setup for plots
Main_phyla <- c('Proteobacteria','Firmicutes', "Actinobacteria", "Bacteroidetes")
cbPalette <- c("navy", "paleturquoise4",  "violetred3", "skyblue1", "goldenrod3")
```

### Gut - Prevalence Trajectory derived variables, creation of a prevalence trajectory feature table
To simplify making of prevalence trajectory scatter plots, first make the 'difference in prevalence' results for every pair of samples for each infant (1Y - 3M, e.g. 0 - 1 = -1; 1 - 0 = 1) - so a "-1" means it's present at 3 months but not 1 year, 1 means 1 year but not 3m and 0 means no change.
```{r} 
## extract otu table and use Participant IDs as row.names instead - will match b/t datasets
G_3m_otu <- data.frame(as.data.frame(t(as(otu_table(G.pa.3m), "matrix"))), row.names=sample_data(G.pa.3m)$CHILDid)
G_1y_otu <- data.frame(as.data.frame(t(as(otu_table(G.pa.1y), "matrix"))), row.names=sample_data(G.pa.1y)$CHILDid)

G_3m_otu <- G_3m_otu[order(row.names(G_3m_otu)),]
G_1y_otu <- G_1y_otu[order(row.names(G_1y_otu)),]

## Checking - colnames and Childid's are in the same order? Yes - good - can move forward
identical(colnames(G_3m_otu), colnames(G_1y_otu))
identical(row.names(G_3m_otu), row.names(G_1y_otu))

## 1 Year minus 3 Month Prevalence For each Taxa
change_3m1y_pa_G <- G_1y_otu - G_3m_otu # -1 = present at 3m not 1y, 1= present at 1y not 3m, 0=either present at both or neither timpoint
dim(change_3m1y_pa_G) # Same order as original physeq obj

sdf_toAdd <- sample_data(G.pa.1y)[order(sample_data(G.pa.1y)$CHILDid),]
sdf_3m <- sample_data(G.pa.3m)[order(sample_data(G.pa.3m)$CHILDid),]
## check that it's in the same order as change_3m1y_pa
identical(row.names(change_3m1y_pa_G), sdf_toAdd$CHILDid) # order matches
identical(row.names(change_3m1y_pa_G), sdf_3m$CHILDid) # order matches

### same order so just merge data with sample metadata using data.frame
change_3m1y_pa_G_sdf <- data.frame(change_3m1y_pa_G, sdf_toAdd, RE_virus_3m = sdf_3m$RE_Virus) # Using 1 year sdf for variables but also adding RE virus at 3 months as "RE_virus_3m". Note that "RE_Virus" is the RE virus at 1 year

N_taxa_G <- ncol(change_3m1y_pa_G) # number of taxa, to use in downstream code
N_subj_G <- nrow(change_3m1y_pa_G) # number of subjects, to use in downstream code
```

## Figure S5A

### Nasal - Asthma at 5 years 
```{r} 
X <- subset(change_3m1y_pa_sdf, asthma_5y == "Probable")
Y <- subset(change_3m1y_pa_sdf, asthma_5y == "No")

# Make data.frame of prevalence like before BUT keeping as proprotions instead of Percentages for now
df.pa.nasal.ast5 <- data.frame(X_P_Change=colSums(X[1:N_taxa])*100/dim(X)[1],Y_P_Change=colSums(Y[1:N_taxa])*100/dim(Y)[1], taxonomy=as(tax_table(Nasal_3M1Y_pa_filt), "matrix"), prevalence=taxa_sums(Nasal_3M1Y_pa_filt)*100/(N_subj*2)) 

# difference in change
df.pa.nasal.ast5$pa.change.diff <- df.pa.nasal.ast5$Y_P_Change - df.pa.nasal.ast5$X_P_Change
```
Aesthetics
```{r} 
## Plot by major phylum & annotate signif taxa
df.pa.nasal.ast5$Phylum <- ifelse(df.pa.nasal.ast5$taxonomy.Phylum %in% Main_phyla, df.pa.nasal.ast5$taxonomy.Phylum, "Other")
#Reorder the phyla 
df.pa.nasal.ast5$Phylum = factor(df.pa.nasal.ast5$Phylum, levels=c(Main_phyla, "Other"))
```
Plot Change in Prevalence
```{r} 
x_lab <- paste("Asthma, 5 Years (n=", nrow(X), ")\n1 Year - 3 Month Prevalence (%)", sep="")
y_lab <- paste("No Asthma (n=", nrow(Y), ")\n1 Year - 3 Month Prevalence (%)", sep="")

Fig_S5A_N <- ggplot() + 
  geom_point(data=df.pa.nasal.ast5, aes(x=X_P_Change, y=Y_P_Change, color=Phylum, size=prevalence), alpha=0.75) + 
  xlab(x_lab) + ylab(y_lab) + 
  scale_color_manual(values = cbPalette) +
  scale_y_continuous(limits=c(-30,60), expand = c(0,0))+
  scale_x_continuous(limits=c(-30,60), expand = c(0,0))+
  annotate("segment", x = -30, xend = 60, y = -30, yend = 60, colour = "red", linewidth=0.5) + ggtitle("Nasal")
Fig_S5A_N
```
Median change in prevalence of an OTU 
```{r} 
# Asthma
summary(df.pa.nasal.ast5$X_P_Change)
# No Asthma
summary(df.pa.nasal.ast5$Y_P_Change)
```

**Permutation test for Prevalence Trajectory analysis - See 'Functions.R' for perm_func_dfs function**
```{r} 
Res_ast5y_N <- perm_func_dfs(df.pa.nasal.ast5, change_3m1y_pa_sdf, 
                           variable = "asthma_5y", Group_X = "Probable", Group_Y = "No", N_taxa = N_taxa)

# Test output
Res_ast5y_N$global_res # Overall significance test; p_value = later in healthy
# Res_per_ast5y_N - p-value assigned to each taxa
Res_per_ast5y_N <- Res_ast5y_N$per_taxa_res
```

### Gut - Asthma at 5 years
```{r} 
X <- subset(change_3m1y_pa_G_sdf, asthma_5y == "Probable")
Y <- subset(change_3m1y_pa_G_sdf, asthma_5y == "No")

# Make data.frame of prevalence like before BUT keeping as proprotions instead of Percentages for now
df.pa.gut.ast5 <- data.frame(X_P_Change=colSums(X[1:N_taxa_G])*100/dim(X)[1],Y_P_Change=colSums(Y[1:N_taxa_G])*100/dim(Y)[1], taxonomy=as(tax_table(Gut_3M1Y_pa_filt), "matrix"), prevalence=taxa_sums(Gut_3M1Y_pa_filt)*100/(N_subj_G*2)) 

# difference in change
df.pa.gut.ast5$pa.change.diff <- df.pa.gut.ast5$Y_P_Change - df.pa.gut.ast5$X_P_Change
```
Aesthetic things, not important
```{r} 
## Plot by major phylum & annotate signif taxa
df.pa.gut.ast5$Phylum <- ifelse(df.pa.gut.ast5$taxonomy.Phylum %in% Main_phyla, df.pa.gut.ast5$taxonomy.Phylum, "Other")
#Reorder the phyla 
df.pa.gut.ast5$Phylum = factor(df.pa.gut.ast5$Phylum, levels=c(Main_phyla, "Other"))
```
Plot Change in Prevalence 
```{r} 
x_lab <- paste("Asthma, 5 Years (n=", nrow(X), ")\n1 Year - 3 Month Prevalence (%)", sep="")
y_lab <- paste("No Asthma (n=", nrow(Y), ")\n1 Year - 3 Month Prevalence (%)", sep="")

Fig_S5A_G <- ggplot() + 
  geom_point(data=df.pa.gut.ast5, aes(x=X_P_Change, y=Y_P_Change, color=Phylum, size=prevalence), alpha=0.75) + 
  xlab(x_lab) +  ylab(y_lab) + 
  scale_color_manual(values = cbPalette) +
  scale_y_continuous(limits=c(-30,60), expand = c(0,0))+
  scale_x_continuous(limits=c(-30,60), expand = c(0,0))+
  annotate("segment", x = -30, xend = 60, y = -30, yend = 60, colour = "red", linewidth=0.25) + ggtitle("Gut")
Fig_S5A_G
```
Median change in prevalence of an OTU 
```{r} 
# Asthma
summary(df.pa.gut.ast5$X_P_Change)
# No Asthma
summary(df.pa.gut.ast5$Y_P_Change)
```

**Permutation test for Prevalence Trajectory analysis**
```{r} 
Res_ast5y_G <- perm_func_dfs(df.pa.gut.ast5, change_3m1y_pa_G_sdf, 
                           variable = "asthma_5y", Group_X = "Probable", Group_Y = "No", N_taxa = N_taxa_G)

# Test output
Res_ast5y_G$global_res # Overall significance test; p_value = later in healthy
# Res_per_ast5y_N - p-value assigned to each taxa
Res_per_ast5y_G <- Res_ast5y_G$per_taxa_res
```


## Figure S5C

### Asthma at 3 years that persists to 5 years vs. no asthma at 3 and no asthma at 5 years
```{r} 
# want probable asthma at 3 years, and probable or possible at 5 years. Those missing 5 year or 3 year asthma data will be NA for this variable
change_3m1y_pa_sdf$asthma_3y_5y <- ifelse(change_3m1y_pa_sdf$asthma_3y == "Probable" & 
                                            change_3m1y_pa_sdf$asthma_5y %in% c("Probable", "Possible"), "Yes", 
                                          ifelse(change_3m1y_pa_sdf$asthma_3y == "No" & 
                                            change_3m1y_pa_sdf$asthma_5y == "No", "No", NA))
table(change_3m1y_pa_sdf$asthma_3y_5y)
```

```{r} 
# want probable asthma at 3 years, and probable or possible at 5 years. Those missing 5 year or 3 year asthma data will be NA for this variable  
change_3m1y_pa_G_sdf$asthma_3y_5y <- ifelse(change_3m1y_pa_G_sdf$asthma_3y == "Probable" & 
                                            change_3m1y_pa_G_sdf$asthma_5y  %in% c("Probable", "Possible"), "Yes", 
                                          ifelse(change_3m1y_pa_G_sdf$asthma_3y == "No" & 
                                            change_3m1y_pa_G_sdf$asthma_5y == "No", "No", NA))

table(change_3m1y_pa_G_sdf$asthma_3y_5y)
```

### Nasal - Asthma at 3 years that persists to 5 years
```{r} 
X <- subset(change_3m1y_pa_sdf, asthma_3y_5y == "Yes")
Y <- subset(change_3m1y_pa_sdf, asthma_3y_5y == "No")

# Make data.frame of prevalence like before BUT keeping as proprotions instead of Percentages for now
df.pa.nasal.ast.3y5y <- data.frame(X_P_Change=colSums(X[1:N_taxa])*100/dim(X)[1],Y_P_Change=colSums(Y[1:N_taxa])*100/dim(Y)[1], taxonomy=as(tax_table(Nasal_3M1Y_pa_filt), "matrix"), prevalence=taxa_sums(Nasal_3M1Y_pa_filt)*100/(N_subj*2)) 

# difference in change
df.pa.nasal.ast.3y5y$pa.change.diff <- df.pa.nasal.ast.3y5y$Y_P_Change - df.pa.nasal.ast.3y5y$X_P_Change
```
Aesthetic things
```{r} 
## Plot by major phylum & annotate signif taxa
df.pa.nasal.ast.3y5y$Phylum <- ifelse(df.pa.nasal.ast.3y5y$taxonomy.Phylum %in% Main_phyla, df.pa.nasal.ast.3y5y$taxonomy.Phylum, "Other")
#Reorder the phyla 
df.pa.nasal.ast.3y5y$Phylum = factor(df.pa.nasal.ast.3y5y$Phylum, levels=c(Main_phyla, "Other"))
```
Plot Change in Prevalence
```{r} 
x_lab <- paste("Asthma, 3 & 5 Years (n=", nrow(X), ")\n1 Year - 3 Month Prevalence (%)", sep="")
y_lab <- paste("No Asthma (n=", nrow(Y), ")\n1 Year - 3 Month Prevalence (%)", sep="")

Fig_S5C_N <- ggplot() + 
  geom_point(data=df.pa.nasal.ast.3y5y, aes(x=X_P_Change, y=Y_P_Change, color=Phylum, size=prevalence), alpha=0.75) + 
  xlab(x_lab) + ylab(y_lab) + 
  scale_color_manual(values = cbPalette) +
  scale_y_continuous(limits=c(-30,60), expand = c(0,0))+
  scale_x_continuous(limits=c(-30,60), expand = c(0,0))+
  annotate("segment", x = -30, xend = 60, y = -30, yend = 60, colour = "red", linewidth=0.5) + ggtitle("Nasal")
Fig_S5C_N
```
Median change in prevalence of an OTU in asthma and non-asthma infants
```{r} 
# Asthma
summary(df.pa.nasal.ast.3y5y$X_P_Change)
# No Asthma
summary(df.pa.nasal.ast.3y5y$Y_P_Change)
```

**Permutation test for Prevalence Trajectory analysis**
```{r} 
Res_ast3y5y_N <- perm_func_dfs(df.pa.nasal.ast.3y5y, change_3m1y_pa_sdf, 
                           variable = "asthma_3y_5y", Group_X = "Yes", Group_Y = "No", N_taxa = N_taxa)

# Test output
Res_ast3y5y_N$global_res # Overall significance test; p_value = later in healthy
# Res_per_ast5y_N - p-value assigned to each taxa
Res_per_ast3y5y_N <- Res_ast3y5y_N$per_taxa_res
```

### Gut - Asthma at 3 years that persists to 5 years
```{r} 
X <- subset(change_3m1y_pa_G_sdf, asthma_3y_5y == "Yes")
Y <- subset(change_3m1y_pa_G_sdf, asthma_3y_5y == "No")

# Make data.frame of prevalence like before BUT keeping as proprotions instead of Percentages for now
df.pa.gut.ast.3y5y <- data.frame(X_P_Change=colSums(X[1:N_taxa_G])*100/dim(X)[1],Y_P_Change=colSums(Y[1:N_taxa_G])*100/dim(Y)[1], taxonomy=as(tax_table(Gut_3M1Y_pa_filt), "matrix"), prevalence=taxa_sums(Gut_3M1Y_pa_filt)*100/(N_subj_G*2)) 

# difference in change
df.pa.gut.ast.3y5y$pa.change.diff <- df.pa.gut.ast.3y5y$Y_P_Change - df.pa.gut.ast.3y5y$X_P_Change
```
Aesthetic things, not important
```{r} 
df.pa.gut.ast.3y5y$Phylum <- ifelse(df.pa.gut.ast.3y5y$taxonomy.Phylum %in% c("Proteobacteria", "Actinobacteria", "Firmicutes", "Bacteroidetes"),
                                      df.pa.gut.ast.3y5y$taxonomy.Phylum, "Other")
#Reorder the phyla 
df.pa.gut.ast.3y5y$Phylum = factor(df.pa.gut.ast.3y5y$Phylum, levels=c('Proteobacteria','Firmicutes', "Actinobacteria", "Bacteroidetes", "Other"))
```
Plot Change in Prevalence 
```{r} 
x_lab <- paste("Asthma, 3 & 5 Years (n=", nrow(X), ")\n1 Year - 3 Month Prevalence (%)", sep="")
y_lab <- paste("No Asthma (n=", nrow(Y), ")\n1 Year - 3 Month Prevalence (%)", sep="")

Fig_S5C_G <- ggplot() + 
  geom_point(data=df.pa.gut.ast.3y5y, aes(x=X_P_Change, y=Y_P_Change, color=Phylum, size=prevalence), alpha=0.75) + 
  xlab(x_lab) + ylab(y_lab) + 
  scale_color_manual(values = cbPalette) +
  scale_y_continuous(limits=c(-32,60), expand = c(0,0))+
  scale_x_continuous(limits=c(-32,60), expand = c(0,0))+
  annotate("segment", x = -32, xend = 60, y = -32, yend = 60, colour = "red", linewidth=0.25) + ggtitle("Gut")
Fig_S5C_G
```
Median change in prevalence of an OTU 
```{r} 
# Asthma
summary(df.pa.gut.ast.3y5y$X_P_Change)
# No Asthma
summary(df.pa.gut.ast.3y5y$Y_P_Change)
```

**Permutation test for Prevalence Trajectory analysis**
```{r} 
Res_ast3y5y_G <- perm_func_dfs(df.pa.gut.ast.3y5y, change_3m1y_pa_G_sdf, 
                           variable = "asthma_3y_5y", Group_X = "Yes", Group_Y = "No", N_taxa = N_taxa_G)

# Test output
Res_ast3y5y_G$global_res # Overall significance test; p_value = later in healthy
# Res_per_ast3y5y_G - p-value assigned to each taxa
Res_per_ast3y5y_G <- Res_ast3y5y_G$per_taxa_res
```

```{r, include=FALSE}
# Save prevalence trajectory scatter plots, remove axis titles and re-add later

## saving a legend (all will be the same)
legend_PT <- plot_grid(get_legend(Fig_S5A_N), ncol = 1)
tiff(file = paste0(path, "Figures/Fig_S5AC_LegendPrevTraj.tiff"), width = 500, height = 1000, units = "px", res = 300) 
legend_PT
dev.off()

;Fig_S5AC <- plot_grid(Fig_S5A_N + theme(legend.position = "none"), Fig_S5A_G + theme(legend.position = "none"),
                      Fig_S5C_N + theme(legend.position = "none"), Fig_S5C_G + theme(legend.position = "none"),
                      ncol = 2, labels = c("A", "", "C", ""))
Fig_S5AC

tiff(file = paste0(path, "Figures/Fig_S5A_S5C_PreTCoAsthma3y5y.tiff"), width = 2500, height = 2500, units = "px", res = 300) 
Fig_S5AC
dev.off()
```


# Figure S5B & S5D - Richness trajectories associated with asthma at 5 years and asthma at 3 years that persists to 5 years

**Similar to scripts for main Figure 2C-2F, but just select asthma at 5 years and 3 years that persists to 5 years**
Also see Fig_2.Rmd for exact p-values in 'Wilcoxon_Res_Fig2D_2F_and_SuppTraj_ColPat_Richness.csv'

### Microbiota Trajectory data subsets
```{r}
# Main metadata variables to test
Main_vars_Traj <- c("asthma_5y", "asthma_3y_5y") 

### List of Trajectory summary variables to be used for these analysis
NG_summary <- c("Observed_N1y3m", "Observed_G1y3m", "Observed_N3m", "Observed_N1y", "Observed_G1y","Observed_G3m") 
```

```{r} 
NG_Traj_sdf <- merge( subset(Traj_data_ls$Trajectories_Nasal_to181_Gut_at182, select = NG_summary),
                      subset(Traj_data_ls$Metadata, select = Main_vars_Traj), 
                      by="row.names" )
```

**Datasets for plots - melt & split -  Trajectories and Individual time-points**
```{r}
NG_Traj_sdf_m <- melt(NG_Traj_sdf, measure.vars = NG_summary)
NG_Traj_sdf_m$Dataset <- ifelse(grepl("_G", NG_Traj_sdf_m$variable), "Gut", "Nasal")
NG_Traj_sdf_m$Dataset <- factor(NG_Traj_sdf_m$Dataset, levels = c("Nasal", "Gut"))

Single_tp <- c("Observed_N3m", "Observed_G3m", "Observed_N1y", "Observed_G1y")
Trajs <- c("Observed_N1y3m", "Observed_G1y3m")

## split data for plots
NG_Traj_sdf_m_Trajs <- subset(NG_Traj_sdf_m, variable %in% Trajs)

NG_Traj_sdf_m_Tp <- subset(NG_Traj_sdf_m, variable %in% Single_tp)
NG_Traj_sdf_m_Tp$Visit <- ifelse(grepl("3", NG_Traj_sdf_m_Tp$variable), "3-Month", "1-Year")
NG_Traj_sdf_m_Tp$Visit <- factor(NG_Traj_sdf_m_Tp$Visit, levels = c("3-Month", "1-Year"))
```

**Asthma 5y, Figure S5B**
```{r, warning=FALSE}
NG_Traj_sdf_m_Trajs_Ast5y <- subset(NG_Traj_sdf_m_Trajs, asthma_5y!="Possible") 

Fig_S5B <- ggplot(NG_Traj_sdf_m_Trajs_Ast5y, aes(x=asthma_5y, y=value, group=asthma_5y)) + 
  facet_wrap(~Dataset, ncol = 1) + 
  geom_jitter(aes(color=Dataset),shape=16, size=1.5, alpha=0.3, position = position_jitter(width = 0.15)) + 
  geom_violin(width=0.5, trim = FALSE, alpha=0) + 
  stat_summary(fun=median, geom="point", shape=23, size=2, fill="black") + 
  scale_y_continuous(expand=c(0,0)) + coord_cartesian(ylim = c(-150, 200))  + 
  theme(axis.text = element_text(size=9), legend.position = "none", strip.text.x = element_text(face = "bold")) + 
  labs(y="Richness Trajectory", x="Asthma at 5 Years") + scale_color_manual(values = Trj_cols) +
  geom_signif(color="black", comparisons = list(c("No", "Yes")), 
              test = "wilcox.test", step_increase = 0.2, vjust = 2, 
              map_signif_level = c("**"=0.001, "*"=0.051,"~"=0.1," "=2), size=0, textsize = 7) 
Fig_S5B
```

**Asthma 3y to 5y, Figure S5D**
```{r, warning=FALSE}
NG_Traj_sdf_m_Trajs_Ast3y5y <- subset(NG_Traj_sdf_m_Trajs, !is.na(asthma_3y_5y)) 

Fig_S5D <- ggplot(NG_Traj_sdf_m_Trajs_Ast3y5y, aes(x=asthma_3y_5y, y=value, group=asthma_3y_5y)) + 
  facet_wrap(~Dataset, ncol= 1) + 
  geom_jitter(aes(color=Dataset),shape=16, size=1.5, alpha=0.3, position = position_jitter(width = 0.15)) + 
  geom_violin(width=0.5, trim = FALSE, alpha=0) +
  stat_summary(fun=median, geom="point", shape=23, size=2, fill="black")  + 
  scale_y_continuous(expand=c(0,0)) + coord_cartesian(ylim = c(-150, 200))  + 
  theme(axis.text = element_text(size=9), legend.position = "none", strip.text.x = element_text(face = "bold")) + 
  labs(y="Richness Trajectory", x="Asthma, 3 & 5 Years") + scale_color_manual(values = Trj_cols) +
  geom_signif(color="black", comparisons = list(c("No", "Yes")), 
              test = "wilcox.test", step_increase = 0.2, vjust = 2, 
              map_signif_level = c("**"=0.001, "*"=0.051,"~"=0.1," "=2), size=0, textsize = 7) 
Fig_S5D
```

```{r, warning=FALSE, include=FALSE} 
# Save Richness Plots for Supplementary
tiff(file = paste0(path, "Figures/Fig_S5B_5yAsth_RichTraj.tiff"), width = 700, height = 1200, units = "px", res = 300)
Fig_S5B
dev.off()

tiff(file = paste0(path, "Figures/Fig_S5D_3y5yAsth_RichTraj.tiff"), width = 700, height = 1200, units = "px", res = 300)
Fig_S5D
dev.off()
```
