---
title: "Figure S4. Distribution of early, persistent, and late colonization patterns in the nasal and gut niches across infant feeding groups and respiratory phenotypes."
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**Similar to scripts for main Figure 2C-2F, but just select colonization patterns**
Also see Fig_2.Rmd for exact p-values in 'Wilcoxon_Res_Fig2D_2F_and_SuppTraj_ColPat_Richness.csv

## Setup

```{r, warning=FALSE, message=FALSE}
R.version.string
library(phyloseq); packageVersion('phyloseq')
library(ggplot2); packageVersion('ggplot2')
library(dplyr); packageVersion('dplyr')  
library(tidyr); packageVersion('tidyr')
library(reshape2); packageVersion('reshape2')
library(ggsignif); packageVersion('ggsignif')
library(cowplot); packageVersion('cowplot')
library(here)

source(paste0(here::here(), "/Functions.R"))

theme_set(theme_bw() + theme(plot.title = element_text(hjust = 0.5,face = "bold"), strip.text.x = element_text(size = 11), strip.background = element_rect(colour="white", fill="white"), panel.grid.major = element_blank(), panel.grid.minor = element_blank(), text = element_text(size=12), axis.text = element_text(color="black", size=12), legend.key.size = unit(0.5,"cm")))

Trj_cols <- c("#f2aeb0", "#f08a0c") 
```

**Data subset used**
```{r} 
# Derived Trajectory variables for the Trajectory subsets
Traj_data_ls <- readRDS(paste0(path, "Data/Data_subsets/ALL_Trajectory_Derived_Data.rds"))
```

### Microbiota Trajectory data subset

```{r}
# Main metadata variables to test
Main_vars_Traj <- c("BF_status_3m_sampling", "asthma_3y") 

### List of Trajectory summary variables to be used for these analysis
NG_summary <- c("Early_N1y3m", "Late_N1y3m", "Persistent_N1y3m", 
                "Early_G1y3m", "Late_G1y3m", "Persistent_G1y3m") 

NG_Traj_sdf <- merge( subset(Traj_data_ls$Trajectories_Nasal_to181_Gut_at182, select = NG_summary),
                      subset(Traj_data_ls$Metadata, select=Main_vars_Traj), 
                      by="row.names" )
```

## Summary tables & Figures

**Datasets for plots - melt & split - Observed Trajectories, Observed Individual timepoint and Colonization patterns**
```{r}
NG_Traj_sdf_m <- melt(NG_Traj_sdf, measure.vars = NG_summary)
NG_Traj_sdf_m$Dataset <- ifelse(grepl("_G", NG_Traj_sdf_m$variable), "Gut", "Nasal")
NG_Traj_sdf_m$Dataset <- factor(NG_Traj_sdf_m$Dataset, levels = c("Nasal", "Gut"))

## split data for plots
##NG_Traj_sdf_m_ColPats <- subset(NG_Traj_sdf_m, variable %in% ColPats)
```

### Colonization patterns
The proportion of total taxa that are early, persistent and late colonizers per-infant
```{r} 
## edit variable names for plot
NG_Traj_sdf_m$variable <- gsub("_.*$", "", NG_Traj_sdf_m$variable)
NG_Traj_sdf_m$variable <- factor(NG_Traj_sdf_m$variable, levels = c("Early", "Persistent", "Late"))

Ast_ColCounts <- subset(NG_Traj_sdf_m, asthma_3y != "Possible")
```

```{r, warning=FALSE, fig.width=9}
Fig_S3AB_BF <- ggplot(NG_Traj_sdf_m, aes(x=BF_status_3m_sampling, y=value, group=BF_status_3m_sampling)) + 
  facet_wrap(~Dataset*variable, ncol=3) + 
  geom_jitter(aes(color=Dataset), shape=16, size=1.5, alpha=0.25, position = position_jitter(width = 0.15)) + 
  stat_summary(fun=median, geom="point", shape=23, size=2, fill="black") + 
  geom_violin(width=0.5, trim = FALSE, alpha=0) + scale_y_continuous(expand=c(0,0)) + 
  theme(axis.text = element_text(size=11), legend.position = "none") + 
  labs(y="Percent of OTUs", x="Breastfeeding at 3 Months") + 
  scale_color_manual(values = Trj_cols) + 
  geom_signif(color="black", comparisons = list(c("Exclusive", "No BM")), test = "wilcox.test", step_increase = 0.2, vjust = 1, map_signif_level = c("**"=0.001, "*"=0.05,"~"=0.1," "=2), textsize = 7, size = 0) 

Fig_S3AB_Ast <- ggplot(Ast_ColCounts, aes(x=asthma_3y, y=value, group=asthma_3y)) + 
  facet_wrap(~Dataset*variable, ncol=3) + 
  geom_jitter(aes(color=Dataset), shape=16, size=1.5, alpha=0.25, position = position_jitter(width = 0.15)) + 
  stat_summary(fun=median, geom="point", shape=23, size=2, fill="black") + 
  geom_violin(width=0.5, trim = FALSE, alpha=0) + scale_y_continuous(expand=c(0,0)) + 
  theme(axis.text = element_text(size=11), legend.position = "none") + 
  labs(y="Percent of OTUs", x="Asthma at 3 Years") + 
  scale_color_manual(values = Trj_cols) + 
  geom_signif(color="black", comparisons = list(c("No", "Yes")), test = "wilcox.test", step_increase = 0.2, vjust = 1, map_signif_level = c("**"=0.001, "*"=0.05,"~"=0.1," "=2), textsize = 7, size = 0) 

Fig_S3 <- plot_grid(Fig_S3AB_Ast, Fig_S3AB_BF, ncol = 2, axis = "tb")
Fig_S3
```

```{r, include=FALSE}
tiff(file = paste0(path, "Figures/Fig_S3_ColPatterns.tiff"), width = 2900, height = 1350, units = "px", res = 300)
Fig_S3
dev.off()
```

