---
title: "Figure S8. The correlation between the colonization patterns and trajectories of functions identified as delayed with breastfeeding at 3 months, and no asthma compared to asthma at 3 years."
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```
 
## Setup 

```{r}
R.version.string
library(phyloseq)
library(ggplot2)
library(tidyr)
library(reshape2)
library(knitr)
library(dplyr)
library(Hmisc)
library(stringr)
library(glue)
library(ggtext)
library(ggsignif)
library(here)

source(paste0(here::here(), "/Functions.R"))

theme_set(theme_bw() + theme(strip.text.x = element_text(size = 11), strip.background = element_rect(colour="white", fill="white"), panel.grid.major = element_blank(), panel.grid.minor = element_blank(), text = element_text(size=12), axis.text = element_text(color="black", size=12), legend.key.size = unit(0.5,"cm")))
```

## Read in Metagenomic data & PreTCO results
```{r}
## cleaned metagenomic taxonomy and sample data, also normalize metagenomic counts
Gut_3M1Y_all_filt <- readRDS(file = paste0(path, "Data/Metagenome_metabolome/final_datasets/03_Abund_metagenomic_Trajectory.rds"))
sdf <- sample_data(Gut_3M1Y_all_filt) # sample dta

funcs_3M1Y_bin_s <- readRDS(file = paste0(path, "Data/Metagenome_metabolome/final_datasets/MainTaxaFunctionsBinaryTraj_sdf_above10EitherVis_v02.rds"))

names_bestPrevalence_s <- readRDS(file = paste0(path, "Data/Metagenome_metabolome/final_datasets/Ls_MainTaxaFuncs_best_Binary_above10_below90EitherVis_v02.rds"))
## some issues with number at beginning of strings
names_bestPrevalence_s$EC <- gsub("^", "X", names_bestPrevalence_s$EC) 
names_bestPrevalence_s$MetaCyc <- gsub("^\\d", "X1", names_bestPrevalence_s$MetaCyc)


# also the results to select specific features
Res_per_taxa_ec_asthma <- read.csv(file = paste0(path, "Tables/Metagenome_metabolome/EC_Strata_Asthma_PreTCo_res.csv"), row.names = 1)
Res_per_taxa_ec_bf <- read.csv(file = paste0(path, "Tables/Metagenome_metabolome/EC_Strata_BF_PreTCo_res.csv"), row.names = 1)


## sig late w/t BF and asthma set w/t p<0.01
late_overlap <- intersect(row.names(subset(Res_per_taxa_ec_bf, p_value < 0.01)),
                          row.names(subset(Res_per_taxa_ec_asthma, p_value < 0.01)))
late_overlap ## see that all R. gnavus - focus of this figure

## check opposite -  select randomly from R. gnavus functions of those not late by BF
ns_late_bf_rg <- row.names(subset(Res_per_taxa_ec_bf, p_value >= 0.01))
ns_late_bf_rg <- ns_late_bf_rg[grepl("ruminococcus.gnavus$", ns_late_bf_rg)]
length(ns_late_bf_rg)

# 30 funcs NS - not late - in breastfeeding -  These are all ALSO not late for asthma. NOTE that  this is of the R. gnavus functions
length(intersect(ns_late_bf_rg, row.names(subset(Res_per_taxa_ec_asthma, p_value >= 0.01))))

## see 12 late for asthma overall, 8 also late for BF (all r. gnavus), 4 not late with BF (all not R. gnavus funcs)
length(row.names(subset(Res_per_taxa_ec_asthma, p_value < 0.01))) 
```


# Figure S8B - Prevalence Plot showing R. gnavus prevalence, and selected EC functions, stratified by breastfeeding and asthma
```{r} 
# taxa - decided to just plot r. gnavus 
taxa_select <- "Ruminococcus_gnavus"
## Gut_3M1Y_all_filt = physeq obj input

# functions - decided to just plot r. gnavus 
strata_select <- "ruminococcus.gnavus"

## Subset/melt: R. gnavus from taxonomic data
sub_taxa <- Gut_3M1Y_all_filt %>% otu_table %>% t %>% as.data.frame %>% 
                  subset(., select = c(taxa_select)) 
# melt(., id.vars = colnames(sdf))

## Subset functions:  R. gnavus functions associated w/t both BF and Asthma - late_overlap
### or not found as late w/t BF (in this case also NS for asthma) - randomly select n
set.seed(1)
ns_late_bf_rg_sub <- sample(ns_late_bf_rg, size = 6)

cols_select <- c(ns_late_bf_rg_sub, late_overlap, colnames(sdf))

### also merge with sub_taxa, then melt all together
melted_funcs_taxa <- funcs_3M1Y_bin_s$EC %>% subset(., select = cols_select) %>%
                      merge(., sub_taxa, by = "row.names") %>% melt(., id.vars = c("Row.names", colnames(sdf))) %>%
                      mutate(data = ifelse(variable %in% taxa_select, "taxonomy", "function"),
                             variable = gsub("_ruminococcus.gnavus$", "", variable) ## simplified version - only ok to do this way for now b/c format differs for function vs taxonomy data
                             ) 
table(melted_funcs_taxa$data)

## make variable name versions to match diff plots - numeric only and alpha only
melted_funcs_taxa$var_alpha <- melted_funcs_taxa$variable %>% gsub("^X\\d\\.\\d", "", .) %>% 
  gsub("^\\d\\.\\d{1,}\\.\\d{1,}\\.|^\\.\\d{1,}\\.\\d{1,}.|^\\.\\d\\.\\d\\.|\\.\\d.\\d{1,}\\.\\d{1,}.", "", .)
table(melted_funcs_taxa$var_alpha)

melted_funcs_taxa$variable_num <- melted_funcs_taxa$variable %>% gsub("^X|[:alpha:].*", "", .) %>%  ## not rm all alpha for some reason
                                                  gsub("c$|g$|n$|dn$|tr$|m$|di$", "", .) %>%  
                                                  gsub("\\.$|\\.o\\.bet|\\.fe\\.3|\\.s\\.r\\.r|\\.de|\\.digu|\\.s\\.formy|\\.s", "", .) %>%
                                                  gsub("Ruminococcus_g", "Ruminococcus\ngnavus", .)
table(melted_funcs_taxa$variable_num)

## summarize functional and taxonomy prevalence
grouping_cols <- c("variable", "var_alpha", "variable_num", "data", "Visit")
BFsumm_prev <- melted_funcs_taxa %>% subset(., BF_status_3m_sampling != "Partial") %>% ## rm Partial
            prev_summary_2(., "value", c(grouping_cols, "BF_status_3m_sampling"))  %>%
            mutate(Group = BF_status_3m_sampling, Comparison = "Breastfeeding")

Astsumm_prev <- melted_funcs_taxa %>% subset(., asthma_3y != "Possible") %>% # rm Possible, and NA
            prev_summary_2(., "value", c(grouping_cols, "asthma_3y")) %>%
            mutate(Group = asthma_3y, Comparison = "Asthma")

#BFsumm_prev$Comparison <- "Breastfeeding" #  Exclusivity at 3 months
BFsumm_prev$BF_status_3m_sampling <- NULL
Astsumm_prev$asthma_3y <- NULL
All_prev_summ <- rbind(BFsumm_prev, Astsumm_prev)
```

```{r, fig.height=7, fig.width=6} 
All_prev_summ$Visit <- factor(All_prev_summ$Visit, levels = c("3 month", "1 year"))
All_prev_summ$Comparison <- factor(All_prev_summ$Comparison, levels = c("Breastfeeding", "Asthma"))

### for asthetics, reordering - R. gnavus, late functions, not late functions, and only using numeric id for non-late functions - just name main ones
table(All_prev_summ$variable_num)
table(All_prev_summ$var_alpha)

num_var_ns <- subset(All_prev_summ, variable %in% gsub("_ruminococcus.gnavus", "", ns_late_bf_rg_sub))$variable_num

unique(subset(All_prev_summ, variable %in% gsub("_ruminococcus.gnavus", "", late_overlap))$variable_num)
## reorder these to match heatmap in main Fig 5A
num_var_sig <- c("2.4.2.18", "1.11.1.1", "3.2.1.22", "2.7.7.7", "5.99.1.2", "3.5.1.2", "1.3.1.76", "3.5.1.24")

All_prev_summ$variable_num <- factor(All_prev_summ$variable_num, levels = c("Ruminococcus\ngnavus", unique(num_var_sig), unique(num_var_ns)))


# Annotate with change in prevalence - Exclusive VS. No BF
## make a df for this
BFsummaryEC_bin_sub_wide <- data.frame(subset(All_prev_summ, Visit == "1 year"),
                                       subset(All_prev_summ, Visit == "3 month"))

## double checks
table(BFsummaryEC_bin_sub_wide$variable == BFsummaryEC_bin_sub_wide$variable.1)
table(BFsummaryEC_bin_sub_wide$Group == BFsummaryEC_bin_sub_wide$Group.1)

## generate summary variable - this will correspond to the prevalence in PreTCoA plots
BFsummaryEC_bin_sub_wide$Change_Prevalence <- round(BFsummaryEC_bin_sub_wide$Prevalence - BFsummaryEC_bin_sub_wide$Prevalence.1, 1)
# annotate Change_Prevalence, link w/t BF_status_3m_sampling & variable

All_prev_summ_sp <- split(All_prev_summ, All_prev_summ$Comparison)
BFsummaryEC_bin_sub_wide_sp <- split(BFsummaryEC_bin_sub_wide, BFsummaryEC_bin_sub_wide$Comparison)

table(BFsummaryEC_bin_sub_wide_sp$Breastfeeding$variable_num)
table(All_prev_summ_sp$Breastfeeding$variable_num)

vjust_n <- 0

tiff(file = paste0(path, "Figures/Fig_S8B_ECStrata_BF.tiff"), width = 1800, height = 2600, units = "px", res = 300)
ggplot(All_prev_summ_sp$Breastfeeding, aes(x = Group, y = Prevalence, fill = Visit)) + 
  geom_bar(stat = "identity", position = "dodge") +
  theme(axis.text.x = element_text(size = 9))+
  labs(y = "Prevalence (%)", x = "Breastfeeding at 3 Months")+
  facet_wrap(~variable_num, ncol = 3, scales = "free_x") + 
  scale_y_continuous(limits = c(0, 115), expand = c(0,0), breaks = c(0,25,50,75,100))+
  geom_text(data = BFsummaryEC_bin_sub_wide_sp$Breastfeeding, aes(label = Change_Prevalence, x = Group), vjust = vjust_n, color = "#2c4382", size = 4) +
      geom_text(data = BFsummaryEC_bin_sub_wide_sp$Breastfeeding, aes(label = Change_Prevalence, x = Group), vjust = vjust_n, color = "#2c4382", size = 4.07) +
  geom_text(data = BFsummaryEC_bin_sub_wide_sp$Breastfeeding, aes(label = Change_Prevalence, x = Group), vjust = vjust_n, color = "#2c4382", size = 4.09) +
  scale_fill_manual(values = c("#c3c4bb", "#8c8c87"))
dev.off()

tiff(file = paste0(path, "Figures/Fig_S8B_ECStrata_Asthma3y.tiff"), width = 1800, height = 2600, units = "px", res = 300)
ggplot(All_prev_summ_sp$Asthma, aes(x = Group, y = Prevalence, fill = Visit)) + 
  geom_bar(stat = "identity", position = "dodge") +
  theme(axis.text.x = element_text(size = 9))+
  labs(y = "Prevalence (%)", x = "Asthma, 3 Years")+
  facet_wrap(~variable_num, ncol = 3, scales = "free_x") + 
  scale_y_continuous(limits = c(0, 115), expand = c(0,0), breaks = c(0,25,50,75,100))+
   # lazy cheating bold rather than parse/ggtext/etc
  geom_text(data = BFsummaryEC_bin_sub_wide_sp$Asthma, aes(label = Change_Prevalence, x = Group), vjust = vjust_n, color = "#2c4382", size = 4) +
    geom_text(data = BFsummaryEC_bin_sub_wide_sp$Asthma, aes(label = Change_Prevalence, x = Group), vjust = vjust_n, color = "#2c4382", size = 4.07) +
  geom_text(data = BFsummaryEC_bin_sub_wide_sp$Asthma, aes(label = Change_Prevalence, x = Group), vjust = vjust_n, color = "#2c4382", size = 4.09) +
  scale_fill_manual(values = c("#c3c4bb", "#8c8c87"))
dev.off()
```


# Figure S8A & S8C - Correlations between colonization patterns and functional prevalence trajectories 

## Clean data, derive Functional Trajectories (used in PreTCO)
To simplify making of prevalence trajectory scatter plots, first make the 'difference in prevalence' results for every pair of samples for each infant (1Y - 3M, e.g. 0 - 1 = -1; 1 - 0 = 1) - so a "-1" means it's present at 3 months but not 1 year, 1 means 1 year but not 3m and 0 means no change.
```{r} 
# Make dataset for Change in Prevalence from 3M to 1y (always TP2 - TP1) - append sdf -  NOTE that this is per-participant 

## proof check they match
### table(subset(funcs_3M1Y_bin_s$EggNog, Visit == "1 year")$subjectnumber == subset(funcs_3M1Y_bin_s$EggNog, Visit == "3 month")$subjectnumber)

# merge difference b/t timepoints - 1 year minus 3 months for same df - with sample data
funcs_3M1Y_bin_diff_s <- list()
for(i in seq(funcs_3M1Y_bin_s)) { 
  funcs_3M1Y_bin_diff_s[[i]] <- merge(subset(funcs_3M1Y_bin_s[[i]],
                                             Visit == "1 year", select =  
                                               intersect(names_bestPrevalence_s[[i]], colnames(funcs_3M1Y_bin_s[[i]]))
                                             ) - subset(funcs_3M1Y_bin_s[[i]], 
                                                        Visit == "3 month", 
                                                        select =  intersect(names_bestPrevalence_s[[i]], colnames(funcs_3M1Y_bin_s[[i]]))
                                                        ), 
                                      sdf, by = "row.names")
  funcs_3M1Y_bin_diff_s[[i]]$Row.names <- NULL
}
names(funcs_3M1Y_bin_diff_s) <- names(funcs_3M1Y_bin_s)
lapply(funcs_3M1Y_bin_diff_s, function(x) ncol(x) - ncol(sdf)) ## good, adds up
#N_vars <- lapply(funcs_3M1Y_bin_diff_s, function(x) ncol(x) - ncol(sdf)) # number of taxa or vars, to use in downstream 
#N_subj <- lapply(funcs_3M1Y_bin_diff_s, nrow) # number of subjects, to use in downstream 
```

## 16S Trajectory dataset - Gut
```{r}
#### Prevalence Trajectory permutation test results generated for individual taxa 
## Results generated in 03_Fig3_4_S4AC_S5_TabS4.Rmd

Per_test_df_ls <- readRDS(paste0(path, "Tables/PermTests/MAIN_Permutation_PrevTraj_Res.rds"))

#### Prevalence Trajectory feature table used for prevalence trajectory analysis 
## Results generated in 03_Fig3_4_S4AC_S5_TabS4.Rmd
change_3m1y_pa_all <- read.csv(file=paste0(path, "Data/Data_subsets/Prevalence_Trajectory_Features_N171_G172.csv"), row.names = 1) # The first 171 columns are nasal taxa, 172 to end of df is gut

#### Trajectory Summary variables derived from trajectory analyses & participant metadata
## generated in 00_Processing_Derived_Trajectories.Rmd
Traj_data_ls <- readRDS(paste0(path, "Data/Data_subsets/ALL_Trajectory_Derived_Data.rds")) 
```

### Prevalence Trajectories of individual taxa significantly delayed with breastfeeding and asthma
```{r}
# Select individual taxa 
Per_test_df_ls_p0.01 <- lapply(Per_test_df_ls, function(x) {subset(x, p_value <0.05)})

BF_asthma_sig_G <- intersect(row.names(Per_test_df_ls_p0.01$BF_Gut),
                             row.names(Per_test_df_ls_p0.01$Ast3y_Gut)) %>% 
                      gsub("^", "G_", .)
BF_asthma_sig_G

change_3m1y_pa_Select <- subset(change_3m1y_pa_all, select = BF_asthma_sig_G)
```

### Add participant metadata
```{r}
# Exposure, outcome, and Covariates used (covariates selected based on associations with Asthma, and with nasal and/or gut microbiota based on Fig. S2A).
Subj_dt <- c("BF_status_3m_sampling", "asthma_3y", "StudyCenter", "atbx", "birthmode2", "prenatal_smoke", "mom_asthma_ever", "coldsnum_0to3m", "older_sibs_living", "RE_Virus_1y", "older_sibs", "RE_Virus_3m", "age_3m_months", "age_1y_months") ## added some extra covariates not included in final analysis but checked in sensitivity analysis too

### Merge Trajectories of individual taxa with Trajectory summary variables and metadata 
## and make for Asthma 3 year analysis only (remove possible asthma)
G_3m1y_asthm_traj_sdf <- merge(subset(Traj_data_ls$Metadata, select = Subj_dt), 
                                change_3m1y_pa_Select, 
                               by="row.names") %>%
                          subset(., asthma_3y != "Possible" & 
                                   !is.na(G_Monoglobus.sp.)) ### Subset for Gut microbome trajectory data only 

names(G_3m1y_asthm_traj_sdf)[names(G_3m1y_asthm_traj_sdf) == "Row.names"] <- "CHILDid" # use correct name for this var

## Extra check, number of non-missing values per variable - should be ~477 for gut 
lapply(as.list(G_3m1y_asthm_traj_sdf), function(x) {sum(!is.na(x))})
```

## Check overlap b/t functional and taxonomic trajectory datasets and merge 
Note that Functional data merged is only related to R. gnavus - may want to make a SEM unique to R. gnavus for functional data integration
```{r}
length(intersect(G_3m1y_asthm_traj_sdf$CHILDid, funcs_3M1Y_bin_diff_s$EggNog$subjectnumber))
### N = 322 for this dataset

# recall from data setup and panel A
## sig late in both: late_overlap
## 29 not sig late: ns_late_bf_rg
#### of these the random 6 plotted in panel A: ns_late_bf_rg_sub
## other labels used for R. gnavus functions: All_prev_summ
## All R.g functions: 
all_rg_func <- row.names(Res_per_taxa_ec_bf)[grepl("ruminococcus.gnavus$", row.names(Res_per_taxa_ec_bf))]

# Taxa with EC trajs
G_taxaEC_3m1y <- merge(G_3m1y_asthm_traj_sdf, 
                                     subset(funcs_3M1Y_bin_diff_s$EC, 
                                            select = c("subjectnumber", all_rg_func)),
                                    by.x = "CHILDid", by.y = "subjectnumber")
dim(G_taxaEC_3m1y) # n = 322


# Also want to add R. gnavus species from metagenomic data to correlation check - quick addition (clean up later?)
## Subset/melt: R. gnavus from taxonomic data
sub_taxa <- Gut_3M1Y_all_filt %>% otu_table %>% t %>% as.data.frame %>% 
                  subset(., select = c(taxa_select)) %>% data.frame(., subset(sample_data(Gut_3M1Y_all_filt), select = c("Visit", "subjectnumber"))) %>%
                  mutate(Ruminococcus_gnavus = ifelse(Ruminococcus_gnavus > 0, 1, 0))
sub_taxa_sp <- split(sub_taxa, sub_taxa$Visit) 

sub_taxa_wide <- merge(sub_taxa_sp$`1 year`, sub_taxa_sp$`3 month`, by = "subjectnumber", suffixes = c("_1y", "_3m")) %>%
                  mutate(Taxa_Rgnavus_traj = ifelse(Ruminococcus_gnavus_1y == 1 & Ruminococcus_gnavus_3m == 0, 1,
                                                           ifelse(Ruminococcus_gnavus_1y == 0 & Ruminococcus_gnavus_3m == 1, -1, 0))
                         )

G_taxaEC_3m1y <- merge(G_taxaEC_3m1y, subset(sub_taxa_wide, select = c("subjectnumber", "Taxa_Rgnavus_traj")),
                       by.x = "CHILDid", by.y = "subjectnumber")
dim(G_taxaEC_3m1y) # same n good
```

# Correlation between colonization patterns of taxa significantly associated with asthma and BF, and EC functional trajectories associated with asthma and BF (all from R. gnavus)
variable is the change in prevalence between timepoints (-1 = early, 0 = no change, 1= late/1 year only) used for PreTCo and SEMs 
```{r}
# note - have a function for this same correlation heatmap, may switch to later

X <- subset(G_taxaEC_3m1y, select = c("Taxa_Rgnavus_traj", BF_asthma_sig_G, all_rg_func))
colnames(X)[colnames(X) %in% BF_asthma_sig_G] <- gsub("^G_|^", "Taxa_", colnames(X)[colnames(X) %in% BF_asthma_sig_G])

### term splitting within vs between group comp 
differentiator <- "^Taxa_"

cor_feat <- rcorr(as.matrix(X), type="spearman") 

## Between-Feature Corr script ##

cor_feat_2 <- cor_feat$r 
cor_feat_2[lower.tri(cor_feat_2)] <- NA # just need upper triangle, removes duplicates

# melt is from reshape2 package
cor_feat_m <- melt(cor_feat_2, value.name = "corr") %>% subset(., !(Var1 == Var2))

table_p_feat <- (as.matrix(cor_feat$P))
table_p_feat[lower.tri(table_p_feat)] <- NA # just need upper triangle of matrix, removes duplicates
#Note that P-val is additionally NA for correlations b/t the same covariate [unlike correlation coefficient which is 1]
table_p_feat_m <- melt(table_p_feat, value.name = "p_val") %>% subset(., !(Var1 == Var2))

table_n_feat <- (as.matrix(cor_feat$n))
table_n_feat[lower.tri(table_n_feat)] <- NA 
table_n_feat_m <- melt(table_n_feat, value.name = "n") %>% subset(., !(Var1 == Var2))


corr_by_feature <- merge(cor_feat_m, table_p_feat_m, by = c("Var1", "Var2")) %>% 
  merge(., table_n_feat_m, by = c("Var1", "Var2")) %>%  
  subset(., !is.na(p_val)) %>% #where p-value = NA remove, not looking at these
  mutate(diss = 1 - abs(corr),
         Type1 = ifelse(grepl(differentiator, Var1), "Taxa", "Function"), ## "differentiator" group first 
         Type2 = ifelse(grepl(differentiator, Var2), "Taxa", "Function"),
         Comparison = ifelse(Type1 == Type2, "WithinGroup", "BetweenGroup"),  
         Sig = ifelse(p_val < 0.001, "**", "") ## For now just using low thresh
         ) 

corr_between <- subset(corr_by_feature, Comparison == "BetweenGroup")
```

**Just check average correlation coefficients per overall group (not per each pairwise comparison!)**
```{r}
## Annotate based on whether function was later in both BF and asthma or not late with BF (Var2)
corr_between$Var2_Cat <- ifelse(corr_between$Var2 %in% late_overlap, "Late, both BM and Asthma",
                                ifelse(corr_between$Var2 %in% ns_late_bf_rg, "Not Late with BM", "Late, BM only"))
corr_between$Var2_Cat  <- factor(corr_between$Var2_Cat,
                                 levels = c("Not Late with BM", "Late, BM only", "Late, both BM and Asthma"))
### n = 8 late w/t BF and Asthma, n = 30 not late with BF, n = 354 late BF only 

# checking correlation w/t our R. gnavus otu specifically 
data_summary_V2(subset(corr_between, Var1 == "Taxa_X.Ruminococcus..gnavus..sp."), varname = "corr", groupnames = c("Var2_Cat"))

## also - checking overall summary 
summary(subset(corr_between, Var1 == "Taxa_X.Ruminococcus..gnavus..sp.")$corr)
boxplot(subset(corr_between, Var1 == "Taxa_X.Ruminococcus..gnavus..sp.")$corr)


# The R. gnavus species in metagen data Taxa_Rgnavus_traj
data_summary_V2(subset(corr_between, Var1 == "Taxa_Rgnavus_traj"), varname = "corr", groupnames = c("Var2_Cat"))

## Also - cross-check contingency table for R. gnavus - 16S OTU and metagenomic species trajectories - line up really well
table(G_taxaEC_3m1y$Taxa_Rgnavus_traj, G_taxaEC_3m1y$G_X.Ruminococcus..gnavus..sp.)
chisq.test(table(G_taxaEC_3m1y$Taxa_Rgnavus_traj, G_taxaEC_3m1y$G_X.Ruminococcus..gnavus..sp.))
```

## Figure S8A - boxplots of correlations
```{r}
func_rgnavus_corr <- subset(corr_between, Var1 == "Taxa_Rgnavus_traj")
pairwise.wilcox.test(func_rgnavus_corr$corr, func_rgnavus_corr$Var2_Cat)

tiff(file = paste0(path, "Figures/Fig_S8A_RGnavusFuncTrajCorr.tiff"), width = 1800, height = 700, units = "px", res = 300)
ggplot(func_rgnavus_corr, aes(x=Var2_Cat, y = corr)) + 
  geom_violin(trim = FALSE, alpha=0) + 
  stat_summary(fun=median, geom="point", shape=23, size=2, fill="black")  +
  theme(axis.text = element_text(size=11), legend.position = "none", 
        axis.title.x = element_text(face = "bold"), axis.title = element_text(size =12))  +
  labs(y="Spearman's\ncorrelation coefficient", x="Delay in R. gnavus EC Functions with Breastfeeding and Asthma") + 
  geom_signif(color="black", comparisons = list(c("Not Late with BM", "Late, BM only"), 
                                                c("Not Late with BM", "Late, both BM and Asthma"),
                                                c("Late, BM only", "Late, both BM and Asthma")),
              test = "wilcox.test", step_increase = 0.2, vjust = 0.6,
              map_signif_level = c("**"=0.001, "*"=0.05,"~"=0.1," "=2), size=0.3, textsize = 7) 
dev.off()
```


## Figure S8C - correlations b/t metagenomic functional trajectories and the 16S colonization patterns found late w/t both asthma and BF
**Move forward with main sub-analysis - selecting just those late w/t both BF and asthma, and the other random set of 6 from the 'Not Late with BF' group shown in Fig S9A**
```{r}
## also adding fdr correction to pvals, and add the cleaned annotation name for Var2 - same used for panel A - All_prev_summ
## levels(All_prev_summ$variable_num)[2:length(levels(All_prev_summ$variable_num))]
# df link - original names to labels
df_link <- data.frame(Var2 = All_prev_summ$variable,
                      variable_num = All_prev_summ$variable_num) %>% subset(., !duplicated(Var2) & Var2 != "Ruminococcus_gnavus")

corr_between_2 <- corr_between %>% subset(., Var2 %in% c(late_overlap, ns_late_bf_rg_sub) & Var1 != "Taxa_Rgnavus_traj") %>% ##
                    mutate(p_val_bh = p.adjust(p_val, method = "BH", n = nrow(corr_between)),
                           Sig_bh = ifelse(p_val_bh < 0.01, "**", 
                                            ifelse(p_val_bh < 0.05, "*", "")),
                           Sig_2 = ifelse(p_val < 0.05 & p_val_bh >= 0.05, "~", ""),
                           Var2 = gsub("_ruminococcus.gnavus$", "", Var2))  %>% 
                    merge(., df_link, by = "Var2") %>%
                    mutate(Lab1 = gsub("Taxa_|X\\.", "", Var1),
                              # levels of funcs to match same as panel A
                           Lab2 = factor(variable_num, 
                                         levels = rev(levels(All_prev_summ$variable_num)[2:length(levels(All_prev_summ$variable_num))])),
                           italic_x = gsub("\\.\\d{1,}$", "", Lab1),
                          non_italic_x = ifelse(is.na(str_extract(Lab1, "\\d{1,}$") ), "", 
                               str_extract(Lab1, "\\d{1,}$") )
                           ) %>%
         mutate(Lab1 = glue("<i>{italic_x}</i> {non_italic_x}"))
table(corr_between_2$Lab1)

# taxa name extra format edit
corr_between_2$Lab1 <- corr_between_2$Lab1 %>% gsub("\\.\\.", " ", .) %>% gsub("\\.", " ", .) #%>% gsub(". ", " ", .)
```

```{r}
## margins
r <- 2 # right margin edit
t <- 0.1 # top margin edit

tiff(file = paste0(path, "Figures/Fig_S8C_FuncTaxaTrajCorr.tiff"), width = 2200, height = 1800, units = "px", res = 300)
ggplot(corr_between_2, aes(Lab1, Lab2, fill=corr)) +
  geom_tile(height=0.8, width=0.8) +
  scale_fill_gradient2(low="#0056cf", mid="white",  high="#d63200", midpoint = 0) +
  geom_text(aes(label=Sig_bh), size=6, nudge_y = -0.14) + 
  geom_text(aes(label=Sig_2), size=5, nudge_y = 0.08) + 
  theme(panel.spacing.x = unit(1, "mm"), axis.title = element_blank(), axis.ticks = element_blank(),
        axis.text.x=element_markdown(size=10, angle=30, vjust=1, hjust=1, color = "black"),
        axis.text.y=element_text(size=12, color = "black"), strip.text.x = element_blank()) + # plot.margin = unit(c(t, r, 0, 0), "cm")
  labs(fill = "Spearman rho") 
dev.off()
```


**Median Correlation values with these R. gnavus functions - per-taxa and stratified by association of functional trajectory with breastfeeding**
```{r}
kable(data_summary_V2(corr_between_2, "corr", "Lab1"))

kable(data_summary_V2(corr_between_2, "corr", c("Var2_Cat", "Lab1")))
```

## Supplementary analysis to obtain p-values for change in prevalene by breastfeeding and asthma for metagenomic species - specifically R. gnavus, highlighted in Fig. S8B
See Fig_3.Rmd and Fig_4.Rmd for main PreTCO analyses - scripts copied from there

```{r}
## Inputs, name of timepoint 1 and 2 variables, name of subject ID variable
T1 <- "3 month"
T2 <- "1 year"
SubjID <- "subjectnumber"

Gut_3M1Y_pa_filt <- transform_sample_counts(Gut_3M1Y_all_filt, function(abund) 1*(abund>0))

G.pa.3m <- subset_samples(Gut_3M1Y_pa_filt, Visit == T1)
G.pa.1y <- subset_samples(Gut_3M1Y_pa_filt, Visit == T2)
```

**Gut - Prevalence Trajectory derived variables, creation of a prevalence trajectory feature table**
```{r} 
# main script
## extract otu table and use Participant IDs as row.names instead - will match b/t datasets
G_3m_otu <- data.frame(as.data.frame(t(as(otu_table(G.pa.3m), "matrix"))), row.names=sample_data(G.pa.3m)[[SubjID]])
G_1y_otu <- data.frame(as.data.frame(t(as(otu_table(G.pa.1y), "matrix"))), row.names=sample_data(G.pa.1y)[[SubjID]])

G_3m_otu <- G_3m_otu[order(row.names(G_3m_otu)),]
G_1y_otu <- G_1y_otu[order(row.names(G_1y_otu)),]

## Checking - colnames and Childid's are in the same order? Yes - good - can move forward
identical(colnames(G_3m_otu), colnames(G_1y_otu))
identical(row.names(G_3m_otu), row.names(G_1y_otu))

## 1 Year minus 3 Month Prevalence For each Taxa
change_3m1y_pa_G <- G_1y_otu - G_3m_otu # -1 = present at 3m not 1y, 1= present at 1y not 3m, 0=either present at both or neither timpoint
dim(change_3m1y_pa_G) # Same order as original physeq obj

sdf_toAdd <- sample_data(G.pa.1y)[order(sample_data(G.pa.1y)[[SubjID]]),]
## check that it's in the same order as change_3m1y_pa
identical(row.names(change_3m1y_pa_G), as.character(sdf_toAdd[[SubjID]])) # order matches

### same order so just merge data with sample metadata using data.frame
change_3m1y_pa_G_sdf <- data.frame(change_3m1y_pa_G, sdf_toAdd) 

N_taxa_G <- ncol(change_3m1y_pa_G) # number of taxa, to use in downstream code
N_subj_G <- nrow(change_3m1y_pa_G) # number of subjects, to use in downstream code
```

### Asthma at 3 years 
```{r} 
## input 
X <- subset(change_3m1y_pa_G_sdf, asthma_3y == "Yes")
Y <- subset(change_3m1y_pa_G_sdf, asthma_3y == "No")

# script

# Make data.frame of prevalence like before BUT keeping as proprotions instead of Percentages for now
df.pa <- data.frame(X_P_Change=colSums(X[1:N_taxa_G])*100/dim(X)[1],
                    Y_P_Change=colSums(Y[1:N_taxa_G])*100/dim(Y)[1],
                    taxonomy=as(tax_table(Gut_3M1Y_pa_filt), "matrix"),
                    prevalence=taxa_sums(Gut_3M1Y_pa_filt)*100/(N_subj_G*2)) 

# difference in change
df.pa$pa.change.diff <- df.pa$Y_P_Change - df.pa$X_P_Change


perm_asthma_metagen_taxa <- perm_func_dfs_v2(X_df = df.pa, 
                                  data = change_3m1y_pa_G_sdf,
                                  variable = "asthma_3y", Group_X = "Yes", Group_Y = "No",
                                  N = N_taxa_G) 

perm_asthma_metagen_taxa$per_taxa_res %>% subset(., row.names(.) == "Ruminococcus_gnavus")
```

### Breastfeeding at 3 months
```{r} 
## input
X <- subset(change_3m1y_pa_G_sdf, BF_status_3m_sampling == "No BM")
Y <- subset(change_3m1y_pa_G_sdf, BF_status_3m_sampling == "Exclusive")

# Make data.frame of prevalence like before BUT keeping as proprotions instead of Percentages for now
df.pa <- data.frame(X_P_Change=colSums(X[1:N_taxa_G])*100/dim(X)[1],
                    Y_P_Change=colSums(Y[1:N_taxa_G])*100/dim(Y)[1],
                    taxonomy=as(tax_table(Gut_3M1Y_pa_filt), "matrix"),
                    prevalence=taxa_sums(Gut_3M1Y_pa_filt)*100/(N_subj_G*2)) # Taxonomy actually maintained compared to physeq obj, great

# difference in change
df.pa$pa.change.diff <- df.pa$Y_P_Change - df.pa$X_P_Change


perm_bf_metagen_taxa <- perm_func_dfs_v2(X_df = df.pa, 
                                  data = change_3m1y_pa_G_sdf,
                                  variable = "BF_status_3m_sampling", 
                                  Group_X = "No BM", Group_Y = "Exclusive",
                                  N = N_taxa_G)

perm_bf_metagen_taxa$per_taxa_res %>% subset(., row.names(.) == "Ruminococcus_gnavus")
```



