---
title: "Figure S3. Replication of Figure 3a, using species-level shotgun metagenomic data."
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```
 
## Setup 

```{r}
R.version.string
library(ggplot2); packageVersion('ggplot2')
library(tidyr); packageVersion('tidyr')
library(reshape2); packageVersion('reshape2')
library(dplyr)
library(here)

source(paste0(here::here(), "/Functions.R"))

theme_set(theme_bw() + theme(strip.text.x = element_text(size = 11), strip.background = element_rect(colour="white", fill="white"), panel.grid.major = element_blank(), panel.grid.minor = element_blank(), text = element_text(size=12), axis.text = element_text(color="black", size=12), legend.key.size = unit(0.5,"cm")))
```

## Read in Metagenomic data & PreTCO results
```{r}
## cleaned metagenomic taxonomy and sample data, also normalize metagenomic counts
Gut_3M1Y_all_filt <- readRDS(file = paste0(path, "Data/Metagenome_metabolome/final_datasets/03_Abund_metagenomic_Trajectory.rds"))
```
Setup input data
```{r}
metagen_taxa <- data.frame(as.data.frame(t(otu_table(Gut_3M1Y_all_filt))),
                           subset(sample_data(Gut_3M1Y_all_filt), select = c("subjectnumber", "Visit"))
                           )
```

# Figure S3 - Volcano plot of change in prevalence and McNemar test to assess change in prevalence between 3 months and 1 year

## Also try Volcano plot for prevalence change over time - see consistent results for presence/absense data? This will also give an indication of whether our PreTCo system will work well or not.
```{r}
# wide df, to make sure 3m and 1y are matched by subject number 
df <- merge( subset(metagen_taxa, Visit == "3 month"),
                        subset(metagen_taxa, Visit == "1 year"),
                        by = "subjectnumber", suffixes = c("_3m", "_1y")
            )
df$Visit_1y <- NULL
df$Visit_3m <- NULL
df$subjectnumber <- NULL

## double checks...
sum(grepl("_3m$|_1y$", colnames(df))) == length(colnames(df))
identical(gsub("_3m", "", colnames(df[,grepl("_3m", colnames(df))])), 
          gsub("_1y", "", colnames(df[,grepl("_1y", colnames(df))])))

# first make binary then apply table to x and y, then apply mcnemar to the table produced
df_pa <- apply(df, MARGIN=2, FUN=make_binary) %>% as.data.frame


## Extra requirment -  there's one taxa only present at 3 months that has to be removed, McNemar test won't work on this one.
df_pa_2 <- df_pa
df_pa_2$Staphylococcus_epidermidis_3m <- NULL 
df_pa_2$Staphylococcus_epidermidis_1y <- NULL 


funcs_mcnemar <- mapply(table, 
                        x = as.list(df_pa_2[,grepl("_3m", colnames(df_pa_2))]),
                        y = as.list(df_pa_2[,grepl("_1y", colnames(df_pa_2))]),
                        SIMPLIFY = FALSE, USE.NAMES = TRUE) %>% 
  lapply(., mcnemar.test) %>%
  lapply(., function(x) {x["p.value"]}) %>% bind_rows(., .id = "variable") %>% 
  mutate(p.fdr = p.adjust(p.value, method="BH", n = length(p.value)))

# For effect sizes, just merging w/t 3m & 1y prevalence previously summarized in funcs_byVisit (just keep those tested)
prev_pertp <- colMeans(df_pa)*100 

prev_pertp_df <- data.frame(Prevalence_3m = prev_pertp[grepl("_3m", names(prev_pertp))],
                            Prevalence_1y = prev_pertp[grepl("_1y", names(prev_pertp))]
                            ) %>% merge(., funcs_mcnemar, by.x = "row.names", by.y = "variable", all = TRUE) %>%
  mutate(ChangeP_1Y3M = (Prevalence_1y - Prevalence_3m), 
         Colonization_cats = factor(ifelse(p.fdr < 0.001 & ChangeP_1Y3M < -7, "Early", 
                                    ifelse(p.fdr < 0.001 & ChangeP_1Y3M > 7, "Late", "Persistent")), 
                                    levels = c("Early", "Persistent", "Late"))
         )
prev_pertp_df$Row.names <- gsub("_3m", "", prev_pertp_df$Row.names)


## reorder and get the top and bottom effect sizes (pos and neg)
prev_pertp_df <- prev_pertp_df[order(-prev_pertp_df$ChangeP_1Y3M),]
top_taxa <- prev_pertp_df$Row.names[1:3]
bot_taxa <- prev_pertp_df$Row.names[(nrow(prev_pertp_df)-2):nrow(prev_pertp_df)]

prev_pertp_df2 <- prev_pertp_df %>% mutate(Label = ifelse(Row.names %in% top_taxa, gsub("_", " ", Row.names), ""), 
                                           Label2 = ifelse(Row.names %in% bot_taxa, gsub("_", " ", Row.names), "") 
                                            )

write.csv(prev_pertp_df2, paste0(path, "Tables/Res_FigS3_Metagen_SpeciesMcNemar.csv"))
```

```{r}
axis_size <- 9
ax_title_s <- 10
ann_size <- 2.8
table(prev_pertp_df2$Label)
table(prev_pertp_df2$Label2)

  volcanos_prev <- ggplot(data=subset(prev_pertp_df2, !is.na(Colonization_cats)), 
                          aes(x=ChangeP_1Y3M, y=-log10(p.fdr), color=Colonization_cats)) + 
  geom_point(size=2) + geom_hline(yintercept = -log10(0.001), color="black") + 
    geom_vline(xintercept = 7, color="black")  + 
  geom_vline(xintercept = -7, color="black") + 
    labs(color="Colonization", x="1 Year - 3 Month Prevalence (%)", y="-log10(BH p-value)") + 
  ggtitle("Species changes in prevalence, metagenomic data") + 
  theme(axis.text = element_text(size=axis_size), axis.title = element_text(size= ax_title_s), title = element_text(size=11)) +
    
    geom_text(aes(label=ifelse(Label != "Blautia wexlerae", Label, "")), 
              nudge_y=0.5, nudge_x=-22, color="black", size=ann_size, fontface=3) + 
    geom_text(aes(label=ifelse(Label == "Blautia wexlerae", Label, "")), 
              nudge_y=0.5, nudge_x=17, color="black", size=ann_size, fontface=3) + 

    geom_text(aes(label=ifelse(Label2 != "Klebsiella pneumoniae", Label2, "")), 
              nudge_x = 20,nudge_y=0.8, color="black", size=ann_size, fontface=3) +
    geom_text(aes(label=ifelse(Label2 == "Klebsiella pneumoniae", Label2, "")), 
              nudge_x = -22,nudge_y=0.8, color="black", size=ann_size, fontface=3)

tiff(file = paste0(path, "Figures/Fig_S3_GutMetagen_Volcano_Overall.tiff"), width = 1900, height = 1300,  units = "px", res = 300) 
volcanos_prev
dev.off()
```

Assess numbers & Largest effect sizes 
```{r}
data_summary_V2(prev_pertp_df, varname = "ChangeP_1Y3M", groupnames = "Colonization_cats") 
```

-most early
```{r}
subset(prev_pertp_df,  ChangeP_1Y3M < -25) 
```

- most late
```{r}
subset(prev_pertp_df,  ChangeP_1Y3M > 50) 
```



