---
title: "The trajectories of nasal and gut microbiota mediate the association between breastfeeding exclusivity and asthma in separate models (Figure S7)"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
 
## Setup 

```{r, message=FALSE, warning=FALSE}
R.version.string
library(lavaan); packageVersion('lavaan')
library(ggplot2); packageVersion('ggplot2')
library(tidyr); packageVersion('tidyr')
library(reshape2); packageVersion('reshape2')

source("Preprocessing/Functions.R")
```

```{r}
theme_set(theme_bw() + theme(strip.text.x = element_text(size = 11), strip.background = element_rect(colour="white", fill="white"), panel.grid.major = element_blank(), panel.grid.minor = element_blank(), text = element_text(size=12), axis.text = element_text(color="black", size=12), legend.key.size = unit(0.5,"cm")))
```


#### Prevalence Trajectory permutation test results generated for individual taxa 
Results generated in 03_Fig3_4_S4AC_S5_TabS4.Rmd
```{r}
Per_test_df_ls <- readRDS("Output/PermTests/ALL_Permutation_PrevTraj_Res.rds")
```

#### Prevalence Trajectory feature table used for prevalence trajectory analysis 
Results generated in 03_Fig3_4_S4AC_S5_TabS4.Rmd
```{r}
change_3m1y_pa_all <- read.csv(file="Output/Prevalence_Trajectory_Features_N171_G172.csv", row.names = 1) # The first 171 columns are nasal taxa, 172 to end of df is gut
```

#### Trajectory Summary variables derived from trajectory analyses & participant metadata
generated in 00_Processing_Derived_Trajectories.Rmd
```{r}
Traj_data_ls <- readRDS("Data_subsets/ALL_Trajectory_Derived_Data.rds") # Derived Trajectory variables 
```


## Variable selection and create data subsets for mediation 
Using a p-value threshold of p<0.01. Mediation is only really possible when path a (BF --> microbiota trajectory) and path b (microbiota trajectory --> Asthma) are significant; so select taxa for which this is true.
### Prevalence Trajectories of individual taxa
```{r}
# Select individual taxa
Per_test_df_ls_p0.01 <- lapply(Per_test_df_ls, function(x) {subset(x, p_value <0.01)})

# Taxa significant for breastfeeding and asthma
## Nasal
BF_asthma_sig <- intersect(row.names(Per_test_df_ls_p0.01$BF_Nasal), row.names(Per_test_df_ls_p0.01$Ast3y_Nasal))
BF_asthma_sig

BF_asthma_sig_G <- intersect(row.names(Per_test_df_ls_p0.01$BF_Gut), row.names(Per_test_df_ls_p0.01$Ast3y_Gut))
BF_asthma_sig_G

# Data subset used from prevalence trajectory feature table
## gut and nasal in same dataset - pre-pended N and G so do the same with sig taxa vectors
BF_asthma_sig <- gsub("^", "N_", BF_asthma_sig)
BF_asthma_sig_G <- gsub("^", "G_", BF_asthma_sig_G)

change_3m1y_pa_Select <- subset(change_3m1y_pa_all, select = c(BF_asthma_sig, BF_asthma_sig_G))
dim(change_3m1y_pa_Select)
```

### Trajectory summary variables and participant metadata
Additionally using the proportion of Late colonizers and the Richness trajectory as summary variables since these were both significantly associated w/t BF and w/t asthma for BOTH nasal and gut (tests in 02_Fig2_S2_S3_S4BD.Rmd)
```{r}
# Trajectory variables
Summ_Trajs_N <- c("Observed_N1y3m", "Late_N1y3m")
Summ_Trajs_G <- c("Observed_G1y3m", "Late_G1y3m")

# Exposure, outcome, and Covariates used (covariates selected based on associations with Asthma, and with nasal and/or gut microbiota based on Fig. S2A).
Subj_dt <- c("BF_status_3m_sampling", "asthma_3y", "StudyCenter", "atbx", "birthmode2", "prenatal_smoke", "mom_asthma_ever", "coldsnum_0to3m", "older_sibs_living", "RE_Virus_1y", "older_sibs", "RE_Virus_3m", "age_3m_months", "age_1y_months") ## added some extra covariates not included in final analysis but checked in sensitivity analysis too


# Merge summary variables and metadata
N_3m1y_otu_sdf3 <- merge( subset(Traj_data_ls$Trajectories_Nasal_to181_Gut_at182, select = c(Summ_Trajs_N, Summ_Trajs_G)), 
                          subset(Traj_data_ls$Metadata, select = Subj_dt), by="row.names" ) 
dim(N_3m1y_otu_sdf3)
```

### Merge Trajectories of individual taxa with Trajectory summary variables and metadata  
```{r}
GN_3m1y_all_traj_sdf2 <- merge(N_3m1y_otu_sdf3, change_3m1y_pa_Select, by.x="Row.names", by.y = "row.names")
nrow(GN_3m1y_all_traj_sdf2) == nrow(N_3m1y_otu_sdf3) # good, still 1707
names(GN_3m1y_all_traj_sdf2)[names(GN_3m1y_all_traj_sdf2) == "Row.names"] <- "CHILDid" # use correct name for this var
```

### Make subset for Asthma at 3 year analysis (remove possible asthma)
```{r}
GN_3m1y_asthm_traj_sdf <- subset(GN_3m1y_all_traj_sdf2, asthma_3y != "Possible")

## Extra check, number of non-missing values per variable - should be 477 for gut trajectories and 1316 for nasal after removing possible asthma
lapply(as.list(GN_3m1y_asthm_traj_sdf), function(x) {sum(!is.na(x))})
```


## Variable modifications for structural equation model and split data for the 3 main SEM models
Note that we are using asthma_3y as the outcome, so the  ordered = argument will be used. In this case using 0/1 version vs. categorical factor doesn't make a difference really. 
From  https://lavaan.ugent.be/ : "When the ordered= argument is used, lavaan will automatically switch to the WLSMV estimator: it will use diagonally weighted least squares (DWLS) to estimate the model parameters, but it will use the full weight matrix to compute robust standard errors, and a mean- and variance-adjusted test statistic." 
```{r}
# Different versions of variables used
GN_3m1y_asthm_traj_sdf$asthma_3y_bin <- ifelse(as.character(GN_3m1y_asthm_traj_sdf$asthma_3y)=="No", 0, 1)

### 0 (no BM), 1 (partial), 2 (exclusive)
GN_3m1y_asthm_traj_sdf$BF_status_3m_sampl_num <- ifelse(as.character(GN_3m1y_asthm_traj_sdf$BF_status_3m_sampling)=="No BM", 0,
                                                        ifelse(as.character(GN_3m1y_asthm_traj_sdf$BF_status_3m_sampling)=="Partial", 1, 2)) 

### covariates
GN_3m1y_asthm_traj_sdf$older_sibs_bin <- ifelse(as.character(GN_3m1y_asthm_traj_sdf$older_sibs)=="No older siblings", 0, 1)
GN_3m1y_asthm_traj_sdf$atbx_bin <- ifelse(GN_3m1y_asthm_traj_sdf$atbx == "Yes", 1, 0)
GN_3m1y_asthm_traj_sdf$prenatal_smoke_num <- ifelse(GN_3m1y_asthm_traj_sdf$prenatal_smoke == "Yes", 1, 0)
GN_3m1y_asthm_traj_sdf$birthmode2_num <- ifelse(GN_3m1y_asthm_traj_sdf$birthmode2 == "Caesarean", 1, 0) # Caesarean ==1
GN_3m1y_asthm_traj_sdf$mom_asthma_num <- ifelse(GN_3m1y_asthm_traj_sdf$mom_asthma_ever == "Yes", 1, 0)

GN_3m1y_asthm_traj_sdf$RE_Virus_1y_num <- ifelse(GN_3m1y_asthm_traj_sdf$RE_Virus_1y == "R/E", 1, 0) 
GN_3m1y_asthm_traj_sdf$RE_Virus_3m_num <- ifelse(GN_3m1y_asthm_traj_sdf$RE_Virus_3m == "R/E", 1, 0) ## RE virus at 3m only used for sensitivity analysis


# Add Study Center too - only including Toronto b/c this is the main center that differs from the others and samples were also analyzed first for Toronto, then other centers (so Toronto vs. other centers also treated as a proxy for analysis batch effect)
GN_3m1y_asthm_traj_sdf$Toronto <- ifelse(GN_3m1y_asthm_traj_sdf$StudyCenter == "Toronto", 1, 0)

# In order to be on the same scale as other variables in the latent construct, must to take Observed and / 10000; for Late, need to /100 
GN_3m1y_asthm_traj_sdf$N_scaled_Late <- GN_3m1y_asthm_traj_sdf$Late_N1y3m/10
GN_3m1y_asthm_traj_sdf$N_scaled_Observed_N1y3m <- GN_3m1y_asthm_traj_sdf$Observed_N1y3m/100

GN_3m1y_asthm_traj_sdf$G_scaled_Late <- GN_3m1y_asthm_traj_sdf$Late_G1y3m/10
GN_3m1y_asthm_traj_sdf$G_scaled_Observed_G1y3m <- GN_3m1y_asthm_traj_sdf$Observed_G1y3m/100
```

### Subset for multiple mediation - Gut and Nasal together - only include infants that have both gut and nasal trajectory data
```{r}
NG_3m1y_asthm_traj_sdf_noNA <- subset(GN_3m1y_asthm_traj_sdf, !is.na(Observed_G1y3m) & !is.na(Observed_N1y3m)) # No NA values for any trajectory variables
dim(NG_3m1y_asthm_traj_sdf_noNA)

### Extra Check -  N per group for ordered / binary / count variables
ord_Vars <- colnames(NG_3m1y_asthm_traj_sdf_noNA)[colnames(NG_3m1y_asthm_traj_sdf_noNA) %ni% c("CHILDid", Summ_Trajs_N, Summ_Trajs_G, "age_3m_months",                 "age_1y_months", "N_scaled_Late", "N_scaled_Observed_N1y3m", "G_scaled_Late", "G_scaled_Observed_G1y3m")]

lapply(as.list(subset(NG_3m1y_asthm_traj_sdf_noNA, select = ord_Vars)), function(x) {table(x)})
```

### Subset for Nasal microbiome trajectory mediation alone 
```{r}
N_3m1y_asthm_traj_sdf <- subset(GN_3m1y_asthm_traj_sdf, !is.na(Observed_N1y3m))
dim(N_3m1y_asthm_traj_sdf)
```

### Subset for Gut microbome trajectory mediation alone
```{r}
G_3m1y_asthm_traj_sdf <- subset(GN_3m1y_asthm_traj_sdf, !is.na(Observed_G1y3m))
dim(G_3m1y_asthm_traj_sdf)
```


# Main models

### Nasal and Gut Microbiome Trajectory Latent Constructs - Trajectory Variables used 
```{r}
## NASAL Latent construct equation
micro.vars_N <- c("N_scaled_Observed_N1y3m", "N_scaled_Late", BF_asthma_sig) 
micro.model_N <- paste0("micro_N =~ ", paste(micro.vars_N, collapse="+"))
micro.model_N


## GUT Latent construct equation
micro.vars_G <- c("G_scaled_Observed_G1y3m","G_scaled_Late", BF_asthma_sig_G) 
micro.model_G <- paste0("micro_G =~ ", paste(micro.vars_G, collapse="+"))
micro.model_G
```


## NASAL Mediation Model (Figure S7)
**Model**
```{r}
N_Mediation_Mod <- ' 
    asthma_3y ~ b1 * micro_N + c1 * BF_status_3m_sampl_num + c2 * coldsnum_0to3m + c3 * older_sibs_living + c4 * atbx_bin + c5 * prenatal_smoke_num + c6 * birthmode2_num + c7 * mom_asthma_num + c8 * Toronto  +  c9 * RE_Virus_1y_num
    
    micro_N ~  a1 * BF_status_3m_sampl_num + a2 * coldsnum_0to3m + a3 * older_sibs_living + a4* atbx_bin + a5 * prenatal_smoke_num + a6 * birthmode2_num + a7 * mom_asthma_num + a8* Toronto  +  a9 * RE_Virus_1y_num
    
#indirect effects
indirect1 := a1 * b1 # MAIN INDIRECT EFFECT
indirect2 := a2 * b1
indirect3 := a3 * b1
indirect4 := a4 * b1
indirect5 := a5 * b1
indirect6 := a6 * b1
indirect7 := a7 * b1
indirect8 := a8 * b1
indirect9 := a9 * b1

# total effects
total1 := c1 + (a1 * b1)  
total2 := c1 + c2 + c3 + c4 + c5 + c6 + c7 + c8 + c9 + b1*(a1 + a2 + a3 + a4 + a5 + a6 + a7 + a8 + a9)
'
```

**Run Model**
```{r}
set.seed(38974)

Med_N <- sem(model = paste(micro.model_N, "\n", N_Mediation_Mod),
            data = N_3m1y_asthm_traj_sdf, ordered = c("asthma_3y")) 

summary(Med_N, fit.measures = TRUE)
```

**Get standardized estimates and other Model results**
```{r}
Med_N_stdEst <- parameterEstimates(Med_N, boot.ci.type = "bca.simple", level = .95, ci = TRUE, standardized = TRUE)
```

## GUT Mediation Model (Figure S7)
**Model**
```{r}
G_Mediation_Mod <- ' 
    asthma_3y ~ b1 * micro_G + c1 * BF_status_3m_sampl_num + c2 * coldsnum_0to3m + c3 * older_sibs_living + c4 * atbx_bin + c5 * prenatal_smoke_num + c6 * birthmode2_num + c7 * mom_asthma_num + c8 * Toronto +  c9 * RE_Virus_1y_num
    
    micro_G ~  a1 * BF_status_3m_sampl_num + a2 * coldsnum_0to3m + a3 * older_sibs_living + a4* atbx_bin + a5 * prenatal_smoke_num + a6 * birthmode2_num + a7 * mom_asthma_num + a8* Toronto +  a9 * RE_Virus_1y_num
    
#indirect effects
indirect1 := a1 * b1 # MAIN INDIRECT EFFECT
indirect2 := a2 * b1
indirect3 := a3 * b1
indirect4 := a4 * b1
indirect5 := a5 * b1
indirect6 := a6 * b1
indirect7 := a7 * b1
indirect8 := a8 * b1
indirect9 := a9 * b1

# total effects
total1 := c1 + (a1 * b1)  
total2 := c1 + c2 + c3 + c4 + c5 + c6 + c7 + c8 + c9 + b1*(a1 + a2 + a3 + a4 + a5 + a6 + a7 + a8 + a9) 
'
```

**Run Model**
```{r}
set.seed(38974)

Med_G <- sem(model = paste(micro.model_G, "\n", G_Mediation_Mod),
            data = G_3m1y_asthm_traj_sdf, ordered = c("asthma_3y")) 

summary(Med_G, fit.measures = TRUE)
```

**Get standardized estimates and other Model results**
```{r}
Med_G_stdEst <- parameterEstimates(Med_G, boot.ci.type = "bca.simple", level = .95, ci = TRUE, standardized = TRUE)
```

## Merge results for Table S5, these results are also used to generate Figure 6 & Figure S7
```{r}
Med_NG_stdEst$Model <- "Main Nasal-Gut model"

Med_N_stdEst$Model <- "Main Nasal model"

Med_G_stdEst$Model <- "Main Gut model"

Update_Combined_SEM_models <- rbind(Med_NG_stdEst, Med_N_stdEst, Med_G_stdEst) # ordered main models first, then extra models

# Terms of interest: Anything with a label, Anything without a label that has a "=~" operation indicating the measurment model, and the only covariance of interest is for the nasal/gut combined model, micro_N	~~	micro_G. 

## SO - Keep only terms of interest in Table S5 (mostly just those with a label)
Combined_SEM_models_V2 <- subset(Update_Combined_SEM_models, label != "" | (label == "" & op == "=~" ) | (label == "" & lhs == "micro_N" & rhs == "micro_G")) # REMOVE either if label is not present, OR if label is Not Present and op == =~ (latent constructs), then Keep; OR keep if label is not present and lhs == "micro_N" & rhs == "micro_G"

Total_mods <- c("c1+c2+c3+c4+c5+c6+c7+c8+c9+b1*(a1+a2+a3+a4+a5+a6+a7+a8+a9)", "c1+c2+c3+c4+c5+c6+c7+c8+c9+b1*(a1+d2+d3+d4+d5+d6+d7+d8+d9)+b2*(a2+e2+e3+e4+e5+e6+e7+e8+e9)")
Combined_SEM_models_V2$rhs <- ifelse(Combined_SEM_models_V2$rhs %in% Total_mods,
                                     "Total model", Combined_SEM_models_V2$rhs)

Combined_SEM_models_V3 <- data.frame(Combined_SEM_models_V2[1:4], round(Combined_SEM_models_V2[5:13], 3), Model=Combined_SEM_models_V2$Model)

library(knitr)
kable(Combined_SEM_models_V3)

write.csv(Combined_SEM_models_V3, file="Output/TableS5_SEM_Model_Res_Fig6_FigS7.csv")
```

#### Extra check of measurement model / nasal and gut latent constructs using confirmatory factor analysis (CFA)
CFI, RMSEA and SRMR are all okay
```{r}
NG_measurment_mod <- cfa(model = paste(micro.model_N,"\n", micro.model_G), data=NG_3m1y_asthm_traj_sdf_noNA)

summary(NG_measurment_mod, fit.measures=TRUE)
```

