---
title: "Additional Data Processing to create Subsets and Derived Variables for Analyses"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Setup

```{r}
R.version.string
library(phyloseq); packageVersion('phyloseq')
library(tidyr); packageVersion('tidyr')
library(tidyverse); packageVersion('tidyverse')
library(ggplot2); packageVersion('ggplot2')
library(reshape2); packageVersion('reshape2')

source("Preprocessing/Functions.R")
```

```{r} 
## used for abundance analyses
UnRarefy <- readRDS("Preprocessing/final_datasets/02_UnRarefied_Merged_Milk_Nasal_Gut.rds")

## used for prevalence-based analyses
Rarefy <- readRDS(file="Preprocessing/final_datasets/02_Rarefied_Merged_Milk_Nasal_Gut.rds")
```

```{r, include=FALSE} 
# post-hoc additions - edit some variables used in analyses of non-rarefied data

sdf <- as(sample_data(UnRarefy), "data.frame")

sdf$Type_visit <- paste(sdf$SampleType, sdf$Visit)
sdf$Fever_3m <- ifelse(sdf$Fever_3m == 0, "No", ifelse(sdf$Fever_3m == 1, "Yes", NA))
sdf$Fever_1y <- ifelse(sdf$Fever_1y == 0, "No", ifelse(sdf$Fever_1y == 1, "Yes", NA))


# need to change factor levels based on which factor we want as reference for downstream models for the factors/characters in this list:
Main_vars <- c("BF_status_3m_sampling", "BF_12m_sampling","RE_Virus", "colds_0to3m", "coldsnum_6to12m","Fever_3m", "Fever_1y", "older_sibs", "AwayfromHome_1y", "StudyCenter", "Season_3m" , "Season_1y", "atbx", "Mother_abs_birth_yn", "prenatal_smoke","mom_smoke_1y", "mom_bmi_best", "birthmode2", "mom_age_birth", "race_mom4", "mom_asthma_ever", "mom_atopy",  "age_3m_months", "age_1y_months", "weight_3", "weight_12", "childgender", "gest_days") 


# split numeric and categorical variables to edit categorical variables, then re-merge with numeric
sdf_3m2_num <- sdf %>% select_if(is.numeric) 
sdf_3m2_cat <- subset(sdf, select = setdiff(Main_vars, colnames(sdf_3m2_num))) # Factors to modify levels for; in Main_vars but not numeric (factor or character)

sdf_other <- subset(sdf, select = setdiff(colnames(sdf), colnames(sdf_3m2_cat))) # DF of Metadata not being edited

# check that all cols are accounted for in two subsets
sum(dim(sdf_other)[2] + dim(sdf_3m2_cat)[2]) == dim(sdf)[2] # TRUE, good

# make all characters factors 
sdf_3m2_cat[] <- lapply(sdf_3m2_cat[], factor)
lapply(sdf_3m2_cat[], levels) # better to swap factor level for most of these... Nearly all need to be reversed

# Want all factor levels to be reversed - this way the "no" group will generally be the last group (ref for lmp is last level)
sdf_3m2_cat[] <- lapply(sdf_3m2_cat[], function(x) {factor(x, levels = rev(levels(x)))})

## A few exceptions to the rule, want a different order for these factor levels
sdf_3m2_cat$BF_status_3m_sampling <- factor(sdf_3m2_cat$BF_status_3m_sampling, levels = c("Exclusive", "Partial", "No BM"))
sdf_3m2_cat$BF_12m_sampling <- factor(sdf_3m2_cat$BF_12m_sampling, levels = c("Yes", "No")) 
sdf_3m2_cat$birthmode2 <- factor(sdf_3m2_cat$birthmode2, levels = c("Caesarean", "Vaginal"))
sdf_3m2_cat$race_mom4 <- factor(sdf_3m2_cat$race_mom4, levels = c("Asian", "FN", "Other", "Caucasian"))
sdf_3m2_cat$StudyCenter <- factor(sdf_3m2_cat$StudyCenter, levels = c("Edmonton", "Toronto", "Vancouver", "Winnipeg"))
sdf_3m2_cat$Season_3m <- factor(sdf_3m2_cat$Season_3m, levels = c("Spring", "Summer", "Fall", "Winter"))
sdf_3m2_cat$Season_1y <- factor(sdf_3m2_cat$Season_1y, levels = c("Spring", "Summer", "Fall", "Winter"))

lapply(as.list(sdf_3m2_cat), levels) # Looks much better

# re-merge data using data.frame function - is maintained... just double check rownames are the same in both first!
identical(rownames(sdf_3m2_cat), rownames(sdf_other))
sdf2 <- data.frame(sdf_3m2_cat, sdf_other)


# re-save sample data to phyloseq object
sample_data(UnRarefy) <- sdf2
```

```{r, include=FALSE} 
# post-hoc additions - edit genus-level taxonomy
# For agglomerating to genus level, there is a small issue with taxonomy where the species level is in the genus level category for some contested taxa with hard brackets - this only affects Genus level plots - Fig1 C & D - won't matter for other analyses - so just edit for this non-rarefied dataset)

taxonomy <- as.data.frame(as(tax_table(UnRarefy), "matrix"))

# two that will be tricky, group into appropriate genera one by one for these two
taxonomy$Genus <- taxonomy$Genus %>% gsub("Unclassified \\[Clostridium\\]_methylpentosum_group", "[Clostridium]", .) %>% 
                  gsub("Unclassified \\[Eubacterium\\]_coprostanoligenes_group", "[Eubacterium]", .)

# remove everything after the space for the rest of taxa with hard brackets (this is the species name)
taxonomy$Hard_bracket_flag <- grepl("\\[", taxonomy$Genus)
taxonomy$Genus <- ifelse(taxonomy$Hard_bracket_flag == TRUE, gsub(" .*", "", taxonomy$Genus), taxonomy$Genus)

# rm hard bracket flag & save taxonomy table back to phyloseq obj
taxonomy$Hard_bracket_flag <- NULL

tax_table(UnRarefy) <- as.matrix(taxonomy)
```


# Sample filters for Non-Rarefied Nasal-Gut dataset

### Remove infants without breastfeeding status information at the time of 3 month sample collection and only keep relevant time-points & sample types
```{r}
#  (exclude viral samples, 3 year and 5 year samples)
Main_data <- subset_samples(UnRarefy, (Name == "Nasal Swab Aliquot" | SampleType == "Gut") & 
             !is.na(BF_status_3m_sampling) & ( SampleType == "Nasal" | SampleType == "Gut" | BF_status_3m_MS_sampling_exact != "No BM") &
             Visit != "3 Years" & Visit != "5 Years") # includes filter that IF sample is a milk sample AND breastfeeding status and milk sampling is "No BM", then exclude.

table(sample_data(Main_data)$Visit, sample_data(Main_data)$SampleType)
```

### Assess infant ages at sampling at each time-point and remove major outliers to avoid overlap between timepoints
```{r, warning=FALSE}
ggplot(sample_data(Main_data), aes(x=Type_visit, y=Age_sampling)) + geom_boxplot() + 
  labs(y="Age at sampling (months)", x="") + geom_hline(yintercept = 8) + geom_hline(yintercept = 21) + 
  geom_hline(yintercept = 72) + scale_y_continuous(breaks = c(0,10,20,30,40,50,60,70)) + 
  scale_x_discrete(labels = c("3 Months Milk" = "Milk","3 Months Nasal" = "3 Months", "1 Year Nasal" = "1 Year","5 Years Nasal" = "5 Years")) 
```

**Remove outliers for age at sampling, important particularly for longitudinal/paired analyses**
```{r}
# check subset if using BF at sampling as filter (done later but should have done here)
Main_data2 <- subset_samples(Main_data, (Age_sampling < 8 & Age_sampling > 1 & Visit == "3 Months") | (Age_sampling < 21 & Age_sampling > 8 & Visit == "1 Year")) 
```

**Remove infants never ever breastfed, even in hospital**
```{r}
Main_data3 <- subset_samples(Main_data2, BF_status_3m_sampling2 != "Never breastfed") 
```

```{r}
data_NG2 <- prune_taxa(taxa_sums(Main_data3)>0, Main_data3) # removing any taxa not present in this subset 
data_NG2

table(sample_data(data_NG2)$Visit, sample_data(data_NG2)$SampleType)

saveRDS(data_NG2, file = "Data_subsets/Main_UnRarefied_NasalGut_Subset.rds")
```



# Sample filters for Main Rarefied Nasal-Gut dataset (Used for Prevalence based analyses, including Trajectory analyses)
Use the same sample filters as above for rarefied data

### Remove infants without breastfeeding status information at the time of 3 month sample collection and only keep relevant time-points & sample types
```{r}
Main_data <- subset_samples(Rarefy, (Name == "Nasal Swab Aliquot" | SampleType == "Gut") & 
             !is.na(BF_status_3m_sampling) & ( SampleType == "Nasal" | SampleType == "Gut" | BF_status_3m_MS_sampling_exact != "No BM") &
             Visit != "3 Years" & Visit != "5 Years") 
```


```{r} 
# Exclude Age at sample collection outliers
Main_data_1 <- subset_samples(Main_data, (Age_sampling < 8 & Age_sampling > 1 & Visit == "3 Months") | (Age_sampling < 21 & Age_sampling > 8  & Visit == "1 Year")) 

# Excludes infants never breastfed
Main_data_2 <- subset_samples(Main_data_1, BF_status_3m_sampling2 != "Never breastfed") # previously named Main_data_NMG2

Main_data_2 <- prune_taxa(taxa_sums(Main_data_2)>0, Main_data_2) # removing any taxa not present in this subset 


table(sample_data(Main_data_2)$Visit, sample_data(Main_data_2)$SampleType)

saveRDS(Main_data_2, file = "Data_subsets/Main_Rarefied_NasalGut_Subset.rds")
```



# Microbiota Trajectory Dataset & Derived Trajectory Variables

## Datasets for 3m1y prevalence trajectory analyses

### Nasal data
```{r} 
# Common Subject IDs between milk and nasal samples at 3 months
Common_3m1yN <- intersect(sample_data(subset_samples(Main_data_2, Visit == "3 Months" & SampleType == "Nasal"))$CHILDid,
                          sample_data(subset_samples(Main_data_2, Visit == "1 Year"  & SampleType == "Nasal"))$CHILDid)
length(Common_3m1yN) # number of infants w/t both 3M and 1Y sample

# Subset Samples for these 1545 infants
Nasal_3m1y <- subset_samples(Main_data_2, SampleType == "Nasal" & (Visit == "3 Months" | Visit == "1 Year") & CHILDid %in% Common_3m1yN) # 1 year nasal

# Taxa wanted for prevalence trajectory analyses - present in >10% of samples at EITHER 3 Months or 1 Year 
nasal_3m_tax_filt = filter_taxa(subset_samples(Nasal_3m1y, Visit =="3 Months"), function(x) sum(x > 0) > (0.10*length(x)), TRUE)
nasal_1y_tax_filt = filter_taxa(subset_samples(Nasal_3m1y, Visit =="1 Year"), function(x) sum(x > 0) > (0.10*length(x)), TRUE)

Nasal_3M1Y_all_filt <- prune_taxa(c(get_taxa_unique(nasal_3m_tax_filt, taxonomic.rank = "ASV_taxonomy"), 
                                    get_taxa_unique(nasal_1y_tax_filt, taxonomic.rank = "ASV_taxonomy")), Nasal_3m1y) # NOTE that there is overlap in taxa b/t these 2 lists so the number in the final DF will be less

Nasal_3M1Y_pa_filt <- transform_sample_counts(Nasal_3M1Y_all_filt, function(abund) 1*(abund>0))
Nasal_3M1Y_pa_filt # 171 taxa & 3090 Samples

saveRDS(Nasal_3M1Y_pa_filt, file = "Data_subsets/Main_Nasal_Trajectory_Subset.rds")
```

### Gut data
```{r} 
# Common Subject IDs between milk and nasal samples at 3 months
Common_3m1yG <- intersect(sample_data(subset_samples(Main_data_2, Visit == "3 Months" & SampleType == "Gut"))$CHILDid,
                          sample_data(subset_samples(Main_data_2, Visit == "1 Year"  & SampleType == "Gut"))$CHILDid)
length(Common_3m1yG) # number of infants w/t both 3M and 1Y sample

Gut_3m1y <- subset_samples(Main_data_2, SampleType == "Gut" & (Visit == "3 Months" | Visit == "1 Year") & CHILDid %in% Common_3m1yG) 
Gut_3m1y 

# Taxa wanted for prevalence trajectory analyses - present in >10% of samples at EITHER 3 Months or 1 Year 
Gut_3m_tax_filt = filter_taxa(subset_samples(Gut_3m1y, Visit =="3 Months"), function(x) sum(x > 0) > (0.10*length(x)), TRUE)
Gut_1y_tax_filt = filter_taxa(subset_samples(Gut_3m1y, Visit =="1 Year"), function(x) sum(x > 0) > (0.10*length(x)), TRUE)

Gut_3M1Y_all_filt <- prune_taxa(c(get_taxa_unique(Gut_3m_tax_filt, taxonomic.rank = "ASV_taxonomy"), 
                                  get_taxa_unique(Gut_1y_tax_filt, taxonomic.rank = "ASV_taxonomy")), Gut_3m1y) 

Gut_3M1Y_pa_filt <- transform_sample_counts(Gut_3M1Y_all_filt, function(abund) 1*(abund>0))
Gut_3M1Y_pa_filt 

saveRDS(Gut_3M1Y_pa_filt, file = "Data_subsets/Main_Gut_Trajectory_Subset.rds")
```


# Derived Variables for Various Trajectory analyses 

## Create Colonization Pattern feature tables (1 = only 3 month, 2 = persistent, 3 = late, 4 = absent)
Used in stacked bar plots in Figure 4B & 4C; and to derive per-infant colonization pattern summaries (proporition of taxa that are early, late, etc.)

### Nasal data
```{r}
N.pa.3m <- subset_samples(Nasal_3M1Y_pa_filt, Visit == "3 Months")
N.pa.1y <- subset_samples(Nasal_3M1Y_pa_filt, Visit == "1 Year")

## extract otu table and add Participant IDs 
N_3m_otu <- data.frame(as.data.frame(t(as(otu_table(N.pa.3m), "matrix"))), CHILDid=sample_data(N.pa.3m)$CHILDid)
N_1y_otu <- data.frame(as.data.frame(t(as(otu_table(N.pa.1y), "matrix"))), CHILDid=sample_data(N.pa.1y)$CHILDid)

## don't like how data.frame function changed the column names, changing back
colnames(N_3m_otu)[1:ntaxa(Nasal_3M1Y_pa_filt)] <- row.names(otu_table(N.pa.3m))
colnames(N_1y_otu)[1:ntaxa(Nasal_3M1Y_pa_filt)] <- row.names(otu_table(N.pa.1y))

N_3m_otu <- N_3m_otu[order(N_3m_otu$CHILDid),]
N_1y_otu <- N_1y_otu[order(N_1y_otu$CHILDid),]

identical(N_3m_otu$CHILDid, N_1y_otu$CHILDid) # good (doesn't really matter for this analysis though)
```

```{r}
N_3m_otu_sdf_m <- melt(N_3m_otu, id.vars = "CHILDid", value.name = "Present_nasal_3m")
N_1y_otu_sdf_m <- melt(N_1y_otu, id.vars = "CHILDid", value.name = "Present_nasal_1y")

N_3m1y_otu_sdf_m <- merge(N_3m_otu_sdf_m, N_1y_otu_sdf_m, by= c("CHILDid", "variable"))
nrow(N_3m1y_otu_sdf_m) == nrow(N_1y_otu_sdf_m) # True, good

# Create Colonization patterns
N_3m1y_otu_sdf_m$Nasal_colonization <- ifelse(N_3m1y_otu_sdf_m$Present_nasal_3m == 1 & N_3m1y_otu_sdf_m$Present_nasal_1y == 1, 2,
                                              ifelse(N_3m1y_otu_sdf_m$Present_nasal_3m == 1 & N_3m1y_otu_sdf_m$Present_nasal_1y == 0, 1,
                                                     ifelse(N_3m1y_otu_sdf_m$Present_nasal_3m == 0 & N_3m1y_otu_sdf_m$Present_nasal_1y == 1, 3, 0))) 
# CODE: 0=absent, 1=early only, 2 = persistent, 3=late,  only 4 possible options

table(N_3m1y_otu_sdf_m$Nasal_colonization)

N3M1Y_colonize <- dcast(N_3m1y_otu_sdf_m, CHILDid~variable, value.var = "Nasal_colonization", mean)
dim(N3M1Y_colonize)  # 1545 infants with both time-points, and  171 OTUs 

# N3M1Y_colonize = The nasal colonization pattern feature table. 
# N_3m1y_otu_sdf_m = use to create per participant summaries (proportion of early, late, etc. per-infant)
```

### Gut data
```{r}
G.pa.3m <- subset_samples(Gut_3M1Y_pa_filt, Visit == "3 Months")
G.pa.1y <- subset_samples(Gut_3M1Y_pa_filt, Visit == "1 Year")

## extract otu table and add Participant IDs 
G_3m_otu <- data.frame(as.data.frame(t(as(otu_table(G.pa.3m), "matrix"))), CHILDid=sample_data(G.pa.3m)$CHILDid)
G_1y_otu <- data.frame(as.data.frame(t(as(otu_table(G.pa.1y), "matrix"))), CHILDid=sample_data(G.pa.1y)$CHILDid)

## don't like how data.frame function changed the column names, changing back
colnames(G_3m_otu)[1:ntaxa(Gut_3M1Y_pa_filt)] <- row.names(otu_table(G.pa.3m))
colnames(G_1y_otu)[1:ntaxa(Gut_3M1Y_pa_filt)] <- row.names(otu_table(G.pa.1y))


G_3m_otu <- G_3m_otu[order(G_3m_otu$CHILDid),]
G_1y_otu <- G_1y_otu[order(G_1y_otu$CHILDid),]

identical(G_3m_otu$CHILDid, G_1y_otu$CHILDid) # good (doesn't really matter for this analysis though)
```

```{r}
G_3m_otu_sdf_m <- melt(G_3m_otu, id.vars = "CHILDid", value.name = "Present_gut_3m")
G_1y_otu_sdf_m <- melt(G_1y_otu, id.vars = "CHILDid", value.name = "Present_gut_1y")

G_3m1y_otu_sdf_m <- merge(G_3m_otu_sdf_m, G_1y_otu_sdf_m, by= c("CHILDid", "variable"))
nrow(G_3m1y_otu_sdf_m) == nrow(G_3m_otu_sdf_m) # True, good

# Create Colonization patterns
G_3m1y_otu_sdf_m$Gut_colonization <- ifelse(G_3m1y_otu_sdf_m$Present_gut_3m == 1 & G_3m1y_otu_sdf_m$Present_gut_1y == 1, 2,
                                            ifelse(G_3m1y_otu_sdf_m$Present_gut_3m == 1 & G_3m1y_otu_sdf_m$Present_gut_1y == 0, 1,
                                                   ifelse(G_3m1y_otu_sdf_m$Present_gut_3m == 0 & G_3m1y_otu_sdf_m$Present_gut_1y == 1, 3, 0))) 
# CODE: 0=absent, 1=early only, 2 = persistent, 3=late,  only 4 possible options

G3M1Y_colonize <- dcast(G_3m1y_otu_sdf_m, CHILDid~variable, value.var = "Gut_colonization", mean)
dim(G3M1Y_colonize)  # 555 infants with both time-points, 115 OTUs

# G3M1Y_colonize =  The gut colonization pattern feature table
# G_3m1y_otu_sdf_m = use to create per participant summaries (proportion of early, late, etc. per-infant)
```

## Summarized Colonization patterns
Calculate for each Infant, the proportion of taxa that are Early (only 3 month), Persistent (3 month and 1 year), Late (only 1 year) or Absent (not at either time-point)  

### Nasal data
```{r, warning=FALSE}
## Make separate columns for Early, Late, etc colonizers (Binary, 1 or 0 for yes or no) 
N_3m1y_otu_sdf_m$Early <- ifelse(N_3m1y_otu_sdf_m$Nasal_colonization== 1, 1, 0)
N_3m1y_otu_sdf_m$Persistent <- ifelse(N_3m1y_otu_sdf_m$Nasal_colonization== 2, 1, 0)
N_3m1y_otu_sdf_m$Late <- ifelse(N_3m1y_otu_sdf_m$Nasal_colonization== 3, 1, 0)
N_3m1y_otu_sdf_m$Absent <- ifelse(N_3m1y_otu_sdf_m$Nasal_colonization== 0, 1, 0)

## Add another melt level to get column differentiating Early/persistent/late
N_3m1y_otu_sdf_m2 <- melt(N_3m1y_otu_sdf_m, measure.vars = c("Early", "Persistent", "Late", "Absent"), variable.name = "Colonization_Pattern")
### "value" is a binary 1, 0 

# Sum number of "1's" for each colonization pattern and each CHILDid - To calculate for each participant - the number of early, late, etc. colonizers.
N_Sum_Patterns_perInfant <- data_sum(N_3m1y_otu_sdf_m2, varname = "value", groupnames = c("CHILDid", "Colonization_Pattern"))

## quick check
nrow(N_Sum_Patterns_perInfant)/4 == nrow(N3M1Y_colonize) # true, good

# The proportion of OTUs will actually be essentially the same variable as the number of OTUs in each of these categories since it's divided by the total number of ASVs in the dataset. I prefer this to using the total number of OTUs a participant has at both time-points since that adds weight to another component (Observed ASVs) other than colonization patterns into the analysis - I would rather look directly at colonization patterns.
N_Sum_Patterns_perInfant$Pct_OTUs_Colonize <- N_Sum_Patterns_perInfant$sum*100/ntaxa(Nasal_3M1Y_pa_filt)

### re-structure data
N_PerInfant_coloniz <- dcast(N_Sum_Patterns_perInfant, CHILDid~Colonization_Pattern, value.var = "Pct_OTUs_Colonize", mean)
dim(N_PerInfant_coloniz)

## Extra - Side-note - can also summarize by taxa - not used in any analyses here but could be used maybe
N_Sum_Patterns_perTaxa <- data_sum(N_3m1y_otu_sdf_m2, varname = "value", groupnames = c("variable", "Colonization_Pattern"))
N_Sum_Patterns_perTaxa$Pct_Taxa_Colonize_Infant <- N_Sum_Patterns_perTaxa$sum*100/nrow(G3M1Y_colonize)
N_PerTaxa_coloniz <- dcast(N_Sum_Patterns_perTaxa, variable~Colonization_Pattern, value.var = "Pct_Taxa_Colonize_Infant", mean)
### write.csv(N_PerTaxa_coloniz, file="Output/Nasal_perTaxa_PctColonizeInfant.csv")
```
N_PerInfant_coloniz, to merge with colonication feature table

### Gut data
```{r, warning=FALSE}

## Make separate columns for Early, Late, etc colonizers (Binary, 1 or 0 for yes or no) 
G_3m1y_otu_sdf_m$Early <- ifelse(G_3m1y_otu_sdf_m$Gut_colonization== 1, 1, 0)
G_3m1y_otu_sdf_m$Persistent <- ifelse(G_3m1y_otu_sdf_m$Gut_colonization== 2, 1, 0)
G_3m1y_otu_sdf_m$Late <- ifelse(G_3m1y_otu_sdf_m$Gut_colonization== 3, 1, 0)
G_3m1y_otu_sdf_m$Absent <- ifelse(G_3m1y_otu_sdf_m$Gut_colonization== 0, 1, 0)

## Add another melt level to get column differentiating Early/persistent/late
G_3m1y_otu_sdf_m2 <- melt(G_3m1y_otu_sdf_m, measure.vars = c("Early", "Persistent", "Late", "Absent"), variable.name = "Colonization_Pattern")
### "value" is a binary 1, 0 

# Sum number of "1's" for each colonization pattern and each CHILDid - To calculate for each participant - the number of early, late, etc. colonizers.
G_Sum_Patterns_perInfant <- data_sum(G_3m1y_otu_sdf_m2, varname = "value", groupnames = c("CHILDid", "Colonization_Pattern"))

## quick check
nrow(G_Sum_Patterns_perInfant)/4 == nrow(G3M1Y_colonize) # true, good

# The proportion of OTUs will actually be essentially the same variable as the number of OTUs in each of these categories since it's divided by the total number of ASVs in the dataset. I prefer this to using the total number of OTUs a participant has at both time-points since that adds weight to another component (Observed ASVs) other than colonization patterns into the analysis - I would rather look directly at colonization patterns.
G_Sum_Patterns_perInfant$Pct_OTUs_Colonize <- G_Sum_Patterns_perInfant$sum*100/ntaxa(Gut_3M1Y_pa_filt)

### re-structure data
G_PerInfant_coloniz <- dcast(G_Sum_Patterns_perInfant, CHILDid~Colonization_Pattern, value.var = "Pct_OTUs_Colonize", mean)
dim(G_PerInfant_coloniz)

## Extra - Side-note - can also summarize by taxa - not used in any analyses here but could be used maybe
G_Sum_Patterns_perTaxa <- data_sum(G_3m1y_otu_sdf_m2, varname = "value", groupnames = c("variable", "Colonization_Pattern"))
G_Sum_Patterns_perTaxa$Pct_Taxa_Colonize_Infant <- G_Sum_Patterns_perTaxa$sum*100/nrow(G3M1Y_colonize)
G_PerTaxa_coloniz <- dcast(G_Sum_Patterns_perTaxa, variable~Colonization_Pattern, value.var = "Pct_Taxa_Colonize_Infant", mean)
### write.csv(G_PerTaxa_coloniz, file="Output/Gut_perTaxa_PctColonizeInfant.csv")
```
G_PerInfant_coloniz, to merge with colonization feature table

### Merge Per-Infant Summarized Colonization patterns with Colonization pattern feature table
```{r, warning=FALSE}
N3M1Y_colonize_all <- merge(N_PerInfant_coloniz, N3M1Y_colonize, by="CHILDid")

G3M1Y_colonize_all <- merge(G_PerInfant_coloniz, G3M1Y_colonize, by="CHILDid")
```
N3M1Y_colonize_all, G3M1Y_colonize_all, to merge with sample metadata / diversity trajectory dataset downstream


## Make sample data used to assess change in diversity and colonization patterns 

#### Main variables to include
```{r} 
# Keeping only main variables that differ b/t 3M and 1Y samples in Trajectory data 
Traj_vars <- c("CHILDid", "Shannon", "Observed", "RE_Virus") 
```

#### Nasal data (n=1545)
```{r} 
# Overall 3 mnth and 1 year sample data (already split up)
sdf_1y_n <- data.frame( subset(sample_data(N.pa.1y), select = Traj_vars), 
                        row.names = sample_data(N.pa.1y)$CHILDid ) 
sdf_3m_n <- data.frame( subset(sample_data(N.pa.3m), select = Traj_vars), 
                        row.names = sample_data(N.pa.3m)$CHILDid )

sdf_N <- merge(sdf_3m_n, sdf_1y_n, by="CHILDid", suffixes = c("_N3m", "_N1y")) # suffices will modify diversity names to indicate sample type and timepoint
dim(sdf_N)

## extra, just want to rename RE virus
colnames(sdf_N)[names(sdf_N) == "RE_Virus_N1y"] <- "RE_Virus_1y"
colnames(sdf_N)[names(sdf_N) == "RE_Virus_N3m"] <- "RE_Virus_3m"
```

#### Gut data (n=555)
```{r} 
# Overall 3 mnth and 1 year sample data (already split up)
sdf_1y_g <- data.frame( subset(sample_data(G.pa.1y), select = Traj_vars), 
                        row.names = sample_data(G.pa.1y)$CHILDid ) 
sdf_3m_g <- data.frame( subset(sample_data(G.pa.3m), select = Traj_vars), 
                        row.names = sample_data(G.pa.3m)$CHILDid )

sdf_G <- merge(sdf_3m_g, sdf_1y_g, by="CHILDid", suffixes = c("_G3m", "_G1y")) # suffices will modify diversity names to indicate sample type and timepoint
dim(sdf_G)

## extra, just want to rename RE virus
colnames(sdf_G)[names(sdf_G) == "RE_Virus_G1y"] <- "RE_Virus_1y"
colnames(sdf_G)[names(sdf_G) == "RE_Virus_G3m"] <- "RE_Virus_3m"
```

### Calculate Change in diversity b/t 3 months and 1 year
```{r} 
sdf_N <- sdf_N %>% mutate(Shannon_N1y3m = Shannon_N1y - Shannon_N3m) %>% 
                   mutate(Observed_N1y3m = Observed_N1y - Observed_N3m)

sdf_G <- sdf_G %>% mutate(Shannon_G1y3m = Shannon_G1y - Shannon_G3m) %>% 
                   mutate(Observed_G1y3m = Observed_G1y - Observed_G3m)
```

## Merge Colonization pattern data with Diversity Trajectory data
```{r}
# Nasal data
N_3m1y_otu_sdf2 <- merge(sdf_N, N3M1Y_colonize_all, by="CHILDid") # merging with sample data
dim(N_3m1y_otu_sdf2)

# Gut data
G_3m1y_otu_sdf2 <- merge(sdf_G, G3M1Y_colonize_all, by="CHILDid") # merging with sample data
dim(G_3m1y_otu_sdf2)
```


## Total Trajectory Dataset - Combine Nasal and Gut data
```{r} 
length(intersect(G_3m1y_otu_sdf2$CHILDid, N_3m1y_otu_sdf2$CHILDid)) # 393 infants with gut-nasal at both time-points 

intersect(colnames(N_3m1y_otu_sdf2), colnames(G_3m1y_otu_sdf2)) # Other than CHILDid, this is the list of variables that will have the suffixes in the below merge script appended to them because they are present in BOTH the nasal and gut data

## Merge nasal and gut datasets
GN_3m1y_all_traj <- merge(N_3m1y_otu_sdf2, G_3m1y_otu_sdf2, 
                           by="CHILDid", all = TRUE, suffixes = c("_N1y3m", "_G1y3m") )

# Make CHILDid's rownames instead
row.names(GN_3m1y_all_traj) <- GN_3m1y_all_traj$CHILDid
GN_3m1y_all_traj$CHILDid <- NULL

# Remove RE Virus variables (will add to Sample Meatadata linking with this dataset instead, too complicated to have 4 versions)
GN_3m1y_all_traj$RE_Virus_1y_G1y3m <- NULL
GN_3m1y_all_traj$RE_Virus_3m_G1y3m <- NULL
GN_3m1y_all_traj$RE_Virus_1y_N1y3m <- NULL
GN_3m1y_all_traj$RE_Virus_3m_N1y3m <- NULL
dim(GN_3m1y_all_traj) # N= 1707 - 1545 + 555 - 393 (this is correct N) - this is the trajectory data to be used

# colnames(G_3m1y_otu_sdf2) ## Gut Traj's start at Shannon_G3m; at column number 182
which(names(GN_3m1y_all_traj) == "Shannon_G3m") # at line 182
```
Compiled Nasal/Gut Trajectory data: GN_3m1y_all_traj


### Create Sample Metadata Subset that Links with the Trajectory Dataset

**Main variables to include**
```{r} 
# Keeping only main variables in dataset (won't use all of these but keeping anyway just in case). Doesn't include "Shannon", "Observed", "RE_Virus", will add these from trajectory datasets
Main_vars <- c("CHILDid", "WheezeTraj", "asthma_3y", "asthma_5y", "atopy3y", "BM_mode_3m", "BF_status_3m_sampling", "BF_12m_sampling", "older_sibs", "older_sibs_living", "AwayfromHome_1y", "childgender", "atbx", "Mom_abs_3mo_IV_oral_supp_yn", "Mother_abs_birth_yn", "StudyCenter", "birthmode2", "mom_bmi_best", "mom_age_birth", "Season_1y", "Season_3m", "age_1y_months", "age_3m_months", "prenatal_smoke", "gest_days", "weight_0", "weight_3","weight_12", "race_mom4", "mom_asthma_ever", "mom_atopy", "colds_0to3m", "Fever_6m", "Fever_3m", "Fever_1y", "Solids_3m_sampling", "Smoked_AtHome3m", "Furry_Pets_3m", "coldsnum_0to3m", "coldsnum_6to12m", "mom_smoke_1y", "BF_duration_imp") 

### Use Participant-Level metadata (not longitudinal version) - this is easier for trajectory data
CHILD_metadata_all <- readRDS(file="Preprocessing/metadata_input/3FINAL_Full_Cleaned_KF_Modified_Updated-05222021_withSprio") 

CHILD_metadata_sub <- subset(CHILD_metadata_all, select = Main_vars)
```

**Additional edits and merge**
```{r} 
### Also Adding RE Virus from Traj datasets (participant level data, not longitudinal - so separate vars for 3m and 1y)
RE_virus_df <- merge(subset(N_3m1y_otu_sdf2, select = c("CHILDid", "RE_Virus_3m", "RE_Virus_1y")),
                     subset(G_3m1y_otu_sdf2, select = c("CHILDid", "RE_Virus_3m", "RE_Virus_1y")), 
                     by="CHILDid", all=TRUE)
#### Need to merge into one - some RE virus values missing from nasal data, some missing from gut data, need to consolidate...
RE_virus_df$RE_Virus_3m <- ifelse(is.na(RE_virus_df$RE_Virus_3m.x), as.character(RE_virus_df$RE_Virus_3m.y), as.character(RE_virus_df$RE_Virus_3m.x))
RE_virus_df$RE_Virus_1y <- ifelse(is.na(RE_virus_df$RE_Virus_1y.x), as.character(RE_virus_df$RE_Virus_1y.y), as.character(RE_virus_df$RE_Virus_1y.x))

# Merge metadata
CHILD_metadata_all2 <- merge(CHILD_metadata_sub, 
                             subset(RE_virus_df, select = c("CHILDid", "RE_Virus_3m", "RE_Virus_1y")), 
                             by="CHILDid", all.x = TRUE) 
row.names(CHILD_metadata_all2) <- CHILD_metadata_all2$CHILDid
```

```{r} 
# ADDED MODIFICATIONS to sample metadata 

## Creating Variable for Asthma at 3 years (Probable) that Persists to 5 years (at least possible or probable at 5 years)
CHILD_metadata_all2$asthma_3y_5y <- ifelse(CHILD_metadata_all2$asthma_3y == "Probable" & 
                                            CHILD_metadata_all2$asthma_5y %in% c("Possible", "Probable"), "Yes", 
                                          ifelse(is.na(CHILD_metadata_all2$asthma_5y) | is.na(CHILD_metadata_all2$asthma_3y) |
                                                   CHILD_metadata_all2$asthma_3y == "Possible", NA, "No"))

table(CHILD_metadata_all2$asthma_3y_5y)


## Make the "Probable" Asthma level "Yes" for asthma
CHILD_metadata_all2$asthma_3y <- gsub("Probable", "Yes", CHILD_metadata_all2$asthma_3y)
CHILD_metadata_all2$asthma_5y <- gsub("Probable", "Yes", CHILD_metadata_all2$asthma_5y)

#### Edit some variables that we decided to use but still need to re-code NA values, etc.
CHILD_metadata_all2$Fever_1y <- as.factor(ifelse(CHILD_metadata_all2$Fever_1y %in% c(8888, 999), NA, CHILD_metadata_all2$Fever_1y))
CHILD_metadata_all2$Fever_1y <- revalue(CHILD_metadata_all2$Fever_1y, c("0"="No", "1"="Yes"))

CHILD_metadata_all2$Fever_3m <- as.factor(ifelse(CHILD_metadata_all2$Fever_3m %in% c(8888, 999), NA, CHILD_metadata_all2$Fever_3m))
CHILD_metadata_all2$Fever_3m <- revalue(CHILD_metadata_all2$Fever_3m, c("0"="No", "1"="Yes"))

CHILD_metadata_all2$older_sibs_living <- ifelse(CHILD_metadata_all2$older_sibs_living %in% c("8888", "999"), NA, CHILD_metadata_all2$older_sibs_living)

### for number of colds if any colds question, 888 indicates no response b/c there were no colds, just make 0
CHILD_metadata_all2$coldsnum_0to3m <- ifelse(CHILD_metadata_all2$coldsnum_0to3m == "8888", NA, 
                                             ifelse(CHILD_metadata_all2$coldsnum_0to3m == "888", 0, CHILD_metadata_all2$coldsnum_0to3m))


# need to change factor levels based on which factor we want as reference for downstream models for the factors/characters in this list:
Main_vars_Traj <- c("BF_status_3m_sampling", "BF_12m_sampling", "RE_Virus_1y", "RE_Virus_3m", "colds_0to3m", "coldsnum_6to12m", "Fever_3m", "Fever_1y", "older_sibs", "AwayfromHome_1y", "StudyCenter","Season_3m", "Season_1y", "atbx", "Mother_abs_birth_yn", "prenatal_smoke",  "mom_smoke_1y", "mom_bmi_best", "birthmode2", "mom_age_birth", "race_mom4", "mom_asthma_ever", "mom_atopy", "age_3m_months", "age_1y_months", "weight_3", "weight_12", "childgender", "gest_days") 

# split numeric and categorical variables to edit categorical variables, then re-merge with numeric
sdf_3m2_num <- CHILD_metadata_all2 %>% select_if(is.numeric) 
sdf_cat <- subset(CHILD_metadata_all2, select = setdiff(Main_vars_Traj, colnames(sdf_3m2_num))) # Factors to modify levels for; in Main_vars_Traj but not numeric

sdf_other <- subset(CHILD_metadata_all2, select = setdiff(colnames(CHILD_metadata_all2), colnames(sdf_cat))) # DF of Metadata not being edited

# check that all cols are accounted for in two subsets
sum(dim(sdf_other)[2] + dim(sdf_cat)[2]) == dim(CHILD_metadata_all2)[2] # TRUE, good

# make all characters factors 
sdf_cat[] <- lapply(sdf_cat[], factor)
lapply(sdf_cat[], levels) # better to swap factor level for most of these... Nearly all need to be reversed

# Want all factor levels to be reversed - this way the "no" group will generally be the last group (ref for lmp is last level)
sdf_cat[] <- lapply(sdf_cat[], function(x) {factor(x, levels = rev(levels(x)))})

## A few exceptions to the rule, want a different order for these factor levels
sdf_cat$BF_status_3m_sampling <- factor(sdf_cat$BF_status_3m_sampling, levels = c("Exclusive", "Partial", "No BM"))
sdf_cat$BF_12m_sampling <- factor(sdf_cat$BF_12m_sampling, levels = c("Yes", "No"))
sdf_cat$birthmode2 <- factor(sdf_cat$birthmode2, levels = c("Caesarean", "Vaginal"))
sdf_cat$race_mom4 <- factor(sdf_cat$race_mom4, levels = c("Asian", "FN", "Other", "Caucasian"))
sdf_cat$StudyCenter <- factor(sdf_cat$StudyCenter, levels = c("Edmonton", "Toronto", "Vancouver", "Winnipeg"))
sdf_cat$Season_3m <- factor(sdf_cat$Season_3m, levels = c("Spring", "Summer", "Fall", "Winter"))
sdf_cat$Season_1y <- factor(sdf_cat$Season_1y, levels = c("Spring", "Summer", "Fall", "Winter"))

lapply(as.list(sdf_cat), levels) # Looks much better

# re-merge data using data.frame function - is maintained... just double check rownames are the same in both first!
identical(rownames(sdf_cat), rownames(sdf_other))
CHILD_metadata_all3 <- data.frame(sdf_cat, sdf_other)

saveRDS(CHILD_metadata_all3, file="Preprocessing/metadata_input/4FINAL_Full_Cleaned_KF_Modified_V2_withRE_forSubj_withNG.rds") ## all CHILD participants, save edited version
```

#### Link Participant metadata with Microbiome Trajectory data and save**
```{r} 
sdf_forAllTrajs <- subset(CHILD_metadata_all3, row.names(CHILD_metadata_all3) %in% row.names(GN_3m1y_all_traj)) 

# Check that row IDs and order is exactly the same - good
identical(row.names(sdf_forAllTrajs), row.names(GN_3m1y_all_traj))

# Won't merge but just combine into a list of dataframes before saving as a Trajectory analysis "package"
## If wanting to split nasal and gut again later, just remove NA rows based on a known nasal-specific variable like Shannon_N1y3m (then any gut-specific taxa colonization features will all be all NA as well and these columns can be removed based on this)
Traj_data_ls <- list(GN_3m1y_all_traj, sdf_forAllTrajs)
names(Traj_data_ls) <- c("Trajectories_Nasal_to181_Gut_at182", "Metadata") # indicating where gut Trajs start
str(Traj_data_ls)

saveRDS(Traj_data_ls, file="Data_subsets/ALL_Trajectory_Derived_Data.rds") 
```



